### 2020-09-09 reports/Project Outline.tex
add analyses and reports folders

\documentclass{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{palatino}

\usepackage{fancyhdr}
 \fancypagestyle{plain}

\usepackage{graphicx}
\usepackage{enumerate}

% Header
\fancyhf{}
\rhead{\begin{picture}(0,0) \put(-56.7,0){\includegraphics[width=20mm]{dbsse}} \end{picture}}
\lhead{\begin{picture}(0,0) \put(0,0){\includegraphics[width=25mm]{ethz}} \end{picture}}
%\chead{Master's Thesis Project Outline}
\renewcommand{\headrulewidth}{0pt}

 % Foot
\rfoot{Basel, \today}

% Title

\usepackage{etoolbox}
\makeatletter
\patchcmd{\@maketitle}{\begin{center}}{\begin{flushleft}}{}{}
\patchcmd{\@maketitle}{\begin{tabular}[t]{c}}{\begin{tabular}[t]{@{}l}}{}{}
\patchcmd{\@maketitle}{\end{center}}{\end{flushleft}}{}{}
\makeatother

\title{A comprehensive study of the phylodynamics of SARS-CoV-2 in Europe }
\author{Cecilia Valenzuela Agüí \\  \\  Supervised by Prof. Dr. Tanja Stadler, Dr. Timothy Vaughan \\ Computational Biology Group }
\date{CBB Master's Thesis Project Outline \\ September 1, 2020 - March 16, 2021 }

% Document
\begin{document}

\maketitle
\begin{flushleft}
\rule{\textwidth}{0.5pt}

\vspace{10mm}

This project will focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe. We will apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences. The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.

\vspace{4mm}

For the analysis, we will use the software package BEAST2, a tool for Bayesian evolutionary analysis of molecular sequences using Markov chain Monte Carlo (MCMC). The sequence data will be gathered from the publicly accesible GISAID database.

\vspace{4mm}

\textbf{Project tasks}

\begin{enumerate}
\item Create a current and representative sequence alignment for the early phase of COVID-19 pandemic from available SARS-CoV sequences . This includes sequence selection, curation and alignment.
\item Use travel and geographical data as the basis for informative priors on migration rates, this could significantly improve the precision and accuracy of the results.
\item Reconstruction of a partial transmission tree of the early pandemic, including inferred geographic location of ancestral lineages.
\item Inference of multi-type epidemic trajectories, allowing us to directly estimate the number SARS-CoV-2 of introductions into a country.
\end{enumerate}

\end{flushleft}
\end{document}

### 2020-11-09 reports/paper/main.tex
workflow, flight matrix and paper

\documentclass[12pt,]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
% \usepackage{authblk}

\usepackage[left]{lineno}
\linenumbers

\title{A comprehensive study of the early phylodynamics of SARS-CoV-2 in Europe}

\begin{document}
\maketitle

\abstract{This project will focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe. We will apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences. The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.}

\section{Results}

blablabl

\end{document}

### 2020-11-09 workflow/scripts/chains_diagnostic.R

@@ -11,6 +11,7 @@
 library(argparse)
 library(dplyr)
 library(stringr)
+library(readr)
 
 # Parser -----------------------------------------------------------------------
 parser <- argparse::ArgumentParser()
@@ -20,36 +21,37 @@ parser$add_argument("--output", type="character", help="Output diagnostic file")
 args <- parser$parse_args()
 
 INPUT <- args$input
-ESS <- args$ess
-OUTPUT <- args$output_log
+ESS_cutoff <- args$ess
+OUTPUT <- args$output
 
 print(paste("log summaries:", INPUT))
-print(paste("ESS cutoff value", ESS))
+print(paste("ESS cutoff value", ESS_cutoff))
 print(paste("output diagnostic file:", OUTPUT))
 
 # Diagnostics ------------------------------------------------------------------
 diagnostic <- data.frame()
 for (fname in INPUT){
-  tb <- read_table2(fname, skip = 1)
+  tb <- read.table(fname, skip = 1, fill = TRUE, header = TRUE, comment.char = "*", flush = TRUE, stringsAsFactors = FALSE) %>%
+    mutate_at(vars(ESS), as.numeric)
   l <- last(strsplit(read_lines(fname, n_max = 1), " ")[[1]])
-  if (all(tb$ESS >= ESS, na.rm = TRUE)) {
-    cat("All ESS values > 200 in log file ", fname, "\nLog file included in analysis.")
-    diagnostic <- rbind(diagnostic, data.frame(analysis = str_split(fname, pattern ="\\.")[[1]][1],
+  if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
+    cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
+    diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".logsummary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
                                                seed = str_split(fname, pattern ="\\.")[[1]][2],
                                                length = l,
-                                               min_ESS = ESS, 
+                                               min_ESS = ESS_cutoff, 
                                                included = 1))
   } else {
-    cat("Items:\n", tb %>% filter(ESS < ESS) %>% pull(statistic),
-        "\nhave ESS value < 200 in log file ", fname, "\nLog file not included in analysis.")
-    diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".logsummary.txt", "", fname), 
[…]

### 2020-11-10 workflow/scripts/analyze_trajectories.R
mapping and analyze trajs wip

@@ -22,62 +22,73 @@ source("./scripts/utils.R")
 
 # Parser -----------------------------------------------------------------------
 parser <- argparse::ArgumentParser()
-parser$add_argument("--input", nargs="+", 
-                    help=".txt with trajectory filenames to be included in the analysis")
+parser$add_argument("--input", type = "character", help="trajectory file")
 parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
 parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
 parser$add_argument("--demes", type = "character", help = "Demes configuration")
 parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
 args <- parser$parse_args()
 
-INPUT <- read_lines(args$input)
-BURNIN <- args$burnin
-METADATA <- args$metadata
-DEMES <- args$demes
-OUTPUT_FIGURE <- args$output_figure
+print(args)
 
-print(paste("Trajectory file:", INPUT))
-print(paste("Burning fraction:", BURNIN))
-print(paste("Alignment sequences metadata: ", METADATA))
-print(paste("Demes: ", "Demes specification in config file"))
-print(paste("Output figures:", OUTPUT_FIGURE))
 
 # Load trajectory data ---------------------------------------------------------
-states <- data.frame()
-events <- data.frame()
-for (filename in INPUT){
-  df <- loadTrajectories(filename, burninFrac=BURNIN, subsample=200)
-  states <- states %>% bind_rows(df$states %>% mutate(file = filename))
-  events <- events %>% bind_rows(df$events %>% mutate(file = filename))
-}
+df <- loadTrajectories(args$input, burninFrac = args$burnin, subsample = 200)
+cat("\nNumber of analyzed trajectories (subsampled):", length(unique(df$states$traj)))
+
+
+# Load demes configuration and sequence metadata -------------------------------
+demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
+                    exclude_country = NA, exclude_division = NA,
+                    min_date = NA, max_date = NA, seq_per_deme = NA) %>%
+  bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(args$demes, collapse = " ")),
[…]

### 2020-11-10 workflow/scripts/subsample.R

@@ -41,7 +41,7 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
            date <= as.Date(to))
   
   # Get case counts for deme
-  cases_deme <- get_cases(region_name, country_name, division_name, 
+  cases_deme <- get_cases_deme(region_name, country_name, division_name, 
                           exclude_country, exclude_division, 
                           from, to)  
 

### 2020-11-10 workflow/scripts/utils.R

@@ -19,52 +19,48 @@ set_plotopts <- function() {
 
 
 
-get_cases <- function(region_name = NA, country_name = NA, division_name = NA, 
+get_cases <- function(demes, from = NA, to = NA) {
+  # Get case data information from ECDC for specific demes (countries and regions)
+  # Set from and to dates with values from the function call instead of the ones from the demes info
+  cat("\nLoading case data counts from ECDC...\n")
+  
+  if (!is.na(from)) demes$min_date <- from
+  if (!is.na(to)) demes$max_date <- to
+  
+  # Filter case count data according to deme info
+  case_data_list <- mapply(get_cases_deme, demes$deme,
+                           demes$region, demes$country, demes$division,
+                           demes$exclude_country, demes$exclude_division,
+                           demes$min_date, demes$max_date)
+  case_data_dfs <- lapply(1:dim(case_data_list)[2], function(i) as.data.frame(case_data_list[,i]))
+  case_data <- bind_rows(case_data_dfs)
+  cat("done!")
+  return(case_data)
+}
+
+
+get_cases_deme <- function(deme_name, region_name = NA, country_name = NA, division_name = NA, 
                       exclude_country = NA, exclude_division = NA,
                       from, to) {
   #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
   #else case_data <- get_dataECDC()
-  
-  # For now, no data 
-  
   case_data <- get_dataECDC()
-  
+  print(deme_name)
   df_cases <- case_data %>%
     filter(if (is.na(region_name)) TRUE else region %in% region_name,
            if (is.na(country_name)) TRUE else country %in% country_name,
            #if (is.na(division_name)) TRUE else division %in% division_name,
-           if (all(is.na(exclude_country))) TRUE else !country %in% exclude_country,
+           if (all(is.na(exclude_country))) TRUE else !country %in% str_split(exclude_country, pattern = ",")[[1]],
            #if (is.na(exclude_division)) TRUE else !division_name %in% exclude_division,
            date >= as.Date(from),
-           date <= as.Date(to))
+           date <= as.Date(to)) %>%
+    mutate(deme = deme_name)
   
   return(df_cases)
 }  
 
-get_dataJH <- function() {
[…]

### 2020-11-11 workflow/scripts/analyze_trajectories.R
rule for getting sampling proportions

@@ -25,7 +25,7 @@ parser <- argparse::ArgumentParser()
 parser$add_argument("--input", type = "character", help="trajectory file")
 parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
 parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
-parser$add_argument("--demes", type = "character", help = "Demes configuration")
+parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
 parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
 args <- parser$parse_args()
 

### 2020-11-11 workflow/scripts/sampling_proportion.R

@@ -1,0 +1,52 @@
+##------------------------------------------------------------------------------
+## Script for plotting and manipulating trajectory data using functions from 
+## BDMM-Prime project from T.Vaughan to parse trajectories.
+## Script for snakemake workflow.
+## 
+## 2020-10-12 Cecilia Valenzuela
+##------------------------------------------------------------------------------
+
+# Load libraries ---------------------------------------------------------------
+library(argparse)
+library(yaml)
+library(tidyverse)
+
+# Source files -----------------------------------------------------------------
+source("./scripts/utils.R")
+
+# Parser -----------------------------------------------------------------------
+parser <- argparse::ArgumentParser()
+parser$add_argument("--input", type = "character", nargs = "+", help="Subsampling demes configuration")
+parser$add_argument("--output", type = "character", help = "Output path for the sampling proportions table")
+args <- parser$parse_args()
+
+print(args)
+
+# Load demes configuration -----------------------------------------------------
+demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
+                    exclude_country = NA, exclude_division = NA,
+                    min_date = NA, max_date = NA, seq_per_deme = NA) %>%
+  bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(args$input, collapse = " ")),
+                             data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
+  group_by(deme) %>% 
+  mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
+  distinct() %>%
+  filter_all(any_vars(!is.na(.)))
+
+cat("\nDeme configuration:\n")
+knitr::kable(demes)
+
+# Case data information from ECDC ----------------------------------------------
+case_data <- get_cases(demes) %>%
+  arrange(date) %>%
+  group_by(deme) %>%
+  summarise(totalcases = sum(cases))
+
+# Compute sampling proportions -------------------------------------------------
+sampling_proportions <- case_data%>%
+  left_join(demes) %>%
+  mutate(sampling = seq_per_deme/totalcases)
+
+# Save table -------------------------------------------------------------------
+write.table(sampling_proportions, args$output, sep = "\t")
+

### 2020-11-12 reports/paper/main.tex
wip results figures #11

@@
 \usepackage{fullpage}
+
 % \usepackage{authblk}
+\usepackage{graphicx}
+\usepackage{subfig}
+\graphicspath{{./figures/}}
 
@@
 
-\abstract{This project will focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe.
-We will apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences.
+\abstract{This project focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe.
+We apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences.
 The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.}
@@
 
-blablabl
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.8\textwidth]{201014_europe2_figtraj01.png}
+    \caption{Inferred population size summary statistics for each deme over time.
+The line represents the median population trajectory and the interval is the interquartile range from a subsampled set of inferred population trajectories.
+Color represents the different demes in the analysis.}
+    \label{fig:gribbon}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.9\textwidth]{201014_europe2_figtraj02.png}
+    \caption{Inferred population size trajectories over time.
+A subsampled set of 200 inferred trajectories is plotted.
+In each subplot, the trajectories (solid lines) are compared with the ECDC total case count data (dashed line).}
+    \label{fig:trajs}
+\end{figure}
+
+\begin{figure}[!tbp]
+  \centering
+  \subfloat[Temporal distribution of first introduction.]{\includegraphics[width=0.5\textwidth]{201014_europe2_figtraj03.png}\label{fig:f1}}
+  \subfloat[Source of first introduction.]{\includegraphics[width=0.5\textwidth]{201014_europe2_figtraj04.png}\label{fig:f2}}
+  \caption{First introductions event time and source in each deme.
+(a) Distribution of the time of the first introduction for each deme from the subsampled set of trajectories.
+Each box and color represent one deme and the X is the first date when cases where reported to ECDC.
+In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
+For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
+(b) Histogram of the source of the first introduction event in each deme for the set of subsampled trajectories.}
+\end{figure}
 

### 2020-11-12 workflow/scripts/flight_data.R
add subsampling pro selection #8

@@ -1,0 +1,151 @@
+ ## Flight data
+
+library(tidyverse)
+library(lubridate)
+
+# Returns dayly passengers in a given month 
+get_flightData <- function(intraUE_data, extraUE_data, year, month) {
+  df_int <- read.csv(intraUE_data, stringsAsFactors = FALSE)
+  df_ext <- read.csv(extraUE_data, stringsAsFactors = FALSE) 
+  
+  df <- bind_rows(df_int, df_ext) %>%
+    # Data from passengers, monthly
+    filter(UNIT == "Passenger", grepl("M", TIME))
+  df1 <- df %>%
+    # Clean dataset
+    mutate(YEAR = as.numeric(substr(TIME,1,4)),
+           MONTH = as.numeric(substr(TIME,6,8)),
+           Value = as.numeric(gsub(" ", "", Value))/lubridate::days_in_month(month),
+           GEO = case_when(
+             GEO == "Germany (until 1990 former territory of the FRG)" ~ "Germany",
+             GEO == "European Union - 27 countries (from 2020)" ~ "EU27",
+             GEO == "European Union - 28 countries (2013-2020)" ~ "EU28",
+             TRUE ~ df$GEO),
+           PARTNER = case_when(
+             PARTNER == "Germany (until 1990 former territory of the FRG)" ~ "Germany",
+             PARTNER == "European Union - 27 countries (from 2020)" ~ "EU27",
+             PARTNER == "European Union - 28 countries (2013-2020)" ~ "EU28",
+             PARTNER == "China including Hong Kong" ~ "China",
+             TRUE ~ df$PARTNER)) %>%
+    # Select passengers carried, departures for all countries (if we consider
+    # departures and arrivals for each country we have duplicated values, one 
+    # for each reporting country. Recommendation from eurostats is to take only
+    # the departures) and departures and arrivals for China
+    filter(TRA_MEAS == "Passengers carried (departures)" | 
+             (TRA_MEAS == "Passengers carried (arrival)" & PARTNER == "China")) %>%
[…]

### 2020-11-12 workflow/scripts/subsample.R

@@ -18,7 +18,7 @@
 subsample <- function(alignment, metadata, include = NA, exclude = NA,
                       region_name = NA, country_name = NA, division_name = NA,
                       exclude_country = NA, exclude_division = NA,
-                      from, to, seq_per_deme, seed) {
+                      from, to, seq_per_deme, seed, prob = "time") {
   set.seed(seed)
   # Read files
   alignment <- read.FASTA(file = alignment)
@@ -40,30 +40,51 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
            date >= as.Date(from),
            date <= as.Date(to))
   
-  # Get case counts for deme
-  cases_deme <- get_cases_deme(region_name, country_name, division_name, 
-                          exclude_country, exclude_division, 
-                          from, to)  
-
-  # Compute weights and probabilities for each date
-  by_date <- metadata_deme %>%
-    count(date, country) %>%
-    left_join(cases_deme, by = c("date", "country")) %>%
-    replace_na(list(cases = 0, cumcases = 0)) %>%
-    mutate(p_case = (cases + 1)/sum(cases),
-           w_date = sum(n)/n,
-           p_date = (p_case * w_date)/sum(p_case * w_date)) %>%
-    select(date, country, p_case, w_date, p_date) 
-
-  #Compute weights for each division
-  by_div <- metadata_deme %>%
-    count(division, country) %>%
-    mutate(w_div = case_when(
-                        is.na(country_name) ~ 1,
-                        TRUE ~ sum(n)/n),
-           p_div = w_div/sum(w_div)) %>%
-    select(division, country, w_div, p_div)
-
+  if (prob %in% c("cases", "cases_and_location")){
+    # Get case counts for deme
+    cases_deme <- get_cases_deme("deme", region_name, country_name, division_name, 
+                            exclude_country, exclude_division, 
+                            from, to)  
+  
+    # Compute weights and probabilities for each date
+    # TODO delay in deaths
+    by_date <- metadata_deme %>%
+      count(date, country) %>%
+      left_join(cases_deme, by = c("date", "country")) %>%
+      replace_na(list(cases = 0, cumcases = 0, deaths = 0, cumdeaths = 0)) %>%
+      mutate(p_case = ifelse(country == "China", (deaths + 1)/sum(deaths), (cases + 1)/sum(cases)),
+             w_date = sum(n)/n,
[…]

### 2020-11-12 workflow/scripts/analyze_trajectories.R
wip trajectory plots #11

@@ -46,7 +46,8 @@ demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
   group_by(deme) %>% 
   mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
   distinct() %>%
-  filter_all(any_vars(!is.na(.)))
+  filter_all(any_vars(!is.na(.))) %>%
+  mutate(deme = ifelse(deme == "OtherEuropean", "Other European", deme))
   
 cat("\nDeme configuration:\n")
 knitr::kable(demes)
@@ -54,7 +55,10 @@ metadata <- read_tsv(args$metadata)
 
 
 # Case data information from ECDC ----------------------------------------------
-case_data <- get_cases(demes, to = MRS) %>%
+# Most recent sample values
+MRS <- max(metadata$date)
+mrs <- ymd_hms(paste(MRS, '23:59:00 UTC'))
+case_data <- get_cases(demes, from = ymd("2019-09-01"), to = MRS) %>%
   arrange(date) %>%
   group_by(deme) %>%
   mutate(cumcases = cumsum(cases),
@@ -63,10 +67,6 @@ case_data <- get_cases(demes, to = MRS) %>%
 
 # Data wrangling ---------------------------------------------------------------
 cat("Data preparation...")
-
-# Most recent sample values
-MRS <- max(metadata$date)
-mrs <- ymd_hms(paste(MRS, '23:59:00 UTC'))
 
 states <- df$states %>%
   mutate(type = factor(type, 
@@ -139,12 +139,17 @@ gribbon <- ggplot(gt_summary) +
   geom_line(aes(date, Imedian, colour = type)) +
   ylab("Population size") +
   xlab("Date") +
-  labs(color = "Median", fill = "IQR", 
-       title = "SARS-CoV-2 early epidemics in Europe and Hubei",
-       subtitle = "Case trajectories Median and IQR by deme") +
+  #labs(title = "SARS-CoV-2 early epidemics in Europe and Hubei",
+       #subtitle = "Case trajectories Median and IQR by deme") +
   scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
-  scale_color_npg() +
-  scale_fill_npg()
+  scale_fill_manual(name = "", values = dcolors) +
+  scale_color_manual(name = "", values = dcolors) +
+  theme(legend.position = c(0.2, 0.7),
+        legend.box="horizontal")
+
[…]

### 2020-11-12 workflow/scripts/utils.R

@@ -5,14 +5,19 @@
 ##------------------------------------------------------
 
 library(countrycode)
+library(ggpubr)
 
 set_plotopts <- function() {
   theme_set(theme_minimal())
-  theme_update(strip.background = element_rect(fill = "grey95", color = "grey95", size = 1),
+  theme_update(#strip.background = element_rect(fill = "grey95", color = "grey95", size = 1),
+               panel.grid = element_blank(),
+               panel.background = element_rect(fill = "white", colour = "grey50"),
+               axis.ticks = element_line(),
                legend.position = "bottom",
                legend.box="vertical",
                #legend.title = element_text(size = 9, face = "bold"),
-               axis.text.x = element_text(angle = 270),
+               #axis.text.x = element_text(angle = 270),
+               #axis.text.y = element_text(),
                axis.title.y = element_text(size = 9, face = "bold", margin = margin(r = 10)),
                axis.title.x = element_text(size = 9, face = "bold", margin = margin(t = 10)))
 }

### 2020-11-13 workflow/scripts/chains_diagnostic.R
solved deleted output of failed beast rule

@@ -31,24 +31,26 @@ print(paste("output diagnostic file:", OUTPUT))
 # Diagnostics ------------------------------------------------------------------
 diagnostic <- data.frame()
 for (fname in INPUT){
-  tb <- read.table(fname, skip = 1, fill = TRUE, header = TRUE, comment.char = "*", flush = TRUE, stringsAsFactors = FALSE) %>%
-    mutate_at(vars(ESS), as.numeric)
-  l <- last(strsplit(read_lines(fname, n_max = 1), " ")[[1]])
-  if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
-    cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
-    diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".logsummary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
-                                               seed = str_split(fname, pattern ="\\.")[[1]][2],
-                                               length = l,
-                                               min_ESS = ESS_cutoff, 
-                                               included = 1))
-  } else {
-    cat("\n\n\nItems:\n\n", tb %>% filter(ESS < ESS_cutoff) %>% pull(statistic),
-        "\n\nhave ESS value < ", ESS_cutoff, " in log file ", fname, "\n\nLog file not included in analysis.\n")
-    diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".logsummary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]), 
-                                               seed = str_split(fname, pattern ="\\.")[[1]][2],
-                                               length = l,
-                                               min_ESS = ESS_cutoff, 
-                                               included = 0))
+  if (file.size(fname) != 0) {
+    tb <- read.table(fname, skip = 1, fill = TRUE, header = TRUE, comment.char = "*", flush = TRUE, stringsAsFactors = FALSE) %>%
+      mutate_at(vars(ESS), as.numeric)
+    l <- last(strsplit(read_lines(fname, n_max = 1), " ")[[1]])
+    if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
+      cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
[…]

### 2020-11-13 workflow/scripts/analyze_subsample.R
debug subsampling, get plots and add deaths prob #8

@@ -65,3 +65,167 @@ cat("Histogram of sequences by time and deme saved in ", OUTPUT_FIGURE)
 mrs <- min(as.Date(metadata%>%dplyr::filter(new_seq_name%in%names(alignment))$date))
 write(mrs, MRS)
 
+
+
+
+## Debuf subsample approach 13 Nov 20
+
+library(tidyverse)
+library(RColorBrewer)
+
+source("scripts/utils.R") 
+subsample_test <- function(metadata, include = NA, exclude = NA,
+                           region_name = NA, country_name = NA, division_name = NA,
+                           exclude_country = NA, exclude_division = NA,
+                           from, to, seq_per_deme, seed, prob = "time") {
+  set.seed(seed)
+  # Read files
+  if (!is.na(exclude)) {
+    exclude <- readLines(include)
+    metadata <- filter(metadata, !strain %in% exclude)
+  }
+  # Filter metadata
+  metadata_deme <- metadata %>%
+    filter(region == region_name,
+           if (is.na(country_name)) TRUE else country %in% country_name,
+           if (is.na(division_name)) TRUE else division %in% division_name,
+           if (all(is.na(exclude_country))) TRUE else !country %in% exclude_country,
+           if (all(is.na(exclude_division))) TRUE else !division %in% exclude_division,
+           date >= as.Date(from),
+           date <= as.Date(to))
+  
+  if (nrow(metadata_deme) == 0) {
+    warning("No sequences for these specifications")
+    return()
+  }
+  
+  cases_deme <- get_cases_deme("deme", region_name, country_name, division_name, 
+                               exclude_country, exclude_division, 
+                               from, to)  
+  
+  if (prob %in% c("cases", "cases_and_location")){
+    # Compute weights and probabilities for each date
+    # TODO delay in deaths
+    by_date <- metadata_deme %>%
+      count(date, country) %>%
+      left_join(cases_deme, by = c("date", "country")) %>%
+      replace_na(list(cases = 0, cumcases = 0, deaths = 0, cumdeaths = 0)) %>%
+      mutate(p_case = ifelse(country == "China", (deaths + 1)/sum(deaths), (cases + 1)/sum(cases)),
+             w_date = sum(n)/n,
[…]

### 2020-11-13 workflow/scripts/subsample.R

@@ -24,12 +24,13 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
   alignment <- read.FASTA(file = alignment)
   metadata <- read.delim(file = metadata, stringsAsFactors = FALSE) %>%
     filter(strain %in% names(alignment)) %>% # Keep only metadata about seqs in the alignment
-    mutate(date = as.Date(date)) 
+    mutate(date = ymd(date)) 
   
   if (!is.null(exclude)) {
     exclude <- readLines(include)
     metadata <- filter(metadata, !strain %in% exclude)
-    }
+  }
+  
   # Filter metadata
   metadata_deme <- metadata %>%
     filter(region == region_name,
@@ -37,40 +38,57 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
            if (is.na(division_name)) TRUE else division %in% division_name,
            if (all(is.na(exclude_country))) TRUE else !country %in% exclude_country,
            if (all(is.na(exclude_division))) TRUE else !division %in% exclude_division,
-           date >= as.Date(from),
-           date <= as.Date(to))
+           date >= ymd(from),
+           date <= ymd(to))
   
-  if (prob %in% c("cases", "cases_and_location")){
-    # Get case counts for deme
-    cases_deme <- get_cases_deme("deme", region_name, country_name, division_name, 
-                            exclude_country, exclude_division, 
-                            from, to)  
+  if (nrow(metadata_deme) == 0) {
+    warning("No sequences for these specifications")
+    return()
+  }
   
+  if (prob == "deaths") to <- ymd(to) + 14
+  
+  # Get case counts for deme
+  cases_deme <- get_cases_deme("deme", region_name, country_name, division_name, 
+                               exclude_country, exclude_division, 
+                               from, to)  
+  
+  # Account for lag cases - deaths
+  if (prob == "deaths") cases_deme <- cases_deme %>% mutate(date = date - 14)
+  
+  if (prob %in% c("cases", "cases_and_location", "deaths")){
     # Compute weights and probabilities for each date
-    # TODO delay in deaths
     by_date <- metadata_deme %>%
       count(date, country) %>%
       left_join(cases_deme, by = c("date", "country")) %>%
[…]

### 2020-11-13 workflow/scripts/adjust_names.R
correction duplicates adjust names

@@ -52,14 +52,15 @@ metadata <- full_metadata %>%
   left_join(demes, by = c("region")) %>%
   filter((country.x == country.y) | 
            (!country.x %in% demes$country & is.na(country.y))) %>%
-# 3. Create names GISAID EPI ISL | Deme | Date
-  mutate(strain_old = strain,
-         strain = paste(gisaid_epi_isl, deme, date, sep = "/")) %>%
-  unique()
+# 3. Create names GISAID EPI ISL / Deme / Date
+  rename(strain_old = strain) %>%
+  mutate(strain = paste(gisaid_epi_isl, deme, date, sep = "/")) %>%
+  distinct(strain, .keep_all = TRUE)
 
 # Adjust alignment names -------------------------------------------------------
 strain_names <- names(alignment)
 full_names <- metadata$strain[match(strain_names, metadata$strain_old)]
+print(full_names)
 if (any(is.na(full_names))) {
   stop("Not all strains have full names in metadata.")
 }

### 2020-11-13 utils/ECDCvsJH.R
update utils

@@ -1,0 +1,35 @@
+library(coronavirus)
+library(dplyr)
+
+jh <- coronavirus%>%filter(province=="Hubei" & type=="confirmed" & as.Date(date) <= "2020-03-08")%>%
+  mutate(source = "jh")%>%
+  select(country, date, cases, source)
+ecdc <- read.csv("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", na.strings = "", fileEncoding = "UTF-8-BOM")%>%
+  dplyr::rename(country = countriesAndTerritories,
+                region = continentExp)%>%
+  dplyr::mutate(date = as.Date(dateRep, "%d/%m/%Y"),
+                source = "ecdc")%>%
+  dplyr::filter(date <= as.Date("2020-03-08") & country == "China")%>%
+  dplyr::select(country, date, cases, source)
+
+summarise(jh, total = sum(cases))
+summarise(ecdc, total = sum(cases))
+
+# No significant change when plotting case count in log scale. 
+# We continue to use ECDC data for China instead of JH data for Hubei
+# knowing that is a overestimation of case count data in Hubei.
+# Easier to work with just one case count dataset and reported case counts 
+# earlied in ECDC dataset
+
+cases_hubei <- dplyr::bind_rows(ecdc, jh)%>%
+  dplyr::arrange(date)%>%
+  dplyr::group_by(source)%>%
+  dplyr::mutate(sumcases = cumsum(cases))
+
+ggplot(cases_hubei) +
+  geom_line(aes(x=date, y=sumcases, linetype=source))+
+  labs(title="Hubei case count")+
+  scale_y_log10(labels= trans_format("log10", math_format(10^.x)))
+
+
+  

### 2020-11-13 utils/check subsamble.R

@@ -1,0 +1,32 @@
+library(dplyr)
+
+SEQS <- "200914_europe1/results/subsampled_alignment.fasta"
+SEQS <- "~/subsampled_alignment.fasta"
+alignment <- ape::read.FASTA(file = SEQS)
+seqs <- names(alignment)
+
+
+INCLUDE <- "200914_europe1/files/include.txt"
+include <- readLines(INCLUDE)
+
+all(include%in%seqs)
+
+include[!include%in%seqs]
+
+METADATA <- "200914_europe1/data/200911_metadata.tsv"
+metadata <- read.delim(file = METADATA)%>%
+  dplyr::mutate(date=as.Date(date))%>%
+  dplyr::filter(strain%in%seqs)
+
+all(metadata$date < as.Date("2020-03-09"))
+
+metadata_hubei <- metadata%>%
+  dplyr::filter(country=="China")
+
+all(metadata_hubei$date < as.Date("2020-01-23"))
+
+summary <- metadata%>%
+  dplyr::count(country, sort=TRUE)
+View(metadata%>%filter(country=="Italy")%>%
+       mutate(include=strain%in%include))
+

### 2020-11-13 utils/explore_seqs.R

@@ -1,190 +1,171 @@
 library(tidyverse)
 library(RColorBrewer)
+library(lubridate)
 
-METADATA <- "200914_europe1/data/200930_metadata.tsv"
-EXCLUDE <- "200914_europe1/files/200930_exclude.txt"
-metadata <- read.delim(file = METADATA)%>%
-  dplyr::mutate(date=as.Date(date))
+source("scripts/utils.R") 
+subsample_test <- function(metadata, include = NA, exclude = NA,
+                      region_name = NA, country_name = NA, division_name = NA,
+                      exclude_country = NA, exclude_division = NA,
+                      from, to, seq_per_deme, seed, prob = "time") {
+  set.seed(seed)
+  # Read files
+  if (!is.na(exclude)) {
+    exclude <- readLines(include)
+    metadata <- filter(metadata, !strain %in% exclude)
+  }
+  # Filter metadata
+  metadata_deme <- metadata %>%
+    filter(region == region_name,
+           if (is.na(country_name)) TRUE else country %in% country_name,
+           if (is.na(division_name)) TRUE else division %in% division_name,
+           if (all(is.na(exclude_country))) TRUE else !country %in% exclude_country,
+           if (all(is.na(exclude_division))) TRUE else !division %in% exclude_division,
+           date >= as.Date(from),
+           date <= as.Date(to))
+  
+  if (nrow(metadata_deme) == 0) {
+    warning("No sequences for these specifications")
+    return()
+  }
+  
+  if (prob == "deaths") to <- ymd(to) + 14
+  
+  # Get case counts for deme
+  cases_deme <- get_cases_deme("deme", region_name, country_name, division_name, 
+                               exclude_country, exclude_division, 
+                               from, to)  
+  
+  # Account for lag cases - deaths
+  if (prob == "deaths") cases_deme <- cases_deme %>% mutate(date = date - 14)
+  
+  if (prob %in% c("cases", "cases_and_location", "deaths")){
+    # Compute weights and probabilities for each date
+    by_date <- metadata_deme %>%
+      count(date, country) %>%
+      left_join(cases_deme, by = c("date", "country")) %>%
[…]

### 2020-11-13 utils/map.R

@@ -1,0 +1,38 @@
+# Plot Europe map
+
+library(tmap)
+library(sf)
+library(rnaturalearth)
+library(RColorBrewer)
+library(maps)
+
+cases <- get_casesECDC(from = "2019-12-01", to = "2020-03-08")%>%
+  filter(date == as.Date("2020-03-08")) %>%
+  rename(cases14 = 12)
+
+Europe <- ne_countries(scale = 'medium', type = 'countries', returnclass = 'sf', 
+                       continent="Europe") %>%
+  rename(country = sovereignt) %>% 
+  left_join(cases, by = 'country')%>%
+  st_crop(xmin = -20, xmax = 40, ymin = 30, ymax = 73)
+ggplot() + 
+  geom_sf(data = Europe, aes(fill = log(sumcases))) +
+  geom_point(data = df %>% filter(region == "Europe"), 
+             aes(y = N, x = W, color = deme), alpha = 0.4)+
+  ggsci::scale_color_npg() +
+  #scale_fill_distiller(palette = "Spectral", limits = c(0, 10))
+  #paletteer::scale_fill_paletteer_c("nord::aurora")#scale_fill_distiller()
+
+# world <- ne_countries(scale = 'medium', type = 'map_units', returnclass = 'sf') %>%
+#   st_crop(y = c(xmin = -10, xmax = 120, ymin = 35, ymax = 70))
+# ggplot() + geom_sf(data=world) +
+#   geom_point(data = df, aes(y = N, x = W, color = deme))
+# 
+
+df <- left_join(metadata_demes, latitudes %>% filter(area == "location"))
+
+tm_shape(Europe) +
+  tm_fill() +
+  tm_borders()+
+  geom_point(aes(x=0,y=40))
+  #geom_point(aes(y = df$N[1:10], x = df$W[1:10]))

### 2020-11-13 utils/processing.sh

@@ -1,0 +1,60 @@
+#!/bin/sh
+WORKDIR="/Users/maceci/code/mt-analysis/"
+ANALYSIS="200910_dsEurope0"
+N=10
+
+# Results Processing 
+# Analysis is run on Euler and entire analysis folder move back to local folder in /mt-analysis
+mkdir $WORKDIR/$ANALYSIS/pResults
+
+
+/Applications/BEAST\ 2.6.3/bin/loganalyser /Users/maceci/code/mt-analysis/200910_dsEurope0/pResults/200910_dsEurope0.log > ~/output.txt 
+
+# 1. Combine log files with Log Combiner 
+# Check first convergence in Tracer
+LOGCOMBINER="/Applications/BEAST 2.6.3/bin/logcombiner"
+$LOGCOMBINER \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.1.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.2.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.3.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.4.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.5.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.6.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.7.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.8.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.9.log \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.10.log \
+-o $WORKDIR/pResults/${ANALYSIS}.log \
+-b 10 
+
+# 3. Log summary table
+# Visualize combined log file and export data table from Tracer
+# Transform table to .md
+Rscript $WORKDIR/txttomd.R \
+--input $WORKDIR/$ANALYSIS/pResults/${ANALYSIS}_logSummary 
+
+
+# 4. Combine typed node trees with Log Combiner
+$LOGCOMBINER \
+-log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.1.typed.node.trees \
[…]

### 2020-11-21 workflow/scripts/analyze_trajectories.R
common config file and minor fixes plots

@@ -120,7 +120,8 @@ ge <- gridEventsByAge(events, ages_1day) %>%
 cat("\nPlotting...")
 set_plotopts()
 dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
-names(dcolors) <- demes$deme
+dcolors <- dcolors[1:nrow(demes)]
+names(dcolors) <- sort(demes$deme)
 dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
 names(dark_dcolors) <- demes$deme
 ecolors <- c("#1e3d59", "#f5f0e1", "#ff6e40", "#ffc13b")
@@ -251,17 +252,18 @@ order_imig_df <-  events %>%
   arrange(traj, time) %>%
   distinct(traj, dest) %>%
   ungroup %>%
-  mutate(order = rep(c("second", "third", "fourth", "fifth", "sixth"), 200)) %>%
+  mutate(order = rep(2:nrow(demes), args$n)) %>%
   pivot_wider(names_from = order, values_from = dest) %>%
-  count(second, third, fourth, fifth, sixth) %>%
-  mutate(first = "Hubei") %>%
-  select(first, everything())
-order_imig_set <- gather_set_data(order_imig_df, 1:6)
-order_imig_set$x <- factor(order_imig_set$x, levels = c("first", "second", "third", "fourth", "fifth", "sixth"))
+  group_by_at(2:nrow(demes)) %>%
+  summarise(n = n(), .groups = "drop") %>%
+  mutate("1" = "Hubei") %>%
+  select("1", everything())
+order_imig_set <- gather_set_data(order_imig_df, 1:nrow(demes))
+order_imig_set$x <- factor(order_imig_set$x, levels = 1:nrow(demes))
 
   # 2. Plot
 order_imig <- ggplot(order_imig_set, aes(x, id = id, split = y, value = n)) +
-  geom_parallel_sets(aes(fill = second), alpha = 0.5, axis.width = 0.22) +
+  geom_parallel_sets(aes_(fill = as.name(2)), alpha = 0.5, axis.width = 0.22) +
   geom_parallel_sets_axes(axis.width = 0.15, fill = "grey80", color = "grey80") +
   geom_parallel_sets_labels(color = 'grey30', size = 9/.pt, angle = 90) +
   scale_x_discrete(breaks = NULL, name = NULL, expand = c(0, 0.2)) +
[…]

### 2020-11-23 workflow/scripts/subsample.R
initial commit europe 3c

@@ -26,7 +26,7 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
     filter(strain %in% names(alignment)) %>% # Keep only metadata about seqs in the alignment
     mutate(date = ymd(date)) 
   
-  if (!is.null(exclude)) {
+  if (!is.na(exclude)) {
     exclude <- readLines(include)
     metadata <- filter(metadata, !strain %in% exclude)
   }

### 2020-11-23 workflow/scripts/subsample.R
add beast action config #19

@@ -26,7 +26,7 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
     filter(strain %in% names(alignment)) %>% # Keep only metadata about seqs in the alignment
     mutate(date = ymd(date)) 
   
-  if (!is.na(exclude)) {
+  if (!is.null(exclude)) {
     exclude <- readLines(include)
     metadata <- filter(metadata, !strain %in% exclude)
   }

### 2020-11-23 workflow/scripts/flight_data.R
fligh matrix and analysis europeGLM 4 and 5

@@ -25,16 +25,16 @@ get_flightData <- function(intraUE_data, extraUE_data, year, month) {
              PARTNER == "Germany (until 1990 former territory of the FRG)" ~ "Germany",
              PARTNER == "European Union - 27 countries (from 2020)" ~ "EU27",
              PARTNER == "European Union - 28 countries (2013-2020)" ~ "EU28",
-             PARTNER == "China including Hong Kong" ~ "China",
+             PARTNER == "China including Hong Kong" ~ "Hubei-China",
              TRUE ~ df$PARTNER)) %>%
     # Select passengers carried, departures for all countries (if we consider
-    # departures and arrivals for each country we have duplicated values, one 
+    # departures and arrivals for each country we have duplicated values, one
     # for each reporting country. Recommendation from eurostats is to take only
     # the departures) and departures and arrivals for China
     filter(TRA_MEAS == "Passengers carried (departures)" | 
-             (TRA_MEAS == "Passengers carried (arrival)" & PARTNER == "China")) %>%
-    mutate(PARTNER = ifelse(TRA_MEAS == "Passengers carried (arrival)" & PARTNER == "China", GEO, PARTNER),
-           GEO = ifelse(TRA_MEAS == "Passengers carried (arrival)", "China", GEO),
+             (TRA_MEAS == "Passengers carried (arrival)" & PARTNER == "Hubei-China")) %>%
+    mutate(PARTNER = ifelse(TRA_MEAS == "Passengers carried (arrival)" & PARTNER == "Hubei-China", GEO, PARTNER),
+           GEO = ifelse(TRA_MEAS == "Passengers carried (arrival)", "Hubei-China", GEO),
            TRA_MEAS = ifelse(TRA_MEAS == "Passengers carried (arrival)", "Passengers carried (departures)", TRA_MEAS)) %>%
     # In 2020 UK is out of the european aggregated value
     filter((YEAR == 2019 & GEO != "EU27") | (YEAR == 2020 & GEO != "EU28")) %>%
[…]

### 2020-11-24 workflow/scripts/analyze_trajectories.R
small fixes workflow #19

@@ -54,7 +54,8 @@ demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
   mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
   distinct() %>%
   filter_all(any_vars(!is.na(.))) %>%
-  mutate(deme = ifelse(deme == "OtherEuropean", "Other European", deme))
+  mutate(deme = str_to_title(deme)) %>%
+  mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme))
   
 cat("\nDeme configuration:\n")
 knitr::kable(demes)
@@ -192,7 +193,7 @@ ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1
 
 
 # 2. Events
-# 2.1 First introduction distribution boxplot by deme and time of first reported cases
+# 2.1 First introduction distribution densuty by deme and time of first reported cases to ECDC
 # TODO Add 12 Dec for Wuhan instead of first reported cases to ECDC?
 first_prob <- events %>%
   # 1. Prepare data
@@ -463,28 +464,21 @@ plot_chord <- function(grid_df, min_date, max_date) {
 }
 
 # Period 1
+png(paste0(args$output_figure, "08a.png"), width = 1000, height = 1000, res = 200)
 plot_chord(ge, ymd("2019-09-01"), ymd("2020-01-23")) 
-text(label = "A", x = -1.1, y = 1, cex = 1.5, font = 2)
-c1 <- recordPlot()
+text(label = "A", x = -1, y = 1, cex = 1.5, font = 2)
+dev.off()
 # Period 2
+png(paste0(args$output_figure, "08b.png"), width = 1000, height = 1000, res = 200)
 plot_chord(ge, ymd("2020-01-23"), ymd("2020-02-28"))
 #mtext(text = "B", at = -1, cex = 1.5, font = 2)
-text(label = "B", x = -1.1, y = 1, cex = 1.5, font = 2)
-c2 <- recordPlot()
+text(label = "B", x = -1, y = 1, cex = 1.5, font = 2)
+dev.off()
 # Period 3
+png(paste0(args$output_figure, "08c.png"), width = 1000, height = 1000, res = 200)
 plot_chord(ge, ymd("2020-02-28"), ymd("2020-03-08"))
[…]

### 2020-11-24 workflow/scripts/chains_diagnostic.R

@@ -37,7 +37,7 @@ for (fname in INPUT){
     l <- last(strsplit(read_lines(fname, n_max = 1), " ")[[1]])
     if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
       cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
-      diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".logsummary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
+      diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
                                                  seed = str_split(fname, pattern ="\\.")[[1]][2],
                                                  length = l,
                                                  min_ESS = ESS_cutoff, 
@@ -45,7 +45,7 @@ for (fname in INPUT){
     } else {
       cat("\n\n\nItems:\n\n", tb %>% filter(ESS < ESS_cutoff) %>% pull(statistic),
           "\n\nhave ESS value < ", ESS_cutoff, " in log file ", fname, "\n\nLog file not included in analysis.\n")
-      diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".logsummary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]), 
+      diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]), 
                                                  seed = str_split(fname, pattern ="\\.")[[1]][2],
                                                  length = l,
                                                  min_ESS = ESS_cutoff, 

### 2020-11-24 workflow/scripts/analyze_trajectories.R
small fixes workflow #19

@@ -369,7 +369,8 @@ srcmig_bar <- ge %>%
   scale_y_continuous(labels = scales::percent_format(), breaks = c(0, 0.5, 1)) +
   scale_x_date(limits = c(ymd("2019-11-01"), ymd("2020-03-08")))+
   scale_fill_manual(values = dcolors) +
-  theme(legend.position = "none",
+  theme(#legend.position = "none",
+        legend.title = element_blank(),
         axis.title.x = element_blank(),
         axis.title.y = element_blank(),
         axis.text.y = element_blank(),
@@ -388,16 +389,17 @@ destmig_bar <- ge %>%
   scale_y_continuous(labels = scales::percent_format(), breaks = c(0, 0.5, 1)) +
   scale_x_date(limits = c(ymd("2019-11-01"), ymd("2020-03-08")))+
   scale_fill_manual(values = dcolors) +
-  theme(legend.position = "none",
+  theme(#legend.position = "none",
         axis.title.x = element_blank(),
         axis.title.y = element_blank(),
         axis.text.y = element_blank(),
         axis.ticks.y = element_blank(),
+        legend.title = element_blank(),
         strip.background.y = element_blank(),
         strip.text.y = element_blank()) +
   facet_grid(src~type)
 
-ggexport(ggarrange(plotlist = list(srcmig_bar, destmig_bar), ncol = 2),
+ggexport(ggarrange(plotlist = list(srcmig_bar, destmig_bar), ncol = 2, common.legend = TRUE),
          filename = paste0(args$output_figure, "07.png"),
          width = 2300, height = 3000, res = 300)
 

### 2020-11-24 workflow/scripts/utils.R

@@ -16,7 +16,7 @@ set_plotopts <- function() {
                legend.position = "bottom",
                legend.box="vertical",
                strip.background = element_rect(fill = "grey80", colour = "grey80"),
-               strip.text = element_text(color = "white"),
+               strip.text = element_text(color = "grey30"),
                #legend.title = element_text(size = 9, face = "bold"),
                #axis.text.x = element_text(angle = 270),
                #axis.text.y = element_text(),

### 2020-11-24 reports/paper/main.tex
wip results paper #11

@@
 \usepackage{graphicx}
+\usepackage{import} 
 \usepackage{subfig}
-\graphicspath{{./figures/}}
+\graphicspath{{../../results/}}
 
@@
 \linenumbers
+
+\usepackage{todonotes}
 
@@
 
-\section{Results}
+\section*{Introduction}
+
+\section*{Methods}
+
+\section*{Results}
+
+\subsubsection*{Epidemiological parameters}
+
+Not sure about this section. But maybe it make sense for completeness to include, briefly, the values of estimate R0, migration and sampling rates.
+
+\subsubsection*{Total case counts}
+
+Figures \ref{fig:gribbon} and \ref{fig:trajs}.\\
+
+Inferred population trajectories that account for all cases in the country.
+It is not an independent estimate since we are using informative priors for the sampling rates (upperbound computed as sequences/reported cases) and the subsampling scheme also includes the number of cases in the country (but that is a weaker link I think.\\
+
+Comparison with ECDC case counts: delay in notifications and underreporting, for some of the  countries at the end of the peioord the reported cases are very closed to the trajectories, e.g.
+Italy, Germany.
+Hubei-China huge difference in tendency and absolute numbers.\\ \todo{Working on these with new analyses}
+
+Measure the (mean?) difference between inferred and reported case counts and compare among countries. By day or in general? Absolute or relative to the number of cases that day?
+
+\subsubsection*{First introductions}
+Figures \ref{fig:first}, \ref{fig:first_src} and \ref{fig:first_dest}.\\
+
+We want to answer the question of when and where the first case in Europe occured.
+No clear, Germany?
+Figure \ref{fig:first} A.
+We can measure the time between first case and first reported case and compare between countries.
+\\
+\todo{ticks for every month}
+And for each country where this first case came from, more support for Hubei introduction in all countries Figure \ref{fig:first} B.\\
+
+In Figure \ref{fig:first_src}, we are looking at the ordered set of introduction intoo european countries.
+For each trajectory, we have the fist introduction into each country happening in a determined order, i.e.
+the first european country with a covid case, the second country with a covid case...
+So we are vissualizing the number of trajectories that share the same order of first cases.
+More trajectories where the first case was to Germany, no clear order after the first country.
+Spain last one.\\
+\todo{Explain this better}
+
+In Figure \ref{fig:first_dest} we are looking at the ordered set of countries spreading covid.
+In the first position we have Hubei for all trajectories since it is the origin of the epidemic.
+In second position we will have the first European country from where a covid case was imported to other country.
+No clear dynamics except that moore support for Spain being the last country having a first exported covid case.
+Germany first one.\\
+
+\subsubsection*{Migration vs within-region transmission}
+
+Figure \ref{fig:events}\\.
+
+This question is addresed on the original paper from Sarah.  Maybe could be interesting to look at this perspective since it is important for border closure measures.
+
+\todo{Think about results}
+ Not sure if we can have any meaningful results insights on the topic.
+Migrations started very early in our analysis to all european countries.
+Withing regions transmision account for most of the cases in the countries from late january onwards (when first cases were being reported in Europe).\\
+
+ Measure the time between the first outcoming mirgration from the country  (specially interesting for Hubei) and first reported case.
+This could be interesting to say if a extreme measure closing borders with the first case could be effective to impede transmission to other countries: percentage of trajectories whre transmission to Europe would have been avoided.
+For other countries we can look at how many migrations events could have been avoided (and how many not) if the country closed borders after first reported case.\\
+
+ Measure the time between first birth/migration and first reported case and compare between countries.
+How well did the countries detecting the first cases, there were already within region transmision?\\
+
+ Measure the time between the first migration into a country and the first birth in the country.
+Or the first outcoming mirgration from the country and first birth.
+This could be interesting to say if we should focus or not the screening and testing capacities to detect incoming migrations or a more general strategy to find cases in the population.
+Is it different for each country?\\
+
+\subsubsection*{Migrations patterns}
+Figures \ref{fig:migs_srcdest} and \ref{fig:migs}.\\
+
+Similar information in both plots, but in the chord plots instead of a dayly evolution the time period is split on three (as in the GLM analysis).
+Chord plots are nicer and easier to understand I think, but barplots shows the great detail of the results of the model.
+Another advantage of the chrod plot is that the ir shows mean absolute values and not only relative values.\\
+
+Hubei-China is the majority source of migrations for all countries till February and then some patterns emerge (we expect more interesting results with the added info in GLM analysis).\\
+
+\todo{Legend in plot 7 and change white text}
+
+\subsubsection*{GLM predictors}
+Which channels were the main sources of transmission across national borders.
+
+\section*{Discussion}
+
+\section*{Conclusion}
 
@@
     \caption{Inferred population size summary statistics for each deme over time.
-The line represents the median population trajectory and the interval is the interquartile range from a subsampled set of inferred population trajectories.
-Color represents the different demes in the analysis.}
+The line represents the median population trajectory and the interval is the interquartile range from a subsampled set of inferred population trajectories.}
     \label{fig:gribbon}
@@
 
-\begin{figure}[ht]
+\begin{figure}[h!]
     \centering
@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj03.png}
+    \caption{First introductions event time and source in each deme.
+\textbf{A} Probability density of the time of the first introduction for each deme from the subsampled set of trajectories.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
+For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
+\textbf{B} Probability density of the source of the first introduction event in each deme for the set of subsampled trajectories.}
+    \label{fig:first}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj04.png}
+    \caption{}
+    \label{fig:first_dest}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj05.png}
+    \caption{}
+    \label{fig:first_src}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj06.png}
+    \caption{Mean inferred cumulative number of events (births and migrations) over time. A subsampled set of 200 inferred trajectories is plotted.}
+    \label{fig:events}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj07.png}
+    \caption{}
+    \label{fig:migs_srcdest}
+\end{figure}
+
 \begin{figure}[!tbp]
   \centering
-  \subfloat[Temporal distribution of first introduction.]{\includegraphics[width=0.5\textwidth]{201014_europe2_figtraj03.png}\label{fig:f1}}
-  \subfloat[Source of first introduction.]{\includegraphics[width=0.5\textwidth]{201014_europe2_figtraj04.png}\label{fig:f2}}
-  \caption{First introductions event time and source in each deme.
-(a) Distribution of the time of the first introduction for each deme from the subsampled set of trajectories.
-Each box and color represent one deme and the X is the first date when cases where reported to ECDC.
-In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
-For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
-(b) Histogram of the source of the first introduction event in each deme for the set of subsampled trajectories.}
+
+  \begin{minipage}[t]{0.4\textwidth}
+  \includegraphics[width=\textwidth]{201014_europe2_figtraj08a.png}
+  \label{fig:migs1}
+  \end{minipage}
+  \begin{minipage}[t]{0.4\textwidth}
+  \includegraphics[width=\textwidth]{201014_europe2_figtraj08b.png}
+  \label{fig:migs2}
+  \end{minipage}
+  \begin{minipage}[t]{0.4\textwidth}
+  \includegraphics[width=\textwidth]{201014_europe2_figtraj08c.png}
+  \label{fig:migs3}
+  \end{minipage}
+  \caption{Migration flux among demes over the three periods defined in the analysis.}
+  \label{fig:migs}
 \end{figure}

### 2020-11-25 workflow/scripts/analyze_trajectories.R
some new data insights

@@ -34,6 +34,7 @@ parser$add_argument("--metadata", type = "character", help = "Alignment sequence
 parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
 parser$add_argument("--n", type = "integer", help = "Number of trajectories to analyze")
 parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
+parser$add_argument("--output_table", type = "character", help = "Output path for the tables")
 args <- parser$parse_args()
 
 print(args)
@@ -70,11 +71,16 @@ case_data <- get_cases(demes, from = ymd("2019-09-01"), to = MRS) %>%
   arrange(date) %>%
   group_by(deme) %>%
   mutate(cumcases = cumsum(cases),
-         cumdeaths = cumsum(deaths))
+         cumdeaths = cumsum(deaths)) %>%
+  group_by(deme, date) %>%
+  mutate(cumcases = max(cumcases),
+         cumdeaths = max(cumdeaths)) %>%
+  select(deme, date, cumcases, cumdeaths)  %>%
+  distinct()
 
 
 # Data wrangling ---------------------------------------------------------------
-cat("Data preparation...")
+cat("\nData preparation...")
 
 states <- df$states %>%
   mutate(type = factor(type, 
@@ -117,6 +123,141 @@ ge <- gridEventsByAge(events, ages_1day) %>%
   mutate(date = date(date_decimal(decimal_date(mrs) - age)),
          dest = ifelse(is.na(dest), as.character(src), as.character(dest)))
 
+# Analyzing --------------------------------------------------------------------
+cat("\nAnalyzing")
+
+# 1. Trajectories - Total case counts
+# 1.1 Median and IQR of Total cases on the full time period. Reporting rate median 
+#    and IQR (cases ecdc/inferred cases) wrt. ECDC reported cases
+
+r_reported <- gt_summary %>%
+  filter(date == MRS) %>%
+  left_join(case_data %>% rename(type = deme), by = c("type", "date")) %>%
+  mutate(reportrate = cumcases/Imedian,
+         lreportrate = cumcases/Ilow,
+         hreportrate = cumcases/Ihigh) %>%
+  mutate(Country = type,
+         `Cases ECDC` = cumcases,
+         `Median Cases (IQR)` = paste0(round(Imedian), " (", round(Ilow), "-", round(Ihigh), ")"),
[…]

### 2020-11-25 reports/paper/main.tex
new insights table

@@
 
-\subsubsection*{Epidemiological parameters}
-
-Not sure about this section. But maybe it make sense for completeness to include, briefly, the values of estimate R0, migration and sampling rates.
-
 \subsubsection*{Total case counts}
@@
 
-Comparison with ECDC case counts: delay in notifications and underreporting, for some of the  countries at the end of the peioord the reported cases are very closed to the trajectories, e.g.
+Comparison with ECDC case counts: delay in notifications and underreporting, for some of the  countries at the end of the periord the reported cases are very closed to the trajectories, e.g.
 Italy, Germany.
@@
 And for each country where this first case came from, more support for Hubei introduction in all countries Figure \ref{fig:first} B.\\
+
+Measure expected day of introduction.
 
@@
 Which channels were the main sources of transmission across national borders.
+
+\subsubsection*{Epidemiological parameters}
+
+Not sure about this section. But maybe it make sense for completeness to include, briefly, the values of estimate R0, migration and sampling rates.
 

### 2020-11-25 workflow/scripts/analyze_trajectories.R

@@ -130,12 +130,14 @@ cat("\nAnalyzing")
 # 1.1 Median and IQR of Total cases on the full time period. Reporting rate median 
 #    and IQR (cases ecdc/inferred cases) wrt. ECDC reported cases
 
-r_reported <- gt_summary %>%
+reported <- gt_summary %>%
   filter(date == MRS) %>%
   left_join(case_data %>% rename(type = deme), by = c("type", "date")) %>%
   mutate(reportrate = cumcases/Imedian,
          lreportrate = cumcases/Ilow,
          hreportrate = cumcases/Ihigh) %>%
+
+r_reported <- reported %>%  
   mutate(Country = type,
          `Cases ECDC` = cumcases,
          `Median Cases (IQR)` = paste0(round(Imedian), " (", round(Ilow), "-", round(Ihigh), ")"),
@@ -145,64 +147,23 @@ r_reported <- gt_summary %>%
 write.csv(r_reported, file = paste0(args$output_table, "01.csv"))
 
 # 2. Events 
-# 2.1. Time interval of first outcoming migration from a country compared to first reported case
-# In which percentage of trajectories, transmission from that country would have been avoided if
-# ways of outcoming migrations werre closed the day of first reported case.
-
-# Measure the time between the first outcoming mirgration from the country  
-# (specially interesting for Hubei) and first reported case. 
-# This could be interesting to say if a extreme measure closing borders with 
-# the first case could be effective to impede transmission to other countries: 
-# percentage of trajectories whre transmission to Europe would have been avoided. 
-# For other countries we can look at how many migrations events could have been avoided 
-# (and how many not) if the country closed borders after first reported case.\\ 
-
-days_firstmig <- events %>%
-  filter(event == "M") %>%
-  group_by(src, traj) %>%
-  arrange(time) %>%
-  slice(1) %>%
-  ungroup %>%
-  left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>% slice(1)  %>% rename(src = deme, min_date = date), by = "src") %>%
-  group_by(src, min_date) %>%
-  summarize(Dmedian = median(date), 
-            Dlow    = quantile(date, 0.25, type = 1), 
-            Dhigh   = quantile(date, 0.75, type = 1),
-            .groups = "drop") %>%
[…]

### 2020-11-27 workflow/scripts/flight_data.R
log and sd fligh matrix #9

@@ -114,20 +114,22 @@ flightsJan <- get_flightData(intraUE_data, extraUE_data, year = 2020, month = 1)
 flightsFeb <- get_flightData(intraUE_data, extraUE_data, year = 2020, month = 2)
 flightsMar <- get_flightData(intraUE_data, extraUE_data, year = 2020, month = 3)
 
+# Log transform and standardization the strictly positive predictors (add pseudocount if needed) (Lemey et al 2014)
 # Constant migration matrix
 flightMatrix_constant <- (flightsDic$matrix + flightsJan$matrix + flightsFeb$matrix + flightsMar$matrix)/120 
-vflightMatrix_constant <- na.exclude(c(t(flightMatrix_constant)))
-nvflightMatrix_constant <- vflightMatrix_constant/mean(vflightMatrix_constant)
-write(nvflightMatrix_constant, "files/flight_matrix_constant.csv", sep = ",")
-  
+v_flightMatrix_constant <- na.exclude(c(t(flightMatrix_constant)))
+lv_flightMatrix_constant  <- log(v_flightMatrix_constant)
+slv_flightMatrix_constant <- scale(lv_flightMatrix_constant)
+write(slv_flightMatrix_constant, "files/flight_matrix_constant.csv", sep = ",")
 
 # 3 epochs migration matrix
 flightMatrix_i0 <- (flightsDic$matrix + flightsJan$matrix)/2
 flightMatrix_i1 <- flightsFeb$matrix
 flightMatrix_i2 <-  flightsMar$matrix
-vflightMatrix_3i <- na.exclude(c(t(flightMatrix_i2),t(flightMatrix_i1),t(flightMatrix_i0)))
-nvflightMatrix_3i <- vflightMatrix_3i/mean(vflightMatrix_3i)
-write(nvflightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
+v_flightMatrix_3i <- na.exclude(c(t(flightMatrix_i2),t(flightMatrix_i1),t(flightMatrix_i0)))
+slv_flightMatrix_3i <- scale(log(v_flightMatrix_3i + 1))
+write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
+
 
 # # Summarise into two periods (Sept - Jan and Feb - March) and calculate # passengers/day
 # first <- Reduce("+", mms[1:5])/(5*30)

### 2020-11-27 reports/paper/main.tex
wip results paper

@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.8\textwidth]{201014_europe2_figtraj01.png}
+    \caption{Inferred population size summary statistics for each deme over time.
+The line represents the median population trajectory and the interval is the interquartile range from a subsampled set of inferred population trajectories.}
+    \label{fig:gribbon}
+\end{figure}
+
 Inferred population trajectories that account for all cases in the country.
 It is not an independent estimate since we are using informative priors for the sampling rates (upperbound computed as sequences/reported cases) and the subsampling scheme also includes the number of cases in the country (but that is a weaker link I think.\\
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.9\textwidth]{201014_europe2_figtraj02.png}
+    \caption{Inferred population size trajectories over time.
+A subsampled set of 200 inferred trajectories is plotted.
+In each subplot, the trajectories (solid lines) are compared with the ECDC total case count data (dashed line).}
+    \label{fig:trajs}
+\end{figure}
 
@@
 Figures \ref{fig:first}, \ref{fig:first_src} and \ref{fig:first_dest}.\\
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj03.png}
+    \caption{First introductions event time and source in each deme.
+\textbf{A} Probability density of the time of the first introduction for each deme from the subsampled set of trajectories.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
+For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
+\textbf{B} Probability density of the source of the first introduction event in each deme for the set of subsampled trajectories.}
+    \label{fig:first}
+\end{figure}
 
@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj04.png}
+    \caption{Ordered set of countries by its first introduction, i.e.
+first case in the country.
+The color corresponds to the first country with a case, and the size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first introductions.}
+    \label{fig:first_dest}
+\end{figure}
+
 In Figure \ref{fig:first_src}, we are looking at the ordered set of introduction intoo european countries.
@@
 More trajectories where the first case was to Germany, no clear order after the first country.
-Spain last one.\\
+Spain and Other European last one.\\
 \todo{Explain this better}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj05.png}
+    \caption{Ordered set of countries by its first migration, i.e.
+spreading event out of the country.
+The color corresponds to the last country with a migration to other countries.
+The size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first migrations.}
+    \label{fig:first_src}
+\end{figure}
 
@@
 Figure \ref{fig:events}\\.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201014_europe2_figtraj06.png}
+    \caption{Mean inferred cumulative number of events (births and migrations) over time. A subsampled set of 200 inferred trajectories is plotted.}
+    \label{fig:events}
+\end{figure}
 
@@
 Figures \ref{fig:migs_srcdest} and \ref{fig:migs}.\\
-
-Similar information in both plots, but in the chord plots instead of a dayly evolution the time period is split on three (as in the GLM analysis).
-Chord plots are nicer and easier to understand I think, but barplots shows the great detail of the results of the model.
-Another advantage of the chrod plot is that the ir shows mean absolute values and not only relative values.\\
-
-Hubei-China is the majority source of migrations for all countries till February and then some patterns emerge (we expect more interesting results with the added info in GLM analysis).\\
-
-\todo{Legend in plot 7 and change white text}
-
-\subsubsection*{GLM predictors}
-Which channels were the main sources of transmission across national borders.
-
-\subsubsection*{Epidemiological parameters}
-
-Not sure about this section. But maybe it make sense for completeness to include, briefly, the values of estimate R0, migration and sampling rates.
-
-\section*{Discussion}
-
-\section*{Conclusion}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.8\textwidth]{201014_europe2_figtraj01.png}
-    \caption{Inferred population size summary statistics for each deme over time.
-The line represents the median population trajectory and the interval is the interquartile range from a subsampled set of inferred population trajectories.}
-    \label{fig:gribbon}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.9\textwidth]{201014_europe2_figtraj02.png}
-    \caption{Inferred population size trajectories over time.
-A subsampled set of 200 inferred trajectories is plotted.
-In each subplot, the trajectories (solid lines) are compared with the ECDC total case count data (dashed line).}
-    \label{fig:trajs}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj03.png}
-    \caption{First introductions event time and source in each deme.
-\textbf{A} Probability density of the time of the first introduction for each deme from the subsampled set of trajectories.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
-For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
-\textbf{B} Probability density of the source of the first introduction event in each deme for the set of subsampled trajectories.}
-    \label{fig:first}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj04.png}
-    \caption{}
-    \label{fig:first_dest}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj05.png}
-    \caption{}
-    \label{fig:first_src}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj06.png}
-    \caption{Mean inferred cumulative number of events (births and migrations) over time. A subsampled set of 200 inferred trajectories is plotted.}
-    \label{fig:events}
-\end{figure}
 
@@
 
+Similar information in both plots, but in the chord plots instead of a dayly evolution the time period is split on three (as in the GLM analysis).
+Chord plots are nicer and easier to understand I think, but barplots shows the great detail of the results of the model.
+Another advantage of the chrod plot is that the ir shows mean absolute values and not only relative values.\\
+
+Hubei-China is the majority source of migrations for all countries till February and then some patterns emerge (we expect more interesting results with the added info in GLM analysis).\\
+
+\todo{Legend in plot 7 and change white text}
+
+\subsubsection*{GLM predictors}
+Which channels were the main sources of transmission across national borders.
+
+\subsubsection*{Epidemiological parameters}
+
+Not sure about this section. But maybe it make sense for completeness to include, briefly, the values of estimate R0, migration and sampling rates.
+
+\section*{Discussion}
+
+\section*{Conclusion}
+
 \end{document}

### 2020-11-30 workflow/scripts/analyze_trajectories.R
analyze traj working workflow #19

@@ -20,6 +20,7 @@ library(hrbrthemes)
 library(circlize)
 library(chorddiag) 
 library(ggforce)
+library(ComplexHeatmap)
 
 # Source files -----------------------------------------------------------------
 source("./scripts/trajProcessing.R")
@@ -57,7 +58,9 @@ demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
   filter_all(any_vars(!is.na(.))) %>%
   mutate(deme = str_to_title(deme)) %>%
   mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme))
-  
+
+china_deme <- demes$deme[demes$region == "Asia"]
+
 cat("\nDeme configuration:\n")
 knitr::kable(demes)
 metadata <- read_tsv(args$metadata)
@@ -113,8 +116,8 @@ gt_summary <- gt %>%
   group_by(date, type) %>%
   summarize(Imean = mean(N),
             Imedian = median(N), 
-            Ilow    = quantile(N, 0.25), 
-            Ihigh   = quantile(N, 0.75),
+            Ilow    = quantile(N, 0.025), 
+            Ihigh   = quantile(N, 0.975),
             .groups = "drop_last") %>%
   ungroup() 
 
@@ -127,21 +130,21 @@ ge <- gridEventsByAge(events, ages_1day) %>%
 cat("\nAnalyzing")
 
 # 1. Trajectories - Total case counts
-# 1.1 Median and IQR of Total cases on the full time period. Reporting rate median 
-#    and IQR (cases ecdc/inferred cases) wrt. ECDC reported cases
+# 1.1 Median and 95% CCI of Total cases on the full time period. Reporting rate median 
+#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases
 
 reported <- gt_summary %>%
   filter(date == MRS) %>%
   left_join(case_data %>% rename(type = deme), by = c("type", "date")) %>%
   mutate(reportrate = cumcases/Imedian,
          lreportrate = cumcases/Ilow,
-         hreportrate = cumcases/Ihigh) %>%
+         hreportrate = cumcases/Ihigh) 
 
 r_reported <- reported %>%  
   mutate(Country = type,
          `Cases ECDC` = cumcases,
-         `Median Cases (IQR)` = paste0(round(Imedian), " (", round(Ilow), "-", round(Ihigh), ")"),
[…]

### 2020-11-30 workflow/scripts/analyze_trajectories.R
small fixes #19

@@ -342,7 +342,8 @@ gribbon <- ggplot(gt_summary) +
   scale_fill_manual(name = "", values = dcolors) +
   scale_color_manual(name = "", values = dcolors) +
   theme(legend.position = c(0.2, 0.7),
-        legend.box="horizontal")
+        legend.box="horizontal",
+        axis.title.x=element_blank())
 
 ggexport(gribbon,
          filename = paste0(args$output_figure, "01.png"),
@@ -398,12 +399,14 @@ reported5days <- gt_summary %>%
   ggplot() +
   geom_histogram(aes(x = date, y = reportrate, fill = type), stat="identity") +
   facet_wrap(~type, scales = "free") +
+  ylab("Reporting rate") +
   scale_fill_manual(values = dcolors) +
   scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %d") +
-  theme(legend.position = "none")
+  theme(legend.position = "none", 
+        axis.title.x=element_blank())
 
 ggexport(reported5days, filename = paste0(args$output_figure, "03.png"),
-         width = 2000, height = 1000, res = 300)
+         width = 1500, height = 1000, res = 300)
 
 # 2. Events
 # 2.1 First introduction distribution densuty by deme and time of first reported cases to ECDC
@@ -575,7 +578,7 @@ migbirths_line <- ge  %>%
 
 ggexport(migbirths_line,
          filename = paste0(args$output_figure, "07.png"),
-         width = 1300, height = 1000, res = 300)
+         width = 1500, height = 1000, res = 300)
 
 
 # 2.6 Proportion barplot source of migrations single deme

### 2020-11-30 workflow/scripts/utils.R

@@ -18,8 +18,8 @@ set_plotopts <- function() {
                strip.background = element_rect(fill = "grey80", colour = "grey80"),
                strip.text = element_text(color = "grey30"),
                #legend.title = element_text(size = 9, face = "bold"),
-               #axis.text.x = element_text(angle = 270),
-               #axis.text.y = element_text(),
+               axis.text.x = element_text(size = 8),
+               axis.text.y = element_text(size = 8),
                axis.title.y = element_text(size = 9, face = "bold", margin = margin(r = 10)),
                axis.title.x = element_text(size = 9, face = "bold", margin = margin(t = 10)))
 }

### 2020-12-02 workflow/scripts/analyze_trajectories.R
small fixes #19

@@ -147,7 +147,7 @@ r_reported <- reported %>%
          `Median Reporting Rate (95% CCI)` = paste0(round(reportrate, 2), " (", round(hreportrate, 2), "-", round(lreportrate, 2), ")")) %>%
   select(-one_of(colnames(reported)))
 
-write.csv(r_reported, file = paste0(args$output_table, "01.csv"))
+write_tsv(r_reported, file = paste0(args$output_table, "01.tsv"))
 
 # 2. Events 
 # 2.1. Time interval of first case (introduction) and first reported case to ECDC.
@@ -313,7 +313,7 @@ r_firsts <- firsts %>%
   left_join(r_firstmig, by = c("Country", "First Reported Case ECDC")) %>%
   select(1,4,5,6,8,2,3,7,9)
 
-write.csv(r_firsts, file = paste0(args$output_table, "02.csv"))
+write_tsv(r_firsts, file = paste0(args$output_table, "02.tsv"))
 
 
 # Plotting ---------------------------------------------------------------------

### 2020-12-02 reports/paper/main.tex
wip paper

@@
 
+\usepackage{pgfplotstable}
+\pgfplotsset{compat=1.16}
+\def\getcell#1#2#3{
+\pgfplotstablegetelem{#1}{[index]#2}\of{#3}\pgfplotsretval%
+}
+
+% Tables 
+\pgfplotstableset{col sep=tab, header=has colnames, string type, text special chars={\%}} 
+\pgfplotstableread{../../results/201030_europe3_table01.tsv}{\casestable}
+\pgfplotstableread{../../results/201030_europe3_table02.tsv}{\firsttable}
+
 \title{A comprehensive study of the early phylodynamics of SARS-CoV-2 in Europe}
@@
 We apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences.
-The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.}
+The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.
+
+The early dynamics of the epidemic in Europe surprised the whole world/ From isolated cases detexcted in some, ina  few weeks we faced an unprecedent epidemic with case numbers and death counts stressing well stalished health infraestructures.
+Understanding what happened during those first month could prepare us to be less surpred in similar future events.
+Complex dynamics, typical epi date, case identification and travel history, very likely not all cases reported.
+Genetic sequences are a valuable source of information than together with popualtion dynamic model could inform us about the firrst steps of the coronavirus in Europe.
+We have integrated sequence data, flisht data and bayesian inference priors?
+to infer population trajectories - case counts and migration dynamics between the most relevant countrries at the start of the European pandemic.
+}
 
@@
 
-\section*{Methods}
+\item Importance of understanding the spread of the virus to prevent future outbreaks
+
+\item About phylodynamics/phylogeographics? Use of genetic sequences as a source of information combined with other sources of information as travel data
+
+\item About what we know of the introduction of sars-cov-2 in Europe and early dynamics till 8 March
+
+\item About what we know of the case counts in Europe? Can sequences help?
+
+\item Introduce/formulate the questions: case counts, first introductions, ?migration vs within region transmission?, migration patterns, ?border closures.
 
@@
 
-Figures \ref{fig:gribbon} and \ref{fig:trajs}.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.8\textwidth]{201014_europe2_figtraj01.png}
-    \caption{Inferred population size summary statistics for each deme over time.
-The line represents the median population trajectory and the interval is the interquartile range from a subsampled set of inferred population trajectories.}
-    \label{fig:gribbon}
-\end{figure}
+Figures \ref{fig:gribbon}, \ref{fig:trajs} and \ref{fig:reported}.\\
 
@@
 It is not an independent estimate since we are using informative priors for the sampling rates (upperbound computed as sequences/reported cases) and the subsampling scheme also includes the number of cases in the country (but that is a weaker link I think.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.9\textwidth]{201014_europe2_figtraj02.png}
-    \caption{Inferred population size trajectories over time.
-A subsampled set of 200 inferred trajectories is plotted.
-In each subplot, the trajectories (solid lines) are compared with the ECDC total case count data (dashed line).}
-    \label{fig:trajs}
-\end{figure}
 
@@
 Figures \ref{fig:first}, \ref{fig:first_src} and \ref{fig:first_dest}.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj03.png}
-    \caption{First introductions event time and source in each deme.
-\textbf{A} Probability density of the time of the first introduction for each deme from the subsampled set of trajectories.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
-For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
-\textbf{B} Probability density of the source of the first introduction event in each deme for the set of subsampled trajectories.}
-    \label{fig:first}
-\end{figure}
 
@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj04.png}
-    \caption{Ordered set of countries by its first introduction, i.e.
-first case in the country.
-The color corresponds to the first country with a case, and the size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first introductions.}
-    \label{fig:first_dest}
-\end{figure}
-
 In Figure \ref{fig:first_src}, we are looking at the ordered set of introduction intoo european countries.
@@
 \todo{Explain this better}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj05.png}
-    \caption{Ordered set of countries by its first migration, i.e.
-spreading event out of the country.
-The color corresponds to the last country with a migration to other countries.
-The size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first migrations.}
-    \label{fig:first_src}
-\end{figure}
 
@@
 Figure \ref{fig:events}\\.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj06.png}
-    \caption{Mean inferred cumulative number of events (births and migrations) over time. A subsampled set of 200 inferred trajectories is plotted.}
-    \label{fig:events}
-\end{figure}
 
@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj07.png}
-    \caption{}
-    \label{fig:migs_srcdest}
-\end{figure}
-
-\begin{figure}[!tbp]
-  \centering
-
-  \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201014_europe2_figtraj08a.png}
-  \label{fig:migs1}
-  \end{minipage}
-  \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201014_europe2_figtraj08b.png}
-  \label{fig:migs2}
-  \end{minipage}
-  \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201014_europe2_figtraj08c.png}
-  \label{fig:migs3}
-  \end{minipage}
-  \caption{Migration flux among demes over the three periods defined in the analysis.}
-  \label{fig:migs}
-\end{figure}
-
 Similar information in both plots, but in the chord plots instead of a dayly evolution the time period is split on three (as in the GLM analysis).
@@
 
+% Figures
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=0.8\textwidth]{201030_europe3_figtraj01.png}
+    \caption{Inferred population size summary statistics for each deme over time.
+The line represents the median population trajectory and the interval is the 95\% center credible interval from a subsampled set of inferred population trajectories.}{}
+    \label{fig:gribbon}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj02.png}
+    \caption{Inferred population size trajectories over time.
+A subsampled set of 200 inferred trajectories is plotted.
+In each subplot, the trajectories (solid lines) are compared with the ECDC total case count data (dashed line).}
+    \label{fig:trajs}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj03.png}
+    \caption{}
+    \label{fig:reported}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj04.png}
+    \caption{First introductions event time and source in each deme.
+\textbf{A} Probability density of the time of the first introduction for each deme from the subsampled set of trajectories.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
+For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
+\textbf{B} Probability density of the source of the first introduction event in each deme for the set of subsampled trajectories.}
+    \label{fig:first}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj06.png}
+    \caption{Ordered set of countries by its first migration, i.e.
+spreading event out of the country.
+The color corresponds to the last country with a migration to other countries.
+The size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first migrations.}
+    \label{fig:first_src}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj05.png}
+    \caption{Ordered set of countries by its first introduction, i.e.
+first case in the country.
+The color corresponds to the first country with a case, and the size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first introductions.}
+    \label{fig:first_dest}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj08.png}
+    \caption{}
+    \label{fig:migs_srcdest}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj07.png}
+    \caption{Mean inferred cumulative number of events (births and migrations) over time. A subsampled set of 200 inferred trajectories is plotted.}
+    \label{fig:events}
+\end{figure}
+
+\begin{figure}[!tbp]
+  \centering
+
+  \begin{minipage}[t]{0.4\textwidth}
+  \includegraphics[width=\textwidth]{201030_europe3_figtraj09a.png}
+  \label{fig:migs1}
+  \end{minipage}
+  \begin{minipage}[t]{0.4\textwidth}
+  \includegraphics[width=\textwidth]{201030_europe3_figtraj09b.png}
+  \label{fig:migs2}
+  \end{minipage}
+  \begin{minipage}[t]{0.4\textwidth}
+  \includegraphics[width=\textwidth]{201030_europe3_figtraj09c.png}
+  \label{fig:migs3}
+  \end{minipage}
+  \caption{Migration flux among demes over the three periods defined in the analysis.}
+  \label{fig:migs}
+\end{figure}
+
 \section*{Discussion}
+
+\item Case counts discussion and reporting rates, second wave in Europe. Other estimations in other studies.
+
+\item First introduction other studies.
+
+\item Migration patterns other studies.
+
+\item Discuss assumptions of the model, caveats and possible improvements.
 
@@
 
+\section*{Material and Methods}
+
+\item Dataset and subsampling. Alignment.
+
+\item Model BDMM-Prime, stochastic mapping. Priors. Bayesian phylogeographic inference. Phylodynamics.
+
+\item GLM model. External data.
+
+\item ?Same analysis but with travel data from last year, can we observe a consequence of Hubei lockdown and reduced travel?
+
+\item Trajectories analysis. Subsampling and summarizing (median, 95\% interval)
+
+\item Snakemake workflow.
+
 \end{document}

### 2020-12-03 reports/paper/main.tex
wip #11

@@
-\documentclass[12pt,]{article}
+\documentclass[12pt]{article}
 \usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
-\usepackage{fullpage}
+\renewcommand*\familydefault{\sfdefault}
+
+\usepackage{anysize}
+\marginsize{2.5cm}{2.5cm}{1cm}{1.5cm}
+\setlength{\marginparwidth}{2cm}
 
@@
 
+\usepackage{xcolor}
+
+\usepackage[font=small,labelfont=bf]{caption}
+
+\def\labelitemi{--}
+
+% Line numbers
 \usepackage[left]{lineno}
-\linenumbers
+%\linenumbers
 
-\usepackage{todonotes}
+\usepackage[color=red!60]{todonotes}
 
@@
 
-\abstract{This project focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe.
+\begin{abstract}
+\todo[inline]{TODO abstract}
+This project focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe.
 We apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences.
@@
 
-The early dynamics of the epidemic in Europe surprised the whole world/ From isolated cases detexcted in some, ina  few weeks we faced an unprecedent epidemic with case numbers and death counts stressing well stalished health infraestructures.
+% The early dynamics of the epidemic in Europe surprised the whole world/ From isolated cases detexcted in some, ina  few weeks we faced an unprecedent epidemic with case numbers and death counts stressing well stalished health infraestructures.
 Understanding what happened during those first month could prepare us to be less surpred in similar future events.
@@
 to infer population trajectories - case counts and migration dynamics between the most relevant countrries at the start of the European pandemic.
-}
+\end{abstract}
 
 \section*{Introduction}
+\todo[inline]{TODO introduction}
 
+\begin{itemize}
 \item Importance of understanding the spread of the virus to prevent future outbreaks
@@
 \item Introduce/formulate the questions: case counts, first introductions, ?migration vs within region transmission?, migration patterns, ?border closures.
+\end{itemize}
 
 \section*{Results}
+% \todo{Modify tables in R to have one value per cell, nr rate and times}
 
-\subsubsection*{Total case counts}
+\todo[inline]{The figures are from analysis Europe3: subsampled datasets according to cases/day and constant migration rate (no GLM), same xml specifications than Sarah's analysis.}
 
-Figures \ref{fig:gribbon}, \ref{fig:trajs} and \ref{fig:reported}.\\
+\subsubsection*{Burden of SARS-CoV-2 infections in Europe}
 
-Inferred population trajectories that account for all cases in the country.
-It is not an independent estimate since we are using informative priors for the sampling rates (upperbound computed as sequences/reported cases) and the subsampling scheme also includes the number of cases in the country (but that is a weaker link I think.\\
+The inferred epidemic trajectories contain the information about the total number of cases until 8th of March.
+For the European countries, we obtain an inferred number of total cases above the number of confirmed cases to ECDC, consistent with known limited test availability of the first wave and  previous studies results \cite{Li2020} \cite{Wu2020}.
+These cases counts correspond to \todo{Include values} x-y times higher than the number of reported cases.
+Italy is the country with the highest infered number of cases x, followed by Spain, France and Germany.
+The infered number of cases for China is below the reported number of cases.
+\todo{limitations of the model, sequence information, partial outbreak dynamics}\\
 
-Comparison with ECDC case counts: delay in notifications and underreporting, for some of the  countries at the end of the periord the reported cases are very closed to the trajectories, e.g.
-Italy, Germany.
-Hubei-China huge difference in tendency and absolute numbers.\\ \todo{Working on these with new analyses}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.8\textwidth]{201030_europe3_figtraj01.png}
+    \caption{Inferred population size summary statistics for each deme over time.
+The line represents the median population trajectory and the interval is the 95\%  credible interval in log scale from a random subsampled set of inferred epidemic trajectories.}
+    \label{fig:gribbon}
+\end{figure}
 
-Measure the (mean?) difference between inferred and reported case counts and compare among countries. By day or in general? Absolute or relative to the number of cases that day?
+\begin{figure}[p]
+    \centering
+    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj02.png}
+    \caption{Inferred epidemic trajectories over time.
+A random subsampled set of 500 trajectories is plotted.
+In each subplot, the trajectories (solid lines) are compared with the ECDC cumulative case count data (dashed line) in log scale.}
+    \label{fig:trajs}
+\end{figure}
 
-\subsubsection*{First introductions}
-Figures \ref{fig:first}, \ref{fig:first_src} and \ref{fig:first_dest}.\\
+In Figure \ref{fig:gribbon} and/or \ref{fig:trajs} we compare the total number of inferred cases by day to the total cumulative number of cases that have been reported to ECDC that same day.
+Our inferred case counts follow a exponential growth earlier in time than the reported curve and with higher number of cases, being the difference bigger for later times in the epidemic.\\
 
-We want to answer the question of when and where the first case in Europe occured.
-No clear, Germany?
-Figure \ref{fig:first} A.
-We can measure the time between first case and first reported case and compare between countries.
-\\
-\todo{ticks for every month}
-And for each country where this first case came from, more support for Hubei introduction in all countries Figure \ref{fig:first} B.\\
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj03.png}
+    \caption{5-days reporting rate, calculated as the cumulative number of ECDC reported cases by the median number of cumulative inferred cases in intervals of 5 days until 8th of March. }
+    \label{fig:reported}
+\end{figure}
 
-Measure expected day of introduction.
+We can think of a reporting rate as the number of reported cases relative to the total number of cases inferred by the model.
+This reporting rate decreases with time for all European countries, except for Italy that increases again around March, Figure \ref{fig:reported}.
 
-In Figure \ref{fig:first_src}, we are looking at the ordered set of introduction intoo european countries.
-For each trajectory, we have the fist introduction into each country happening in a determined order, i.e.
-the first european country with a covid case, the second country with a covid case...
-So we are vissualizing the number of trajectories that share the same order of first cases.
-More trajectories where the first case was to Germany, no clear order after the first country.
-Spain and Other European last one.\\
-\todo{Explain this better}
+\subsubsection*{First introductions into European countries}
 
-In Figure \ref{fig:first_dest} we are looking at the ordered set of countries spreading covid.
-In the first position we have Hubei for all trajectories since it is the origin of the epidemic.
-In second position we will have the first European country from where a covid case was imported to other country.
-No clear dynamics except that moore support for Spain being the last country having a first exported covid case.
-Germany first one.\\
+\begin{figure}[p]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj04.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:first}
+    \todo[inline]{ticks for every month}
+\end{figure}
+
+We would like to answer the question of when and where the first case in Europe occured.
+We can look to the model inferred epidemic trajectories and analyze the time distribution of the first case in Europe and in each deme.
+China is defined as the origin of the epidemic in the analysis and this is reflected in Figure \ref{fig:first} A.
+The occurrence of the origin of the epidemic, i.e.
+first case in China, is almost one month before the epidemic starts in any European country, ranging \todo{Include values}  x-y.
+The inferred date of the first case in Europe is inferred to be between a 95\% interval x-y.
+Among the European countries, France and Germany have the earlier time of introduction, followed by Italy, Other European deme and Spain.
+We can compare this inferred dates of introduction with the day each country reported its first case to ECDC.
+In all cases, the reported day was later than the median of the inferred distribution, but it is inside the 95\% interval.\\
+
+The first case in Europe for each trajectoy corresponds to a migration from China to any European country.
+We can analyze the destination of this first migration into Europe as is shown in Figure \ref{fig:first} B.
+According to our model, the probability that this first European case was in Spain is very low.
+The destination of the first case amonth the rest of European demes is equally probable and there is no clear difference in the time distribution for each destination country, with a maximum probability around x.\\
+\todo{explain this better, compute the probability of the first case in Europe destination}
+
+The first case in each European country could have been imported from China or from other European country with an ongoing epidemic that started earlier.
+However, in our analysis we obtain a much higher support for China being the most probable source of the first case for all European demes, Figure \ref{fig:first} C.\\
+
+Even if the first case in each European country came from China, the timing of the introductions in the European countries relative to each other probablu expanded across several days, defining and order of countries with the time of their first case.
+
+We can analyze if this order of countries, defined by the time in which the first case occured, is shared among the majority of the inferred population trajectories, Figure \ref{fig:first_imig}.
+We obtain that the first European country with a SARS-CoV-2 case was more likely France or Germany, followed by Italy and Other European deme.
+While Spain was more likely the last European country with a case among the ones included in the analysis.\\
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj05.png}
+    \caption{Order of countries by the time of its first introduction, i.e.
+first case in the country.
+Each row is the order for one of the subsampled epidemic trajectories and each column represents the position relative to the other countries first introduction, e.g.
+in first position for all epidemic trajectories is China since it is the origin of the epidemic.}
+    \label{fig:first_imig}
+\end{figure}
+
+Along the same lines, we could ask if this order of countries is mantained when instead of looking at the first case in the country we look into at first case exported (migration) from that country to other European country.
+In Figure \ref{fig:first_omig} we observe a similar pattern to the first cases order, with Germany, France and Italy being the countries in the first positions in more than half of the inferred trajectories and never in the last position, and Spain as the country in last position in almost every epidemic trajectory.\\
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj06.png}
+    \caption{Order of countries by the time of its first migration out of the country, i.e.
+first exported case to other country.
+Each row is the order for one of the subsampled epidemic trajectories and each column represents the position relative to the other countries first migration, e.g.
+in first position for all epidemic trajectories is China since it was the first country with exported cases of SARS-CoV-2 to other regions.}
+    \label{fig:first_omig}
+    \todo[inline]{I don't like much these "order" plots but maybe they could be useful to detect interesting patterns?}
+\end{figure}
 
@@
 
-Figure \ref{fig:events}\\.
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj07.png}
+    \caption{Median and 95\% credible interval for the cumulative number of events (within-region transmissions and migrations to the country (incoming) and from the country (outgoing) over time.}
+    \label{fig:events}
+    \todo[inline]{change legend outgoing migration}
+\end{figure}
 
-This question is addresed on the original paper from Sarah.  Maybe could be interesting to look at this perspective since it is important for border closure measures.
+\todo[inline]{TODO}
 
-\todo{Think about results}
- Not sure if we can have any meaningful results insights on the topic.
-Migrations started very early in our analysis to all european countries.
-Withing regions transmision account for most of the cases in the countries from late january onwards (when first cases were being reported in Europe).\\
+From the epidemic trajectories, we can extract the information about how many cases are within-region transmission and how many are migrations from other countries.
+The cumulative number of transmission events and migrations, represented in Figure \ref{fig:events} increases exponentially over time.
+An incomming migration into every European deme happens always before within-region transmission, seeding the epidemic.
+Within-region transmision accounts for most of the cases in the countries from late january onwards (when first cases were being reported in Europe).
 
- Measure the time between the first outcoming mirgration from the country  (specially interesting for Hubei) and first reported case.
-This could be interesting to say if a extreme measure closing borders with the first case could be effective to impede transmission to other countries: percentage of trajectories whre transmission to Europe would have been avoided.
-For other countries we can look at how many migrations events could have been avoided (and how many not) if the country closed borders after first reported case.\\
+\todo{get the proportion of within-region transmission and incoming migrations?}
 
- Measure the time between first birth/migration and first reported case and compare between countries.
-How well did the countries detecting the first cases, there were already within region transmision?\\
+The time between the first case in the country, i.e.
+first incoming migration event and the first case from within-region transmission is of x  (y-z) with similar values for all demes?.
+The time from the first incoming migration to the first outgoing migration is longer with a median of x (y-z).
+(This could be interesting to say if we should focus or not the screening and testing capacities to detect incoming migrations or if when we have evidence of cases in the population we shoud follow a more general strategy to find cases in the population according to the model.
+Is it different for each country?)\\
 
- Measure the time between the first migration into a country and the first birth in the country.
-Or the first outcoming mirgration from the country and first birth.
-This could be interesting to say if we should focus or not the screening and testing capacities to detect incoming migrations or a more general strategy to find cases in the population.
-Is it different for each country?\\
+(How well did the countries detecting the first cases, there were already within region transmision?) We compare the date of the first reported rate of each country with the date in which within-region transmission for that country started according to the model.
+We see some differences among countries, France and Germany had in all inferred epidemic trajectories ongoing within-region transmission when first cases werre reported, while x\% of epidemic trajectories did not have had a within region transmission case when the first case in Spain was reported.\\
+
+We can also compare the date of the first reported case with the date of the first outgoing migration from the country.
+(This could be interesting to say if a extreme measure closing borders with the first case could be effective to impede transmission to other countries: percentage of trajectories whre transmission to Europe would have been avoided.
+For other countries we can look at how many migrations events could have been avoided (and how many not) if the country closed borders after first reported case according to the model.
+Not realistic measure, extreme case.)\\
 
@@
 Figures \ref{fig:migs_srcdest} and \ref{fig:migs}.\\
+\todo[inline]{TODO}
 
-Similar information in both plots, but in the chord plots instead of a dayly evolution the time period is split on three (as in the GLM analysis).
+Similar information in both plots, but in the chord plots instead of a daily evolution the time period is split on three (as in the GLM analysis).
 Chord plots are nicer and easier to understand I think, but barplots shows the great detail of the results of the model.
-Another advantage of the chrod plot is that the ir shows mean absolute values and not only relative values.\\
+Another advantage of the chrod plot is that the it shows mean absolute values and not only relative values.\\
 
 Hubei-China is the majority source of migrations for all countries till February and then some patterns emerge (we expect more interesting results with the added info in GLM analysis).\\
-
-\todo{Legend in plot 7 and change white text}
-
-\subsubsection*{GLM predictors}
-Which channels were the main sources of transmission across national borders.
-
-\subsubsection*{Epidemiological parameters}
-
-Not sure about this section. But maybe it make sense for completeness to include, briefly, the values of estimate R0, migration and sampling rates.
-
-% Figures
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=0.8\textwidth]{201030_europe3_figtraj01.png}
-    \caption{Inferred population size summary statistics for each deme over time.
-The line represents the median population trajectory and the interval is the 95\% center credible interval from a subsampled set of inferred population trajectories.}{}
-    \label{fig:gribbon}
-\end{figure}
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj02.png}
-    \caption{Inferred population size trajectories over time.
-A subsampled set of 200 inferred trajectories is plotted.
-In each subplot, the trajectories (solid lines) are compared with the ECDC total case count data (dashed line).}
-    \label{fig:trajs}
-\end{figure}
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj03.png}
-    \caption{}
-    \label{fig:reported}
-\end{figure}
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj04.png}
-    \caption{First introductions event time and source in each deme.
-\textbf{A} Probability density of the time of the first introduction for each deme from the subsampled set of trajectories.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
-For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
-\textbf{B} Probability density of the source of the first introduction event in each deme for the set of subsampled trajectories.}
-    \label{fig:first}
-\end{figure}
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj06.png}
-    \caption{Ordered set of countries by its first migration, i.e.
-spreading event out of the country.
-The color corresponds to the last country with a migration to other countries.
-The size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first migrations.}
-    \label{fig:first_src}
-\end{figure}
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj05.png}
-    \caption{Ordered set of countries by its first introduction, i.e.
-first case in the country.
-The color corresponds to the first country with a case, and the size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first introductions.}
-    \label{fig:first_dest}
-\end{figure}
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj08.png}
-    \caption{}
-    \label{fig:migs_srcdest}
-\end{figure}
-
-\begin{figure}[]
-    \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj07.png}
-    \caption{Mean inferred cumulative number of events (births and migrations) over time. A subsampled set of 200 inferred trajectories is plotted.}
-    \label{fig:events}
-\end{figure}
 
@@
 
+\begin{figure}[]
+    \centering
+    \includegraphics[width=\textwidth]{201030_europe3_figtraj08.png}
+    \caption{}
+    \label{fig:migs_srcdest}
+\end{figure}
+
+\subsubsection*{GLM predictors}
+\todo[inline]{TODO}
+
+Which channels were the main sources of transmission across national borders.
+
+\subsubsection*{Epidemiological parameters}
+\todo[inline]{TODO}
+
+Not sure if it is necessary. But maybe include, briefly, the values of estimate R0, migration and sampling rates?
+
 \section*{Discussion}
+\todo[inline]{TODO discussion}
 
+\begin{itemize}
 \item Case counts discussion and reporting rates, second wave in Europe. Other estimations in other studies.
+
+% Inferred population trajectories that account for all cases in the country.
+It is not an independent estimate since we are using informative priors for the sampling rates (upperbound computed as sequences/reported cases) and the subsampling scheme also includes the number of cases in the country (but that is a weaker link I think.\\
 
@@
 \item Discuss assumptions of the model, caveats and possible improvements.
+\end{itemize}
 
 \section*{Conclusion}
+\todo[inline]{TODO conclusion}
 
 \section*{Material and Methods}
-
+\todo[inline]{TODO methods}
+\begin{itemize}
 \item Dataset and subsampling. Alignment.
@@
 \item Snakemake workflow.
+\end{itemize}
+
+\bibliographystyle{unsrt}
+\bibliography{references-mt}
+
+\listoftodos
 

### 2020-12-03 workflow/scripts/analyze_trajectories.R
new reported plot #19

@@ -147,7 +147,7 @@ r_reported <- reported %>%
          `Median Reporting Rate (95% CCI)` = paste0(round(reportrate, 2), " (", round(hreportrate, 2), "-", round(lreportrate, 2), ")")) %>%
   select(-one_of(colnames(reported)))
 
-write_tsv(r_reported, file = paste0(args$output_table, "01.tsv"))
+write_tsv(r_reported, paste0(args$output_table, "01.tsv"))
 
 # 2. Events 
 # 2.1. Time interval of first case (introduction) and first reported case to ECDC.
@@ -313,7 +313,7 @@ r_firsts <- firsts %>%
   left_join(r_firstmig, by = c("Country", "First Reported Case ECDC")) %>%
   select(1,4,5,6,8,2,3,7,9)
 
-write_tsv(r_firsts, file = paste0(args$output_table, "02.tsv"))
+write_tsv(r_firsts, paste0(args$output_table, "02.tsv"))
 
 
 # Plotting ---------------------------------------------------------------------
@@ -393,16 +393,22 @@ ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1
 
 # 1.3. Reporting rate 5 days cumulative cases
 reported5days <- gt_summary %>%
+  filter(type != china_deme) %>%
   left_join(case_data %>% rename(type = deme), by = c("type", "date"))  %>%
   replace_na(list(cumcases = 0)) %>% 
-  mutate(reportrate = cumcases/Imedian) %>%
+  mutate(reportrate = ifelse(is.infinite(cumcases/Imedian), 0, cumcases/Imedian),
+         rlow = ifelse(is.infinite(cumcases/Ihigh), 0, cumcases/Ihigh),
+         rhigh = ifelse(is.infinite(cumcases/Ilow), 0, cumcases/Ilow)) %>%
+  filter(!is.na(reportrate)) %>%
   ggplot() +
-  geom_histogram(aes(x = date, y = reportrate, fill = type), stat="identity") +
-  facet_wrap(~type, scales = "free") +
+  geom_line(aes(x = date, y = reportrate, color = type), stat="identity") +
+  geom_point(aes(x = date, y = reportrate, color = type), stat="identity") +
+  #geom_ribbon(aes(date, ymin = rlow, ymax = rhigh, fill = type), alpha = 0.5) +
   ylab("Reporting rate") +
-  scale_fill_manual(values = dcolors) +
[…]

### 2020-12-04 workflow/scripts/flight_data.R
new flight matrix pop #9

@@ -130,6 +130,39 @@ v_flightMatrix_3i <- na.exclude(c(t(flightMatrix_i2),t(flightMatrix_i1),t(flight
 slv_flightMatrix_3i <- scale(log(v_flightMatrix_3i + 1))
 write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
 
+# 3 epochs migration matrix divided by population
+
+# Find better way to load population data by deme
+demes <- read.csv("/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/workflow/files/demes.csv")
+cases <- get_cases(demes)
+pops <- cases %>%
+  select(deme, population_deme) %>%
+  mutate(GEO = as.character(deme)) %>%
+  distinct
+pops$GEO[pops$deme == "Hubei"] <- "Hubei-China"
+
+# flightMatrix_i0 <- sweep((flightsDic$matrix + flightsJan$matrix)/2, 2, pops$population_deme, `/`)
+# flightMatrix_i1 <- sweep(flightsFeb$matrix, 2, pops$population_deme, `/`)
+# flightMatrix_i2 <-  sweep(flightsMar$matrix, 2, pops$population_deme, `/`)
+nflightMatrix_i0 <- flightMatrix_i0 / pops$population_deme
+nflightMatrix_i1 <- flightMatrix_i1 / pops$population_deme
+nflightMatrix_i2 <- flightMatrix_i2 / pops$population_deme
+vn_flightMatrix_3i <- na.exclude(c(t(nflightMatrix_i2),t(nflightMatrix_i1),t(nflightMatrix_i0)))
+slnv_flightMatrix_3i <- scale(log(vn_flightMatrix_3i + 1))
+write(slnv_flightMatrix_3i, "files/flight_matrix_3i_pop.csv", sep = ",")
+
+# Population source matrix
+pop_source <- (diag(6) - 1) * (-1)
+colnames(pop_source) <- pops$deme
+rownames(pop_source) <- pops$deme
+pop_source <- pop_source * pops$population_deme
+
+# Population destination matrix
+pop_dest <- (diag(6) - 1) * (-1)
+colnames(pop_dest) <- pops$deme
+rownames(pop_dest) <- pops$deme
+pop_dest <- sweep(pop_dest, 2, pops$population_deme, `*`)
+
 
[…]

### 2020-12-04 workflow/scripts/utils.R

@@ -30,12 +30,15 @@ get_cases <- function(demes, from = NA, to = NA) {
   # Get case data information from ECDC for specific demes (countries and regions)
   # Set from and to dates with values from the function call instead of the ones from the demes info
   cat("\nLoading case data counts from ECDC...\n")
+  #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
+  #else case_data <- get_dataECDC()
+  case_data <- get_dataECDC()
   
   if (!is.na(from)) demes$min_date <- from
   if (!is.na(to)) demes$max_date <- to
   
   # Filter case count data according to deme info
-  case_data_list <- mapply(get_cases_deme, demes$deme,
+  case_data_list <- mapply(get_cases_deme, list(case_data), demes$deme,
                            demes$region, demes$country, demes$division,
                            demes$exclude_country, demes$exclude_division,
                            demes$min_date, demes$max_date)
@@ -46,12 +49,9 @@ get_cases <- function(demes, from = NA, to = NA) {
 }
 
 
-get_cases_deme <- function(deme_name, region_name = NA, country_name = NA, division_name = NA, 
+get_cases_deme <- function(case_data, deme_name, region_name = NA, country_name = NA, division_name = NA, 
                       exclude_country = NA, exclude_division = NA,
                       from, to) {
-  #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
-  #else case_data <- get_dataECDC()
-  case_data <- get_dataECDC()
   print(deme_name)
   df_cases <- case_data %>%
     filter(if (is.na(region_name)) TRUE else region %in% region_name,
@@ -61,7 +61,14 @@ get_cases_deme <- function(deme_name, region_name = NA, country_name = NA, divis
            #if (is.na(exclude_division)) TRUE else !division_name %in% exclude_division,
            date >= as.Date(from),
            date <= as.Date(to)) %>%
-    mutate(deme = deme_name)
+    mutate(deme = deme_name) %>%
+    arrange(date) %>%
+    mutate(cumcases = cumsum(cases),
+          cumdeaths = cumsum(deaths),
+          population_deme = sum(unique(population))) %>%
+    group_by(date) %>%
+    mutate(cumcases = max(cumcases),
+           cumdeaths = max(cumdeaths))
   
   return(df_cases)
 }  
[…]

### 2020-12-09 workflow/plotscasesandflights.R
beast v2.6.3 update, lagged 10 days analyze traj

@@ -1,0 +1,34 @@
+library(tidyverse)
+
+source("./scripts/utils.R")
+
+demes <- read.csv("/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/workflow/files/demes.csv")
+cases <- get_cases(demes, to="2020-03-18")
+  
+
+ggplot(cases) +
+  geom_line(aes(x = date, y = cumcases, color = deme)) +
+  ggsci::scale_color_jco()
+  
+ggplot(cases) +
+  geom_line(aes(x = date, y = cumcases/population_deme, color = deme)) +
+  ggsci::scale_color_jco()
+
+pops <- cases %>%
+  select(deme, population_deme) %>%
+  mutate(GEO = as.character(deme)) %>%
+  distinct
+pops$GEO[pops$deme == "Hubei"] <- "Hubei-China"
+
+flightsDic$df %>% filter (!is.na(VALUE)) %>%
+  mutate(svalue = scale(log(VALUE))) %>%
+  ggplot() +
+  geom_point(aes(x = GEO, y = svalue, color = PARTNER)) +
+  ggsci::scale_color_jco()
+
+flightsDic$df %>% filter (!is.na(VALUE)) %>% 
+  left_join(pops) %>%
+  mutate(svalue = scale(log(VALUE/population_deme))) %>%
+  ggplot() +
+  geom_point(aes(x = GEO, y = svalue, color = PARTNER)) +
+  ggsci::scale_color_jco()

### 2020-12-09 workflow/scripts/analyze_trajectories.R

@@ -89,7 +89,8 @@ states <- df$states %>%
   mutate(type = factor(type, 
                        levels = 0:(nrow(demes) - 1), 
                        labels = sort(demes$deme)),
-         date = date(date_decimal(decimal_date(ymd(MRS)) - age)))
+         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
+  mutate(date = date_model + 10)
 
 events <- df$events %>%
   mutate(src = factor(src, 
@@ -99,7 +100,8 @@ events <- df$events %>%
                        as.character(factor(dest, 
                                            levels = 0:(nrow(demes) - 1),
                                            labels = sort(demes$deme)))),
-         date = date(date_decimal(decimal_date(ymd(MRS)) - age)))
+         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
+  mutate(date = date_model + 10)
 
 # Date grids
 max_age <- max(states$age)

### 2020-12-09 workflow/scripts/chains_diagnostic.R

@@ -17,29 +17,26 @@ library(readr)
 parser <- argparse::ArgumentParser()
 parser$add_argument("--input", nargs="+", help="Input log summary file")
 parser$add_argument("--ess", type="integer", help="ESS value cutoff")
+parser$add_argument("--burnin", type="integer", help="Burning percentage")
+parser$add_argument("--length", type="integer", help="Chain length")
 parser$add_argument("--output", type="character", help="Output diagnostic file")
 args <- parser$parse_args()
 
-INPUT <- args$input
 ESS_cutoff <- args$ess
-OUTPUT <- args$output
 
-print(paste("log summaries:", INPUT))
-print(paste("ESS cutoff value", ESS_cutoff))
-print(paste("output diagnostic file:", OUTPUT))
+
 
 # Diagnostics ------------------------------------------------------------------
 diagnostic <- data.frame()
-for (fname in INPUT){
+for (fname in args$input){
   if (file.size(fname) != 0) {
-    tb <- read.table(fname, skip = 1, fill = TRUE, header = TRUE, comment.char = "*", flush = TRUE, stringsAsFactors = FALSE) %>%
-      mutate_at(vars(ESS), as.numeric)
-    l <- last(strsplit(read_lines(fname, n_max = 1), " ")[[1]])
+    tb <- read_table(fname) 
     if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
       cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
       diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
                                                  seed = str_split(fname, pattern ="\\.")[[1]][2],
-                                                 length = l,
+                                                 burnin = args$burnin,
+                                                 length = args$length,
                                                  min_ESS = ESS_cutoff, 
                                                  included = 1))
     } else {
@@ -47,7 +44,8 @@ for (fname in INPUT){
           "\n\nhave ESS value < ", ESS_cutoff, " in log file ", fname, "\n\nLog file not included in analysis.\n")
       diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]), 
                                                  seed = str_split(fname, pattern ="\\.")[[1]][2],
-                                                 length = l,
+                                                 burnin = args$burnin,
[…]

### 2020-12-11 workflow/scripts/predictor_matrices.Rmd
predictor matrices rmd

---
title: "GLM Predictor Matrices"
subtitle: "Analyses Europe6 and Europe7"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dove"
    fonttheme: "structurebold"
    highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, echo = FALSE, results ='hide', message = FALSE}
library(knitr)
library(wpp2019)
library(tidyverse)
library(readr)
source("./scripts/flight_data.R")
```

## Description

Generation of the predictor matrices for the GLM model from different sources 
of information:

- Eurostats, European data and statistics, transport data: <https://ec.europa.eu/eurostat/web/transport/data/database>
- United Nations Population Division, World Population Prospects 2019: R library `wpp2019, data(pop)`
- Distance

```{r}
## Daily flight passengers

demes <- c(China = "CN", France = "FR", Germany = "DE", Italy = "IT", OtherEuropean = "OE", Spain = "ES")

# These datasets are dowload from eurostats 
intraUE_data <- "./data/avia_paincc.tsv"
extraUE_data <- "./data/avia_paexcc.tsv"

fligh_data <- get_flightData(intraUE_data, extraUE_data, countries = demes)

fligh_data_3i <- fligh_data %>%
  filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
  mutate(EPOCH = case_when(
    MONTH == 2 ~ 2,
    MONTH == 3 ~ 3,
    TRUE ~ 1)) %>%
  group_by(GEO, PARTNER, EPOCH) %>%
  mutate(N = n(),
         EVALUE = sum(DVALUE/N)) %>%
  ungroup %>%
  # Transformations
  mutate(LVALUE = log(EVALUE + 1),
         SLVALUE = scale(LVALUE))

write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
```

## Population counts

Population counts from World Population Prospects 2019, United Nations Population Division,  `wpp2019` R library. Values for **China**, **France**, **Germany**, **Italy**, **Other European** and **Spain**.

A predictor matrix for source population and a predictor matrix for destination population are generated, log transformed and standardized. Constant across demes.

\small
```{r, echo=FALSE}
data(pop)
population <- pop %>% filter(name %in% labels(demes)) %>% select(name, "2020") 
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
  arrange(name)
```

```{r, echo=FALSE}
# Population source matrix
pop_source <- matrix(1, 6, 6) * population$`2020`
diag(pop_source) <- NA
vslpop_source <- scale(log(c(pop_source)))
slpop_source <- matrix(vslpop_source, 6, 6)
colnames(slpop_source) <- demes[population$name]
rownames(slpop_source) <- demes[population$name]

# Population destination matrix
pop_dest <- sweep(matrix(1, 6, 6), 2, population$`2020`, `*`)
diag(pop_dest) <- NA
vslpop_dest <- scale(log(c(pop_dest)))
slpop_dest <- matrix(vslpop_dest, 6, 6)
colnames(slpop_dest) <- demes[population$name]
rownames(slpop_dest) <- demes[population$name]

knitr::kable(list(round(slpop_source, 2), round(slpop_dest, 2)), booktabs = TRUE, caption = "Population counts source (left) and destination (right)")
```

\normalsize
Other European population count is computed as the difference between the Europe population and the sum of the populations from France, Germany, Italy and Spain.

## Distance between countries

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

### 2020-12-11 workflow/scripts/flight_data.R

@@ -2,212 +2,234 @@
 
 library(tidyverse)
 library(lubridate)
+source("./scripts/utils.R")
 
 # Returns dayly passengers in a given month 
-get_flightData <- function(intraUE_data, extraUE_data, year, month) {
-  df_int <- read.csv(intraUE_data, stringsAsFactors = FALSE)
-  df_ext <- read.csv(extraUE_data, stringsAsFactors = FALSE) 
+get_flightData <- function(intraUE_data, extraUE_data, countries) {
+  df_int <- read_tsv(intraUE_data) 
+  df_ext <- read_tsv(extraUE_data)
+  if ("OE" %in% countries) countries <- c(countries, "EU27_2020", "EU28", "UK")
   
   df <- bind_rows(df_int, df_ext) %>%
+    pivot_longer(!1, names_to = "TIME", values_to = "VALUE") %>%
+    separate(1, into = c("UNIT", "TRA_MEAS", "PARTNER", "GEO"), sep = ',') %>%
     # Data from passengers, monthly
-    filter(UNIT == "Passenger", grepl("M", TIME))
+    filter(UNIT == "PAS", GEO %in% countries, PARTNER %in% countries, grepl("M", TIME))
+    
   df1 <- df %>%
-    # Clean dataset
+    # Add year and month
     mutate(YEAR = as.numeric(substr(TIME,1,4)),
            MONTH = as.numeric(substr(TIME,6,8)),
-           Value = as.numeric(gsub(" ", "", Value))/lubridate::days_in_month(month),
-           GEO = case_when(
-             GEO == "Germany (until 1990 former territory of the FRG)" ~ "Germany",
-             GEO == "European Union - 27 countries (from 2020)" ~ "EU27",
-             GEO == "European Union - 28 countries (2013-2020)" ~ "EU28",
-             TRUE ~ df$GEO),
-           PARTNER = case_when(
-             PARTNER == "Germany (until 1990 former territory of the FRG)" ~ "Germany",
-             PARTNER == "European Union - 27 countries (from 2020)" ~ "EU27",
-             PARTNER == "European Union - 28 countries (2013-2020)" ~ "EU28",
-             PARTNER == "China including Hong Kong" ~ "Hubei-China",
[…]

### 2020-12-12 workflow/scripts/predictor_matrices.Rmd
wip predictor matrices

@@
 library(readr)
+library(ggsci)
+library(scales)
+source("./scripts/utils.R")
 source("./scripts/flight_data.R")
@@
 
-```{r}
+## GLM Model
+
+We define the migration rates $m_{ij}^{e}$ from deme $i$ to deme $j$ in epoch $e$ as a log linear function of a set of predictors $\boldsymbol{X}$, log transformed and scaled, such that
+
+$$
+m_{ij}^{e} = c \exp(\beta_{pas} \delta_{pas} x_{ij.pas}^e + \beta_{pop} \delta_{pop }x_{ij.pop} + \beta_{dist} \delta_{dist}x_{ij.dist})
+$$
+
+where $\boldsymbol{\beta}$ are the coefficients for the predictors that can be between $-\infty$ and $\infty$ (prior  $\mathcal{N}(0,4)$) and $\boldsymbol{\delta}$ are the indicators that can be 0 or 1 and denote if a predictor contributes at all.
+
 ## Daily flight passengers
 
+Passengers carried data from Eurostats.
+Reported _departures_ values are taken for European countries (recommendation from Eurostats to avoid duplicates).
+For China, arrivals and departures values are considered.
+
+- **Epoch 1:** December 2019 - January 2020 (Wuhan lockdown 23 Jan)
+- **Epoch 2:** February 2020 
+- **Epoch 3:** March 2020
+
+\tiny
+```{r, echo = FALSE, message = FALSE, warning = FALSE}
 demes <- c(China = "CN", France = "FR", Germany = "DE", Italy = "IT", OtherEuropean = "OE", Spain = "ES")
@@
   group_by(GEO, PARTNER, EPOCH) %>%
-  mutate(N = n(),
+  summarise(N = n(),
          EVALUE = sum(DVALUE/N)) %>%
@@
   mutate(LVALUE = log(EVALUE + 1),
-         SLVALUE = scale(LVALUE))
+         SLVALUE = scale(LVALUE)) %>%
+  arrange(EPOCH, factor(GEO, levels = demes), factor(PARTNER, levels = demes))
 
-write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
+#write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
+```
+
+```{r, echo = FALSE}
+flight_matrices <- lapply(1:3, function (i) {
+  fligh_data_3i %>%
+  filter(EPOCH == i) %>%
+  select(GEO, PARTNER, SLVALUE) %>%
+  pivot_wider(names_from = "PARTNER", values_from = "SLVALUE") %>%
+  mutate_if(is.numeric, ~round(.,2)) %>%
+  column_to_rownames("GEO") %>%
+  select(unname(demes))
+})
+
+knitr::kable(flight_matrices, booktabs = TRUE, caption = "Flight passengers between demes, epochs 1-3")
+```
+
+## Daily flight passengers
+
+```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 3}
+set_plotopts()
+dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
+dcolors <- dcolors[1:length(demes)]
+names(dcolors) <- demes[sort(labels(demes))]
+#dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
+#names(dark_dcolors) <- names(dcolors)
+
+# Fligh data plot
+ggplot(fligh_data_3i, aes(x = EPOCH, y = EVALUE, color = PARTNER)) +
+  geom_line() +
+  geom_point(size = 1) +
+  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
+  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
+  facet_grid(~GEO) +
+  xlab("Epoch") +
+  ylab("# Daily Flight Passengers") + 
+  #theme(axis.title.x = element_blank()) + 
+  scale_color_manual(values = dcolors, name = "Destination")
 ```
@@
 
-Population counts from World Population Prospects 2019, United Nations Population Division,  `wpp2019` R library.
-Values for **China**, **France**, **Germany**, **Italy**, **Other European** and **Spain**.
-
-A predictor matrix for source population and a predictor matrix for destination population are generated, log transformed and standardized. Constant across demes.
-
-\small
 ```{r, echo=FALSE}
@@
 
+Population counts from World Population Prospects 2019, United Nations Population Division,  `wpp2019` R library: **China** - `r population[1,2] * 1e3`, **France** - `r population[2,2] * 1e3`, **Germany** - `r population[3,2] * 1e3`, **Italy** - `r population[4,2] * 1e3`, **Other European** - `r population[5,2] * 1e3` and **Spain** - `r population[6,2] * 1e3`.
+
+A predictor matrix for source population and a predictor matrix for destination population are generated, log transformed and standardized. Constant across demes.
+
+\tiny
 ```{r, echo=FALSE}
+# TODO use long format, easy to combine later to export to xml, like with the flight data
 # Population source matrix
@@
 ## Distance between countries
-
-Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.
-
-When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).
-
-The preview shows you a rendered HTML copy of the contents of the editor.
-Consequently, unlike *Knit*, *Preview* does not run any R code chunks.
-Instead, the output of the chunk when it was last run in the editor is displayed.
+Options:
+Great circle distance between centroids of each country - Which distance for other european deme? Could be the average
+Average pairwaise distance between airports - considers that only flight transportation matters, more work
+Distance taking into account population distribution inside the country.
+Minimum distance.

### 2020-12-12 workflow/scripts/flight_data.R

@@ -14,7 +14,8 @@ get_flightData <- function(intraUE_data, extraUE_data, countries) {
     pivot_longer(!1, names_to = "TIME", values_to = "VALUE") %>%
     separate(1, into = c("UNIT", "TRA_MEAS", "PARTNER", "GEO"), sep = ',') %>%
     # Data from passengers, monthly
-    filter(UNIT == "PAS", GEO %in% countries, PARTNER %in% countries, grepl("M", TIME))
+    filter(UNIT == "PAS", GEO %in% countries, 
+           PARTNER %in% countries, grepl("M", TIME))
     
   df1 <- df %>%
     # Add year and month
@@ -31,44 +32,49 @@ get_flightData <- function(intraUE_data, extraUE_data, countries) {
            GEO = ifelse(TRA_MEAS == "PAS_CRD_ARR", "CN", GEO),
            TRA_MEAS = ifelse(TRA_MEAS == "PAS_CRD_ARR", "PAS_CRD_DEP", TRA_MEAS)) %>%
     # In 2020 UK is out of the european aggregated value
-    filter((YEAR == 2019 & GEO != "EU27") | (YEAR == 2020 & GEO != "EU28")) 
-  
+    filter(!is.na(VALUE)) 
   
   # Correct for UK Brexit, EU28 -> EU27 in 2020
-  # Only for months 2020:
-  # - GEO EU28 = GEO EU27 + GEO United Kingdom to Partner European Countries
-  # - GEO European countries = Partner EU27 + Partner United kingdom
-  
-  eu28_geo <-  left_join(df1 %>% filter(GEO == "EU27_2020"), 
-                         df1 %>% filter(GEO == "UK") %>% select(-GEO), 
-                         by = c("TIME", "UNIT", "TRA_MEAS", "PARTNER", "YEAR", "MONTH")) %>%
-    mutate(VALUE = VALUE.x + VALUE.y,
-           GEO = "EU28")
-  
-  eu28_partner = left_join(df1 %>% filter(PARTNER == "EU27_2020"), 
-                           df1 %>% filter(PARTNER == "UK") %>% select(-PARTNER), 
-                           by = c("TIME", "UNIT", "TRA_MEAS", "GEO", "YEAR", "MONTH")) %>%
-    filter(YEAR == 2020) %>%
[…]

### 2020-12-14 workflow/scripts/predictor_matrices.Rmd
covariate matrix and analysis files europe6

@@
 library(scales)
+library(countrycode)
+library(geosphere)
+library(maps)
+library(sf)
 source("./scripts/utils.R")
@@
 - United Nations Population Division, World Population Prospects 2019: R library `wpp2019, data(pop)`
-- Distance
+- World map from the Natural Earth project (2013), loaded with R library `maps`.
 
@@
 
-fligh_data <- get_flightData(intraUE_data, extraUE_data, countries = demes)
+flight_df <- get_flightData(intraUE_data, extraUE_data, countries = demes)
 
-fligh_data_3i <- fligh_data %>%
+flight_df_3e <- flight_df %>%
   filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
@@
     MONTH == 3 ~ 3,
-    TRUE ~ 1)) %>%
-  group_by(GEO, PARTNER, EPOCH) %>%
+    TRUE ~ 1),
+    PRED = 1) %>%
+  rename(SRC = GEO, DEST = PARTNER) %>%
+  group_by(SRC, DEST, EPOCH, PRED) %>%
   summarise(N = n(),
@@
          SLVALUE = scale(LVALUE)) %>%
-  arrange(EPOCH, factor(GEO, levels = demes), factor(PARTNER, levels = demes))
+  arrange(EPOCH, factor(SRC, levels = demes), factor(DEST, levels = demes))
 
-#write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
+#write(slv_flightMatrix_3e, "files/flight_matrix_3e.csv", sep = ",")
 ```
@@
 flight_matrices <- lapply(1:3, function (i) {
-  fligh_data_3i %>%
+  flight_df_3e %>%
   filter(EPOCH == i) %>%
-  select(GEO, PARTNER, SLVALUE) %>%
-  pivot_wider(names_from = "PARTNER", values_from = "SLVALUE") %>%
+  select(SRC, DEST, SLVALUE) %>%
+  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
   mutate_if(is.numeric, ~round(.,2)) %>%
-  column_to_rownames("GEO") %>%
-  select(unname(demes))
+  column_to_rownames("SRC") %>%
+  select(all_of(unname(demes)))
 })
@@
 names(dcolors) <- demes[sort(labels(demes))]
-#dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
-#names(dark_dcolors) <- names(dcolors)
 
 # Fligh data plot
-ggplot(fligh_data_3i, aes(x = EPOCH, y = EVALUE, color = PARTNER)) +
+ggplot(flight_df_3e, aes(x = EPOCH, y = EVALUE, color = DEST)) +
   geom_line() +
@@
   scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
-  facet_grid(~GEO) +
+  facet_grid(~SRC) +
   xlab("Epoch") +
@@
   tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
-  arrange(name)
+  arrange(name) %>%
+  rename(POP = 2) %>%
+  mutate(SLVALUE = scale(log(POP)))
+pop_df <- population %>%
+  mutate(n = 6, SRC = demes) %>%
+  uncount(n) %>%
+  mutate(DEST = rep(demes, 6),
+         POP = ifelse(SRC == DEST, NA, POP),
+         SLVALUE = ifelse(SRC == DEST, NA, SLVALUE)) %>%
+  select(-name) %>%
+  filter(!is.na(POP))
+pop_df_3e <- bind_rows(replicate(3, pop_df %>% mutate(PRED = 2), simplify = FALSE)) %>% 
+  mutate(EPOCH = rep(1:3, each = 30, length = 90)) %>%
+  bind_rows(bind_rows(replicate(3, pop_df %>% 
+                        arrange(factor(DEST, levels = demes),
+                                factor(SRC, levels = demes)) %>%
+                          mutate(TEMP = SRC,
+                                 SRC = DEST,
+                                 DEST = TEMP,
+                                 PRED = 3),
+                      simplify = FALSE)) %>%
+  mutate(EPOCH = rep(1:3, each = 30, length = 90)))
 ```
@@
 
-A predictor matrix for source population and a predictor matrix for destination population are generated, log transformed and standardized. Constant across demes.
+A predictor matrix for source population and a predictor matrix for destination population are generated, log transformed and standardized. Constant across epochs.
 
@@
 ```{r, echo=FALSE}
-# TODO use long format, easy to combine later to export to xml, like with the flight data
-# Population source matrix
-pop_source <- matrix(1, 6, 6) * population$`2020`
-diag(pop_source) <- NA
-vslpop_source <- scale(log(c(pop_source)))
-slpop_source <- matrix(vslpop_source, 6, 6)
-colnames(slpop_source) <- demes[population$name]
-rownames(slpop_source) <- demes[population$name]
+pop_matrices <- lapply(c("DEST", "SRC"), function (x) {
+  pop_df %>%
+  select(-POP) %>%
+  pivot_wider(names_from = all_of(x), values_from = "SLVALUE") %>%
+  mutate_if(is.numeric, ~round(., 2)) %>%
+  mutate_at(1, ~factor(., levels = demes)) %>%
+  arrange_at(1) %>%
+  select(1, all_of(unname(demes)))
+})
 
-# Population destination matrix
-pop_dest <- sweep(matrix(1, 6, 6), 2, population$`2020`, `*`)
-diag(pop_dest) <- NA
-vslpop_dest <- scale(log(c(pop_dest)))
-slpop_dest <- matrix(vslpop_dest, 6, 6)
-colnames(slpop_dest) <- demes[population$name]
-rownames(slpop_dest) <- demes[population$name]
-
-knitr::kable(list(round(slpop_source, 2), round(slpop_dest, 2)), booktabs = TRUE, caption = "Population counts source (left) and destination (right)")
+knitr::kable(pop_matrices, booktabs = TRUE, caption = "Population counts source (left) and destination (right)")
 ```
@@
 
-## Distance between countries
-Options:
-Great circle distance between centroids of each country - Which distance for other european deme? Could be the average
-Average pairwaise distance between airports - considers that only flight transportation matters, more work
-Distance taking into account population distribution inside the country.
-Minimum distance.
+## Geographical distance
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE}
+map <- st_as_sf(map("world", plot = FALSE, fill = TRUE)) %>%
+  mutate(country = ifelse(ID == "UK", "United Kingdom", as.character(ID)),
+         continent = countrycode(sourcevar = country,
+                             origin = "country.name",
+                             destination = "continent")) %>%
+  filter(country %in% labels(demes) | continent == "Europe") %>%
+  mutate(deme = case_when(
+    country %in% labels(demes) ~ country,
+    TRUE ~ "OtherEuropean"))
+map <- cbind(map, st_coordinates(st_centroid(map))) %>%
+  rename(longitude = X, latitude = Y) %>%
+  mutate(deme = demes[deme])
+
+centroids <- data.frame(map) %>%
+  select(country, deme, longitude, latitude)
+
+dist_df <- centroids %>%
+  full_join(centroids, by = character())
+dist_df <- dist_df%>%
+  mutate(DIST = distHaversine(select(dist_df, longitude.x, latitude.x),
+                              select(dist_df, longitude.y, latitude.y))) %>%
+  group_by(deme.x, deme.y) %>%
+  summarise(DIST = mean(DIST), .groups = "drop") %>%
+  filter(deme.x != deme.y) %>%
+  rename(SRC = deme.x, DEST = deme.y) %>% 
+  mutate(SLVALUE = scale(log(DIST))) %>%
+  arrange(factor(SRC, levels = demes), factor(DEST, levels = demes))
+dist_df_3e <- bind_rows(replicate(3, dist_df, simplify = FALSE)) %>%
+  mutate(EPOCH = rep(1:3, each = 30, length = 90), PRED = 4)
+```
+
+Great circle pairwise distance between centroids of each country.
+Centroids are computed  from the Natural Earth project (the 1:50m resolution version) world map 2013.
+Constant across epochs.
+Symmetric.
+
+\tiny
+```{r, echo = FALSE}
+dist_matrix <- dist_df %>%
+  select(-DIST) %>%
+  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
+  mutate_if(is.numeric, ~round(., 2)) %>%
+  column_to_rownames("SRC") %>%
+  select(all_of(unname(demes)))
+
+knitr::kable(dist_matrix, booktabs = TRUE, caption = "Great circle pairwise distance between countries geographic centroid.")
+```
+
+\normalsize
+The distance from one deme to the Other European deme is the average distance between the deme centroid and all the Other European countries centroids.
+
+## Geographical distance
+
+```{r, echo = FALSE, message =  FALSE, fig.align="center"}
+
+ggplot() +
+    geom_sf(data = map, aes(fill = deme), alpha = 0.7) + 
+    geom_point(data = map %>% filter(country %in% labels(demes)), aes(x = longitude, y = latitude), shape = 4, size = 2) +
+    coord_sf(xlim = c(-15, 120), ylim = c(25, 60), expand = FALSE) +
+  scale_fill_manual(values = dcolors, na.value = "grey") +
+  theme(legend.position = "none",
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+```
+
+Map of Europe and part of Asia showing the demes included in the analysis (in color).
+The X represents the centroid of each deme.
+The centroids of each European country fromm the Other European deme are not drawed.
+
+<!-- Other distance ptions: -->
+<!-- Average pairwaise distance between airports - considers that only flight transportation matters, more work -->
+<!-- Distance taking into account population distribution inside the country. CEPPI dataset -->
+<!-- Minimum distance. -->
+
+## Covariate list BEAST BDMM-Prime
+
+The three predictors are saved in the following order: **Flight data, population count source, population count destination and geographical distance.**
+
+Each of the predictor is composed of three matrices, one per epoch, with the first matrix being the most recent epoch, e.g. the first 30 values correponds to the fligh data for March.
+
+Diagonal values are skipped.
+
+Therefore the final size of the covariate list is of dimension $(30 \cdot 3) \cdot 4 = 360$.
+
+```{r, echo = FALSE}
+pred_3e <- bind_rows(flight_df_3e, pop_df_3e) %>%
+  bind_rows(dist_df_3e) %>%
+  arrange(PRED, desc(EPOCH), factor(SRC, levels = demes), factor(DEST, levels = demes)) %>%
+  select(PRED, EPOCH, SRC, DEST, SLVALUE)
+
+if (nrow(pred_3e) != 360) warning("Something went wrong, the size of the combined predictors is ", nrow(pred_3e))
+
+mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(SLVALUE), ncol = 5, byrow = TRUE))
+write_csv(mpred_3e, "./files/covariates_4p_3e.csv", col_names = FALSE)
+```
+
+\
+
+Covariate list saved as `files/covariates_4p_3e.csv`

### 2020-12-16 workflow/scripts/analyze_trajectories_v2.Rmd
analyze traj rmd

---
title: "Trajectory Mapping Results"
subtitle: "analysis x"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dove"
    fonttheme: "structurebold"
    highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(knitr.kable.NA = '')
```

```{r libraries}
# Load libraries ---------------------------------------------------------------
library(argparse)
library(yaml)
library(tidyverse)
library(lubridate)
library(scales)
library(ggpubr)
library(ggsci)
library(ggridges)
library(patchwork)
library(hrbrthemes)
library(circlize)
library(chorddiag) 
library(ggforce)
library(ComplexHeatmap)
library(maps)
library(sf)
library(wpp2019)
library(gridExtra)
library(grid)
# Source files -----------------------------------------------------------------
source("./scripts/trajProcessing.R")
source("./scripts/utils.R")
```

```{r parser}
# Parser -----------------------------------------------------------------------
parser <- argparse::ArgumentParser()
parser$add_argument("--input", type = "character", help="trajectory file")
parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
parser$add_argument("--n", type = "integer", help = "Number of trajectories to analyze")
parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
parser$add_argument("--output_table", type = "character", help = "Output path for the tables")
args <- parser$parse_args()
```

```{r debugging}
# Debugging
args$burnin = 0.1
args$demes = read.csv("/Users/maceci/code/mt-analysis/201014_europe2/files/demes.csv", stringsAsFactors = FALSE)
args$input = "/Users/maceci/code/mt-analysis/201030_europe3/results/europe3c/201030_europe3.TL.traj"
args$metadata =  "/Users/maceci/code/mt-analysis/201014_europe2/results/europe2/final_metadata.tsv"
args$n = 50
demes = args$demes
demes$exclude_country[5] = "France,Germany,Italy,Spain"
demes$deme[3] = "China"
```

```{r plotopts}
# Plot options -----------------------------------------------------------------------
set_plotopts()
dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
dcolors <- dcolors[1:nrow(demes)]
names(dcolors) <- sort(demes$deme)
dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
names(dark_dcolors) <- names(dcolors)
ecolors <- c("#1e3d59", "#f5f0e1", "#ff6e40", "#ffc13b")
names(ecolors) <- c("B", "M", "IM", "OM")
```

## Deme configuration
\tiny
```{r demeconfig}
# Load demes configuration and sequence metadata -------------------------------
# demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
#                     exclude_country = NA, exclude_division = NA,
#                     min_date = NA, max_date = NA, seq_per_deme = NA) %>%
#   bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(args$demes, collapse = " ")),
#                           data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
#   group_by(deme) %>% 
#   mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
#   distinct() %>%
#   filter_all(any_vars(!is.na(.))) %>%
#   mutate(deme = str_to_title(deme)) %>%
#   mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme))

china_deme <- demes$deme[demes$region == "Asia"]
if (china_deme == "Hubei-China") china_deme = "China"
```

```{r map, fig.height=2, fig.align="center"}
# https://www.r-bloggers.com/2019/04/zooming-in-on-maps-with-sf-and-ggplot2/
worldmap <- st_as_sf(map("world", plot = FALSE, fill = TRUE)) %>%
  mutate(country = ifelse(ID == "UK", "United Kingdom", as.character(ID)),
         continent = countrycode(sourcevar = country,
                             origin = "country.name",
                             destination = "continent")) %>%
  #filter(country %in% demes$country | continent == "Europe") %>%
  mutate(deme = case_when(
    country %in% demes$country ~ country,
    continent == "Europe" ~ "OtherEuropean"))

zoom_to <- c(20, 50) 
zoom_level <- 2.6
target_crs <- sprintf('+proj=aeqd +lon_0=%f +lat_0=%f',
                      zoom_to[1], zoom_to[2])
C <- 40075016.686   # ~ circumference of Earth in meters
x_span <- C / 2^zoom_level
y_span <- C / 2^(zoom_level+1)

zoom_to_xy <- st_transform(st_sfc(st_point(zoom_to), crs = 4326),
                           crs = target_crs)
disp_window <- st_sfc(
    st_point(st_coordinates(zoom_to_xy - c(x_span/2, y_span/2))),
    st_point(st_coordinates(zoom_to_xy + c(x_span *1.2, y_span *1.2))),
    crs = target_crs
)
ggplot() + geom_sf(data = worldmap, aes(fill = deme), alpha = 0.8, size = 0) +
    scale_x_continuous(breaks = seq(-100, 200, by = 10)) +
    scale_y_continuous(breaks = seq(0, 100, by = 10)) +
    #geom_sf(data = zoom_to_xy, color = 'red') +
    coord_sf(xlim = st_coordinates(disp_window)[,'X'],
             ylim = st_coordinates(disp_window)[,'Y'],
             crs = target_crs, expand = FALSE) +
  scale_fill_manual(values = dcolors, na.value = "grey90") +
  theme_void() +
  theme(panel.grid.major = element_line(colour = "grey", size = 0.1),
        legend.position = "none")
        # panel.ontop = TRUE,
        # panel.background = element_blank())

knitr::kable(demes %>% select(-c("demeID", "exclude_division", "prob")), booktabs = TRUE)
```

## ECDC Case count data

```{r ecdc-data, results='hide'}
# Case data information from ECDC ----------------------------------------------
# Most recent sample value
metadata <- read_tsv(args$metadata)
MRS <- max(metadata$date)
mrs <- ymd_hms(paste(MRS, '23:59:00 UTC'))

case_data <- get_cases(demes, from = ymd("2019-09-01"), to = MRS + 10) %>%
  arrange(date) %>%
  group_by(deme) %>%
  mutate(cumcases = cumsum(cases),
         cumdeaths = cumsum(deaths)) %>%
  group_by(deme, date) %>%
  mutate(cumcases = max(cumcases),
         cumdeaths = max(cumdeaths)) %>%
  select(deme, date, cumcases, cumdeaths, cases)  %>%
  distinct()

# Population data
data(pop)
population <- pop %>% filter(name %in% demes$deme) %>% select(name, "2020")
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>% rename(deme = name, pop = 2)

case_data_pop <- case_data %>% left_join(population, by="deme") %>% mutate(cas100 = cumcases * 1e5 / (pop * 1e3))
```

::: {.columns}

:::: {.column width="40%"}
\small
```{r ecdc-table, echo=FALSE}
knitr::kable(case_data_pop %>% ungroup %>% filter(date == MRS + 10) %>% select(deme, cumcases, pop, cas100) %>% mutate_if(is.numeric, function (x) round(x,1)), booktabs = TRUE, caption = "Total number of cases reported to ECDC 18th March")
```
::::

:::: {.column width="60%"}
```{r ecdc-plot, fig.height = 3, fig.width=4, fig.cap="ECDC case counts for each deme from the beginning of the pandemic to March 18"}
ggplot(case_data) +
  geom_line(aes(x = date, y = cumcases, color = deme)) +
   scale_color_manual(values = dcolors, name = "") +
   scale_x_date(limits = c(ymd("2019-12-30"), ymd("2020-03-18")), date_breaks = "1 month", date_labels = "%b %d") +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    ylab("Case count") +
    theme(axis.title.x=element_blank(),
          legend.position = c(0.2, 0.7),
          legend.box="horizontal",
          panel.grid.major.x =  element_line(colour = "grey80", size = 0.3),
          panel.grid.minor.x =  element_line(colour = "grey80", size = 0.2))
```

::::

:::

## Epidemic trajectory data

From the Stochastic Trajectory Mapping analysis, we obtain one epidemic trajectory per set of parameters + typed node tree.

The processing of the trajectory data includes the generation of two different datasets:

- `states`: We have the total number of inferred cases by trajectory, deme and time.
- `events`: We have each event that happened in a epidemic trajectory, with its type (origin, birth, death or migration), the source/destination deme and time.

```{r traj-data, warnings = FALSE, message = FALSE}
# Load trajectory data ---------------------------------------------------------
df <- loadTrajectories(args$input, burninFrac = args$burnin, subsample = args$n)

# Data wrangling ---------------------------------------------------------------
states <- df$states %>%
  mutate(type = factor(type,
                       levels = 0:(nrow(demes) - 1),
                       labels = sort(demes$deme)),
         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
  mutate(date = date_model + 10)

events <- df$events %>%
  mutate(src = factor(src,
                      levels = 0:(nrow(demes) - 1),
                      labels = sort(demes$deme)),
         dest = ifelse(is.na(dest), as.character(src),
                       as.character(factor(dest,
                                           levels = 0:(nrow(demes) - 1),
                                           labels = sort(demes$deme)))),
         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
  mutate(date = date_model + 10)
```

```{r table-states}
knitr::kable(head(states, 5), booktabs = TRUE, caption = "States dataset")
```

## Epidemic trajectory data

```{r table-events}
knitr::kable(head(events, 5), booktabs = TRUE, caption = "Events dataset")
```

\normalsize
To have a feasible time of analysis of the epidemic trajectories we take a random subsample of 500 trajectories.

To facilitate visualization and summarise the results, we take a grid time of 1 day and summarise the number of events that day as the sum of the events in the corresponding time interval; and the number of inferred cases as the maximum of the interval.

```{r traj-process, warning=FALSE}
# Date grids
max_age <- max(states$age)
min_date <- date_decimal(decimal_date(mrs) - max_age)
#ages_5day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-5 days')))
ages_1day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-1 days')))

# Grid trajectories, time grid of 5 days taking maximum value
gt <- gridTrajectoriesByAge(states, ages_1day) %>%
  replace_na(list(N = 0)) %>%
  mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
         date = date_model + 10)

# Summary statistics
gt_summary <- gt %>%
  group_by(date, type) %>%
  summarize(Imean = mean(N),
            Imedian = median(N),
            Ilow    = quantile(N, 0.025), #HPD??
            Ihigh   = quantile(N, 0.975),
            .groups = "drop_last") %>%
  ungroup()

# Grid events, time grid of 1 day, taking the sum of events
ge <- gridEventsByAge(events, ages_1day) %>%
  mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
         date = date_model + 10,
         dest = ifelse(is.na(dest), as.character(src), as.character(dest)))

```

<!-- ## Inferred case counts -->

```{r}
# 1. Trajectories - Total case counts
# 1.1 Median and 95% CCI of Total cases on the full time period. Fraction median
#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases

reported <- gt_summary %>%
  rename(deme = type) %>%
  filter(date == MRS + 10) %>%
  left_join(case_data_pop , by = c("deme", "date")) %>%
  mutate(fc = cumcases/Imedian,
         lfc = cumcases/Ihigh,
         hfc = cumcases/Ilow,
         rc = Imedian/cumcases,
         lrc = Ilow/cumcases,
         hrc = Ihigh/cumcases,
         mcas100 = Imedian * 100 / pop,
         lcas100 = Ilow * 100 / pop,
         hcas100 = Ihigh * 100 / pop) %>%
  #select(deme, Imedian, Ilow, Ihigh, rc, lrc, hrc, cumcases, pop, cas100) %>%
  mutate_if(is.numeric, function(x) round(x,2))

# r_reported <- reported %>%
#   mutate(Country = type,
#          `Cases ECDC` = cumcases,
#          `Median Cases (95% CCI)` = paste0(round(Imedian), " (", round(Ilow), "-", round(Ihigh), ")"),
#          `Median Reporting Rate (95% CCI)` = paste0(round(reportrate, 2), " (", round(hreportrate, 2), "-", round(lreportrate, 2), ")")) %>%
#   select(-one_of(colnames(reported)))
#
# write_tsv(r_reported, paste0(args$output_table, "01.tsv"))
```

```{r trajs}
# 1.2 All trajectories single deme in log scale with ECDC case count data
trajs <- lapply(demes$deme, function(deme) {
  gt %>%
    # 1. Prepare data
    filter(type == deme) %>%
    mutate(traj = as.character(traj)) %>%
    bind_rows(case_data %>%
                rename(type = deme,
                       N = cumcases) %>%
                filter(type == deme) %>%
                mutate(traj = "ecdc") %>%
                select(type, traj, date, N) %>%
                distinct()) %>%
    # 2. Plot
    ggplot() +
    geom_line(aes(date, N, group = traj,
                  color = traj == "ecdc",
                  alpha = traj == "ecdc",
                  linetype = traj == "ecdc",
                  size = traj == "ecdc")) +
    scale_color_manual(name = "",
                       labels = c(paste0(deme, " inferred\n population trajectories"),
                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                         " ECDC total case count")),
                       values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
    scale_linetype_manual(name = "", values = c(1, 2),
                          labels = c(paste0(deme, " inferred\n population trajectories"),
                                     paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                            " ECDC total case count"))) +
    scale_size_manual(name = "", values = c(0.5, 1),
                      labels = c(paste0(deme, " inferred\n population trajectories"),
                                 paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                        " ECDC total case count"))) +
    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
                 date_labels = "%b %d") +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    ylab("Population size") +
    theme(legend.position = c(0.4, 0.9),
          axis.title.x=element_blank())
  })

names(trajs) <- demes$deme
# ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1:length(trajs)), nrow = 3, ncol = 2),
#          filename = paste0(args$output_figure, "02.png"),
#          width = 2300, height = 3000, res = 300)
```

\tiny
```{r trajs-plotv2, results='asis', fig.height = 4}
for (d in demes$deme) {
  cat("\n## Inferred case counts - ", d, "\n")
  grid.arrange(trajs[[d]], tableGrob(reported %>% filter(deme == d) %>% t, theme = ttheme_minimal()), ncol = 2, widths = c(2,1))
  cat("\n")
}
```
```{r}
gt_v2 <- ge %>%
  ungroup %>%
  filter(event %in% c("B", "M")) %>%
  group_by(traj, dest, date) %>%
  arrange(date) %>%
  summarise(Nt = sum(N), .groups = 'drop_last') %>%
  mutate(cumNt = cumsum(Nt)) %>%
  rename(type = dest)

gt_summary_v2 <- gt_v2 %>%
  group_by(date, type) %>%
  summarize(Imean = mean(cumNt),
            Imedian = median(cumNt),
            Ilow    = quantile(cumNt, 0.025), #HPD??
            Ihigh   = quantile(cumNt, 0.975),
            Dmedian = median(Nt),
            Dlow    = quantile(Nt, 0.025), #HPD??
            Dhigh   = quantile(Nt, 0.975),
            .groups = "drop_last") %>%
  ungroup()

```

```{r trajs-infov2}
# 1. Trajectories - Total case counts
# 1.1 Median and 95% CCI of Total cases on the full time period. Fraction median
#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases

reported <- gt_summary_v2 %>%
  rename(deme = type) %>%
  filter(date == MRS + 10) %>%
  left_join(case_data_pop , by = c("deme", "date")) %>%
  mutate(fc = cumcases/Imedian,
         lfc = cumcases/Ihigh,
         hfc = cumcases/Ilow,
         rc = Imedian/cumcases,
         lrc = Ilow/cumcases,
         hrc = Ihigh/cumcases,
         mcas100 = Imedian * 100 / pop,
         lcas100 = Ilow * 100 / pop,
         hcas100 = Ihigh * 100 / pop) %>%
  #select(deme, Imedian, Ilow, Ihigh, rc, lrc, hrc, cumcases, pop, cas100) %>%
  mutate_if(is.numeric, function(x) round(x,2))

# r_reported <- reported %>%
#   mutate(Country = type,
#          `Cases ECDC` = cumcases,
#          `Median Cases (95% CCI)` = paste0(round(Imedian), " (", round(Ilow), "-", round(Ihigh), ")"),
#          `Median Reporting Rate (95% CCI)` = paste0(round(reportrate, 2), " (", round(hreportrate, 2), "-", round(lreportrate, 2), ")")) %>%
#   select(-one_of(colnames(reported)))
#
# write_tsv(r_reported, paste0(args$output_table, "01.tsv"))
```

```{r trajsv2}
# 1.2 All trajectories single deme in log scale with ECDC case count data
trajs <- lapply(demes$deme, function(deme) {
  gt_v2 %>%
    # 1. Prepare data
    filter(type == deme) %>%
    mutate(traj = as.character(traj)) %>%
    bind_rows(case_data %>%
                rename(type = deme,
                       cumNt = cumcases) %>%
                filter(type == deme) %>%
                mutate(traj = "ecdc") %>%
                select(type, traj, date, cumNt) %>%
                distinct()) %>%
    # 2. Plot
    ggplot() +
    geom_line(aes(date, cumNt, group = traj,
                  color = traj == "ecdc",
                  alpha = traj == "ecdc",
                  linetype = traj == "ecdc",
                  size = traj == "ecdc")) +
    scale_color_manual(name = "",
                       labels = c(paste0(deme, " inferred\n population trajectories"),
                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                         " ECDC total case count")),
                       values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
    scale_linetype_manual(name = "", values = c(1, 2),
                          labels = c(paste0(deme, " inferred\n population trajectories"),
                                     paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                            " ECDC total case count"))) +
    scale_size_manual(name = "", values = c(0.5, 1),
                      labels = c(paste0(deme, " inferred\n population trajectories"),
                                 paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                        " ECDC total case count"))) +
    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
                 date_labels = "%b %d") +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    ylab("Population size") +
    theme(legend.position = c(0.4, 0.9),
          axis.title.x=element_blank())
  })

names(trajs) <- demes$deme
# ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1:length(trajs)), nrow = 3, ncol = 2),
#          filename = paste0(args$output_figure, "02.png"),
#          width = 2300, height = 3000, res = 300)
```

\tiny
```{r trajs-plot, results='asis', fig.height = 4}
for (d in demes$deme) {
  cat("\n## Inferred case counts v2- ", d, "\n")
  grid.arrange(trajs[[d]], tableGrob(reported %>% filter(deme == d) %>% t, theme = ttheme_minimal()), ncol = 2, widths = c(2,1))
  cat("\n")
}
```

## Reporting rate

```{r report-rate}
# 1.3. Reporting rate by day by case
reportrate <- gt_summary_v2 %>%
  filter(!type %in% c(china_deme, "OtherEuropean")) %>%
  left_join(case_data %>% rename(type = deme), by = c("type", "date"))  %>%
  replace_na(list(cases = 0)) %>%
  mutate(reportrate = ifelse(is.infinite(cases/Dmedian), NA, cases/Dmedian),
         rlow = ifelse(is.infinite(cases/Dhigh), NA, cases/Dhigh),
         rhigh = ifelse(is.infinite(cases/Dlow), NA, cases/Dlow)) %>%
  filter(!is.na(reportrate)) %>%
  # mutate(week = lubridate::week(date)) %>%
  # group_by(week, type) %>%
  # mutate(w_reportrate = mean(reportrate)) %>%
  ggplot() +
  geom_line(aes(x = date, y = reportrate, color = type), stat="identity") +
  geom_point(aes(x = date, y = reportrate, color = type), stat="identity") +
  geom_ribbon(aes(date, ymin = rlow, ymax = rhigh, fill = type), alpha = 0.1) +
  ylab("Reporting rate") +
  scale_color_manual(values = dcolors, name = "") +
  scale_fill_manual(values = dcolors, name = "") +
  scale_x_date(limits = c(ymd("2020-01-20"), ymd(MRS + 10)), date_breaks = "1 week", date_labels = "%b %d") +
  theme(#legend.position = "none",
        axis.title.x=element_blank()) +
  coord_cartesian(ylim = c(0,5))

reportrate
#
# ggexport(reported5days, filename = paste0(args$output_figure, "03.png"),
#          width = 1500, height = 1000, res = 300)

```

## Inferred case counts - Additional figures
```{r}
# 1.1 Median and 95% CCI by deme in log scale
gribbon <- ggplot(gt_summary) +
  geom_ribbon(aes(date, ymin = Ilow, ymax = Ihigh, fill = type), alpha = 0.5) +
  geom_line(aes(date, Imedian, colour = type)) +
  ylab("Population size") +
  xlab("Date") +
  #labs(title = "SARS-CoV-2 early epidemics in Europe and Hubei",
       #subtitle = "Case trajectories Median and 95% CCI by deme") +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_x_date(limits = c(ymd("2019-10-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %d") +
  scale_fill_manual(name = "", values = dcolors) +
  scale_color_manual(name = "", values = dcolors) +
  theme(legend.position = c(0.2, 0.7),
        legend.box="horizontal",
        axis.title.x=element_blank())
#
# ggexport(gribbon,
#          filename = paste0(args$output_figure, "01.png"),
#          width = 1500, height = 1000, res = 300)
```

### 2020-12-16 workflow/scripts/predictor_matrices.Rmd
analyze traj rmd

@@
 where $\boldsymbol{\beta}$ are the coefficients for the predictors that can be between $-\infty$ and $\infty$ (prior  $\mathcal{N}(0,4)$) and $\boldsymbol{\delta}$ are the indicators that can be 0 or 1 and denote if a predictor contributes at all.
+
+Analysis Europe 6 without indicator variables, Europe 7 with indicator variables, same predictors.
 
@@
 The X represents the centroid of each deme.
-The centroids of each European country fromm the Other European deme are not drawed.
+The centroids of each European country from Other European deme are not drawed.
 
-<!-- Other distance ptions: -->
+<!-- Other distance options: -->
 <!-- Average pairwaise distance between airports - considers that only flight transportation matters, more work -->

### 2020-12-16 workflow/scripts/subsample.R

@@ -20,6 +20,7 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
                       exclude_country = NA, exclude_division = NA,
                       from, to, seq_per_deme, seed, prob = "time") {
   set.seed(seed)
+  
   # Read files
   alignment <- read.FASTA(file = alignment)
   metadata <- read.delim(file = metadata, stringsAsFactors = FALSE) %>%
@@ -49,7 +50,8 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
   if (prob == "deaths") to <- ymd(to) + 14
   
   # Get case counts for deme
-  cases_deme <- get_cases_deme("deme", region_name, country_name, division_name, 
+  case_data <- get_dataECDC()
+  cases_deme <- get_cases_deme(case_data, "deme", region_name, country_name, division_name, 
                                exclude_country, exclude_division, 
                                from, to)  
   

### 2020-12-16 workflow/scripts/analyze_trajectories_v2.Rmd
wip analyze traj Rmd

@@
   select(deme, date, cumcases, cumdeaths, cases)  %>%
-  distinct()
+  distinct_at(vars(deme, date, cumcases, cumdeaths), .keep_all = TRUE)
 
@@
     ylab("Population size") +
-    theme(legend.position = c(0.4, 0.9),
+    theme(legend.position = c(0.3, 0.9),
           axis.title.x=element_blank())
@@
 
-```{r report-rate}
+```{r report-rate, fig.height = 4}
 # 1.3. Reporting rate by day by case
@@
 ## Inferred case counts - Additional figures
-```{r}
+```{r traj-sum, fig.height = 4}
 # 1.1 Median and 95% CCI by deme in log scale
@@
         axis.title.x=element_blank())
+
+gribbon
+
 #

### 2020-12-23 scripts/analyze_trajectories_v2.Rmd
restructuting

--- 2020-12-16 workflow/scripts/analyze_trajectories_v2.Rmd
@@
 args$demes = read.csv("/Users/maceci/code/mt-analysis/201014_europe2/files/demes.csv", stringsAsFactors = FALSE)
-args$input = "/Users/maceci/code/mt-analysis/201030_europe3/results/europe3c/201030_europe3.TL.traj"
-args$metadata =  "/Users/maceci/code/mt-analysis/201014_europe2/results/europe2/final_metadata.tsv"
-args$n = 50
+args$input = "/Users/maceci/code/mt-analysis/europeGLM/results/europeC/201106_europe5.TL.traj"
+args$metadata =  "/Users/maceci/code/mt-analysis/europeGLM/results/europeC/final_metadata.tsv"
+args$n = 200
 demes = args$demes
@@
 demes$deme[3] = "China"
+demes$division[3] = NA
+demes <- demes %>% arrange(deme)
 ```
@@
                        levels = 0:(nrow(demes) - 1),
-                       labels = sort(demes$deme)),
+                       #labels = sort(demes$deme)),
+                       labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
          date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
@@
                       levels = 0:(nrow(demes) - 1),
-                      labels = sort(demes$deme)),
+                      #labels = sort(demes$deme)),
+                      labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
          dest = ifelse(is.na(dest), as.character(src),
@@
 ```
+
+<!-- ## First introduction -->
+<!-- ```{r} -->
+<!-- # 2. Events  -->
+<!-- # 2.1. Time interval of first case (introduction) and first reported case to ECDC. -->
+
+<!-- first_ecdc <- case_data %>%  -->
+<!--   group_by(deme) %>%  -->
+<!--   filter(cumcases != 0) %>%  -->
+<!--   arrange(date) %>%  -->
+<!--   slice(1)  %>%  -->
+<!--   rename(dest = deme, min_date = date) -->
+
+<!-- firstcase <- events %>% -->
+<!--   filter(event == "M" | event == "O") %>% -->
+<!--   mutate(dest = ifelse(is.na(dest), china_deme, dest)) %>% -->
+<!--   group_by(dest, traj) %>% -->
+<!--   arrange(time) %>% -->
+<!--   slice(1) %>% -->
+<!--   ungroup %>% -->
+<!--   # Add row for Europe (all countries) -->
+<!--   bind_rows(events %>% -->
+<!--               filter(event == "M") %>% -->
+<!--               group_by(traj) %>% -->
+<!--               arrange(time) %>% -->
+<!--               slice(1)  %>% -->
+<!--               mutate(dest = "Europe")) -->
+
+<!-- dates_firstcase <-  firstcase %>% -->
+<!--   group_by(dest, min_date) %>% -->
+<!--   summarize(Dmedian = median(date),  -->
+<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
+<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
+<!--             .groups = "drop") %>% -->
+<!--   mutate(difmedian = min_date - Dmedian, -->
+<!--          diflow = min_date - Dlow, -->
+<!--          difhigh = min_date - Dhigh)  -->
+
+<!-- # 2.2. Time interval of first bith event (within region transmission) and first reported case -->
+<!-- # to ECDC. -->
+<!-- # How well did the countries detecting the first cases, there was already within region transmision? -->
+
+<!-- # First introduction: Measure expected day of introduction. -->
+<!-- firstbirth <- events %>% -->
+<!--   filter(event == "B" | event == "O") %>% -->
+<!--   group_by(dest, traj) %>% -->
+<!--   arrange(time) %>% -->
+<!--   slice(1) %>% -->
+<!--   ungroup %>% -->
+<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>%slice(1)  %>% rename(dest = deme, min_date = date), by = "dest") %>% -->
+<!--   bind_rows(events %>% -->
+<!--               filter(dest != china_deme, event == "B") %>% -->
+<!--               group_by(traj) %>% -->
+<!--               arrange(time) %>% -->
+<!--               slice(1)  %>% -->
+<!--               mutate(dest = "Europe")) -->
+
+<!-- dates_firstbirth <- firstbirth %>% -->
+<!--   group_by(dest, min_date) %>% -->
+<!--   summarize(Dmedian = median(date),  -->
+<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
+<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
+<!--             .groups = "drop") %>% -->
+<!--   mutate(difmedian = min_date - Dmedian, -->
+<!--          diflow = min_date - Dlow, -->
+<!--          difhigh = min_date - Dhigh) -->
+
+<!-- p_firstbirth <- events %>% -->
+<!--   filter(event == "B" | event == "O") %>% -->
+<!--   bind_rows(events %>% -->
+<!--               filter(dest != china_deme, event == "B") %>% -->
+<!--               group_by(traj) %>% -->
+<!--               arrange(time) %>% -->
+<!--               slice(1)  %>% -->
+<!--               mutate(dest = "Europe")) %>% -->
+<!--   group_by(dest, traj) %>% -->
+<!--   arrange(time) %>% -->
+<!--   slice(1) %>% -->
+<!--   ungroup %>% -->
+<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>%  -->
+<!--               arrange(date) %>% slice(1) %>% rename(dest = deme, min_date = date), by = "dest") %>% -->
+<!--   replace_na(list(min_date = ymd("2020-01-25"))) %>% -->
+<!--   filter(date <= min_date) %>% -->
+<!--   count(dest) %>% -->
+<!--   mutate(p = n/args$n * 100) -->
+
+<!-- r_firstcases <- dates_firstcase%>% -->
+<!--   left_join(dates_firstbirth, by = c("dest", "min_date")) %>% -->
+<!--   left_join(p_firstbirth, by = "dest") %>% -->
+<!--   replace_na(list(min_date = dates_firstcase$min_date[dates_firstcase$dest == "France"])) %>% -->
+<!--   mutate(Country = dest, -->
+<!--          `First Reported Case ECDC` = min_date, -->
+<!--          `Median date (95% CCI) first introduction` = paste0(Dmedian.x, " (", Dlow.x, ", ", Dhigh.x, ")"), -->
+<!--          `Median start date (95% CCI) within region transmission` = paste0(Dmedian.y, " (", Dlow.y, ", ", Dhigh.y, ")"), -->
+<!--          `Percentage of trajectories within region transmission when first reported case` = paste0(p, "%")) %>% -->
+<!--   select(c("Country", "First Reported Case ECDC", "Median date (95% CCI) first introduction", "Median start date (95% CCI) within region transmission", -->
+<!--            "Percentage of trajectories within region transmission when first reported case")) -->
+
+<!-- # 2.1. Time interval of first outcoming migration from a country compared to first reported case -->
+<!-- # In which percentage of trajectories, transmission from that country would have been avoided if -->
+<!-- # any ways of outcoming migrations were closed the day of first reported case. -->
+
+<!-- firstmig <- events %>% -->
+<!--   filter(event == "M") %>% -->
+<!--   group_by(src, traj) %>% -->
+<!--   arrange(time) %>% -->
+<!--   slice(1) %>% -->
+<!--   ungroup %>% -->
+<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>% slice(1)  %>% rename(src = deme, min_date = date), by = "src")  -->
+
+<!-- dates_firstmig <- firstmig%>% -->
+<!--   group_by(src, min_date) %>% -->
+<!--   summarize(Dmedian = median(date),  -->
+<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
+<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
+<!--             .groups = "drop") %>% -->
+<!--   mutate(difmedian = min_date - Dmedian, -->
+<!--          diflow = min_date - Dlow, -->
+<!--          difhigh = min_date - Dhigh)  -->
+
+<!-- p_firstmig <- events %>% -->
+<!--   filter(event == "M") %>% -->
+<!--   group_by(src, traj) %>% -->
+<!--   arrange(time) %>% -->
+<!--   slice(1) %>% -->
+<!--   ungroup %>% -->
+<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>% slice(1)  %>% rename(src = deme, min_date = date), by = "src") %>% -->
+<!--   filter(date >= min_date) %>% -->
+<!--   count(src) %>% -->
+<!--   mutate(p = n/args$n * 100) -->
+
+<!-- r_firstmig <- dates_firstmig %>% -->
+<!--   left_join(p_firstmig, by = "src") %>% -->
+<!--   mutate(Country = src, -->
+<!--          `First Reported Case ECDC` = min_date, -->
+<!--          `Median date (95% CCI) first outgoing migration` = paste0(Dmedian, " (", Dlow, ", ", Dhigh, ")"), -->
+<!--          `Percentage of trajectories first migration after first reported case` = paste0(p, "%")) %>% -->
+<!--   select(-one_of(c(colnames(dates_firstmig), colnames(p_firstmig)))) -->
+
+<!-- # Measure the time between the first migration into a country and the first birth in the country.  -->
+<!-- # Or the first outcoming mirgration from the country and first birth.  -->
+<!-- # This could be interesting to say if we should focus or not the screening and testing  -->
+<!-- # capacities to detect incoming migrations or a more general strategy to find cases in  -->
+<!-- # the population. Is it different for each country? -->
+
+<!-- firsts <- firstcase %>% -->
+<!--   select(dest, date, min_date) %>% -->
+<!--   rename(deme = dest, introduction = date) %>% -->
+<!--   left_join(firstbirth %>% select(date, dest) %>% rename(deme = dest, birth = date), by = "deme") %>% -->
+<!--   left_join(firstmig %>% select(date, src) %>% rename(deme = src, mig = date), by = "deme") %>% -->
+<!--   filter(deme != "Europe") %>% -->
+<!--   mutate(dif1 = birth - introduction, -->
+<!--          dif2 = mig - introduction) %>% -->
+<!--   group_by(deme) %>% -->
+<!--   summarize(dif1median = median(dif1),  -->
+<!--             dif1low    = quantile(dif1, 0.025, type = 1),  -->
+<!--             dif1high   = quantile(dif1, 0.975, type = 1), -->
+<!--             dif2median = median(dif2),  -->
+<!--             dif2low    = quantile(dif2, 0.025, type = 1),  -->
+<!--             dif2high   = quantile(dif2, 0.975, type = 1), -->
+<!--             .groups = "drop")  -->
+
+<!-- r_firsts <- firsts %>% -->
+<!--   mutate(Country = deme, -->
+<!--          `Days between introduction and first within region transmision` = paste0(dif1median, " (", dif1low, ", ", dif1high, ")"), -->
+<!--          `Days between introduction and first outgoing migration` = paste0(dif2median, " (", dif2low, ", ", dif2high, ")")) %>% -->
+<!--   select(-one_of(colnames(firsts))) %>% -->
+<!--   left_join(r_firstcases, by = "Country") %>% -->
+<!--   left_join(r_firstmig, by = c("Country", "First Reported Case ECDC")) %>% -->
+<!--   select(1,4,5,6,8,2,3,7,9) -->
+
+<!-- write_tsv(r_firsts, paste0(args$output_table, "02.tsv")) -->
+
+<!-- ``` -->

### 2020-12-23 scripts/predictor_matrices.Rmd
restructuting

--- 2020-12-16 workflow/scripts/predictor_matrices.Rmd
@@
 ---
-title: "GLM Predictor Matrices"
+title: "GLM predictor Matrices"
 subtitle: "Analyses Europe6 and Europe7"
@@
 $$
-m_{ij}^{e} = c \exp(\beta_{pas} \delta_{pas} x_{ij.pas}^e + \beta_{pop} \delta_{pop }x_{ij.pop} + \beta_{dist} \delta_{dist}x_{ij.dist})
+m_{ij}^{e} = c \exp(\beta_{pas} \delta_{pas} x_{ij.pas}^e + \beta_{popsrc} \delta_{popsrc}x_{ij.popsrc} + \beta_{popdes} \delta_{popdes}x_{ij.popdes} + \beta_{dist} \delta_{dist}x_{ij.dist})
 $$
 
-where $\boldsymbol{\beta}$ are the coefficients for the predictors that can be between $-\infty$ and $\infty$ (prior  $\mathcal{N}(0,4)$) and $\boldsymbol{\delta}$ are the indicators that can be 0 or 1 and denote if a predictor contributes at all.
-
-Analysis Europe 6 without indicator variables, Europe 7 with indicator variables, same predictors.
+where $\boldsymbol{\beta}$ are the coefficients for the predictors that can be between $-\infty$ and $\infty$ (prior  $\mathcal{N}(0,4)$, random walk operator).
+The indicator $\boldsymbol{\delta}$ can be 0 or 1 and denote if a predictor contributes at all (Bit Flip Operator).
 
@@
 
+Daily number of passengers betwwen demes.
 Passengers carried data from Eurostats.
@@
 
-# These datasets are dowload from eurostats 
-intraUE_data <- "./data/avia_paincc.tsv"
-extraUE_data <- "./data/avia_paexcc.tsv"
-
-flight_df <- get_flightData(intraUE_data, extraUE_data, countries = demes)
+flight_df <- get_flightData(countries = demes)
 
 flight_df_3e <- flight_df %>%
-  filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
-  mutate(EPOCH = case_when(
-    MONTH == 2 ~ 2,
-    MONTH == 3 ~ 3,
+  filter((month(time) %in% 1:3 & year(time) == 2020) | (month(time) == 12 & year(time) == 2019)) %>%
+  mutate(epoch = case_when(
+    month(time) == 2 ~ 2,
+    month(time) == 3 ~ 3,
     TRUE ~ 1),
-    PRED = 1) %>%
-  rename(SRC = GEO, DEST = PARTNER) %>%
-  group_by(SRC, DEST, EPOCH, PRED) %>%
+    pred = 1) %>%
+  rename(src = geo, dest = partner) %>%
+  group_by(src, dest, epoch, pred) %>%
   summarise(N = n(),
-         EVALUE = sum(DVALUE/N)) %>%
+         evalues = sum(dvalues/N)) %>%
   ungroup %>%
   # Transformations
-  mutate(LVALUE = log(EVALUE + 1),
-         SLVALUE = scale(LVALUE)) %>%
-  arrange(EPOCH, factor(SRC, levels = demes), factor(DEST, levels = demes))
-
-#write(slv_flightMatrix_3e, "files/flight_matrix_3e.csv", sep = ",")
+  mutate(lvalues = log(evalues + 1),
+         slvalues = scale(lvalues)) %>%
+  arrange(epoch, factor(src, levels = demes), factor(dest, levels = demes))
 ```
@@
   flight_df_3e %>%
-  filter(EPOCH == i) %>%
-  select(SRC, DEST, SLVALUE) %>%
-  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
+  filter(epoch == i) %>%
+  select(src, dest, slvalues) %>%
+  pivot_wider(names_from = "dest", values_from = "slvalues") %>%
   mutate_if(is.numeric, ~round(.,2)) %>%
-  column_to_rownames("SRC") %>%
+  column_to_rownames("src") %>%
   select(all_of(unname(demes)))
@@
 # Fligh data plot
-ggplot(flight_df_3e, aes(x = EPOCH, y = EVALUE, color = DEST)) +
+ggplot(flight_df_3e, aes(x = epoch, y = evalues, color = dest)) +
   geom_line() +
@@
   scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
-  facet_grid(~SRC) +
+  facet_grid(~src) +
   xlab("Epoch") +
@@
   arrange(name) %>%
-  rename(POP = 2) %>%
-  mutate(SLVALUE = scale(log(POP)))
+  rename(popvalue = 2) %>%
+  mutate(slvalues = scale(log(popvalue)))
 pop_df <- population %>%
-  mutate(n = 6, SRC = demes) %>%
+  mutate(n = 6, src = demes) %>%
   uncount(n) %>%
-  mutate(DEST = rep(demes, 6),
-         POP = ifelse(SRC == DEST, NA, POP),
-         SLVALUE = ifelse(SRC == DEST, NA, SLVALUE)) %>%
+  mutate(dest = rep(demes, 6),
+         popvalue = ifelse(src == dest, NA, popvalue),
+         slvalues = ifelse(src == dest, NA, slvalues)) %>%
   select(-name) %>%
-  filter(!is.na(POP))
-pop_df_3e <- bind_rows(replicate(3, pop_df %>% mutate(PRED = 2), simplify = FALSE)) %>% 
-  mutate(EPOCH = rep(1:3, each = 30, length = 90)) %>%
+  filter(!is.na(popvalue))
+pop_df_3e <- bind_rows(replicate(3, pop_df %>% mutate(pred = 2), simplify = FALSE)) %>% 
+  mutate(epoch = rep(1:3, each = 30, length = 90)) %>%
   bind_rows(bind_rows(replicate(3, pop_df %>% 
-                        arrange(factor(DEST, levels = demes),
-                                factor(SRC, levels = demes)) %>%
-                          mutate(TEMP = SRC,
-                                 SRC = DEST,
-                                 DEST = TEMP,
-                                 PRED = 3),
+                        arrange(factor(dest, levels = demes),
+                                factor(src, levels = demes)) %>%
+                          mutate(tmp = src,
+                                 src = dest,
+                                 dest = tmp,
+                                 pred = 3),
                       simplify = FALSE)) %>%
-  mutate(EPOCH = rep(1:3, each = 30, length = 90)))
+  mutate(epoch = rep(1:3, each = 30, length = 90)))
 ```
@@
 ```{r, echo=FALSE}
-pop_matrices <- lapply(c("DEST", "SRC"), function (x) {
+pop_matrices <- lapply(c("dest", "src"), function (x) {
   pop_df %>%
-  select(-POP) %>%
-  pivot_wider(names_from = all_of(x), values_from = "SLVALUE") %>%
+  select(-popvalue) %>%
+  pivot_wider(names_from = all_of(x), values_from = "slvalues") %>%
   mutate_if(is.numeric, ~round(., 2)) %>%
@@
 dist_df <- dist_df%>%
-  mutate(DIST = distHaversine(select(dist_df, longitude.x, latitude.x),
+  mutate(dist = distHaversine(select(dist_df, longitude.x, latitude.x),
                               select(dist_df, longitude.y, latitude.y))) %>%
   group_by(deme.x, deme.y) %>%
-  summarise(DIST = mean(DIST), .groups = "drop") %>%
+  summarise(dist = mean(dist), .groups = "drop") %>%
   filter(deme.x != deme.y) %>%
-  rename(SRC = deme.x, DEST = deme.y) %>% 
-  mutate(SLVALUE = scale(log(DIST))) %>%
-  arrange(factor(SRC, levels = demes), factor(DEST, levels = demes))
+  rename(src = deme.x, dest = deme.y) %>% 
+  mutate(slvalues = scale(log(dist))) %>%
+  arrange(factor(src, levels = demes), factor(dest, levels = demes))
 dist_df_3e <- bind_rows(replicate(3, dist_df, simplify = FALSE)) %>%
-  mutate(EPOCH = rep(1:3, each = 30, length = 90), PRED = 4)
+  mutate(epoch = rep(1:3, each = 30, length = 90), pred = 4)
 ```
@@
 dist_matrix <- dist_df %>%
-  select(-DIST) %>%
-  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
+  select(-dist) %>%
+  pivot_wider(names_from = "dest", values_from = "slvalues") %>%
   mutate_if(is.numeric, ~round(., 2)) %>%
-  column_to_rownames("SRC") %>%
+  column_to_rownames("src") %>%
   select(all_of(unname(demes)))
@@
 
+## Daily flight passengers by source population
+
+Daily number of passengers betwwen demes normalized by the population of the source deme.
+Passengers carried data from Eurostats.
+Reported _departures_ values are taken for European countries (recommendation from Eurostats to avoid duplicates).
+For China, arrivals and departures values are considered.
+
+- **Epoch 1:** December 2019 - January 2020 (Wuhan lockdown 23 Jan)
+- **Epoch 2:** February 2020 
+- **Epoch 3:** March 2020
+
+\tiny
+```{r, echo = FALSE, message = FALSE, warning = FALSE}
+demes <- c(China = "CN", France = "FR", Germany = "DE", Italy = "IT", OtherEuropean = "OE", Spain = "ES")
+
+flight_df <- get_flightData(countries = demes)
+
+flight_df_3e_n <- flight_df %>%
+  filter((month(time) %in% 1:3 & year(time) == 2020) | (month(time) == 12 & year(time) == 2019)) %>%
+  mutate(epoch = case_when(
+    month(time) == 2 ~ 2,
+    month(time) == 3 ~ 3,
+    TRUE ~ 1),
+    pred = 1) %>%
+  rename(src = geo, dest = partner) %>%
+  left_join(population %>% mutate(src = demes[name]), by = "src") %>%
+  group_by(src, dest, epoch, pred, popvalue) %>%
+  summarise(N = n(),
+         evalues = sum(dvalues/N)) %>%
+  mutate(nevalues = evalues/(popvalue*1000)) %>%
+  ungroup %>%
+  # Transformations
+  mutate(#lvalues = log(nevalues + 1),
+         slvalues = scale(nevalues)) %>%
+  arrange(epoch, factor(src, levels = demes), factor(dest, levels = demes))
+```
+
+```{r, echo = FALSE}
+flight_matrices <- lapply(1:3, function (i) {
+  flight_df_3e_n %>%
+  filter(epoch == i) %>%
+  select(src, dest, slvalues) %>%
+  pivot_wider(names_from = "dest", values_from = "slvalues") %>%
+  mutate_if(is.numeric, ~round(.,2)) %>%
+  column_to_rownames("src") %>%
+  select(all_of(unname(demes)))
+})
+
+knitr::kable(flight_matrices, booktabs = TRUE, caption = "Flight passengers between demes, epochs 1-3")
+```
+
+## Daily flight passengers
+
+```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 3}
+set_plotopts()
+dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
+dcolors <- dcolors[1:length(demes)]
+names(dcolors) <- demes[sort(labels(demes))]
+
+# Fligh data plot
+ggplot(flight_df_3e_n, aes(x = epoch, y = nevalues, color = dest)) +
+  geom_line() +
+  geom_point(size = 1) +
+  #scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
+  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
+  facet_grid(~src) +
+  xlab("Epoch") +
+  ylab("# Daily Flight Passengers") + 
+  #theme(axis.title.x = element_blank()) + 
+  scale_color_manual(values = dcolors, name = "Destination")
+```
+
 ## Covariate list BEAST BDMM-Prime
@@
   bind_rows(dist_df_3e) %>%
-  arrange(PRED, desc(EPOCH), factor(SRC, levels = demes), factor(DEST, levels = demes)) %>%
-  select(PRED, EPOCH, SRC, DEST, SLVALUE)
+  arrange(pred, desc(epoch), factor(src, levels = demes), factor(dest, levels = demes)) %>%
+  select(pred, epoch, src, dest, slvalues)
 
@@
 
-mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(SLVALUE), ncol = 5, byrow = TRUE))
+mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(slvalues), ncol = 5, byrow = TRUE))
 write_csv(mpred_3e, "./files/covariates_4p_3e.csv", col_names = FALSE)
+```
+
+```{r, echo = FALSE}
+pred_3e <- bind_rows(flight_df_3e_n, pop_df_3e) %>%
+  bind_rows(dist_df_3e) %>%
+  arrange(pred, desc(epoch), factor(src, levels = demes), factor(dest, levels = demes)) %>%
+  select(pred, epoch, src, dest, slvalues)
+
+if (nrow(pred_3e) != 360) warning("Something went wrong, the size of the combined predictors is ", nrow(pred_3e))
+
+mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(slvalues), ncol = 5, byrow = TRUE))
+write_csv(mpred_3e, "./files/covariates_4p_3e_n.csv", col_names = FALSE)
 ```
@@
 
-Covariate list saved as `files/covariates_4p_3e.csv`
+Covariate list saved as `files/covariates_4p_3e.csv` and `files/covariates_4p_3e_n.csv` (normalized by population passenger data).
+
+## Analyses
+
+- **Europe 4:** GLM with daily flight data as predictor, constant over all the analysis time.
+- **Europe 5:** GLM with daily flight data as predictor, 3 epochs.
+- **Europe 5b:** GLM with daily flight data normalized by population of the deme, 3 epochs.
+- **Europe 6** GLM with the 4 predictors described here, without indicator variables 
+- **Europe 6b** GLM with the 4 predictors described here (fligh data normalized by population), without indicator variables 
+- **Europe 7** with indicator variables, same predictors than eu6.
+<!-- - **Europe 7b** with indicator variables, same predictors than eu7. -->

### 2020-12-23 scripts/utils.R

@@ -1,0 +1,195 @@
+##-----------------------------------------------------
+## Util functions for analysis workflow
+## 
+## 2020-10-14 Cecilia Valenzuela
+##------------------------------------------------------
+
+library(countrycode)
+library(ggpubr)
+library(lubridate)
+
+set_plotopts <- function() {
+  theme_set(theme_minimal())
+  theme_update(#strip.background = element_rect(fill = "grey95", color = "grey95", size = 1),
+               panel.grid = element_blank(),
+               panel.background = element_rect(fill = "white", colour = "grey80"),
+               axis.ticks = element_line(),
+               legend.position = "bottom",
+               legend.box="vertical",
+               strip.background = element_rect(fill = "grey80", colour = "grey80"),
+               strip.text = element_text(color = "grey30"),
+               #legend.title = element_text(size = 9, face = "bold"),
+               axis.text.x = element_text(size = 8),
+               axis.text.y = element_text(size = 8),
+               axis.title.y = element_text(size = 9, face = "bold", margin = margin(r = 10)),
+               axis.title.x = element_text(size = 9, face = "bold", margin = margin(t = 10)))
+}
+
+
+
+get_cases <- function(demes, from = NA, to = NA) {
+  # Get case data information from ECDC for specific demes (countries and regions)
+  # Set from and to dates with values from the function call instead of the ones from the demes info
+  cat("\nLoading case data counts from ECDC...\n")
+  #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
+  #else case_data <- get_dataECDC()
+  case_data <- get_dataECDC()
+  
+  if (!is.na(from)) demes$min_date <- from
+  if (!is.na(to)) demes$max_date <- to
+  
+  # Filter case count data according to deme info
+  case_data_list <- mapply(get_cases_deme, list(case_data), demes$deme,
+                           demes$region, demes$country, demes$division,
+                           demes$exclude_country, demes$exclude_division,
+                           demes$min_date, demes$max_date)
+  case_data_dfs <- lapply(1:dim(case_data_list)[2], function(i) as.data.frame(case_data_list[,i]))
+  case_data <- bind_rows(case_data_dfs)
+  cat("done!")
+  return(case_data)
+}
+
+
[…]

### 2020-12-23 utils/ECDCvsJH.R

@@ -1,35 +1,0 @@
-library(coronavirus)
-library(dplyr)
-
-jh <- coronavirus%>%filter(province=="Hubei" & type=="confirmed" & as.Date(date) <= "2020-03-08")%>%
-  mutate(source = "jh")%>%
-  select(country, date, cases, source)
-ecdc <- read.csv("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", na.strings = "", fileEncoding = "UTF-8-BOM")%>%
-  dplyr::rename(country = countriesAndTerritories,
-                region = continentExp)%>%
-  dplyr::mutate(date = as.Date(dateRep, "%d/%m/%Y"),
-                source = "ecdc")%>%
-  dplyr::filter(date <= as.Date("2020-03-08") & country == "China")%>%
-  dplyr::select(country, date, cases, source)
-
-summarise(jh, total = sum(cases))
-summarise(ecdc, total = sum(cases))
-
-# No significant change when plotting case count in log scale. 
-# We continue to use ECDC data for China instead of JH data for Hubei
-# knowing that is a overestimation of case count data in Hubei.
-# Easier to work with just one case count dataset and reported case counts 
-# earlied in ECDC dataset
-
-cases_hubei <- dplyr::bind_rows(ecdc, jh)%>%
-  dplyr::arrange(date)%>%
-  dplyr::group_by(source)%>%
-  dplyr::mutate(sumcases = cumsum(cases))
-
-ggplot(cases_hubei) +
-  geom_line(aes(x=date, y=sumcases, linetype=source))+
-  labs(title="Hubei case count")+
-  scale_y_log10(labels= trans_format("log10", math_format(10^.x)))
-
-
-  

### 2020-12-23 utils/check subsamble.R

@@ -1,32 +1,0 @@
-library(dplyr)
-
-SEQS <- "200914_europe1/results/subsampled_alignment.fasta"
-SEQS <- "~/subsampled_alignment.fasta"
-alignment <- ape::read.FASTA(file = SEQS)
-seqs <- names(alignment)
-
-
-INCLUDE <- "200914_europe1/files/include.txt"
-include <- readLines(INCLUDE)
-
-all(include%in%seqs)
-
-include[!include%in%seqs]
-
-METADATA <- "200914_europe1/data/200911_metadata.tsv"
-metadata <- read.delim(file = METADATA)%>%
-  dplyr::mutate(date=as.Date(date))%>%
-  dplyr::filter(strain%in%seqs)
-
-all(metadata$date < as.Date("2020-03-09"))
-
-metadata_hubei <- metadata%>%
-  dplyr::filter(country=="China")
-
-all(metadata_hubei$date < as.Date("2020-01-23"))
-
-summary <- metadata%>%
-  dplyr::count(country, sort=TRUE)
-View(metadata%>%filter(country=="Italy")%>%
-       mutate(include=strain%in%include))
-

### 2020-12-23 utils/downsample_seq.R

@@ -1,29 +1,0 @@
-library(ape)
-library(tidyverse)
-
-ALIGNMENT <- "200904_sarah/sarah_europe_demes.fasta"
-OUTPUT <- "200910_ds_sarah/200910_ds_sarah.fasta"
-SEED <- 1234
-
-set.seed(seed = SEED)
-
-alignment <- ape::read.FASTA(file = ALIGNMENT)
-
-labs <- labels(alignment)
-demes <- unlist(lapply(labs, function(x) str_split(x, "\\|")[[1]][2]))
-df <- tibble::tibble(labs, demes)%>%
-  tibble::rowid_to_column("id")
-
-df_s <- tibble::tibble()
-for (deme in unique(df$demes)){
-  ids <- df%>%
-    dplyr::filter(deme == demes)%>%
-    dplyr::pull(id)
-  rows <- df%>%dplyr::filter(id%in%sample(ids, 5))
-  df_s <- dplyr::bind_rows(df_s, rows)
-}
-
-alignment_s <- alignment[df_s$id]
-
-ape::write.FASTA(x = alignment_s, file = OUTPUT)
-

### 2020-12-23 utils/explore_seqs.R

@@ -1,171 +1,0 @@
-library(tidyverse)
-library(RColorBrewer)
-library(lubridate)
-
-source("scripts/utils.R") 
-subsample_test <- function(metadata, include = NA, exclude = NA,
-                      region_name = NA, country_name = NA, division_name = NA,
-                      exclude_country = NA, exclude_division = NA,
-                      from, to, seq_per_deme, seed, prob = "time") {
-  set.seed(seed)
-  # Read files
-  if (!is.na(exclude)) {
-    exclude <- readLines(include)
-    metadata <- filter(metadata, !strain %in% exclude)
-  }
-  # Filter metadata
-  metadata_deme <- metadata %>%
-    filter(region == region_name,
-           if (is.na(country_name)) TRUE else country %in% country_name,
-           if (is.na(division_name)) TRUE else division %in% division_name,
-           if (all(is.na(exclude_country))) TRUE else !country %in% exclude_country,
-           if (all(is.na(exclude_division))) TRUE else !division %in% exclude_division,
-           date >= as.Date(from),
-           date <= as.Date(to))
-  
-  if (nrow(metadata_deme) == 0) {
-    warning("No sequences for these specifications")
-    return()
-  }
-  
-  if (prob == "deaths") to <- ymd(to) + 14
-  
-  # Get case counts for deme
-  cases_deme <- get_cases_deme("deme", region_name, country_name, division_name, 
-                               exclude_country, exclude_division, 
-                               from, to)  
-  
-  # Account for lag cases - deaths
-  if (prob == "deaths") cases_deme <- cases_deme %>% mutate(date = date - 14)
-  
-  if (prob %in% c("cases", "cases_and_location", "deaths")){
-    # Compute weights and probabilities for each date
-    by_date <- metadata_deme %>%
-      count(date, country) %>%
-      left_join(cases_deme, by = c("date", "country")) %>%
-      replace_na(list(cases = 0, cumcases = 0, deaths = 0, cumdeaths = 0)) %>%
-      mutate(p_case = (cases + 1)/sum(cases),
-             p_death = (deaths + 1)/sum(deaths),
-             w_date = sum(n)/n,
-             p_date_cases = (p_case * w_date)/sum(p_case * w_date),
[…]

### 2020-12-23 utils/map.R

@@ -1,38 +1,0 @@
-# Plot Europe map
-
-library(tmap)
-library(sf)
-library(rnaturalearth)
-library(RColorBrewer)
-library(maps)
-
-cases <- get_casesECDC(from = "2019-12-01", to = "2020-03-08")%>%
-  filter(date == as.Date("2020-03-08")) %>%
-  rename(cases14 = 12)
-
-Europe <- ne_countries(scale = 'medium', type = 'countries', returnclass = 'sf', 
-                       continent="Europe") %>%
-  rename(country = sovereignt) %>% 
-  left_join(cases, by = 'country')%>%
-  st_crop(xmin = -20, xmax = 40, ymin = 30, ymax = 73)
-ggplot() + 
-  geom_sf(data = Europe, aes(fill = log(sumcases))) +
-  geom_point(data = df %>% filter(region == "Europe"), 
-             aes(y = N, x = W, color = deme), alpha = 0.4)+
-  ggsci::scale_color_npg() +
-  #scale_fill_distiller(palette = "Spectral", limits = c(0, 10))
-  #paletteer::scale_fill_paletteer_c("nord::aurora")#scale_fill_distiller()
-
-# world <- ne_countries(scale = 'medium', type = 'map_units', returnclass = 'sf') %>%
-#   st_crop(y = c(xmin = -10, xmax = 120, ymin = 35, ymax = 70))
-# ggplot() + geom_sf(data=world) +
-#   geom_point(data = df, aes(y = N, x = W, color = deme))
-# 
-
-df <- left_join(metadata_demes, latitudes %>% filter(area == "location"))
-
-tm_shape(Europe) +
-  tm_fill() +
-  tm_borders()+
-  geom_point(aes(x=0,y=40))
-  #geom_point(aes(y = df$N[1:10], x = df$W[1:10]))

### 2020-12-23 utils/processing.sh

@@ -1,60 +1,0 @@
-#!/bin/sh
-WORKDIR="/Users/maceci/code/mt-analysis/"
-ANALYSIS="200910_dsEurope0"
-N=10
-
-# Results Processing 
-# Analysis is run on Euler and entire analysis folder move back to local folder in /mt-analysis
-mkdir $WORKDIR/$ANALYSIS/pResults
-
-
-/Applications/BEAST\ 2.6.3/bin/loganalyser /Users/maceci/code/mt-analysis/200910_dsEurope0/pResults/200910_dsEurope0.log > ~/output.txt 
-
-# 1. Combine log files with Log Combiner 
-# Check first convergence in Tracer
-LOGCOMBINER="/Applications/BEAST 2.6.3/bin/logcombiner"
-$LOGCOMBINER \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.1.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.2.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.3.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.4.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.5.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.6.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.7.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.8.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.9.log \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.10.log \
--o $WORKDIR/pResults/${ANALYSIS}.log \
--b 10 
-
-# 3. Log summary table
-# Visualize combined log file and export data table from Tracer
-# Transform table to .md
-Rscript $WORKDIR/txttomd.R \
---input $WORKDIR/$ANALYSIS/pResults/${ANALYSIS}_logSummary 
-
-
-# 4. Combine typed node trees with Log Combiner
-$LOGCOMBINER \
--log $WORKDIR/$ANALYSIS/rResults/${ANALYSIS}.1.typed.node.trees \
[…]

### 2020-12-23 utils/traj_processing.R

@@ -1,285 +1,0 @@
-##-----------------------------------------------------
-## Script for plotting and manipulating trajectory data 
-## using functions from BDMM-Prime project from Tim Vaughan 
-## to parse trajectories.
-## Script for snakemake workflow.
-## 
-## 2020-10-12 Cecilia Valenzuela
-##------------------------------------------------------
-
-library(argparse)
-library(tidyverse)
-library(scales)
-library(ggpubr)
-library(ggthemes)
-source("~/code/mt-beast-dev/BDMM-Prime/scripts/trajProcessing.R")
-
-## Parser
-parser <- argparse::ArgumentParser()
-parser$add_argument("--input", nargs="+",
-                    help="txt file with the names of the trajectory documents to be included in the analysis")
-parser$add_argument("--burnin", type = "double",
-                    help = "Burning fraction for trajectory files")
-parser$add_argument("--mrs", type = "character",
-                    help = "Most recent sample date file")
-parser$add_argument("--output_figure", type = "character",
-                    help = "Output path for the figures")
-args <- parser$parse_args()
-
-INPUT <- read_lines(args$input)
-BURNIN <- args$burnin
-MRS <- read(args$mrs)
-OUTPUT_FIGURE <- args$output_figure
-
-print(paste("Trajectory file:", INPUT))
-print(paste("Burning fraction:", BURNIN))
-print(paste("Most recent sample date in the alignment: ", MRS))
-print(paste("output figure:", OUTPUT_FIGURE))
-
-## Read trajectories and events information
-states <- data.frame()
-events <- data.frame()
-for (filename in INPUT){
-  df <- loadTrajectories(filename, burninFrac=BURNIN, subsample=100)
-  states <- states%>%dplyr::bind_rows(df$states%>%dplyr::mutate(file = filename))
-  events <- events%>%dplyr::bind_rows(df$events%>%dplyr::mutate(file = filename))
-}
-states$type <- factor(states$type, levels=c(0,1,2,3,4), labels=c("France", "Germany", "Hubei", "Italy", "Other European"))
-states$date <- as.Date(-states$age*365, origin=as.Date(MRS))
[…]

### 2020-12-23 workflow/plotscasesandflights.R

@@ -1,34 +1,0 @@
-library(tidyverse)
-
-source("./scripts/utils.R")
-
-demes <- read.csv("/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/workflow/files/demes.csv")
-cases <- get_cases(demes, to="2020-03-18")
-  
-
-ggplot(cases) +
-  geom_line(aes(x = date, y = cumcases, color = deme)) +
-  ggsci::scale_color_jco()
-  
-ggplot(cases) +
-  geom_line(aes(x = date, y = cumcases/population_deme, color = deme)) +
-  ggsci::scale_color_jco()
-
-pops <- cases %>%
-  select(deme, population_deme) %>%
-  mutate(GEO = as.character(deme)) %>%
-  distinct
-pops$GEO[pops$deme == "Hubei"] <- "Hubei-China"
-
-flightsDic$df %>% filter (!is.na(VALUE)) %>%
-  mutate(svalue = scale(log(VALUE))) %>%
-  ggplot() +
-  geom_point(aes(x = GEO, y = svalue, color = PARTNER)) +
-  ggsci::scale_color_jco()
-
-flightsDic$df %>% filter (!is.na(VALUE)) %>% 
-  left_join(pops) %>%
-  mutate(svalue = scale(log(VALUE/population_deme))) %>%
-  ggplot() +
-  geom_point(aes(x = GEO, y = svalue, color = PARTNER)) +
-  ggsci::scale_color_jco()

### 2020-12-23 workflow/scripts/flight_data.R

@@ -1,241 +1,0 @@
- ## Flight data
-
-library(tidyverse)
-library(lubridate)
-source("./scripts/utils.R")
-
-# Returns dayly passengers in a given month 
-get_flightData <- function(intraUE_data, extraUE_data, countries) {
-  df_int <- read_tsv(intraUE_data) 
-  df_ext <- read_tsv(extraUE_data)
-  if ("OE" %in% countries) countries <- c(countries, "EU27_2020", "EU28", "UK")
-  
-  df <- bind_rows(df_int, df_ext) %>%
-    pivot_longer(!1, names_to = "TIME", values_to = "VALUE") %>%
-    separate(1, into = c("UNIT", "TRA_MEAS", "PARTNER", "GEO"), sep = ',') %>%
-    # Data from passengers, monthly
-    filter(UNIT == "PAS", GEO %in% countries, 
-           PARTNER %in% countries, grepl("M", TIME))
-    
-  df1 <- df %>%
-    # Add year and month
-    mutate(YEAR = as.numeric(substr(TIME,1,4)),
-           MONTH = as.numeric(substr(TIME,6,8)),
-           VALUE = as.numeric(VALUE)) %>%
-    # Select passengers carried, departures for all countries (if we consider
-    # departures and arrivals for each country we have duplicated values, one
-    # for each reporting country. Recommendation from eurostats is to take only
-    # the departures) and departures and arrivals for China
-    filter(TRA_MEAS == "PAS_CRD_DEP" | 
-             (TRA_MEAS == "PAS_CRD_ARR" & PARTNER == "CN")) %>%
-    mutate(PARTNER = ifelse(TRA_MEAS == "PAS_CRD_ARR" & PARTNER == "CN", GEO, PARTNER),
-           GEO = ifelse(TRA_MEAS == "PAS_CRD_ARR", "CN", GEO),
-           TRA_MEAS = ifelse(TRA_MEAS == "PAS_CRD_ARR", "PAS_CRD_DEP", TRA_MEAS)) %>%
-    # In 2020 UK is out of the european aggregated value
-    filter(!is.na(VALUE)) 
-  
-  # Correct for UK Brexit, EU28 -> EU27 in 2020
-  # Only for months 2020
-  
-  df2 <- df1 %>%
[…]

### 2020-12-23 workflow/scripts/utils.R

@@ -1,193 +1,0 @@
-##-----------------------------------------------------
-## Util functions for analysis workflow
-## 
-## 2020-10-14 Cecilia Valenzuela
-##------------------------------------------------------
-
-library(countrycode)
-library(ggpubr)
-
-set_plotopts <- function() {
-  theme_set(theme_minimal())
-  theme_update(#strip.background = element_rect(fill = "grey95", color = "grey95", size = 1),
-               panel.grid = element_blank(),
-               panel.background = element_rect(fill = "white", colour = "grey80"),
-               axis.ticks = element_line(),
-               legend.position = "bottom",
-               legend.box="vertical",
-               strip.background = element_rect(fill = "grey80", colour = "grey80"),
-               strip.text = element_text(color = "grey30"),
-               #legend.title = element_text(size = 9, face = "bold"),
-               axis.text.x = element_text(size = 8),
-               axis.text.y = element_text(size = 8),
-               axis.title.y = element_text(size = 9, face = "bold", margin = margin(r = 10)),
-               axis.title.x = element_text(size = 9, face = "bold", margin = margin(t = 10)))
-}
-
-
-
-get_cases <- function(demes, from = NA, to = NA) {
-  # Get case data information from ECDC for specific demes (countries and regions)
-  # Set from and to dates with values from the function call instead of the ones from the demes info
-  cat("\nLoading case data counts from ECDC...\n")
-  #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
-  #else case_data <- get_dataECDC()
-  case_data <- get_dataECDC()
-  
-  if (!is.na(from)) demes$min_date <- from
-  if (!is.na(to)) demes$max_date <- to
-  
-  # Filter case count data according to deme info
-  case_data_list <- mapply(get_cases_deme, list(case_data), demes$deme,
-                           demes$region, demes$country, demes$division,
-                           demes$exclude_country, demes$exclude_division,
-                           demes$min_date, demes$max_date)
-  case_data_dfs <- lapply(1:dim(case_data_list)[2], function(i) as.data.frame(case_data_list[,i]))
-  case_data <- bind_rows(case_data_dfs)
-  cat("done!")
-  return(case_data)
-}
-
-
[…]

### 2020-12-30 scripts/analyze_trajectories_v2.Rmd
wip dataset trajectories

@@
 args$burnin = 0.1
-args$demes = read.csv("/Users/maceci/code/mt-analysis/201014_europe2/files/demes.csv", stringsAsFactors = FALSE)
-args$input = "/Users/maceci/code/mt-analysis/europeGLM/results/europeC/201106_europe5.TL.traj"
-args$metadata =  "/Users/maceci/code/mt-analysis/europeGLM/results/europeC/final_metadata.tsv"
-args$n = 200
+args$demes = read.csv("./files/demes.csv", stringsAsFactors = FALSE)
+args$input = "./results/europeC/201106_europe5.TL.traj"
+args$metadata =  "./results/europeC/final_metadata.tsv"
+args$n = 50
 demes = args$demes
@@
 zoom_to <- c(20, 50) 
-zoom_level <- 2.6
+zoom_level <- 2.5
 target_crs <- sprintf('+proj=aeqd +lon_0=%f +lat_0=%f',
@@
                                            levels = 0:(nrow(demes) - 1),
-                                           labels = sort(demes$deme)))),
+                                           #labels = sort(demes$deme),
+                                           labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain")))),
          date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
-  mutate(date = date_model + 10)
+  mutate(date = date_model + 10,
+         src = ifelse(event == "O", "China", src),
+         dest = ) #TODO
 ```
@@
             Imedian = median(N),
-            Ilow    = quantile(N, 0.025), #HPD??
+            Ilow    = quantile(N, 0.025), #95% CI, compute HPD??
             Ihigh   = quantile(N, 0.975),
@@
          date = date_model + 10,
-         dest = ifelse(is.na(dest), as.character(src), as.character(dest)))
+         dest = ifelse(is.na(dest), as.character(src), as.character(dest))) %>%
+  ungroup %>%
+  filter(N != 0)
 
@@
 #          width = 2300, height = 3000, res = 300)
+```
+
+```{r}
+# Active cases
+gt %>%
+    # 1. Prepare data
+    filter(type == deme) %>%
+    mutate(traj = as.character(traj)) %>%
+    bind_rows(case_data %>%
+                rename(type = deme,
+                       N = cumcases) %>%
+                filter(type == deme) %>%
+                mutate(traj = "ecdc") %>%
+                select(type, traj, date, N) %>%
+                distinct()) %>%
+    # 2. Plot
+    ggplot() +
+    geom_line(aes(date, N, group = traj,
+                  color = traj == "ecdc",
+                  alpha = traj == "ecdc",
+                  linetype = traj == "ecdc",
+                  size = traj == "ecdc")) +
+    scale_color_manual(name = "",
+                       labels = c(paste0(deme, " inferred\n population trajectories"),
+                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+                                         " ECDC total case count")),
+                       values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
+    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
+    scale_linetype_manual(name = "", values = c(1, 2),
+                          labels = c(paste0(deme, " inferred\n population trajectories"),
+                                     paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+                                            " ECDC total case count"))) +
+    scale_size_manual(name = "", values = c(0.5, 1),
+                      labels = c(paste0(deme, " inferred\n population trajectories"),
+                                 paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+                                        " ECDC total case count"))) +
+    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
+                 date_labels = "%b %d") +
+    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
+    ylab("Population size") +
+    theme(legend.position = c(0.4, 0.9),
+          axis.title.x=element_blank())
+
+```
+
+```{r}
+# Cumulative cases
+events <- df$events
+
+events[events$event == "O", "src"] = 2
+events[events$event == "O", "mult"] =  1
+
+df_cases1 <- events %>%
+  group_by_at(vars(-mult)) %>%
+  summarise(N = sum(mult), .groups = "drop") %>%
+  mutate(event2 = ifelse(event == "M", paste0("IM_", src), NA),
+         event3 = ifelse(event == "M", paste0("OM_", dest), NA)) %>%
+  pivot_longer(src:dest, names_to = "role", values_to = "type") %>%
+  mutate(event = ifelse(event == "M" & role == "src", event3,
+                        ifelse(event == "M" & role == "dest", event2, event))) %>% 
+  select(-c("role", "event2", "event3")) %>%
+  filter(!is.na(type)) %>%
+  group_by_at(vars(-N)) %>%
+  summarise(value = sum(N), .groups = "drop") %>%
+  pivot_wider(names_from = event, values_from = value, values_fill = list(value = 0)) %>%
+  mutate(IM = IM_0 + IM_1 + IM_2 + IM_3 + IM_4 + IM_5,
+         OM = OM_0 + OM_1 + OM_2 + OM_3 + OM_4 + OM_5) %>%
+  group_by(traj, type) %>%
+  arrange(traj, type, time) %>%
+  mutate_at(tail(names(.), -4), list(c = cumsum)) %>%
+  mutate(activePop = O_c + B_c + IM_c - D_c - OM_c - S_c,
+         cumPop = O_c + B_c + IM_c,
+         inPop = O + B + IM,
+         outPop = D + OM + S) %>% 
+  pivot_longer(O:outPop, names_to = "var", values_to = "value")
+  # pivot_longer(O:outPop, names_to = c("value", "set"),
+  #  names_pattern = "(.*)_(.)")
+
+df_cases <- events %>%
+  group_by_at(vars(-mult)) %>%
+  summarise(N = sum(mult), .groups = "drop") %>%
+  mutate(srctmp = src, destmp = dest) %>%
+  pivot_longer(src:dest, names_to = "role", values_to = "type") %>%
+  mutate(event = ifelse(event == "M" & role == "src", "OM",
+                        ifelse(event == "M" & role == "dest", "IM", event)),
+         partner = ifelse(event == "OM", destmp,
+                        ifelse(event == "IM", srctmp, NA))) %>% 
+  select(-c("role", "srctmp", "destmp")) %>%
+  filter(!is.na(type)) %>%
+  group_by_at(vars(-N)) %>%
+  summarise(value = sum(N), .groups = "drop") %>%
+  group_by(traj, type, partner, event) %>%
+  arrange(traj, time, type) %>%
+  mutate(cumvalue = cumsum(value))
+
+mig_cum = df_cases %>%
+  filter(event %in% c("IM", "OM")) %>%
+  group_by(traj, event, time, age, type) %>%
+  summarise(value1 = sum(value), .groups = "drop") %>%
+  group_by(traj, event, type) %>%
+  arrange(traj, time, type) %>%
+  mutate(cumvalue1 = cumsum(value1),
+            partner1 = "all")
+
+inout_pop = df_cases %>% 
+  mutate(var = ifelse(event %in% c("O", "B", "IM"), "in_pop", "out_pop")) %>%
+  group_by(traj, type, var, time, age) %>%
+  summarise(value = sum(value), .groups = "drop") %>%
+  group_by(traj, type, var) %>%
+  arrange(traj, time, type) %>%
+  mutate(cumvalue = cumsum(value))
+
+active_pop = inout_pop %>%
+  pivot_wider(names_from = var, values_from = value, values_fill = list(value = 0)) %>%
+  mutate(active_pop = in_pop - out_pop) %>%
+  pivot_longer(active_pop, names_to = "var", values_to = "value") %>%
+  group_by(traj, type) %>%
+  arrange(traj, time, type) %>%
+  mutate(cumvalue = cumsum(value)) %>%
+  select(-c("in_pop", "out_pop"))
+
+df_cases <- df_cases %>%
+  mutate(src = factor(src,
+                      levels = 0:(nrow(demes) - 1),
+                      #labels = sort(demes$deme)),
+                      labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
+         dest = ifelse(is.na(dest), as.character(src),
+                       as.character(factor(dest,
+                                           levels = 0:(nrow(demes) - 1),
+                                           #labels = sort(demes$deme),
+                                           labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain")))))
+
+ge %>%
+    # 1. Prepare data
+    filter((dest == deme & event == "B") | (dest == deme & event == "M")) %>%
+  filter(traj == 1) %>%
+    mutate(traj = as.character(traj)) %>%
+    bind_rows(case_data %>%
+                rename(dest = deme,
+                       N = cumcases) %>%
+                filter(dest == deme) %>%
+                mutate(traj = "ecdc") %>%
+                select(dest, traj, date, N) %>%
+                distinct()) %>%
+    # 2. Plot
+    ggplot() +
+    geom_line(aes(date, N, group = traj,
+                  color = traj == "ecdc",
+                  alpha = traj == "ecdc",
+                  linetype = traj == "ecdc",
+                  size = traj == "ecdc")) +
+    scale_color_manual(name = "",
+                       labels = c(paste0(deme, " inferred\n population trajectories"),
+                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+                                         " ECDC total case count")),
+                       values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
+    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
+    scale_linetype_manual(name = "", values = c(1, 2),
+                          labels = c(paste0(deme, " inferred\n population trajectories"),
+                                     paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+                                            " ECDC total case count"))) +
+    scale_size_manual(name = "", values = c(0.5, 1),
+                      labels = c(paste0(deme, " inferred\n population trajectories"),
+                                 paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+                                        " ECDC total case count"))) +
+    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
+                 date_labels = "%b %d") +
+    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
+    ylab("Population size") +
+    theme(legend.position = c(0.4, 0.9),
+          axis.title.x=element_blank())
+
 ```

### 2020-12-30 scripts/analyze_trajectories_v2.Rmd
wip dataset trajectories

--- 2020-12-23 scripts/analyze_trajectories_v2.Rmd
@@
   group_by(deme) %>%
-  mutate(cumcases = cumsum(cases),
-         cumdeaths = cumsum(deaths)) %>%
+  rename(value = cases) %>%
+  mutate(cumvalue = cumsum(value)) %>%
   group_by(deme, date) %>%
-  mutate(cumcases = max(cumcases),
-         cumdeaths = max(cumdeaths)) %>%
-  select(deme, date, cumcases, cumdeaths, cases)  %>%
-  distinct_at(vars(deme, date, cumcases, cumdeaths), .keep_all = TRUE)
+  mutate(cumvalue = max(cumvalue)) %>%
+  select(deme, date, cumvalue, value)  %>%
+  distinct_at(vars(deme, date, cumvalue), .keep_all = TRUE)
 
@@
 
-case_data_pop <- case_data %>% left_join(population, by="deme") %>% mutate(cas100 = cumcases * 1e5 / (pop * 1e3))
+case_data_pop <- case_data %>% left_join(population, by="deme") %>% mutate(ECDCcas100 = cumcases * 1e5 / (pop * 1e3)) 
 ```
@@
 
+# events <- df$events %>%
+#   mutate(src = factor(src,
+#                       levels = 0:(nrow(demes) - 1),
+#                       #labels = sort(demes$deme)),
+#                       labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
+#          dest = ifelse(is.na(dest), as.character(src),
+#                        as.character(factor(dest,
+#                                            levels = 0:(nrow(demes) - 1),
+#                                            #labels = sort(demes$deme),
+#                                            labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain")))),
+#          date_model = date(date_decimal(decimal_date(ymd(MRS)) - age)))
+
+# Do this in a function, useful also for Claire
 events <- df$events %>%
@@
                       labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
-         dest = ifelse(is.na(dest), as.character(src),
-                       as.character(factor(dest,
+         dest = (factor(dest,
                            levels = 0:(nrow(demes) - 1),
-                                           labels = sort(demes$deme)))),
-         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
-  mutate(date = date_model + 10)
+                           #labels = sort(demes$deme),
+                           labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain"))))
+events[events$event == "O", "src"] = "China"
+events[events$event == "O", "mult"] =  1
+
+df_traj0 <- events %>%
+  mutate(date = date(date_decimal(decimal_date(mrs) - age))) %>%
+  select(-time, -age) %>%
+  group_by_at(vars(-mult)) %>%
+  summarise(N = sum(mult), .groups = "drop") %>%
+  mutate(srctmp = src, destmp = dest) %>%
+  pivot_longer(src:dest, names_to = "role", values_to = "deme") %>%
+  mutate(event = ifelse(event == "M" & role == "src", "OM",
+                        ifelse(event == "M" & role == "dest", "IM", event)),
+         partner = ifelse(event == "OM", as.character(destmp),
+                        ifelse(event == "IM", as.character(srctmp), NA))) %>% 
+  select(-c("role", "srctmp", "destmp")) %>%
+  filter(!is.na(deme)) %>%
+  group_by_at(vars(-N)) %>%
+  summarise(value = sum(N), .groups = "drop") %>%
+  group_by(traj, deme, partner, event) %>%
+  arrange(traj, date, deme) %>%
+  mutate(cumvalue = cumsum(value)) %>%
+  rename(var = event)
+
+mig_cum = df_traj0 %>%
+  filter(var %in% c("IM", "OM")) %>%
+  group_by(traj, var, date, deme) %>%
+  summarise(value = sum(value), .groups = "drop") %>%
+  group_by(traj, var, deme) %>%
+  arrange(traj, date, deme) %>%
+  mutate(cumvalue = cumsum(value),
+         partner = NA)
+
+inout_pop = df_traj0 %>% 
+  mutate(var = ifelse(var %in% c("O", "B", "IM"), "in_pop", "out_pop")) %>%
+  group_by(traj, deme, var, date) %>%
+  summarise(value = sum(value), .groups = "drop_last") %>%
+  arrange(traj, date, deme) %>%
+  mutate(cumvalue = cumsum(value))
+
+active_pop = inout_pop %>%
+  select(-cumvalue) %>%
+  pivot_wider(names_from = var, values_from = value, values_fill = list(value = 0)) %>%
+  mutate(active_pop = in_pop - out_pop) %>%
+  pivot_longer(active_pop, names_to = "var", values_to = "value") %>%
+  group_by(traj, deme, var) %>%
+  arrange(traj, date, deme) %>%
+  mutate(cumvalue = cumsum(value)) %>%
+  select(-c("in_pop", "out_pop"))
+
+df_traj <- bind_rows(df_traj0, mig_cum) %>%
+  bind_rows(inout_pop) %>%
+  bind_rows(active_pop)
+
+df_traj_cases <- df_traj %>%
+  mutate(traj = as.character(traj)) %>%
+  bind_rows(case_data %>%
+            filter(cumvalue != 0) %>%
+            mutate(var = "ECDC"))
+                  # traj = "ecdc"))
 ```
@@
 ```{r traj-process, warning=FALSE}
-# Date grids
-max_age <- max(states$age)
-min_date <- date_decimal(decimal_date(mrs) - max_age)
-#ages_5day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-5 days')))
-ages_1day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-1 days')))
-
-# Grid trajectories, time grid of 5 days taking maximum value
-gt <- gridTrajectoriesByAge(states, ages_1day) %>%
-  replace_na(list(N = 0)) %>%
-  mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
-         date = date_model + 10)
-
-# Summary statistics
-gt_summary <- gt %>%
-  group_by(date, type) %>%
-  summarize(Imean = mean(N),
-            Imedian = median(N),
-            Ilow    = quantile(N, 0.025), #HPD??
-            Ihigh   = quantile(N, 0.975),
-            .groups = "drop_last") %>%
-  ungroup()
-
-# Grid events, time grid of 1 day, taking the sum of events
-ge <- gridEventsByAge(events, ages_1day) %>%
-  mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
-         date = date_model + 10,
-         dest = ifelse(is.na(dest), as.character(src), as.character(dest)))
+# # Date grids
+# max_age <- max(states$age)
+# min_date <- date_decimal(decimal_date(mrs) - max_age)
+# #ages_5day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-5 days')))
+# ages_1day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-1 days')))
+# 
+# # Grid trajectories, time grid of 5 days taking maximum value
+# gt <- gridTrajectoriesByAge(states, ages_1day) %>%
+#   replace_na(list(N = 0)) %>%
+#   mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
+#          date = date_model + 10)
+# 
+# # Summary statistics
+# gt_summary <- gt %>%
+#   group_by(date, type) %>%
+#   summarize(Imean = mean(N),
+#             Imedian = median(N),
+#             Ilow    = quantile(N, 0.025), #95% CI, compute HPD??
+#             Ihigh   = quantile(N, 0.975),
+#             .groups = "drop_last") %>%
+#   ungroup()
+# 
+# # Grid events, time grid of 1 day, taking the sum of events
+# ge <- gridEventsByAge(events, ages_1day) %>%
+#   mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
+#          date = date_model + 10,
+#          dest = ifelse(is.na(dest), as.character(src), as.character(dest))) %>%
+#   ungroup %>%
+#   filter(N != 0)
 
@@
 #          width = 2300, height = 3000, res = 300)
+```
+
+```{r}
+# Active cases
+d = "France"
+df_traj_cases %>%
+    # 1. Prepare data
+    filter(deme == d, var %in% c("active_pop", "in_pop", "ECDC"), is.na(partner)) %>%
+    # 2. Plot
+    ggplot() +
+    geom_line(aes(date, cumvalue, group = traj,
+                  color = var)) +
+                  # color = traj == "ecdc",
+                  # alpha = traj == "ecdc",
+                  # linetype = traj == "ecdc",
+                  # size = traj == "ecdc")) +
+    # scale_color_manual(name = "",
+    #                    labels = c(paste0(deme, " inferred\n population trajectories"),
+    #                               paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+    #                                      " ECDC total case count")),
+    #                    values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
+    # scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
+    # scale_linetype_manual(name = "", values = c(1, 2),
+    #                       labels = c(paste0(deme, " inferred\n population trajectories"),
+    #                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+    #                                         " ECDC total case count"))) +
+    # scale_size_manual(name = "", values = c(0.5, 1),
+    #                   labels = c(paste0(deme, " inferred\n population trajectories"),
+    #                              paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
+    #                                     " ECDC total case count"))) +
+    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
+                 date_labels = "%b %d") +
+    scale_y_log10(expand = c(0, 0),labels = trans_format("log10", math_format(10^.x))) +
+    ylab("Population size") +
+    theme(legend.position = c(0.4, 0.9),
+          axis.title.x=element_blank())
 ```

### 2021-02-02 scripts/analyze_trace.Rmd
eu10 and use jar for beast apps

---
title: "Trace Results"
subtitle: "analysis 3,6 and 10"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
  theme: "Pittsburgh"
colortheme: "dove"
fonttheme: "structurebold"
highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(knitr.kable.NA = '')
```

```{r libraries}
# Load libraries ---------------------------------------------------------------
library(argparse)
library(yaml)
library(tidyverse)
library(lubridate)
library(scales)
library(ggpubr)
library(ggsci)
library(ggridges)
library(patchwork)
library(hrbrthemes)
library(circlize)
library(chorddiag) 
library(ggforce)
library(ComplexHeatmap)
library(maps)
library(sf)
library(wpp2019)
library(gridExtra)
library(grid)
# Source files -----------------------------------------------------------------
source("./scripts/trajProcessing.R")
source("./scripts/utils.R")
```

```{r data}
# Analyze trace with priors
eu3 <- read.table("/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/201030_europe3_comb.log", header = TRUE) %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value") %>% 
  mutate(analysis = "eu3") #%>%
# rename(R0SVChina = R0SVHubei.China)
eu6 <- read.table("/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/201211_europe6p2_comb.log", header = TRUE) %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value") %>% 
  mutate(analysis = "eu6")
eu10 <- read.table("/Users/maceci/chains/210120_europe10a_comb.log", header = TRUE) %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value") %>% 
  mutate(analysis = "eu10")
         # R0SVFrance = R0SVi1_France,
         # R0SVItaly = R0SVi1_Italy,
         # R0SVGermany = R0SVi1_Germany,
         # R0SVSpain = R0SVi1_Spain,
         # R0SVOtherEuropean = R0SVi1_OtherEuropean)

trace <- bind_rows(bind_rows(eu3, eu6), eu10) %>%
  mutate(param = gsub("Hubei.", "", param)) %>%
  separate(param, into = c("param", "deme"), "SV") %>%
  separate(param, into = c("param", "mig"), "SM") %>%
  separate(deme, into = c("epoch1", "deme"), "_", fill = "left") %>%
  separate(mig, into = c("src", "dest"), "_to_") %>%
  separate(src, into = c("epoch2", "src"), "_", fill = "left") %>%
  unite(epoch, c("epoch1", "epoch2"), na.rm = TRUE)
```

## Origin

```{r origin}

prior <- dlnorm(seq(0,1,0.01), meanlog = -1.0, sdlog = 0.2)
b <- date(date_decimal(decimal_date(ymd("2020-03-08")) - seq(0,1,0.05)))
ggplot() +
  geom_density(data = trace %>% filter(param == "originBDMMPrime"),
               aes(value, color = analysis, fill = analysis), alpha = 0.3) +
  geom_line(aes(seq(0,1,0.01),prior)) +
  scale_x_continuous(labels = b, breaks = seq(0,1, 0.05)) +
  theme(axis.text = element_text(angle=270)) +
  scale_color_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed")) +
  scale_fill_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed"))
```

## R0

```{r r0}
prior <- dlnorm(seq(0,6,0.01), meanlog = 0.8, sdlog = 0.5)
ggplot() +
  geom_density(data = trace %>% filter(param == "R0", deme != "endtime"),
               aes(value, color = analysis, fill = analysis), alpha = 0.3) +
  facet_wrap(~deme) +
  #geom_line(aes(seq(0,1,0.01),prior)) +
  scale_color_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed")) +
  scale_fill_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed"))
```

## Migration rate
```{r migrate}

ggplot() +
  geom_density(data = trace %>% filter(param == "migrationRate", !is.na(dest)) %>% 
                 select(value, src, dest, analysis, epoch) %>% group_by(analysis),
               aes(value, color = interaction(epoch, analysis), fill = interaction(epoch, analysis)), alpha = 0.3) +
  facet_grid(src~dest, scales = "free") +
  scale_x_continuous(limits = c(0,4)) +
  scale_y_continuous(limits = c(0,20)) +
  scale_color_viridis_d() +
  scale_fill_viridis_d()
```

## Sampling proportion
```{r sp}

ggplot() +
  geom_density(data = trace %>% filter(param == "samplingProportion", deme != "endtime", epoch != "i0"),
               aes(value, color = interaction(epoch, analysis), fill = interaction(epoch, analysis)), alpha = 0.3) +
  facet_wrap(~deme, scales = "free") +
  scale_x_continuous(limits = c(0,0.1)) +
  #scale_y_continuous(limits = c(0,20)) +
  scale_color_viridis_d() +
  scale_fill_viridis_d()
```

## GLM 1 - FLight data
```{r glm1}

prior <- dnorm(seq(-10,10,0.01), mean = 0, sd = 4)
ggplot() +
  geom_density(data = trace %>% filter(param %in% c("scaleParamGLM.0", "scalerParamGLM.0")),
               aes(value, color = analysis, fill = analysis), alpha = 0.3) +
  geom_line(aes(seq(-10, 10,0.01), prior)) +
  #scale_x_continuous(labels = b, breaks = seq(0,1, 0.05)) +
  #theme(axis.text = element_text(angle=270)) +
  scale_color_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed")) +
  scale_fill_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed"))
```

## GLM 2 - distance
```{r glm2}

prior <- dnorm(seq(-10,10,0.01), mean = 0, sd = 4)
ggplot() +
  geom_density(data = trace %>% filter(param %in% c("scaleParamGLM.1", "scalerParamGLM.1")),
               aes(value, color = analysis, fill = analysis), alpha = 0.3) +
  geom_line(aes(seq(-10, 10,0.01), prior)) +
  #scale_x_continuous(labels = b, breaks = seq(0,1, 0.05)) +
  #theme(axis.text = element_text(angle=270)) +
  scale_color_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed")) +
  scale_fill_viridis_d(labels = c("eu3 - initial setup", "eu6 - GLM 2p no ind", "eu10 - GLM 2p ind\n R0 China fixed"))
```

### 2021-02-02 scripts/analyze_trajectories_v2.Rmd
eu10 and use jar for beast apps

@@
 args$burnin = 0.1
-args$demes = read.csv("./files/demes.csv", stringsAsFactors = FALSE)
-args$input = "./results/europeC/201106_europe5.TL.traj"
+args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
+#args$input = "./results/europeC/201211_europe6p2.TL.traj"
+args$input = "/Users/maceci/traj_10_test/notimeparameter/210122_europe10a.TL.traj"
 args$metadata =  "./results/europeC/final_metadata.tsv"
-args$n = 50
+args$n = 200
 demes = args$demes
 demes$exclude_country[5] = "France,Germany,Italy,Spain"
-demes$deme[3] = "China"
-demes$division[3] = NA
+#demes$deme[3] = "China"
+#demes$division[3] = NA
 demes <- demes %>% arrange(deme)
@@
 dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
-dcolors <- dcolors[1:nrow(demes)]
+dcolors <- c("#E64B35FF", "#4DBBD5FF", "#00A087FF", "#3C5488FF", "#ffaa4fFF", "#7E6148FF")
+#dcolors <- c("#E64B35FF", "#7E6148FF", "#00A087FF", "#3C5488FF", "#4DBBD5FF", "#ffc954")
+#dcolors <- dcolors[1:nrow(demes)]
 names(dcolors) <- sort(demes$deme)
@@
 names(dark_dcolors) <- names(dcolors)
-ecolors <- c("#1e3d59", "#f5f0e1", "#ff6e40", "#ffc13b")
-names(ecolors) <- c("B", "M", "IM", "OM")
+ecolors <- c(B = "#1e3d59", M = "#f5f0e1", IM = "#ff6e40", OM = "#ffc13b")
 ```
@@
     st_point(st_coordinates(zoom_to_xy - c(x_span/2, y_span/2))),
-    st_point(st_coordinates(zoom_to_xy + c(x_span *1.2, y_span *1.2))),
+    st_point(st_coordinates(zoom_to_xy + c(x_span *1.2, y_span *1.25))),
     crs = target_crs
 )
-ggplot() + geom_sf(data = worldmap, aes(fill = deme), alpha = 0.8, size = 0) +
+ggplot() + geom_sf(data = worldmap, aes(fill = deme), alpha = 0.9, size = 0) +
     scale_x_continuous(breaks = seq(-100, 200, by = 10)) +
@@
   scale_fill_manual(values = dcolors, na.value = "grey90") +
+  #scale_fill_manual(values = c("#E64B35FF", "grey50", "grey50", "grey50", "grey50", "grey50"), na.value = "grey90") +
+  #ggsci::scale_fill_tron()
   theme_void() +
@@
   arrange(date) %>%
+  group_by(deme, date) %>%
+  rename(value = new_cases) %>%
+  summarise(value = sum(value)) %>%
+  mutate(cumvalue = cumsum(value))
+
+library(zoo)
+
+case_data <- case_data %>%
   group_by(deme) %>%
-  rename(value = cases) %>%
-  mutate(cumvalue = cumsum(value)) %>%
-  group_by(deme, date) %>%
-  mutate(cumvalue = max(cumvalue)) %>%
-  select(deme, date, cumvalue, value)  %>%
-  distinct_at(vars(deme, date, cumvalue), .keep_all = TRUE)
+  mutate(rollavg3 = zoo::rollmean(value, k = 3, fill = NA),
+         #rollavg7 = zoo::rollmean(value, k = 7, fill = NA),
+         datereport = date,
+         date = date - 10)
 
@@
 
-case_data_pop <- case_data %>% left_join(population, by="deme") %>% mutate(ECDCcas100 = cumcases * 1e5 / (pop * 1e3)) 
+case_data_pop <- case_data %>% left_join(population, by="deme") %>% mutate(ECDCcas100 = cumvalue * 1e5 / (pop * 1e3)) 
 ```
@@
 ```{r ecdc-table, echo=FALSE}
-knitr::kable(case_data_pop %>% ungroup %>% filter(date == MRS + 10) %>% select(deme, cumcases, pop, cas100) %>% mutate_if(is.numeric, function (x) round(x,1)), booktabs = TRUE, caption = "Total number of cases reported to ECDC 18th March")
+knitr::kable(case_data_pop %>% ungroup %>% filter(date == MRS + 10) %>% select(deme, cumvalue, pop, ECDCcas100) %>% mutate_if(is.numeric, function (x) round(x,1)), booktabs = TRUE, caption = "Total number of cases reported to ECDC 18th March")
 ```
@@
 ```{r ecdc-plot, fig.height = 3, fig.width=4, fig.cap="ECDC case counts for each deme from the beginning of the pandemic to March 18"}
+
 ggplot(case_data) +
-  geom_line(aes(x = date, y = cumcases, color = deme)) +
+  geom_line(aes(x = date, y = cumvalue, color = deme)) +
+  geom_point(aes(x = date, y = cumvalue, color = deme)) +
    scale_color_manual(values = dcolors, name = "") +
@@
   scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
+  #scale_y_continuous(expand = c(0, 0)) +
     ylab("Case count") +
@@
           legend.box="horizontal",
-          panel.grid.major.x =  element_line(colour = "grey80", size = 0.3),
-          panel.grid.minor.x =  element_line(colour = "grey80", size = 0.2))
+          panel.grid.major.y =  element_line(colour = "grey80", size = 0.3, linetype = 2),
+          panel.grid.minor.y =  element_line(colour = "grey80", size = 0.2,  linetype = 2),
+          panel.background =  element_blank())
 ```
@@
                       levels = 0:(nrow(demes) - 1),
-                      #labels = sort(demes$deme)),
-                      labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
+                      labels = sort(demes$deme)),
+                      #labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
          dest = (factor(dest,
                            levels = 0:(nrow(demes) - 1),
-                           #labels = sort(demes$deme),
-                           labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain"))))
+                           labels = sort(demes$deme))))
+                           #labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain"))))
 events[events$event == "O", "src"] = "China"
@@
   bind_rows(case_data %>%
-            filter(cumvalue != 0) %>%
+            #filter(cumvalue != 0) %>%
             mutate(var = "ECDC"))
                   # traj = "ecdc"))
+
+min_date <- min(df_traj_cases$date)
+df_traj_cases_complete <- df_traj_cases %>%
+  group_by(traj, var, deme, partner) %>%
+  complete(date = seq.Date(min_date, MRS, by = "day")) %>%
+  replace_na(list(value = 0)) %>%
+  arrange(traj, date, deme) %>%
+  mutate(cumvalue = cumsum(value))
 ```
@@
 # write_tsv(r_reported, paste0(args$output_table, "01.tsv"))
+
+reported1 <- df_traj_cases %>%
+  filter(var == "in_pop", date == MRS) %>%
+  left_join(case_data_pop %>% rename(cumcases = cumvalue) %>% select(deme, date, cumcases, pop), by = c("deme", "date")) %>%
+  group_by(var, deme, cumcases, pop) %>%
+  summarise(median_in = median(cumvalue),
+            l95_in = HDInterval::hdi(cumvalue, 0.95)[1],
+            h95_in = HDInterval::hdi(cumvalue, 0.95)[2]) %>%
+  mutate(times_med = median_in/cumcases,
+         times_l = l95_in/cumcases,
+         times_h = h95_in/cumcases,
+         p_med = median_in/(pop*1e3) * 100,
+         p_l = l95_in/(pop*1e3) * 100,
+         p_h = h95_in/(pop*1e3) * 100)
+
+reported2 <- df_traj_cases %>%
+  filter(var == "in_pop") %>%
+  left_join(case_data_pop %>% rename(cases = value, ecdcavg3=rollavg3) %>% select(deme, date, cases, pop, ecdcavg3), by = c("deme", "date")) %>%
+  group_by(var, deme, date, cases, ecdcavg3) %>%
+  summarise(median_in = median(value),
+            l95_in = HDInterval::hdi(value, 0.95)[1],
+            h95_in = HDInterval::hdi(value, 0.95)[2]) %>%
+  mutate(report_rate = cases / median_in)
+
+reported3 <- df_traj_cases %>%
+  filter(var == "in_pop") %>%
+  left_join(case_data_pop %>% rename(cases = value, ecdcavg3=rollavg3) %>% select(deme, date, cases, pop, ecdcavg3), by = c("deme", "date")) %>%
+  group_by(var, deme, week = week(date), cases, ecdcavg3) %>%
+  summarise(median_in = median(value),
+            l95_in = HDInterval::hdi(value, 0.95)[1],
+            h95_in = HDInterval::hdi(value, 0.95)[2]) %>%
+  mutate(report_rate = cases / median_in)
+
+reported3 <- df_traj_cases_complete %>%
+  group_by(var, deme, week = week(date), cases, ecdcavg3) %>%
+  summarise(median_in = median(value),
+            l95_in = HDInterval::hdi(value, 0.95)[1],
+            h95_in = HDInterval::hdi(value, 0.95)[2]) %>%
+  mutate(report_rate = cases / median_in)
+
+ggplot(reported3 %>% filter(deme != "China")) +
+  geom_line(aes(week, report_rate, color = deme)) +
+  geom_point(aes(week, report_rate, color = deme))
+
 ```
@@
 # Active cases
-d = "France"
+library(gganimate)
+d = "Germany"
 df_traj_cases %>%
     # 1. Prepare data
-    filter(deme == d, var %in% c("active_pop", "in_pop", "ECDC"), is.na(partner)) %>%
+    filter(deme == d, 
+           #traj %in% c(1, NA), 
+           var %in% c("in_pop", "ECDC"), 
+           is.na(partner)) %>%
     # 2. Plot
@@
     geom_line(aes(date, cumvalue, group = traj,
-                  color = var)) +
-                  # color = traj == "ecdc",
-                  # alpha = traj == "ecdc",
-                  # linetype = traj == "ecdc",
-                  # size = traj == "ecdc")) +
-    # scale_color_manual(name = "",
-    #                    labels = c(paste0(deme, " inferred\n population trajectories"),
-    #                               paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
-    #                                      " ECDC total case count")),
-    #                    values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
-    # scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
-    # scale_linetype_manual(name = "", values = c(1, 2),
-    #                       labels = c(paste0(deme, " inferred\n population trajectories"),
-    #                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
-    #                                         " ECDC total case count"))) +
-    # scale_size_manual(name = "", values = c(0.5, 1),
-    #                   labels = c(paste0(deme, " inferred\n population trajectories"),
-    #                              paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
-    #                                     " ECDC total case count"))) +
+                  color = var == "ECDC",
+                  alpha = var == "ECDC",
+                  linetype = var == "ECDC",
+                  size = var == "ECDC")) +
+    scale_color_manual(name = "",
+                       labels = c(paste0(d, " inferred\n population trajectories"),
+                                  paste0(ifelse(d == china_deme, "China", as.character(d)),
+                                         " ECDC total case count")),
+                       values = c(dcolors[[d]], dark_dcolors[[d]])) +
+                       #values = c("#4ab9ed", dark_dcolors[[d]])) +
+    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
+    scale_linetype_manual(name = "", values = c(1, 2),
+                          labels = c(paste0(d, " inferred\n population trajectories"),
+                                     paste0(ifelse(d == china_deme, "China", as.character(d)),
+                                            " ECDC total case count"))) +
+    scale_size_manual(name = "", values = c(0.5, 1),
+                      labels = c(paste0(d, " inferred\n population trajectories"),
+                                 paste0(ifelse(d == china_deme, "China", as.character(d)),
+                                        " ECDC total case count"))) +
     scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks",
@@
           axis.title.x=element_blank())
+  #theme_void() #+
+  #transition_reveal(date)
 ```

### 2021-02-02 scripts/case_data.R

@@ -1,0 +1,85 @@
+source("./scripts/utils.R")
+library(tidyverse)
+
+ecdc <- get_dataECDC()
+who <- get_dataWHO()
+jh <- get_dataJH()
+owid <- get_dataOWID() #%>% select(-new_cases) %>% rename(new_cases = new_cases_smoothed)
+
+ecdc2 <- ecdc %>%
+  mutate(new_cases = zoo::rollmean(new_cases, k = 7, fill = NA, align = "center"),
+         source = "ecdc2")
+case_data <- bind_rows(bind_rows(bind_rows(ecdc, jh), owid), ecdc2)
+
+
+case_data2 <- case_data %>%
+  group_by(country, source) %>%
+  mutate(new_cases2 = zoo::rollmean(new_cases, k = 7, fill = NA)) %>%
+  select(country, date, source, new_cases, new_cases2)
+
+#bind_rows(case_data, china) %>%
+case_data  %>%
+  filter(date > ymd("2019-12-01"),
+         date <= ymd("2020-03-18"),
+         country %in% c("China", "France", "Germany", "Italy", "Spain")) %>%
+  ggplot() +
+  geom_line(aes(date, new_cases, color = source, linetype = source)) +
+  facet_wrap(~country, scales = "free")
+
+samples <- tibble(country = c("China", "France", "Germany", "Italy", "OtherEuropean", "Spain"), s = c(60, 60, 50, 60, 50, 60))
+sp <- case_data %>%
+  filter(date > ymd("2019-12-01"),
+         (date <= ymd("2020-03-08") & country != "China")
+         | (date <= ymd("2020-01-23") & country == "China"),
+         country %in% c("China", "France", "Germany", "Italy", "Spain")) %>%
+  group_by(country, source) %>%
+  summarise(cases = sum(new_cases, na.rm = TRUE)) %>%
+  left_join(samples) %>%
+  mutate(sp = s/cases)
+
+spain <- read.csv("https://cnecovid.isciii.es/covid19/resources/casos_tecnica_ccaa.csv") %>%
+  group_by(fecha) %>%
+  summarise(new_cases = sum(num_casos)) %>%
+  mutate(country = "Spain",
+         date = ymd(fecha) + 6,
+         source = "RENAVE")
+
+italy <- read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-andamento-nazionale/dpc-covid19-ita-andamento-nazionale.csv") %>%
[…]

### 2021-02-02 scripts/chains_diagnostic.R

@@ -30,7 +30,8 @@ ESS_cutoff <- args$ess
 diagnostic <- data.frame()
 for (fname in args$input){
   if (file.size(fname) != 0) {
-    tb <- read_table(fname) 
+    tb <- read_table(fname) %>%
+    filter(`95%HPDlo` != `95%HPDup`) # Filter fixed values
     if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
       cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
       diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
@@ -40,7 +41,7 @@ for (fname in args$input){
                                                  min_ESS = ESS_cutoff, 
                                                  included = 1))
     } else {
-      cat("\n\n\nItems:\n\n", tb %>% filter(ESS < ESS_cutoff) %>% pull(statistic),
+      cat("\n\n\nItems:\n\n", tb %>% filter(ESS < ESS_cutoff) %>% pull(item),
           "\n\nhave ESS value < ", ESS_cutoff, " in log file ", fname, "\n\nLog file not included in analysis.\n")
       diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]), 
                                                  seed = str_split(fname, pattern ="\\.")[[1]][2],

### 2021-02-02 scripts/utils.R

@@ -30,7 +30,7 @@ set_plotopts <- function() {
 get_cases <- function(demes, from = NA, to = NA) {
   # Get case data information from ECDC for specific demes (countries and regions)
   # Set from and to dates with values from the function call instead of the ones from the demes info
-  cat("\nLoading case data counts from ECDC...\n")
+  #cat("\nLoading case data counts from ECDC...\n")
   #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
   #else case_data <- get_dataECDC()
   case_data <- get_dataECDC()
@@ -64,33 +64,88 @@ get_cases_deme <- function(case_data, deme_name, region_name = NA, country_name 
            date <= as.Date(to)) %>%
     mutate(deme = deme_name) %>%
     arrange(date) %>%
-    mutate(cumcases = cumsum(cases),
-          cumdeaths = cumsum(deaths),
+    mutate(cumulative_cases = cumsum(new_cases),
+          cumulative_deaths = cumsum(new_deaths),
           population_deme = sum(unique(population))) %>%
     group_by(date) %>%
-    mutate(cumcases = max(cumcases),
-           cumdeaths = max(cumdeaths))
+    mutate(cumulative_cases = max(cumulative_cases),
+           cumulative_deaths = max(cumulative_deaths))
   
   return(df_cases)
 }  
 
 get_dataECDC <- function() {
-  #' Get case data information from ECDC for specific demes (countries and regions)
   # case_data <-  read.csv("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", 
   #                        na.strings = "", fileEncoding = "UTF-8-BOM") # From 14 Dec, it is weekly counts
+  cat("\n Loading case data counts from ECDC...")
   case_data <-  readxl::read_excel("data/ECDC_covidcases.xlsx")
   case_data1 <- case_data %>%
     rename(country = countriesAndTerritories,
            region = continentExp,
            date = dateRep,
            population = popData2019,
+           new_cases = cases,
+           new_deaths = deaths,
            cases14days = 12,
            ) %>%
-    mutate(date = ymd(date),
+    mutate(source = "ecdc",
+           date = ymd(date),
            country = ifelse(country == "Czechia", "Czech Republic", gsub("_", " ", country))) %>%
[…]

### 2021-02-21 reports/mt-thesis/introduction.tex
initial commit thesis

\chapter{Introduction}

We have many sources of information, but all of them are imperfect in some aspect. Bayesian phylodynamics incorporates different beliefs and data to get a detailed picture of the epidemic. Genetic sequences as an objective source of information about virus evolution and transmision together with viral phylogenetics, transpotation data and geographic data to understand the dynamics and case counts to guide the magnitutde of the numbers.

We lack absolute numbers of the pandemic time resoluted without reporting issues dependent. We lack understading of the dynamics, from where to where, how often, how long.

We can use travel data as a proxy, or we can incorporate it in a model. Lemey et al. Difference we get absolute numbers, not only the rates. And we use BD model instead of coalescent, better to describe beginning of a epidemic. While due to the complexite and time consuming model we dow scaled to europe and only thhe initial phase with the first countries to experiment relevant sars cov 2 epidemics.

### 2021-02-21 reports/mt-thesis/methods.tex
initial commit thesis

\chapter{Material and Methods}

\section{The multytipe birth death model}

We have used the multitype birth-death phylodynamic model. Phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process. A multitype birth death process describes the evolution of an structured population. The process starts with one individual in one of the subpopulations (or demes) and it can give rise to another individual at rate $\beta$, die at death $\delta$, migrate to other deme or be sampled at rate $\phi$ to become part of the phylogeny.

Since the rate of the evolution of RNA virus, in specific SARS-CoV-2 is in the same scale than the epidemic spread, we can study the transmission dynamics by looking at the virus sequences.

The parameters that we will consider are the R0, the becoming uninfectious rate, the migration rates and the sampling proportion (proportion of no longer infectious individual that arer sampled and included in the phylogenetic tree).

Thus, the probability that we want to know is:

$P(T, P | D)$

This process describes the generation of the tree as part of the dynamics of the whole population, it is very rich and detailed in information. Therefore it is also possible to obtain information about the epidemic trajectory of the whole population. We can obtain the full sequence of events that describes the epidemic in each of the demes.

From the phylogenetic tree and the phylodynamics parameters we use Stochastic mapping to infer the ancestral traits and then to infer the epidemic trajectories.

$P(Tc | T, P)$

$P(E | Tc, P)$

\section{Infering epidemic trajectories}

Stochastic mapping of ancestral traits

Read paper and tim presentation

Stochastic mapping of epidemic trajectories

From the phylogenie with ancestral locations and the set of phylodynamic parameters we can simulate epidemic trajectories using an algorithm similar to Gillespie (with tau leaping and particle filtering? the subsampling). And most improtant, we can compute the probability of that specific trajectory given those parrameters and tree.

$P(E|Tc, P)$

Any trajectories without the events that are respresent in the tree has probability 0, so to avoid the simualtion of this trajectories, no time effcieint, we will enforce the events that arer represented in the tree (and will include a weigth). Also to avoid low probabilities trajectories, we will importance subsample one trajectory after a certain time of simulation according to the weigths detemrined by its probabilities.

\section{Modeling the early dynamics of SARS-CoV-2 in Europe}

In our case, we have 6 demes, one for each of the countries/geographic location. A birth represents a transmission event of covid-19, a death event is an individual becoming uninfectious by overcoming the disease, being isolated or dying. From the birth rate and death rate we can estuimate the R0, key vallue in the epidemic understanding and control. A migration event happens when an individual travels from one country to another, we know that is an important way of covid-19 spreading and we are interested on the dynamics.

We focus our analysis in the early spread in Europe of SARS-CoV-2, therefore we constrain the analysis to the period from the origin of the epidemic to March 8, when the first European lockdown in the Lombardy region started. During this time, for the European countries we expect an unimpeded spread of the virus that could be described by a constant R0 particular to each deme. However, the migration rates could have changed during these months due to the increase public awareness, so we will consider a different migration rate for origin to 23 jan, 23 jan to end of february and 1 to 8 of March. We assume a constant sampling proportion different for each of the demes, to account for difference in the sequencing efforts. The becoming uninfectious rate we will fixed in 36.5, assuming that one individual is infectious for a exponential time distirbuted with mean 10.

\section{Incorporating travel data}

To incorporatet travel data in the model we have use a GLM model described in Lemey et al 2009. We use a generalized linear model to describe the migration matrix KxK, where the rpedictors are the number of dayly flight from one country to another from EUROSTATS, the average distance between the centrroids of the country and the population sizes from origin and destinations. The predictors are log transformed, we add a pseudocount to make them all positive, and standardized as described in ..

$mij = c exp(\sum_{i = 1}^p \delta_i \beta_i xij)$

The coefficients beta describe the effect size of the predictors in the migrations rate and the delta coefficients act as a model selection variables, taking 0-1 values including orr not that specific predictor in the model.

\section{SARS-CoV 2 genome data set}

GISAID. Sequences from France, Germany, Italy, Spain and other European countries. Subsampling a sequence with probability equal to the probability of having a case that day in the country, inverseley weigthed by the probability of being sequenced that day.

\section{GLM predictos data}

EUROSTAT, countries considers. Distance and population datasets.

\section{Bayesian inference of phylodynamic parameters}

We use bayesian inference and Metropolis Hasting Monte Carlo algorith to infer our phylodynamics parameters and phylogeneit tree. We use the following priors

Clock rate and substitution model. Fixed to 8x10-4, HKY 4 gamma categories with priors...
BDMM-Prime epi parameterization:
R0
Become uninfectious rate
Sampling rates
Migration rates

\subsubsection{R0}
R0 constant except for China (and Italy? maybe not in the main analysis).

Prior Log Normal (0.8, 0.5) Median 2.2, 95IQR 0.8 to 5.9.
(Lai A. 2020 prior log normal (0,1) median 1.0 0.1-7.10)

- Main analysis: We fix China R0 to 2.7 (23 jan) 1.3 (10 feb) 0.8
- Use a decreasing paramater as Sarah with migration rate? But then we would need to add sequences.
- Or instead of fixing it we can put a strong informative prior.
- Allow Italy R0 to change March 1.

Meta-analyses R0:

R0 in China (3.14,95\%CI,2.40–4.09).\cite{Billah2020}

R0value 2.90 (2.32, 3.63
 2.9 (95\% CI: 2.1–4.5)\cite{Park2020}

Periods:
until 23 jan R0 2.9 
23 jan - 10 feb 1.1 dashboard cevo
10 feb - 8 mar 0.43 dashboard cevo

Become uninfectious rate same for all demes, fixed to 36.5.
Migrations among demes, 3 epochs.
Sampling proportion 0 before first sample. Upperbounded by number of cases.

Implementation BEAST 2.6.3 and BDMM-Prime package.
We run 5 chains of 10e7 iterations witth different seeds.

\section{Processing of epidemic trajectories}
In our case, for each set of parameters and tree we will simulate 10000 trajectories with an epsilon of 0.3 and select one of them with importance sampling.
We have one trtajectory for each step in the MCMC. We subsampled x.

\section{Data availability}
GISAID data table.

\section{Code availability and analyses reproducibility}
All the code is availables at ...
We have implement the full workflow with Snakemake so its fully reproducible.

### 2021-02-21 reports/mt-thesis/thesis.tex
initial commit thesis

\documentclass[a4paper, oneside]{discothesis}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DOCUMENT METADATA

\thesistype{Master Thesis} % Master's Thesis, Bachelor's Thesis, Semester Thesis, Group Project

\title{\fontsize{18}{22}\selectfont A comprehensive study of the phylodynamics of SARS-CoV-2 in Europe}

\author{Cecilia Valenzuela Agüí}
\email{ceciliav@student.ethz.ch}

\institute{Department of Biosystems Science and Engineering \\[2pt]
Computational Evolution \\[2pt]
ETH Zürich}

% Optionally, you can put in your own logo here
%\logo{\includegraphics[width=0.2\columnwidth]{figures/disco_logo_faded}}

\supervisors{Prof.\ Dr.\ Tanja
  Stadler, Dr. Timothy Vaughan, Sarah Nadeau}

% Optionally, keywords and categories of the work can be shown (on the Abstract page)
%\keywords{Keywords go here.}
%\categories{ACM categories go here.}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\frontmatter % do not remove this line
\maketitle

\cleardoublepage

\begin{acknowledgements}
	I thank Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet.
\end{acknowledgements}

\begin{abstract}
    The abstract should be short, stating what you did and what the most important result is.
	Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet.
\end{abstract}

\tableofcontents

\mainmatter % do not remove this line

\include{introduction}

\include{methods}

% \begin{theorem}[First Theorem] \label{thm:first theorem}
% 	This is our first theorem.
% \end{theorem}

% \begin{proof}
% 	And this is the proof of the first theorem with a complicated formula and a reference to Theorem \ref{thm:first theorem}. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua. Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua.
% 	\begin{equation}
% 		{\frac {\mathrm d}{\mathrm dx}}\arctan(\sin({x}^{2}))=-2 \cdot {\frac {\cos({x}^{2})x}{-2+\left (\cos({x}^{2})\right )^{2}}}
% 	\end{equation}	
% \end{proof}

%% \begin{figure}
%%     \centering
%%     \includegraphics[width=0.2\columnwidth]{figures/disco_logo_faded}
%%     \caption{This is an example graphic.}
%%     \label{fig:example_figure}
%% \end{figure}

% And here we cite some external documents~\cite{TestReference, TestReference2}.
% An example of an included graphic can be found in Figure~%\ref{fig:example_figure}.
% Note that in \LaTeX, ``quotes'' do not use the usual double quote characters.

\chapter{Discussion}

% This displays the bibliography for all cited external documents. All references have to be defined in the file references.bib and can then be cited from within this document.
\bibliographystyle{IEEEtran}
\bibliography{references}

% This creates an appendix chapter, comment if not needed.
\appendix
\chapter{Data}
\chapter{Priors}
\chapter{Supplementary Figures}

\end{document}

### 2021-02-22 reports/mt-thesis/methods.tex
wip methods

@@
 
-We have used the multitype birth-death phylodynamic model.
-Phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process.
-A multitype birth death process describes the evolution of an structured population.
-The process starts with one individual in one of the subpopulations (or demes) and it can give rise to another individual at rate $\beta$, die at death $\delta$, migrate to other deme or be sampled at rate $\phi$ to become part of the phylogeny.
+\todo{Introduction?}
+In a phylodynamic analysis, we aim to integrate genetic and epidemiological information to understand pathogen evolution and transmission dynamics.
+This is possible for RNA viruses, and in specific for SARS-CoV-2, because population genetic and epidemic processes are in the same time scale \cite{Grenfell2013}.
+Thus, from virus sequences we can gain information about the interactions between hosts and the epidemic spread \cite{Volz2014}.
+More precisely, the field of phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process.
 
-Since the rate of the evolution of RNA virus, in specific SARS-CoV-2 is in the same scale than the epidemic spread, we can study the transmission dynamics by looking at the virus sequences.
+To study the early dynamics of SARS-CoV-2 in the European countries we have used a simplified version of the multitype birth-death model described in \cite{Kuhnert2016}.
+Birth death models are classical compartmental population models with high flexibility.
+In the multitype version, we can describe a structured population in types or subpopulations.
+The process starts with one individual in one of the subpopulations, e.g in subpopulation $i$, who can give rise to another individual at rate $\lambda_i$, die at death $\mu_i$, migrate to another deme $j$ at rate $m_ij$ or be sampled at rate $\phi_i$ to become part of the phylogenectic tree.
+
+This process depicts the full transmission dynamics and specifically, the generation of the sampled transmission tree that we observed from our sequence data.
+From the subpopulation rates we parameterize the effective reproductive number $R_i = \frac{\lambda_i}{\mu_i + \phi_i}$, a key value in epidemic control and understanding, the rate of becoming non infectious $\delta+i = \mu_i + \phi_i$ and the probability of and individual to be sampled $s_i = \frac{\phi_i}{\mu_i + \phi_i}$
 
 The parameters that we will consider are the R0, the becoming uninfectious rate, the migration rates and the sampling proportion (proportion of no longer infectious individual that arer sampled and included in the phylogenetic tree).
+
+As describen in JKuhner based in the work of stadler we can compute the tree likelihood of the multi type birth death model.
 

### 2021-03-01 reports/mt-thesis/methods.tex
wip methods sequences

--- 2021-02-21 reports/mt-thesis/methods.tex
@@
 
-\section{The multytipe birth death model}
+\section{SARS-CoV 2 genome data set}
 
-We have used the multitype birth-death phylodynamic model.
-Phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process.
-A multitype birth death process describes the evolution of an structured population.
-The process starts with one individual in one of the subpopulations (or demes) and it can give rise to another individual at rate $\beta$, die at death $\delta$, migrate to other deme or be sampled at rate $\phi$ to become part of the phylogeny.
+We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{GISAID} (accessed on November 2020).
+We follow the Nextstrain workflow for the curation of the sequences \cite{nextstrain}.
+Sequences with incomplete collection date, less than 27.000 bases in length or with more than 3.000 unknown bases are omitted.
+Also, sequences from known clusters of transmission or from the same patient are excluded from the analysis.
+The resulting dataset of x sequences are aligned with MAFFT.
+The beggining and the end of the alignment are masked respectively by 100 and 50 sequences as well as sites  13402, 24389 and 24390, identified by Nextstrain as prone to sequencing errors.
 
-Since the rate of the evolution of RNA virus, in specific SARS-CoV-2 is in the same scale than the epidemic spread, we can study the transmission dynamics by looking at the virus sequences.
+To focus on the early dynamics in Europe we select sequences from China, origin of the epidemic, France and Germany, the European countries with the earliest cases, and Italy and Spain, the European countries with the biggest outbreaks in March.
+To take into account the dynamics in other areas of Europe we include a group of 50 sequences from other European countries.
+We limit our sample of Chinese genomes to sequences until January 23, the starting date of the lockdown in Hubei.
+Due to the large (and unprecendented) number of available genetic sequences for SARS-CoV-2, we need to subsample the alignment.
+Each sequence is subsampled with a probability of having a reported case in that country the day of sample collection, inversely weigted by the probability of having a sequence in GISAID that day from the country.
+With this subsampling protocol, we aim to get a constant sampling proportion accross the full period for each country.
 
-The parameters that we will consider are the R0, the becoming uninfectious rate, the migration rates and the sampling proportion (proportion of no longer infectious individual that arer sampled and included in the phylogenetic tree).
+\section{The multitype birth death model}
 
-Thus, the probability that we want to know is:
+In a phylodynamic analysis, we aim to integrate genetic and epidemiological information to understand pathogen evolution and transmission dynamics.
+We can learn about the epidemic spread and the interactions between hosts from the imprint that these events leave in virus phylogenies.
+This is possible for RNA viruses, and in specific for SARS-CoV-2, because virus genetic evolution and epidemic processes are in the same time scale \cite{Grenfell2013} \cite{Volz2014}.
+More precisely, the field of phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process.
+\todo{Introduction?}
 
-$P(T, P | D)$
+To study the early dynamics of SARS-CoV-2 in the European countries, we use a simplified version of the multitype birth-death model described in \cite{Kuhnert2016}, following the analysis in \cite{Nadeau2020}.
+Birth death models are compartmental population models with high flexibility that describe the process of epidemiological transmission.
+The stochastic formulation of these models are used in phylodynamic analyses \cite{Stadler2012}.
+In the multitype version, we consider a structured population in types or subpopulations with characteristic within-subpopulation dynamics and migrations between them.
 
-This process describes the generation of the tree as part of the dynamics of the whole population, it is very rich and detailed in information.
-Therefore it is also possible to obtain information about the epidemic trajectory of the whole population.
-We can obtain the full sequence of events that describes the epidemic in each of the demes.
+The process starts with one infected host in one of the subpopulations, e.g in subpopulation $i$, who can infect another individual at rate $\lambda_i$, become uninfectious at rate $\mu_i$ by death or recovery, migrate to another deme $j$ at rate $m_{ij}$ or be sequenced at rate $\psi_i$ to become part of the phylogenectic tree.
+This process depicts the full transmission dynamics and specifically, the generation of the transmission tree that we observed from our sequence data.
+
+We parameterize our model in terms of the effective reproductive number $R_i = \frac{\lambda_i}{\mu_i + \psi_i}$, a key value in epidemic control and understanding, the rate of becoming uninfectious $\delta_i = \mu_i + \psi_i$ and the probability of and individual to be sequenced $s_i = \frac{\psi_i}{\mu_i + \psi_i}$.
+
+\section{Multi-typed trees and epidemic trajectories}
+
+\section{Inference of epidemic dynamics and trajectories}
+
+Under this model, we are able to compute the likelihood of the multitype birth-death parameters for a given phylogenetic tree.
+This likelihood is derived in \cite{Kuhnert} by considering the probability of an individual evolved as observed in the tree.This derivation is analogous to the work in Stadler and Bonhoeffer (2013), which is based on ideas from (Maddison et al.
+2007)
+
+This phylodynamic likelihood is the tree prior in our Bayesian inference.
+We want to estimate the overall posterior distribution of the phylogenetic tree with ancestral locations $\mathcal{T}_c$, the epidemic trajectory $\mathcal{E}$, the substitution model parameters $\theta$ and the multi-type birth death parameters $\eta$:
+
+\begin{equation}
+P(\mathcal{T}_c, \mathcal{E}, \theta, \eta | A, S) = \frac{\overbrace{P(A | \tau, \theta)}^{\text{tree likelihood}} \overbrace{P(\tau, S | \eta)}^{\text{tree prior}} \overbrace{P(\theta, \eta)}^{\text{parameters prior}}}{\underbrace{P(A, S)}_{\text{marginal likelihood}}}
+\end{equation}
+
+where A is the sequence alignment and S the observed sequence times.
 
@@
 
+New method to infer epidemic trajectories!!
+
 $P(E | Tc, P)$
-
-\section{Infering epidemic trajectories}
 
@@
 The coefficients beta describe the effect size of the predictors in the migrations rate and the delta coefficients act as a model selection variables, taking 0-1 values including orr not that specific predictor in the model.
-
-\section{SARS-CoV 2 genome data set}
-
-GISAID.
-Sequences from France, Germany, Italy, Spain and other European countries.
-Subsampling a sequence with probability equal to the probability of having a case that day in the country, inverseley weigthed by the probability of being sequenced that day.
 

### 2021-03-03 reports/mt-thesis/methods.tex
wip methods

@@
 We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{GISAID} (accessed on November 2020).
-We follow the Nextstrain workflow for the curation of the sequences \cite{nextstrain}.
+We follow the Nextstrain workflow for the curation of the dataset \cite{nextstrain}.
 Sequences with incomplete collection date, less than 27.000 bases in length or with more than 3.000 unknown bases are omitted.
 Also, sequences from known clusters of transmission or from the same patient are excluded from the analysis.
-The resulting dataset of x sequences are aligned with MAFFT.
+The resulting dataset of x \todo{look number of seqs} sequences is aligned with MAFFT.
 The beggining and the end of the alignment are masked respectively by 100 and 50 sequences as well as sites  13402, 24389 and 24390, identified by Nextstrain as prone to sequencing errors.
@@
 To focus on the early dynamics in Europe we select sequences from China, origin of the epidemic, France and Germany, the European countries with the earliest cases, and Italy and Spain, the European countries with the biggest outbreaks in March.
-To take into account the dynamics in other areas of Europe we include a group of 50 sequences from other European countries.
+To take into account the dynamics in other regions of Europe we include a group of 50 sequences from other European countries.
 We limit our sample of Chinese genomes to sequences until January 23, the starting date of the lockdown in Hubei.
+
 Due to the large (and unprecendented) number of available genetic sequences for SARS-CoV-2, we need to subsample the alignment.
@@
 
+This dataset of genetic sequences is the main source of information for our phylodynamic analysis.
+The goal is to infer the phylogenetic tree, i.e.
+the evolutionary tree-shaped relationship among the sequences, together with the epidemiological transmission parameters that gave rise to it.
+These parameters are defined within a population dynamic model and will inform us about the epidemic in the population that the sequences come from.
+
 \section{The multitype birth death model}
-
-In a phylodynamic analysis, we aim to integrate genetic and epidemiological information to understand pathogen evolution and transmission dynamics.
-We can learn about the epidemic spread and the interactions between hosts from the imprint that these events leave in virus phylogenies.
-This is possible for RNA viruses, and in specific for SARS-CoV-2, because virus genetic evolution and epidemic processes are in the same time scale \cite{Grenfell2013} \cite{Volz2014}.
-More precisely, the field of phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process.
-\todo{Introduction?}
 
@@
 In the multitype version, we consider a structured population in types or subpopulations with characteristic within-subpopulation dynamics and migrations between them.
+In our case, the subpopulations are the different locations of the samples.
 
 The process starts with one infected host in one of the subpopulations, e.g in subpopulation $i$, who can infect another individual at rate $\lambda_i$, become uninfectious at rate $\mu_i$ by death or recovery, migrate to another deme $j$ at rate $m_{ij}$ or be sequenced at rate $\psi_i$ to become part of the phylogenectic tree.
-This process depicts the full transmission dynamics and specifically, the generation of the transmission tree that we observed from our sequence data.
+This process depicts the full transmission dynamics and specifically, the generation of the phylogenetic tree that we observed from our sequence data.
+
+Under this model, we are able to compute the likelihood of the multitype birth-death parameters for a given tree.
+This likelihood is derived in \cite{Kuhnert} by considering the probability of an individual evolved as observed in the tree.
+This derivation is analogous to the work in Stadler and Bonhoeffer (2013), which is based on ideas from (Maddison et al.
+2007)
 
@@
 
-\section{Multi-typed trees and epidemic trajectories}
+\section{Epidemic trajectories and structured trees}
 
-\section{Inference of epidemic dynamics and trajectories}
+We will refer to the full sequence of transmissions, recoveries/deaths, migrations and sampling events as an epidemic trajectory.
+One set of sequences, and therefore one phylogenetic tree, is the product of one epidemic trajectory.
+Moreover, the tree represents a fraction of the events in the epidemic trajectory, those that involve the sampled individuals.
+In epidemiology, we aim to learn about the true epidemic trajectory since we usually have incomplete information caused by unreported cases or unknown transmission chains.
 
-Under this model, we are able to compute the likelihood of the multitype birth-death parameters for a given phylogenetic tree.
-This likelihood is derived in \cite{Kuhnert} by considering the probability of an individual evolved as observed in the tree.This derivation is analogous to the work in Stadler and Bonhoeffer (2013), which is based on ideas from (Maddison et al.
-2007)
+From the sequences metadata, we know the location of the tree tips.
+However, if we know the epidemic trajectory we also know the location of the lineage at any point in time in the tree.
+We will refer to the phylogenies with ancentral locations mapped onto the tree as structured trees \cite{Vaughan2014}.
+In Fig \ref{fig:epitrajs} we show the epidemic trajectories corresponding to two different subpopulations and the structured tree of a set of samples.
+The change of ancestral location, represented by a change in color from blue to red in the tree, is caused by a migration event from one subpopulation to the other depicted in the epidemic trajectories.
 
-This phylodynamic likelihood is the tree prior in our Bayesian inference.
-We want to estimate the overall posterior distribution of the phylogenetic tree with ancestral locations $\mathcal{T}_c$, the epidemic trajectory $\mathcal{E}$, the substitution model parameters $\theta$ and the multi-type birth death parameters $\eta$:
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{figures/epitrajs.png}
+    \caption{Example of structured tree (up) and epidemic trajectories (bottom) for two types.
+The horizontal axis represents time, and for the epidemic trajectories the vertical axis is the population size N(t).
+Figure from Tim Vaughan.}
+    \label{fig:epitrajs}
+\end{figure}
+
+\section{BDMM-Prime Inference}
+
+We want to infer the epidemic trajectories under the multitype birth-death model fit to SARS-CoV-2 early epidemics in Europe.
+In order to do this, we have to estimate the overall posterior distribution of the structured tree $\mathcal{T}_c$, the epidemic trajectory $\mathcal{E}$, the substitution model parameters $\theta$ and the multi-type birth death parameters $\eta$.
+This posterior probability can be expressed with Bayes' formula as:
 
 \begin{equation}
-P(\mathcal{T}_c, \mathcal{E}, \theta, \eta | A, S) = \frac{\overbrace{P(A | \tau, \theta)}^{\text{tree likelihood}} \overbrace{P(\tau, S | \eta)}^{\text{tree prior}} \overbrace{P(\theta, \eta)}^{\text{parameters prior}}}{\underbrace{P(A, S)}_{\text{marginal likelihood}}}
+P(\mathcal{T}_c, \mathcal{E}, \theta, \eta | A ) = \frac{P(A | \mathcal{T}_c, \mathcal{E}, \theta, \eta ) P(\mathcal{T}_c, \mathcal{E}, \theta, \eta)}{P(A)}
+\label{eq:posterior1}
 \end{equation}
 
-where A is the sequence alignment and S the observed sequence times.
+where A is the sequence alignment.
+
+We make the following independence assumptions:
+
+\begin{align}
+P(A | \mathcal{T}_c, \mathcal{E}, \theta, \eta ) &= P(A | \mathcal{T},\theta )\\
+P(\mathcal{T}_c, \mathcal{E}, \theta, \eta) &= P(\mathcal{T}_c|\mathcal{T}, \eta) P(\mathcal{E}|\mathcal{T}_c, \eta) P(\mathcal{T} | \eta ) P(\theta) P(\eta)
+\label{eq:assumptions}
+\end{align}
+
+Where $\mathcal{T}$ is the rooted time tree without ancestral locations. Thus,
+
+\begin{equation}
+P(\mathcal{T}_c, \mathcal{E}, \theta, \eta | A) = P(\mathcal{T}_c|\mathcal{T}, \eta) P(\mathcal{E}|\mathcal{T}_c, \eta) \frac{P(A | \mathcal{T}, \theta) P(\mathcal{T} | \eta) P(\theta) P(\eta)}{P(A)}
+\label{eq:posterior2}
+\end{equation}
+
+We use a MCMC Metropolis hastings algorithm implemented in BEAST 2 to smaple from this distribution. No marginal likelihood P(A) calculation.
+
+The factorization of the posterior probability in this way allow us to use the fast solution and The third term in equation \ref{eq:posterior2} is the standard posterior probability infer in BDMM.
+SO we can use the fast solutions from ..
+{}
+The tree likelihood $P(A | \mathcal{T},\theta )$ is the probability of the sequence alignment and can be efficiently evaluated using Felsenstein’s pruning algorithm (Felsenstein, 1981).
+The tree prior, $P(\mathcal{T} | \eta)$ also called phylodynamic likelihood is derived from the multitype birth-death model.
+$P(\theta)$ and $P(\eta)$ represent our prior belief in the distribution of the population and substitution model parameters.
 
@@
 
-In our case, we have 6 demes, one for each of the countries/geographic location.
-A birth represents a transmission event of covid-19, a death event is an individual becoming uninfectious by overcoming the disease, being isolated or dying.
-From the birth rate and death rate we can estuimate the R0, key vallue in the epidemic understanding and control.
-A migration event happens when an individual travels from one country to another, we know that is an important way of covid-19 spreading and we are interested on the dynamics.
-
-We focus our analysis in the early spread in Europe of SARS-CoV-2, therefore we constrain the analysis to the period from the origin of the epidemic to March 8, when the first European lockdown in the Lombardy region started.
-During this time, for the European countries we expect an unimpeded spread of the virus that could be described by a constant R0 particular to each deme.
+The time range of our analysis lasts from the origin of SARS-CoV-2 to March 8, when the first European lockdown in the Lombardy region started.
+This period in the European countries we expect an unimpeded spread of the virus that could be described by a constant R0 particular to each deme.
 However, the migration rates could have changed during these months due to the increase public awareness, so we will consider a different migration rate for origin to 23 jan, 23 jan to end of february and 1 to 8 of March.
@@
 The becoming uninfectious rate we will fixed in 36.5, assuming that one individual is infectious for a exponential time distirbuted with mean 10.
+
+Since we focus on the early epidemic outbreak, we expect un unimpeded spread of the virus that can be described by the exponential growth of the infected population.
+Thus, we assume a constant basic reproductive number $R_0$ for each of the European countries.
+We assume that the $R_0$ for China changes two times during this period, on January 23 and March 1.
+
+We focus on a special scenario, where only the early epidemic outbreak, i.e.
+exponential growth of the infected population, is being considered.
+We can simplify the model to a constant rate birth-death process, where we assume no significant decrease in the number of susceptibles over time \cite{TanjaBonhoeffer2014}
 

### 2021-03-03 reports/mt-thesis/methods.tex
wip methods

@@
 
-\section{SARS-CoV 2 genome data set}
+\textbf{SARS-CoV 2 genome data set}
 
@@
 
-\section{The multitype birth death model}
+\textbf{The multitype birth death model}
 
@@
 
-We parameterize our model in terms of the effective reproductive number $R_i = \frac{\lambda_i}{\mu_i + \psi_i}$, a key value in epidemic control and understanding, the rate of becoming uninfectious $\delta_i = \mu_i + \psi_i$ and the probability of and individual to be sequenced $s_i = \frac{\psi_i}{\mu_i + \psi_i}$.
+We parameterize our model in terms of the effective reproductive number $R_i = \frac{\lambda_i}{\mu_i + \psi_i}$, a key value in epidemic control and understanding, the rate of becoming uninfectious $\delta_i = \mu_i + \psi_i$ and the probability of and individual to be sequenced $s_i = \frac{\psi_i}{\mu_i + \psi_i}$.\\
 
-\section{Epidemic trajectories and structured trees}
+\textbf{Epidemic trajectories and structured trees}
 
@@
 
-\section{BDMM-Prime Inference}
+\textbf{BDMM-Prime Inference}
 
@@
 
+While it is possible to do it in the same step \cite{epiestim} it is very slow and computationally expensive.
+With these two step analyses implmented in BDMM-Prime \cite{bdmm-primegit} the epidemic trajectory inference has almost no overcost and we obtain the same result.
+
 From the phylogenetic tree and the phylodynamics parameters we use Stochastic mapping to infer the ancestral traits and then to infer the epidemic trajectories.
@@
 
-\section{Modeling the early dynamics of SARS-CoV-2 in Europe}
+\textbf{Incorporating travel data}
+
+To incorporatet travel data in the model we have use a GLM model described in Lemey et al 2009.
+We use a generalized linear model to describe the migration matrix KxK, where the rpedictors are the number of dayly flight from one country to another from EUROSTATS, the average distance between the centrroids of the country and the population sizes from origin and destinations.
+The predictors are log transformed, we add a pseudocount to make them all positive, and standardized as described in ..
+
+$mij = c exp(\sum_{i = 1}^p \delta_i \beta_i xij)$
+
+The coefficients beta describe the effect size of the predictors in the migrations rate and the delta coefficients act as a model selection variables, taking 0-1 values including orr not that specific predictor in the model.
+
+GLM predictos data
+
+EUROSTAT, countries considers. Distance and population datasets.
+
+\textbf{Model specifications and priors}
 
@@
 
-\section{Incorporating travel data}
-
-To incorporatet travel data in the model we have use a GLM model described in Lemey et al 2009.
-We use a generalized linear model to describe the migration matrix KxK, where the rpedictors are the number of dayly flight from one country to another from EUROSTATS, the average distance between the centrroids of the country and the population sizes from origin and destinations.
-The predictors are log transformed, we add a pseudocount to make them all positive, and standardized as described in ..
-
-$mij = c exp(\sum_{i = 1}^p \delta_i \beta_i xij)$
-
-The coefficients beta describe the effect size of the predictors in the migrations rate and the delta coefficients act as a model selection variables, taking 0-1 values including orr not that specific predictor in the model.
-
-\section{GLM predictos data}
-
-EUROSTAT, countries considers. Distance and population datasets.
-
-\section{Bayesian inference of phylodynamic parameters}
-
 We use bayesian inference and Metropolis Hasting Monte Carlo algorith to infer our phylodynamics parameters and phylogeneit tree. We use the following priors
@@
 
-\subsubsection{R0}
 R0 constant except for China (and Italy? maybe not in the main analysis).
@@
 
-\section{Processing of epidemic trajectories}
+\textbf{Processing of epidemic trajectories}
 In our case, for each set of parameters and tree we will simulate 10000 trajectories with an epsilon of 0.3 and select one of them with importance sampling.
@@
 
-\section{Data availability}
+\textbf{Data availability}
 GISAID data table.
 
-\section{Code availability and analyses reproducibility}
+\textbf{Code availability and analyses reproducibility}
 All the code is availables at ...

### 2021-03-05 reports/mt-thesis/introduction.tex
methods wip

@@
 While due to the complexite and time consuming model we dow scaled to europe and only thhe initial phase with the first countries to experiment relevant sars cov 2 epidemics.
+
+the first reported cases were in China in early December \cite{},
+
+In a phylodynamic analysis, we aim to integrate genetic and epidemiological information to understand pathogen evolution and transmission dynamics.
+We can learn about the epidemic spread and the interactions between hosts from the imprint that these events leave in virus phylogenies.
+This is possible for RNA viruses, and in specific for SARS-CoV-2, because virus genetic evolution and epidemic processes are in the same time scale \cite{Grenfell2013} \cite{Volz2014}.
+More precisely, the field of phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process.
+\todo{Better in the Introduction?}

### 2021-03-05 reports/mt-thesis/methods.tex
methods wip

@@
 
-We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{GISAID} (accessed on November 2020).
+We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{gisaid} (accessed on November 2020).
 We follow the Nextstrain workflow for the curation of the dataset \cite{nextstrain}.
 Sequences with incomplete collection date, less than 27.000 bases in length or with more than 3.000 unknown bases are omitted.
-Also, sequences from known clusters of transmission or from the same patient are excluded from the analysis.
+Also, sequences from known clusters of transmission or from the same patient are excluded.
 The resulting dataset of x \todo{look number of seqs} sequences is aligned with MAFFT.
-The beggining and the end of the alignment are masked respectively by 100 and 50 sequences as well as sites  13402, 24389 and 24390, identified by Nextstrain as prone to sequencing errors.
+The begining and the end of the alignment are masked respectively by 100 and 50 sequences as well as sites  13402, 24389 and 24390, identified by Nextstrain as prone to sequencing errors.
 
-To focus on the early dynamics in Europe we select sequences from China, origin of the epidemic, France and Germany, the European countries with the earliest cases, and Italy and Spain, the European countries with the biggest outbreaks in March.
+To focus on the early dynamics in Europe we select sequences from China, origin of the epidemic; France and Germany, the European countries with the earliest cases; and Italy and Spain, the European countries with the biggest outbreaks in March.
 To take into account the dynamics in other regions of Europe we include a group of 50 sequences from other European countries.
@@
 Due to the large (and unprecendented) number of available genetic sequences for SARS-CoV-2, we need to subsample the alignment.
-Each sequence is subsampled with a probability of having a reported case in that country the day of sample collection, inversely weigted by the probability of having a sequence in GISAID that day from the country.
-With this subsampling protocol, we aim to get a constant sampling proportion accross the full period for each country.
+Each sequence is subsampled with a probability equal to the probability of having a reported case in that country the day of sample collection, inversely weighted by the probability of having a sequence in GISAID that day in the country.
+With this subsampling protocol, we aim to get a constant sampling proportion across the full period for each country.
 
-This dataset of genetic sequences is the main source of information for our phylodynamic analysis.
+This dataset of genetic sequences is the main source of information in our phylodynamic analysis.
 The goal is to infer the phylogenetic tree, i.e.
 the evolutionary tree-shaped relationship among the sequences, together with the epidemiological transmission parameters that gave rise to it.
-These parameters are defined within a population dynamic model and will inform us about the epidemic in the population that the sequences come from.
+These parameters are defined within a population dynamic model and will inform us about the epidemic that the viral genetic sequences come from.
 
@@
 Under this model, we are able to compute the likelihood of the multitype birth-death parameters for a given tree.
-This likelihood is derived in \cite{Kuhnert} by considering the probability of an individual evolved as observed in the tree.
-This derivation is analogous to the work in Stadler and Bonhoeffer (2013), which is based on ideas from (Maddison et al.
-2007)
+This likelihood is derived in \cite{Kuhnert2016} by considering the probability of an individual evolved as observed in the tree.
+This derivation is analogous to the work in \cite{Stadler2013}, which is based on ideas from \cite{Maddison2007}.
 
-We parameterize our model in terms of the effective reproductive number $R_i = \frac{\lambda_i}{\mu_i + \psi_i}$, a key value in epidemic control and understanding, the rate of becoming uninfectious $\delta_i = \mu_i + \psi_i$ and the probability of and individual to be sequenced $s_i = \frac{\psi_i}{\mu_i + \psi_i}$.\\
+We parameterize our model in terms of the effective reproductive number $R_i = \frac{\lambda_i}{\mu_i + \psi_i}$, a key value in epidemic control and understanding, the rate of becoming uninfectious $\delta_i = \mu_i + \psi_i$, the probability of and individual to be sequenced $s_i = \frac{\psi_i}{\mu_i + \psi_i}$ and the migration rates between locations $m_ij$.\\
 
@@
 In Fig \ref{fig:epitrajs} we show the epidemic trajectories corresponding to two different subpopulations and the structured tree of a set of samples.
-The change of ancestral location, represented by a change in color from blue to red in the tree, is caused by a migration event from one subpopulation to the other depicted in the epidemic trajectories.
+The change of ancestral location, represented by a change in color from blue to red in the tree, is caused by a migration event from one subpopulation to the other, depicted in the epidemic trajectories.
 
@@
 The horizontal axis represents time, and for the epidemic trajectories the vertical axis is the population size N(t).
-Figure from Tim Vaughan.}
+Figure from Tim Vaughan.
+\todo[inline]{Use figure from EpinInf paper?}}
     \label{fig:epitrajs}
@@
 
-\textbf{BDMM-Prime Inference}
+\textbf{Bayesian Inference with BDMM-Prime}
 
-We want to infer the epidemic trajectories under the multitype birth-death model fit to SARS-CoV-2 early epidemics in Europe.
-In order to do this, we have to estimate the overall posterior distribution of the structured tree $\mathcal{T}_c$, the epidemic trajectory $\mathcal{E}$, the substitution model parameters $\theta$ and the multi-type birth death parameters $\eta$.
+We want to infer the epidemic trajectories under the multitype birth-death model fit to a set of samples collected throughout SARS-CoV-2 early epidemics in Europe.
+In order to do this, we have to estimate the joint posterior distribution of the structured tree $\mathcal{T}_c$, the epidemic trajectory $\mathcal{E}$, the substitution model parameters $\theta$ and the multi-type birth death parameters $\eta$.
 This posterior probability can be expressed with Bayes' formula as:
@@
 
-Where $\mathcal{T}$ is the rooted time tree without ancestral locations. Thus,
+Where $\mathcal{T}$ is the rooted time tree without ancestral locations. Thus, we can express Equation \ref{eq:posterior1} in terms of the conditional distributions:
+
+\begin{align}
+P(\mathcal{T}_c, \mathcal{E}, \theta, \eta | A) &= P(\mathcal{T}_c|\mathcal{T}, \eta) P(\mathcal{E}|\mathcal{T}_c, \eta) \frac{P(A | \mathcal{T}, \theta) P(\mathcal{T} | \eta) P(\theta) P(\eta)}{P(A)}\\
+&= P(\mathcal{T}_c|\mathcal{T}, \eta) P(\mathcal{E}|\mathcal{T}_c, \eta) P(\mathcal{T}, \theta, \eta | A)
+\label{eq:posterior2}
+\end{align}
+
+We use a Markov chain Monte Carlo (MCMC) Metropolis-Hastings algorihtm to approximate this posterior.
+Since the MCMC only uses ratios of posterior probabilities, we avoid calculating the marginal likelihood $P(A)$.
+This algorithm is implemented in BDMM-Prime package \cite{bdmmp} for BEAST 2.6.3 \cite{Bouckaert2019}.
+BDMM-Prime first samples from $P(\mathcal{T}, \theta, \eta | A)$, then in a second step these samples are augmented by sampling from $P(\mathcal{T}_c|\mathcal{T}, \eta)$ and $P(\mathcal{E}|\mathcal{T}_c, \eta) $ and adding these variables to obtain the overall posterior $P(\mathcal{T}_c, \mathcal{E}, \theta, \eta | A)$.
+In this way, $P(\mathcal{T}_c|\mathcal{T}, \eta)$ and $P(\mathcal{E}|\mathcal{T}_c, \eta) $  are only calculated for a subset of samples instead of every MCMC step.
+
+While it is also possible to include $\mathcal{T}_c$ and $\mathcal{E}$ directly into the Bayes' rule as implemented in EpiInf package\cite{Vaughan2019}, the factorization of the posterior in these three terms allow us to use the standard birth-death-sampling tree prior implementations in BDMM \cite{Kuhnert2016} \cite{Scire2020} to compute $P(\mathcal{T}, \theta, \eta | A)$.
+This will speedup the analysis with almost no overhead compared to the standard BDMM inference and can be used in pre-existing multitype analyses withoud additional MCMC.\\
+
+\textbf{Stochastic mapping of ancestral locations and epidemic trajectories}
+
+To sample from $P(\mathcal{T}_c|\mathcal{T}, \eta)$, BDMM-Prime implements a stochastic mapping algorithm based on the work by \cite{Hohna2019}.
+A set of differential equations is numerically integrated over $\mathcal{T}$ to obtain the marginal probability of a location value at any point along each branch.
+From this probabilities, we can derive time-dependent rates that define the changes of the location along the tree according to a continuous-time Markov process.
+Then, we can simulate forward in time from the root of the tree the location trajectories down the tree edges.
+For a detailed derivation of the stochastic mapping of the algorithm refer to \cite{VaughanWIP}.\\
+
+In the case of $P(\mathcal{E}|\mathcal{T}_c, \eta)$, following Bayes'rule:
 
 \begin{equation}
-P(\mathcal{T}_c, \mathcal{E}, \theta, \eta | A) = P(\mathcal{T}_c|\mathcal{T}, \eta) P(\mathcal{E}|\mathcal{T}_c, \eta) \frac{P(A | \mathcal{T}, \theta) P(\mathcal{T} | \eta) P(\theta) P(\eta)}{P(A)}
-\label{eq:posterior2}
+P(\mathcal{E}|\mathcal{T}_c, \eta) = \frac{P(\mathcal{T}_c, \eta|\mathcal{E}) P(\mathcal{E})}{P(\mathcal{T}_c, \eta)} = \frac{P(\mathcal{T}_c|\mathcal{E}) P(\mathcal{E}| \eta)}{P(\mathcal{T}_c | \eta)}
+\label{trajectories}
 \end{equation}
 
-We use a MCMC Metropolis hastings algorithm implemented in BEAST 2 to smaple from this distribution. No marginal likelihood P(A) calculation.
+We can easily compute the probability of a structured tree for a given epidemic trajectory $P(\mathcal{T}_c|\mathcal{E})$ as described in \cite{Vaughan2019}.
+Each node in the structured tree must correspond to a compatible event in the epidemic trajectory for this probability to be nonzero.
+If we simulate an epidemic trajectory directly from $P(\mathcal{E}| \eta)$, it is very likely that the simulated events will not match with the events in the tree and $P(\mathcal{T}_c|\mathcal{E})$ will be 0.
+To avoid this problem, we simulate from $P^*(\mathcal{E}| \eta)$, which guarantees trajectories with non-zero probabilities by enforcing the tree events.
+Provided we weight these trajectories accordingly we can efficiently sample from $P(\mathcal{E}|\mathcal{T}_c, \eta)$:
 
-The factorization of the posterior probability in this way allow us to use the fast solution and The third term in equation \ref{eq:posterior2} is the standard posterior probability infer in BDMM.
-SO we can use the fast solutions from ..
-{}
-The tree likelihood $P(A | \mathcal{T},\theta )$ is the probability of the sequence alignment and can be efficiently evaluated using Felsenstein’s pruning algorithm (Felsenstein, 1981).
-The tree prior, $P(\mathcal{T} | \eta)$ also called phylodynamic likelihood is derived from the multitype birth-death model.
-$P(\theta)$ and $P(\eta)$ represent our prior belief in the distribution of the population and substitution model parameters.
+\begin{equation}
+P(\mathcal{E}^{(a)}|\mathcal{T}_c, \eta) = w_a^* P^*(\mathcal{E}^{(a)}| \eta) \propto P(\mathcal{T}_c | \mathcal{E}^{(a)}) \frac{P(\mathcal{E}^{(a)} | \eta)}{P^*(\mathcal{E}^{(a)} | \eta)} P^*(\mathcal{E}^{(a)}| \eta)
+\label{eq:traj}
+\end{equation}
 
-While it is possible to do it in the same step \cite{epiestim} it is very slow and computationally expensive.
-With these two step analyses implmented in BDMM-Prime \cite{bdmm-primegit} the epidemic trajectory inference has almost no overcost and we obtain the same result.
+The epidemic trajectories are simulated with an adaptive tau-leaping algorithm \cite{Gillespie2000} in BDMM-Prime.
+This method is based on the Gillespie algorithm \cite{Gillespie1977} but the simulation time is divided in small intervals and the number of events is drawn from a poisson distribution for each interval.
+This algorithm is more efficient that Gillespie when we have a high number of individuals since we have to update the rates less often.
 
-From the phylogenetic tree and the phylodynamics parameters we use Stochastic mapping to infer the ancestral traits and then to infer the epidemic trajectories.
+We simulate a fixed number of trajectories, called particles, for each set of population parameters $\eta$ and structured tree $\mathcal{T}_c$.
+The trajectories are simulated until some point and are weigthed as described in Equation \ref{eq:traj}.
+We resample using importance sampling to get rid of low-weighted trajectories and then resume the simulation of all the particles from the sampled trajectory ending point.
+This process is repeated until the end of the simulation.
+Then, we sample one single trajectory from the set of final trajectories with important sampling and record each event and its timing to future analysis \cite{VaughanWIP}.
+All these steps are implemented by BDMM-Prime trajectory logger and we use it in our analysis.\\
 
-$P(Tc | T, P)$
+We conducted two different analyses.
+One with constant migration rates as in the standard multi-type birth death model and the other one with time-changing migration rates informed by travel data.
+We use the BDMM-Prime package \cite{bdmmp} implemented in BEAST 2.6.3 \cite{beast}.
+We run 10 parallel chains for analysis, of $10^7$ iterations each with different initial values.
+These chains are assesed for convergence using Tracer v.1.7.1 and then combined after removing a 10\% burnin.
+We check that the effective sample size (ESS) is greater than 200 for all parameters.
+The stochastic mapping of structured trees and epidemic trajectories is performed in a second step using the trace and tree logs from the MCMC.
+We use 300, 1000, 3000 and 10000 particles in the tau-leaping simulation of epidemic trajectories, and a tolerance of 0.03 for selecting tau leap length.\\
 
-New method to infer epidemic trajectories!!
+\textbf{Incorporating air travel data and geographic distances}
 
-$P(E | Tc, P)$
+In order to inform our migration rates with external information, and to avoid having to estimate a large number of migration rates we implement a generalized linear model (GLM) for the migration matrix based on \cite{Lemey2014} and \cite{Nicola2019}.
+This GLM model parametrizes the migration rates as a log linear function of a number of potential predictors, Equation \ref{eq:glm}.
+For each predictor $x_i$, the GLM parameterization includes a coefficient $\beta_i$, which quantifies the effect size of the predictor, and a binary indicator variable $\delta_i$, that allows the predictor to be included or excluded from the model.
+The parameter $c$ represents the overal magnitude of migration.
+This means that if every indicator is 0, every migration rate will be equal to this parameter.
+In the MCMC, we estimate the effect size of each of the predictors, as well as their inclusion probability $\mathbb{E}[\delta_i]$.
 
-Stochastic mapping of ancestral traits
+\begin{equation}
+m_{ij} = c \exp(\sum_{i = 1}^p \delta_i \beta_i xij)
+\label{eq:glm}
+\end{equation}
 
-Read paper and tim presentation
+In a similar way to the GLM formulation in \cite{Lemey2020} we consider as predictors air travel data and geographical distance between locations.
+We define three time periods in our analysis: from the origin of the epidemic to January 23 (Hubei lockdown), from January 23 to end of February, from March 1 to March 8.
+The international travel during these months decreased drastically due to the increasing awareness about the SARS-CoV-2 pandemic as is shown in Figure \ref{fig:travel}.
+Thus, we would expect the migration rates to change during the time of the analysis.
 
-Stochastic mapping of epidemic trajectories
+The air travel data is obtained from EUROSTATS transport datasets \textit{avia\_paexcc} and \textit{avia\_paincc}, in particular the passengers carried departures values for December 2019, January, February and March 2020.
+We compute the average daily flux of passengers for each of the time periods.
+The geographical distance is defined as the great circle pairwise distance between the centroids of the countries, computed based on the Natural Earth project (the 1:50m resolution version) world map 2013.
+The predictors are log transformed, a pseudocount is added to make all values positive, and then they are standardized, following the description in \cite{Lemey2014}.
 
-From the phylogenie with ancestral locations and the set of phylodynamic parameters we can simulate epidemic trajectories using an algorithm similar to Gillespie (with tau leaping and particle filtering?
-the subsampling).
-And most improtant, we can compute the probability of that specific trajectory given those parrameters and tree.
-
-$P(E|Tc, P)$
-
-Any trajectories without the events that are respresent in the tree has probability 0, so to avoid the simualtion of this trajectories, no time effcieint, we will enforce the events that arer represented in the tree (and will include a weigth).
-Also to avoid low probabilities trajectories, we will importance subsample one trajectory after a certain time of simulation according to the weigths detemrined by its probabilities.
-
-\textbf{Incorporating travel data}
-
-To incorporatet travel data in the model we have use a GLM model described in Lemey et al 2009.
-We use a generalized linear model to describe the migration matrix KxK, where the rpedictors are the number of dayly flight from one country to another from EUROSTATS, the average distance between the centrroids of the country and the population sizes from origin and destinations.
-The predictors are log transformed, we add a pseudocount to make them all positive, and standardized as described in ..
-
-$mij = c exp(\sum_{i = 1}^p \delta_i \beta_i xij)$
-
-The coefficients beta describe the effect size of the predictors in the migrations rate and the delta coefficients act as a model selection variables, taking 0-1 values including orr not that specific predictor in the model.
-
-GLM predictos data
-
-EUROSTAT, countries considers. Distance and population datasets.
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{figures/flight_data.png}
+    \caption{\todo[inline]{todo}}
+    \label{fig:epitrajs}
+\end{figure}
 
@@
 
-The time range of our analysis lasts from the origin of SARS-CoV-2 to March 8, when the first European lockdown in the Lombardy region started.
-This period in the European countries we expect an unimpeded spread of the virus that could be described by a constant R0 particular to each deme.
-However, the migration rates could have changed during these months due to the increase public awareness, so we will consider a different migration rate for origin to 23 jan, 23 jan to end of february and 1 to 8 of March.
-We assume a constant sampling proportion different for each of the demes, to account for difference in the sequencing efforts.
-The becoming uninfectious rate we will fixed in 36.5, assuming that one individual is infectious for a exponential time distirbuted with mean 10.
+Since we focus on the early epidemic outbreak, we expect un unimpeded spread of the virus that can be described by the exponential growth of the infected population and no significant decrease in the number of susceptibles over time \cite{Boskova2014}.
+Thus, we assume a constant basic reproductive number $R_0$ for each of the European countries before the lockdown in the Lombardy region in March 8.
+However, in China the first wave of the epidedmic had almost concluded in March so we can not make the same assumption.
+We fixed the reproductive number in China and include three different periods, changing on January 23 and February 10 based on \cite{Pan2020} \cite{Wu2020} \cite{Xiao2020}.
+Before January 23, we use the estimation in \cite{Park2020} \cite{Billah2020} to fix the basic reproductive number $R_0$ in China to $2.9$ .
+From January 23 to February 10 and from February 11 to March 8, we calculate the average effective reproductive number based on the estimations in \cite{covidre} \cite{Huisman2020} and obtain the $R_e$ in China to be $1.1$ and $0.43$ respectively.
 
-Since we focus on the early epidemic outbreak, we expect un unimpeded spread of the virus that can be described by the exponential growth of the infected population.
-Thus, we assume a constant basic reproductive number $R_0$ for each of the European countries.
-We assume that the $R_0$ for China changes two times during this period, on January 23 and March 1.
+In Table \ref{table:priors} we show the values and prior distributions used in the Bayesian inference of the multitype birth death model and substitution parameters.
+Several of these model specifications are identical to those from \cite{Nadeau2020}.
 
-We focus on a special scenario, where only the early epidemic outbreak, i.e.
-exponential growth of the infected population, is being considered.
-We can simplify the model to a constant rate birth-death process, where we assume no significant decrease in the number of susceptibles over time \cite{TanjaBonhoeffer2014}
+\renewcommand{\arraystretch}{1.2}
+\begin{table}[h]
+\footnotesize
+\begin{tabular}{@{}lll@{}}
+\toprule
+\textbf{Parameter}  & \textbf{Value or Prior} & \textbf{Rationale}                                              \\ \hline
+\begin{tabular}[c]{@{}l@{}}Nucleotide substitution \\ model\end{tabular} &
+  HKY + Gamma &
+  \begin{tabular}[c]{@{}l@{}}Unequal transition/transversion rates, \\ unequal base frequencies, rate\\ heterogeneity among sites with 4 categories\end{tabular} \\
+Clock rate          & 0.0008                  & Approximatelly 24 mutations per year \textbackslash{}cite\{10\} \\
+Location of origin  & Hubei                   & Putative pandemic origin                                        \\
+Time of origin      & Lognormal (-1, 0.2)     & \begin{tabular}[c]{@{}l@{}}Median 26 October, 95\% IQR 22 August \\ to 8 December 2019 \end{tabular}       \\
+\begin{tabular}[c]{@{}l@{}}European countries \\ reproductive number\end{tabular} &
+  Lognormal (0.8, 0.5) &
+  Median 2.2, 95\% IQR 0.8 to 5.9 \\
+\begin{tabular}[c]{@{}l@{}}China reproductive \\ number\end{tabular} &
+  \begin{tabular}[c]{@{}l@{}}Origin - Jan 23: 2.9\\ Jan 23 - Feb 10: 1.1\\ Feb 11 - Mar 8:  0.43\end{tabular} &
+  Fixed based on \textbackslash{}cite\{\}. \\
+\begin{tabular}[c]{@{}l@{}}Becoming uninfectious \\ rate \end{tabular} &
+  36.5 y-1 &
+  \begin{tabular}[c]{@{}l@{}}Period between infections and becoming \\ uninfectious assumed exponentially \\ distributed with a mean of 10 days\end{tabular} \\
+Sampling proportion &                         & Upper bounds based on reported cases:                           \\
+China               & Uniform (0, 0.12)       &                                                                 \\
+France              & Uniform (0, 0.07)       &                                                                 \\
+Germany             & Uniform (0, 0.07)       &                                                                 \\
+Italy               & Uniform (0, 0.01)       &                                                                 \\
+Spain               & Uniform (0, 0.08)       &                                                                 \\
+Other European      & Uniform (0, 0.03)       &                                                                 \\
+Sampling start date & December 23, 2019       & Just before first sample from China                             \\
+\begin{tabular}[c]{@{}l@{}}Sampling end date \\ (only China)\end{tabular} &
+  January 23, 2020 &
+  \begin{tabular}[c]{@{}l@{}} Only included Chinese sequences before \\ Lockdown in Hubei \end{tabular} 		\\
 
-We use bayesian inference and Metropolis Hasting Monte Carlo algorith to infer our phylodynamics parameters and phylogeneit tree. We use the following priors
+\begin{tabular}[c]{@{}l@{}}Migration rates \\ (no GLM analysis)\end{tabular} & Lognormal(0,1) & Median, IQR                            \\
+\begin{tabular}[c]{@{}l@{}}Migration rates \\ (GLM analysis)\end{tabular} & Uniform(0,50) & \begin{tabular}[c]{@{}l@{}}7.3 days as mean of exponential time \\ to travel for every individual \end{tabular}\\
+\begin{tabular}[c]{@{}l@{}}Migration rates \\ change times \\(GLM analysis)\end{tabular} & December 23, 2019 and March 1, 2020       & \\
+\begin{tabular}[c]{@{}l@{}}Global scaler \\ GLM parameter\end{tabular} & December 23, 2019 & Median, IQR                              \\
+\begin{tabular}[c]{@{}l@{}}Coefficients \\ GLM predictors\end{tabular} & Normal (0, 2) & Median 0, IQR                            							\\
+\begin{tabular}[c]{@{}l@{}}Binary indicators \\ GLM predictors\end{tabular} & 0 - 1 equiprobable & BitFlipOperator.  			\\
 
-Clock rate and substitution model. Fixed to 8x10-4, HKY 4 gamma categories with priors...
-BDMM-Prime epi parameterization:
-R0
-Become uninfectious rate
-Sampling rates
-Migration rates
-
-R0 constant except for China (and Italy? maybe not in the main analysis).
-
-Prior Log Normal (0.8, 0.5) Median 2.2, 95IQR 0.8 to 5.9.
-(Lai A. 2020 prior log normal (0,1) median 1.0 0.1-7.10)
-
-- Main analysis: We fix China R0 to 2.7 (23 jan) 1.3 (10 feb) 0.8
-- Use a decreasing paramater as Sarah with migration rate? But then we would need to add sequences.
-- Or instead of fixing it we can put a strong informative prior.
-- Allow Italy R0 to change March 1.
-
-Meta-analyses R0:
-
-R0 in China (3.14,95\%CI,2.40–4.09).\cite{Billah2020}
-
-R0value 2.90 (2.32, 3.63
- 2.9 (95\% CI: 2.1–4.5)\cite{Park2020}
-
-Periods:
-until 23 jan R0 2.9 
-23 jan - 10 feb 1.1 dashboard cevo
-10 feb - 8 mar 0.43 dashboard cevo
-
-Become uninfectious rate same for all demes, fixed to 36.5.
-Migrations among demes, 3 epochs.
-Sampling proportion 0 before first sample. Upperbounded by number of cases.
-
-Implementation BEAST 2.6.3 and BDMM-Prime package.
-We run 5 chains of 10e7 iterations witth different seeds.
-
-\textbf{Processing of epidemic trajectories}
-In our case, for each set of parameters and tree we will simulate 10000 trajectories with an epsilon of 0.3 and select one of them with importance sampling.
-We have one trtajectory for each step in the MCMC. We subsampled x.
+\bottomrule
+\end{tabular}
+\caption{Caption. Source of case data. \todo[inline]{TODO. Vertical alignment}}
+\label{table:priors}
+\end{table}
 
 \textbf{Data availability}
-GISAID data table.
+The SARS-CoV-2 genetic data can be downloaded from www.gisaid.org.
+Supplementary Table \ref{stable:seqs} lists accession numbers for the genetic sequences used in this study.
+Reported SARS-CoV-2 case counts are obtained from the country official agencies indicated in Table \ref{table:cases}.
+Flight data can be obtained from EUROSTATS https://ec.europa.eu/eurostat/.
 
 \textbf{Code availability and analyses reproducibility}
-All the code is availables at ...
-We have implement the full workflow with Snakemake so its fully reproducible.
+All the code is available at ...
+To ensure the reproducibility of the analysis, we have implemented the whole workflow with Snakemake \cite{snakemake}.
+The workflow is a modified version of Nextrain workflow \cite{nextstrain} with the additional rules for BEAST 2 analysis and results processing.

### 2021-03-05 reports/mt-thesis/methods.tex
wip methods almost finished

@@
 
-\textbf{SARS-CoV 2 genome data set}
-
-We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{gisaid} (accessed on November 2020).
+\textbf{SARS-CoV 2 genome data set.} 
+We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{Shu2017} (accessed on November 2020).
 We follow the Nextstrain workflow for the curation of the dataset \cite{nextstrain}.
@@
 Also, sequences from known clusters of transmission or from the same patient are excluded.
-The resulting dataset of x \todo{look number of seqs} sequences is aligned with MAFFT.
+The resulting worldwide sequence dataset is aligned with MAFFT.
 The begining and the end of the alignment are masked respectively by 100 and 50 sequences as well as sites  13402, 24389 and 24390, identified by Nextstrain as prone to sequencing errors.
@@
 
-\textbf{The multitype birth death model}
-
+\textbf{The multitype birth death model.} 
 To study the early dynamics of SARS-CoV-2 in the European countries, we use a simplified version of the multitype birth-death model described in \cite{Kuhnert2016}, following the analysis in \cite{Nadeau2020}.
@@
 
-\textbf{Epidemic trajectories and structured trees}
-
+\textbf{Epidemic trajectories and structured trees.} 
 We will refer to the full sequence of transmissions, recoveries/deaths, migrations and sampling events as an epidemic trajectory.
@@
 
-\textbf{Bayesian Inference with BDMM-Prime}
-
+\textbf{Bayesian Inference with BDMM-Prime.} 
 We want to infer the epidemic trajectories under the multitype birth-death model fit to a set of samples collected throughout SARS-CoV-2 early epidemics in Europe.
@@
 
+$P(\mathcal{T}_c|\mathcal{T}, \eta)$, $P(\mathcal{E}|\mathcal{T}_c, \eta)$ and $P(\mathcal{T}, \theta, \eta | A)$ are the posterior probabilities of the structured tree, the epidemic trajectory and the joint posterior of the phylogenetic tree (without ancestral locations) and the model parameters.
+The tree likelihood $P(A | \mathcal{T},\theta )$ is the probability of the sequence alignment and can be efficiently evaluated using Felsenstein’s pruning algorithm \cite{Felsenstein1981}.
+The tree prior, $P(\mathcal{T} | \eta)$ also called phylodynamic likelihood is derived from the multitype birth-death model.
+$P(\theta)$ and $P(\eta)$ represent our prior belief in the distribution of the population and substitution model parameters.
+
 We use a Markov chain Monte Carlo (MCMC) Metropolis-Hastings algorihtm to approximate this posterior.
@@
 
-While it is also possible to include $\mathcal{T}_c$ and $\mathcal{E}$ directly into the Bayes' rule as implemented in EpiInf package\cite{Vaughan2019}, the factorization of the posterior in these three terms allow us to use the standard birth-death-sampling tree prior implementations in BDMM \cite{Kuhnert2016} \cite{Scire2020} to compute $P(\mathcal{T}, \theta, \eta | A)$.
+While it is also possible to include $\mathcal{T}_c$ and $\mathcal{E}$ directly into the Bayes' rule as implemented in EpiInf package \cite{Vaughan2019}, the factorization of the posterior in these three terms allow us to use the standard birth-death-sampling tree prior implementations in BDMM to compute $P(\mathcal{T}, \theta, \eta | A)$ \cite{Kuhnert2016} \cite{Scire2020}.
 This will speedup the analysis with almost no overhead compared to the standard BDMM inference and can be used in pre-existing multitype analyses withoud additional MCMC.\\
@@
 \textbf{Stochastic mapping of ancestral locations and epidemic trajectories}
-
 To sample from $P(\mathcal{T}_c|\mathcal{T}, \eta)$, BDMM-Prime implements a stochastic mapping algorithm based on the work by \cite{Hohna2019}.
@@
 
-\textbf{Incorporating air travel data and geographic distances}
-
+\textbf{Incorporating air travel data and geographic distances.}
  In order to inform our migration rates with external information, and to avoid having to estimate a large number of migration rates we implement a generalized linear model (GLM) for the migration matrix based on \cite{Lemey2014} and \cite{Nicola2019}.
@@
     \includegraphics[width=\textwidth]{figures/flight_data.png}
-    \caption{\todo[inline]{todo}}
+    \caption{\todo[inline]{improve figure}}
     \label{fig:epitrajs}
@@
 
-\textbf{Model specifications and priors}
-
+\textbf{Model specifications and priors.} 
 Since we focus on the early epidemic outbreak, we expect un unimpeded spread of the virus that can be described by the exponential growth of the infected population and no significant decrease in the number of susceptibles over time \cite{Boskova2014}.
@@
 \end{tabular}
-\caption{Caption. Source of case data. \todo[inline]{TODO. Vertical alignment}}
+\caption{Caption \todo[inline]{finish table and caption}}
 \label{table:priors}
@@
 
-\textbf{Data availability}
-The SARS-CoV-2 genetic data can be downloaded from www.gisaid.org.
-Supplementary Table \ref{stable:seqs} lists accession numbers for the genetic sequences used in this study.
-Reported SARS-CoV-2 case counts are obtained from the country official agencies indicated in Table \ref{table:cases}.
+\textbf{Data availability.}
+The SARS-CoV-2 genetic data can be downloaded from www.gisaid.org \cite{Shu2017}.
+Supplementary Table \ref{stable:seqs} \todo{todo} lists accession numbers for the genetic sequences used in this study.
+Reported SARS-CoV-2 case counts are obtained from the country official agencies indicated in Table \ref{table:cases} \todo{todo}.
 Flight data can be obtained from EUROSTATS https://ec.europa.eu/eurostat/.
 
-\textbf{Code availability and analyses reproducibility}
+\textbf{Code availability and analyses reproducibility.}
 All the code is available at ...

### 2021-03-19 reports/mt-thesis/discussion.tex
wip mt

--- 2020-11-27 reports/paper/main.tex
@@
-\documentclass[12pt,]{article}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{fullpage}
+\chapter{Discussion}
+\todo[inline]{TODO discussion}
 
-% \usepackage{authblk}
-\usepackage{graphicx}
-\usepackage{import} 
-\usepackage{subfig}
-\graphicspath{{../../results/}}
+\begin{itemize}
+\item Case counts discussion and reporting rates, second wave in Europe. Other estimations in other studies.
 
-\usepackage[left]{lineno}
-\linenumbers
-
-\usepackage{todonotes}
-
-\title{A comprehensive study of the early phylodynamics of SARS-CoV-2 in Europe}
-
-\begin{document}
-\maketitle
-
-\abstract{This project focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe.
-We apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences.
-The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.}
-
-\section*{Introduction}
-
-\section*{Methods}
-
-\section*{Results}
-
-\subsubsection*{Total case counts}
-
-Figures \ref{fig:gribbon} and \ref{fig:trajs}.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.8\textwidth]{201014_europe2_figtraj01.png}
-    \caption{Inferred population size summary statistics for each deme over time.
-The line represents the median population trajectory and the interval is the interquartile range from a subsampled set of inferred population trajectories.}
-    \label{fig:gribbon}
-\end{figure}
-
-Inferred population trajectories that account for all cases in the country.
+% Inferred population trajectories that account for all cases in the country.
 It is not an independent estimate since we are using informative priors for the sampling rates (upperbound computed as sequences/reported cases) and the subsampling scheme also includes the number of cases in the country (but that is a weaker link I think.\\
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.9\textwidth]{201014_europe2_figtraj02.png}
-    \caption{Inferred population size trajectories over time.
-A subsampled set of 200 inferred trajectories is plotted.
-In each subplot, the trajectories (solid lines) are compared with the ECDC total case count data (dashed line).}
-    \label{fig:trajs}
-\end{figure}
+\item First introduction other studies.
 
-Comparison with ECDC case counts: delay in notifications and underreporting, for some of the  countries at the end of the periord the reported cases are very closed to the trajectories, e.g.
-Italy, Germany.
-Hubei-China huge difference in tendency and absolute numbers.\\ \todo{Working on these with new analyses}
+\item Migration patterns other studies.
 
-Measure the (mean?) difference between inferred and reported case counts and compare among countries. By day or in general? Absolute or relative to the number of cases that day?
+\item Discuss assumptions of the model, caveats and possible improvements.
 
-\subsubsection*{First introductions}
-Figures \ref{fig:first}, \ref{fig:first_src} and \ref{fig:first_dest}.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj03.png}
-    \caption{First introductions event time and source in each deme.
-\textbf{A} Probability density of the time of the first introduction for each deme from the subsampled set of trajectories.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of Hubei, the distribution of the time of the origin is plotted, since it was defined as the origin of the epidemic in the analysis with probability 1.
-For the other five demes, the distribution of the time of the first migration into the corresponding deme is plotted.
-\textbf{B} Probability density of the source of the first introduction event in each deme for the set of subsampled trajectories.}
-    \label{fig:first}
-\end{figure}
-
-We want to answer the question of when and where the first case in Europe occured.
-No clear, Germany?
-Figure \ref{fig:first} A.
-We can measure the time between first case and first reported case and compare between countries.
-\\
-\todo{ticks for every month}
-And for each country where this first case came from, more support for Hubei introduction in all countries Figure \ref{fig:first} B.\\
-
-Measure expected day of introduction.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj04.png}
-    \caption{Ordered set of countries by its first introduction, i.e.
-first case in the country.
-The color corresponds to the first country with a case, and the size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first introductions.}
-    \label{fig:first_dest}
-\end{figure}
-
-In Figure \ref{fig:first_src}, we are looking at the ordered set of introduction intoo european countries.
-For each trajectory, we have the fist introduction into each country happening in a determined order, i.e.
-the first european country with a covid case, the second country with a covid case...
-So we are vissualizing the number of trajectories that share the same order of first cases.
-More trajectories where the first case was to Germany, no clear order after the first country.
-Spain and Other European last one.\\
-\todo{Explain this better}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj05.png}
-    \caption{Ordered set of countries by its first migration, i.e.
-spreading event out of the country.
-The color corresponds to the last country with a migration to other countries.
-The size of the box with the country name / ribbon / flux is the number / proportion of trajectories that have that particular order of first migrations.}
-    \label{fig:first_src}
-\end{figure}
-
-In Figure \ref{fig:first_dest} we are looking at the ordered set of countries spreading covid.
-In the first position we have Hubei for all trajectories since it is the origin of the epidemic.
-In second position we will have the first European country from where a covid case was imported to other country.
-No clear dynamics except that moore support for Spain being the last country having a first exported covid case.
-Germany first one.\\
-
-\subsubsection*{Migration vs within-region transmission}
-
-Figure \ref{fig:events}\\.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj06.png}
-    \caption{Mean inferred cumulative number of events (births and migrations) over time. A subsampled set of 200 inferred trajectories is plotted.}
-    \label{fig:events}
-\end{figure}
-
-This question is addresed on the original paper from Sarah.  Maybe could be interesting to look at this perspective since it is important for border closure measures.
-
-\todo{Think about results}
- Not sure if we can have any meaningful results insights on the topic.
-Migrations started very early in our analysis to all european countries.
-Withing regions transmision account for most of the cases in the countries from late january onwards (when first cases were being reported in Europe).\\
-
- Measure the time between the first outcoming mirgration from the country  (specially interesting for Hubei) and first reported case.
-This could be interesting to say if a extreme measure closing borders with the first case could be effective to impede transmission to other countries: percentage of trajectories whre transmission to Europe would have been avoided.
-For other countries we can look at how many migrations events could have been avoided (and how many not) if the country closed borders after first reported case.\\
-
- Measure the time between first birth/migration and first reported case and compare between countries.
-How well did the countries detecting the first cases, there were already within region transmision?\\
-
- Measure the time between the first migration into a country and the first birth in the country.
-Or the first outcoming mirgration from the country and first birth.
-This could be interesting to say if we should focus or not the screening and testing capacities to detect incoming migrations or a more general strategy to find cases in the population.
-Is it different for each country?\\
-
-\subsubsection*{Migrations patterns}
-Figures \ref{fig:migs_srcdest} and \ref{fig:migs}.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{201014_europe2_figtraj07.png}
-    \caption{}
-    \label{fig:migs_srcdest}
-\end{figure}
-
-\begin{figure}[!tbp]
-  \centering
-
-  \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201014_europe2_figtraj08a.png}
-  \label{fig:migs1}
-  \end{minipage}
-  \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201014_europe2_figtraj08b.png}
-  \label{fig:migs2}
-  \end{minipage}
-  \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201014_europe2_figtraj08c.png}
-  \label{fig:migs3}
-  \end{minipage}
-  \caption{Migration flux among demes over the three periods defined in the analysis.}
-  \label{fig:migs}
-\end{figure}
-
-Similar information in both plots, but in the chord plots instead of a dayly evolution the time period is split on three (as in the GLM analysis).
-Chord plots are nicer and easier to understand I think, but barplots shows the great detail of the results of the model.
-Another advantage of the chrod plot is that the ir shows mean absolute values and not only relative values.\\
-
-Hubei-China is the majority source of migrations for all countries till February and then some patterns emerge (we expect more interesting results with the added info in GLM analysis).\\
-
-\todo{Legend in plot 7 and change white text}
-
-\subsubsection*{GLM predictors}
-Which channels were the main sources of transmission across national borders.
-
-\subsubsection*{Epidemiological parameters}
-
-Not sure about this section. But maybe it make sense for completeness to include, briefly, the values of estimate R0, migration and sampling rates.
-
-\section*{Discussion}
-
-\section*{Conclusion}
-
-\end{document}
+\item Number of particles
+\end{itemize}

### 2021-03-19 reports/mt-thesis/introduction.tex
wip mt

@@
 More precisely, the field of phylodynamics studies how phylogenetic trees are being generated and infers the population parameters behind that process.
-\todo{Better in the Introduction?}
+
+\todo[inline]{TODO introduction}
+
+\begin{itemize}
+\item Importance of understanding the spread of the virus to prevent future outbreaks
+
+\item About phylodynamics/phylogeographics? Use of genetic sequences as a source of information combined with other sources of information as travel data
+
+\item About what we know of the introduction of sars-cov-2 in Europe and early dynamics till 8 March
+
+\item About what we know of the case counts in Europe? Can sequences help?
+
+\item Introduce/formulate the questions: case counts, first introductions, ?migration vs within region transmission?, migration patterns, ?border closures.
+\end{itemize}

### 2021-03-19 reports/mt-thesis/methods.tex
wip mt

@@
 \textbf{SARS-CoV 2 genome data set.} 
-We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{Shu2017} (accessed on November 2020).
+We assemble a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{Shu2017} (accessed on November 2020).
 We follow the Nextstrain workflow for the curation of the dataset \cite{nextstrain}.
@@
 
-To focus on the early dynamics in Europe we select sequences from China, origin of the epidemic; France and Germany, the European countries with the earliest cases; and Italy and Spain, the European countries with the biggest outbreaks in March.
+To focus on the early dynamics in Europe we select sequences from China, the origin of the epidemic; France and Germany, the European countries with the earliest cases; and Italy and Spain, the European countries with the biggest outbreaks in March.
 To take into account the dynamics in other regions of Europe we include a group of 50 sequences from other European countries.
-We limit our sample of Chinese genomes to sequences until January 23, the starting date of the lockdown in Hubei.
+We limit our sample of Chinese genomes to sequences collected before January 23, the starting date of the lockdown in Hubei.
 
@@
 
-Where $\mathcal{T}$ is the rooted time tree without ancestral locations. Thus, we can express Equation \ref{eq:posterior1} in terms of the conditional distributions:
+where $\mathcal{T}$ is the rooted time tree without ancestral locations. Thus, we can express Equation \ref{eq:posterior1} in terms of the conditional distributions:
 
@@
 While it is also possible to include $\mathcal{T}_c$ and $\mathcal{E}$ directly into the Bayes' rule as implemented in EpiInf package \cite{Vaughan2019}, the factorization of the posterior in these three terms allow us to use the standard birth-death-sampling tree prior implementations in BDMM to compute $P(\mathcal{T}, \theta, \eta | A)$ \cite{Kuhnert2016} \cite{Scire2020}.
-This will speedup the analysis with almost no overhead compared to the standard BDMM inference and can be used in pre-existing multitype analyses withoud additional MCMC.\\
+This will speed up the analysis with almost no overhead compared to the standard BDMM inference and can be used in pre-existing multitype analyses withoud additional MCMC.\\
 
@@
 The epidemic trajectories are simulated with an adaptive tau-leaping algorithm \cite{Gillespie2000} in BDMM-Prime.
-This method is based on the Gillespie algorithm \cite{Gillespie1977} but the simulation time is divided in small intervals and the number of events is drawn from a poisson distribution for each interval.
-This algorithm is more efficient that Gillespie when we have a high number of individuals since we have to update the rates less often.
+This method is based on the Gillespie algorithm \cite{Gillespie1977} but the simulation time is divided in small intervals and the number of events is drawn from a Poisson distribution for each interval.
+This algorithm is more efficient than Gillespie when we have a high number of individuals since we have to update the rates less often.
 
@@
 \textbf{Model specifications and priors.} 
-Since we focus on the early epidemic outbreak, we expect un unimpeded spread of the virus that can be described by the exponential growth of the infected population and no significant decrease in the number of susceptibles over time \cite{Boskova2014}.
+Since we focus on the early epidemic outbreak, we expect unimpeded spread of the virus that can be described by the exponential growth of the infected population and no significant decrease in the number of susceptibles over time \cite{Boskova2014}.
 Thus, we assume a constant basic reproductive number $R_0$ for each of the European countries before the lockdown in the Lombardy region in March 8.
@@
 The workflow is a modified version of Nextrain workflow \cite{nextstrain} with the additional rules for BEAST 2 analysis and results processing.
+\todo{add snakemake graph}

### 2021-03-19 reports/mt-thesis/results.tex
wip mt

--- 2020-12-03 reports/paper/main.tex
@@
-\documentclass[12pt]{article}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\renewcommand*\familydefault{\sfdefault}
-
-\usepackage{anysize}
-\marginsize{2.5cm}{2.5cm}{1cm}{1.5cm}
-\setlength{\marginparwidth}{2cm}
-
-% \usepackage{authblk}
-\usepackage{graphicx}
-\usepackage{import} 
-\usepackage{subfig}
 \graphicspath{{../../results/}}
 
-\usepackage{xcolor}
+\chapter{Results}
 
-\usepackage[font=small,labelfont=bf]{caption}
-
-\def\labelitemi{--}
-
-% Line numbers
-\usepackage[left]{lineno}
-%\linenumbers
-
-\usepackage[color=red!60]{todonotes}
-
-\usepackage{pgfplotstable}
-\pgfplotsset{compat=1.16}
-\def\getcell#1#2#3{
-\pgfplotstablegetelem{#1}{[index]#2}\of{#3}\pgfplotsretval%
-}
-
-% Tables 
-\pgfplotstableset{col sep=tab, header=has colnames, string type, text special chars={\%}} 
-\pgfplotstableread{../../results/201030_europe3_table01.tsv}{\casestable}
-\pgfplotstableread{../../results/201030_europe3_table02.tsv}{\firsttable}
-
-\title{A comprehensive study of the early phylodynamics of SARS-CoV-2 in Europe}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-\todo[inline]{TODO abstract}
-This project focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe.
-We apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences.
-The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.
-
-% The early dynamics of the epidemic in Europe surprised the whole world/ From isolated cases detexcted in some, ina  few weeks we faced an unprecedent epidemic with case numbers and death counts stressing well stalished health infraestructures.
-Understanding what happened during those first month could prepare us to be less surpred in similar future events.
-Complex dynamics, typical epi date, case identification and travel history, very likely not all cases reported.
-Genetic sequences are a valuable source of information than together with popualtion dynamic model could inform us about the firrst steps of the coronavirus in Europe.
-We have integrated sequence data, flisht data and bayesian inference priors?
-to infer population trajectories - case counts and migration dynamics between the most relevant countrries at the start of the European pandemic.
-\end{abstract}
-
-\section*{Introduction}
-\todo[inline]{TODO introduction}
-
-\begin{itemize}
-\item Importance of understanding the spread of the virus to prevent future outbreaks
-
-\item About phylodynamics/phylogeographics? Use of genetic sequences as a source of information combined with other sources of information as travel data
-
-\item About what we know of the introduction of sars-cov-2 in Europe and early dynamics till 8 March
-
-\item About what we know of the case counts in Europe? Can sequences help?
-
-\item Introduce/formulate the questions: case counts, first introductions, ?migration vs within region transmission?, migration patterns, ?border closures.
-\end{itemize}
-
-\section*{Results}
 % \todo{Modify tables in R to have one value per cell, nr rate and times}
@@
     \label{fig:gribbon}
+    \todo[inline]{Specify that the random subset are 500 trajectories in all figure legends}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{201030_europe3_figtraj05.png}
-    \caption{Order of countries by the time of its first introduction, i.e.
+    \caption{Countries ordered by the time of its first introduction, i.e.
 first case in the country.
@@
 
-\section*{Discussion}
-\todo[inline]{TODO discussion}
-
-\begin{itemize}
-\item Case counts discussion and reporting rates, second wave in Europe. Other estimations in other studies.
-
-% Inferred population trajectories that account for all cases in the country.
-It is not an independent estimate since we are using informative priors for the sampling rates (upperbound computed as sequences/reported cases) and the subsampling scheme also includes the number of cases in the country (but that is a weaker link I think.\\
-
-\item First introduction other studies.
-
-\item Migration patterns other studies.
-
-\item Discuss assumptions of the model, caveats and possible improvements.
-\end{itemize}
-
-\section*{Conclusion}
-\todo[inline]{TODO conclusion}
-
-\section*{Material and Methods}
-\todo[inline]{TODO methods}
-\begin{itemize}
-\item Dataset and subsampling. Alignment.
-
-\item Model BDMM-Prime, stochastic mapping. Priors. Bayesian phylogeographic inference. Phylodynamics.
-
-\item GLM model. External data.
-
-\item ?Same analysis but with travel data from last year, can we observe a consequence of Hubei lockdown and reduced travel?
-
-\item Trajectories analysis. Subsampling and summarizing (median, 95\% interval)
-
-\item Snakemake workflow.
-\end{itemize}
-
-\bibliographystyle{unsrt}
-\bibliography{references-mt}
-
-\listoftodos
-
-\end{document}
+\subsubsection*{Number of particles}
+\todo[inline]{TODO, or into the discussion?? }

### 2021-03-19 reports/mt-thesis/thesis.tex
wip methods almost finished
wip mt

@@
     The abstract should be short, stating what you did and what the most important result is.
-	Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua.
-At vero eos et accusam et justo duo dolores et ea rebum.
-Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet.
-Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat, sed diam voluptua.
-At vero eos et accusam et justo duo dolores et ea rebum.
-Stet clita kasd gubergren, no sea takimata sanctus est Lorem ipsum dolor sit amet.
+	\todo[inline]{TODO abstract}
+	This project focus on the spatial dynamics of the early spread of SARS-CoV-2 in Europe.
+We apply a novel approach based on the Multi-type Birth Death phylodynamic model to infer structured population dynamics jointly with between-subpopulation transmission rates from viral genome sequences.
+The inferred epidemic trajectories for the combined outbreak responsible for the observed sequence data will allow us to better understand the entry into and early spread of SARS-CoV-2 in Europe.
+
+	% The early dynamics of the epidemic in Europe surprised the whole world/ From isolated cases detexcted in some, ina  few weeks we faced an unprecedent epidemic with case numbers and death counts stressing well stalished health infraestructures.
+Understanding what happened during those first month could prepare us to be less surpred in similar future events.
+Complex dynamics, typical epi date, case identification and travel history, very likely not all cases reported.
+Genetic sequences are a valuable source of information than together with popualtion dynamic model could inform us about the firrst steps of the coronavirus in Europe.
+We have integrated sequence data, flisht data and bayesian inference priors?
+to infer population trajectories - case counts and migration dynamics between the most relevant countrries at the start of the European pandemic.
 \end{abstract}
@@
 \include{methods}
+
+\include{results}
+
+\include{discussion}
+
+\include{conclusion}
 
@@
 
-\chapter{Discussion}
-
 % This displays the bibliography for all cited external documents. All references have to be defined in the file references.bib and can then be cited from within this document.
-\bibliographystyle{IEEEtran}
+\bibliographystyle{apalike}
 \bibliography{references}

### 2021-03-22 scripts/analyze_trace_v2.Rmd
analyze traj v2

---
title: "Parameter Inference Results"
subtitle: "Analysis Europe 10d"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dove"
    fonttheme: "structurebold"
    highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(knitr.kable.NA = '')
```

```{r libraries}
# Load libraries ---------------------------------------------------------------
library(argparse)
library(yaml)
library(tidyverse)
library(lubridate)
library(scales)
# library(ggpubr)
library(ggsci)
# library(ggridges)
# library(patchwork)
# library(hrbrthemes)
# library(circlize)
# library(chorddiag) 
# library(ggforce)
# library(ComplexHeatmap)
# library(maps)
# library(sf)
# library(wpp2019)
# library(gridExtra)
# library(grid)
# Source files -----------------------------------------------------------------
source("./scripts/utils.R")
```

```{r parser}
# Parser -----------------------------------------------------------------------
parser <- argparse::ArgumentParser()
parser$add_argument("--diagnostic", type = "character", help="chains diagnostic file")
parser$add_argument("--trace", type = "character", help="trace file")
parser$add_argument("--trace_summary", type = "character", help="trace summary  file")
parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
parser$add_argument("--analysis",  type = "character", help = "Analysis name")
parser$add_argument("--n", type = "integer", help = "Number of trajectories to analyze")
parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
parser$add_argument("--output_table", type = "character", help = "Output path for the tables")
args <- parser$parse_args()
```

```{r debugging}
# Debugging
args$burnin = 0.1
args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
args$metadata =  "./results/europeC/final_metadata.tsv"
args$n = 200
demes = args$demes
demes$exclude_country[5] = "France,Germany,Italy,Spain"
#demes$deme[3] = "China"
#demes$division[3] = NA
demes <- demes %>% arrange(deme)
args$diagnostic = "/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/210129_europe10d_chains_diagnostic.txt"
args$trace = "/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/210129_europe10d_comb.log"
args$trace_summary = "/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/210129_europe10d_comb.summary.tsv"
```

```{r plotopts}
# Plot options -----------------------------------------------------------------------
set_plotopts()
#dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
dcolors <- c("#E64B35FF", "#4DBBD5FF", "#00A087FF", "#3C5488FF", "#ffaa4fFF", "#7E6148FF")
# #dcolors <- c("#E64B35FF", "#7E6148FF", "#00A087FF", "#3C5488FF", "#4DBBD5FF", "#ffc954")
# #dcolors <- dcolors[1:nrow(demes)]
names(dcolors) <- sort(demes$deme)
# dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
# names(dark_dcolors) <- names(dcolors)
# ecolors <- c(B = "#1e3d59", M = "#f5f0e1", IM = "#ff6e40", OM = "#ffc13b")
colors = viridis_pal()(10)
```

## Deme configuration
\tiny
```{r demeconfig}
# Load demes configuration and sequence metadata -------------------------------
# demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
#                     exclude_country = NA, exclude_division = NA,
#                     min_date = NA, max_date = NA, seq_per_deme = NA) %>%
#   bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(args$demes, collapse = " ")),
#                           data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
#   group_by(deme) %>% 
#   mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
#   distinct() %>%
#   filter_all(any_vars(!is.na(.))) %>%
#   mutate(deme = str_to_title(deme)) %>%
#   mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme))
# 
china_deme <- demes$deme[demes$region == "Asia"]
if (china_deme == "Hubei-China") china_deme = "China"

knitr::kable(demes)
```

```{r}
# Read files ------------------------------------------------------------------
analysis <- gsub("_comb.log", "", tail(str_split(args$trace, "/")[[1]], 1))
analysis_specs <- read_tsv("analyses/analyses_specs.tsv") %>%
  select(1:4, analysis)

diagnostic <- read_table2(args$diagnostic)
trace <- read_trace(args$trace, burninFrac = 0)
trace_summary <- read_table(args$trace_summary)

first_sample <- analysis_specs %>%
  filter(input == "firstSample") %>%
  pull(analysis)

last_sample <- analysis_specs %>%
  filter(input == "mostRecentSampleTime") %>%
  pull(analysis)
```

## Chains diagnostic

```{r diagnostic}
knitr::kable(diagnostic)
```

## Convergence

- **Mean ESS value:** `r mean(trace_summary$ESS, na.rm = TRUE)`
- **Median ESS value:** `r median(trace_summary$ESS, na.rm = TRUE)`
- **Standard deviation ESS value:** `r sd(trace_summary$ESS, na.rm = TRUE)`

```{r convergence, fig.height=4.5}
trace %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value") %>%
  filter(param %in% c("posterior", "prior", "treeLikelihood")) %>%
  ggplot() +
  geom_line(aes(x = Sample, y = value, color = param), alpha = 0.5) +
  facet_wrap(~param, ncol = 1, scales = "free") +
  scale_color_viridis_d() +
  theme(legend.position = "none")

stats <- trace_summary %>%
  filter(item %in%  c("posterior", "prior", "treeLikelihood")) %>%
  select(5:7, ESS)
knitr::kable(stats)
```

## TreeHeight

**First sample date:** `r first_sample`

```{r treeheight2, fig.height=3}
param = "TreeHeight"

ggplot(trace) +
  geom_density(aes_string(param), color = colors[1], fill = colors[1], alpha = 0.3) +
  scale_x_continuous(labels = to_date(seq(0,1,0.01), mrs = last_sample), breaks = seq(0,1,0.01)) +
  theme(axis.text = element_text(angle=270))

stats <- trace_summary %>%
  filter(item == param) %>%
  mutate(median = to_date(median, mrs = last_sample),
         `95%HPDlo` = to_date(`95%HPDlo`, mrs = last_sample),
         `95%HPDup` = to_date(`95%HPDup`, mrs = last_sample)) %>%
  select(5:7, ESS)
knitr::kable(stats)

```

## gammaShape

```{r gammaShapePrior}
param = "gammaShape"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", BEASTobject == param) 
```

**Prior:** `r t(prior %>% pull(analysis))`

```{r gammaShapePosterior, fig.height=3}
ggplot(trace) +
  geom_density(aes_string(param), color = colors[2], fill = colors[2], alpha = 0.3) +
  stat_function(aes(x = seq(0,0.8)), fun = function(x) dexp(x, 2), linetype = 2)

stats <- trace_summary %>%
  filter(item == param) %>%
  select(5:7, ESS)
knitr::kable(stats)
```

## kappa

```{r kappaPrior}
param = "kappa"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", BEASTobject == param)
```

**Prior:** `r t(prior %>% pull(analysis))`

```{r kappaPosterior, fig.height=3}
ggplot(trace) +
  geom_density(aes_string(param), color = colors[3], fill = colors[3], alpha = 0.3) +
  stat_function(aes(x = seq(0,0.8)), fun = function(x) dlnorm(x, 
                                                              meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                                              sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2)

stats <- trace_summary %>%
  filter(item == param) %>%
  select(5:7, ESS)
knitr::kable(stats)
```

## originBDMMPrime

```{r originBDMMPrimePrior}
param = "originBDMMPrime"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", BEASTobject == param)
```

**Prior:** `r t(prior %>% pull(analysis))`

```{r originBDMMPrimePosterior, fig.height=3}
ggplot(trace) +
  geom_density(aes_string(param), color = colors[4], fill = colors[4], alpha = 0.3) +
  stat_function(aes(x = seq(0.2,1)), fun = function(x) dlnorm(x,
                                                              meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                                              sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))),
                linetype = 2) +
  scale_x_continuous(limits = c(0.2, 0.6), labels = to_date(seq(0,1,0.01), mrs = last_sample), breaks = seq(0,1,0.01)) +
  theme(axis.text = element_text(angle=270))

stats <- trace_summary %>%
  filter(item == param) %>%
  mutate(median = to_date(median, mrs = last_sample),
         `95%HPDlo` = to_date(`95%HPDlo`, mrs = last_sample),
         `95%HPDup` = to_date(`95%HPDup`, mrs = last_sample)) %>%
  select(5:7, ESS)
knitr::kable(stats)
```

## R0

```{r R0Prior}
param = "R0"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", BEASTobject == param, input != "Spec")
```

**Prior:** `r t(prior %>% pull(analysis))`

```{r R0Posterior, fig.height=5}
posterior <- trace %>%
  select(Sample, contains(param)) %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value") %>%
  separate(param, into = c("param", "deme"), "SV") %>%
  separate(deme, into = c("epoch", "deme"), "_", fill = "left") %>%
  filter(deme != "endtime")

ggplot(posterior) +
  geom_density(aes(value, color = deme, fill = deme), alpha = 0.3) +
  facet_wrap(~epoch, scales = "free", ncol = 1) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2)
```

## R0
```{r R0 stats}
stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), !grepl("Among", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## becomeUninfectiousRate
Fixed to 36.5 year^-1^ (time exponentially distributed mean 10 days)

## samplingProportion

```{r samplingProportionPrior}
param = "samplingProportion"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", BEASTobject == param, input != "Spec")
```

**Prior:** `r knitr::kable(prior %>% select(input, deme, analysis))`

## samplingProportion

```{r samplingProportionPosterior, fig.height=5}
posterior <- trace %>%
  select(Sample, contains(param)) %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value") %>%
  separate(param, into = c("param", "deme"), "SV") %>%
  separate(deme, into = c("epoch", "deme"), "_", fill = "left") %>%
  filter(deme != "endtime")

ggplot(posterior) +
  geom_density(aes(value, color = deme, fill = deme), alpha = 0.3) +
  facet_wrap(~epoch, scales = "free", ncol = 1) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2)
```

## samplingProportion
```{r samplingProportion stats}
stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), !grepl("Among", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## removalProb
Fixed to 1.

## migrationRate China

```{r migrationRatePrior}
param = "migrationRate"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", BEASTobject == param)
```

**Prior:** `r t(prior %>% pull(analysis))`

```{r migrationRatePosterior, fig.height=3}
posterior <- trace %>%
  select(Sample, contains(param)) %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value") %>%
  separate(param, into = c("param", "mig"), "SM") %>%
  separate(mig, into = c("src", "dest"), "_to_") %>%
  separate(src, into = c("epoch", "src"), "_", fill = "left") %>%
  filter(!is.na(dest))
```

\tiny
```{r migrationRateChina, fig.height=2.5}
ggplot(posterior %>% filter(src == "China")) +
  geom_density(aes(value, color = dest, fill = dest), alpha = 0.3) +
  facet_grid(~epoch) +
  coord_cartesian(xlim = c(0,0.3), ylim = c(0,20)) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2) +
  theme(legend.position = "right")

stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), grepl("China_to", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## migrationRate France

**Prior:** `r t(prior %>% pull(analysis))`

\tiny
```{r migrationRateFrance, fig.height=2.5}
ggplot(posterior %>% filter(src == "France")) +
  geom_density(aes(value, color = dest, fill = dest), alpha = 0.3) +
  facet_grid(~epoch) +
  coord_cartesian(xlim = c(0,5), ylim = c(0,20)) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2) +
  theme(legend.position = "right")

stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), grepl("France_to", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## migrationRate Germany

**Prior:** `r t(prior %>% pull(analysis))`

\tiny
```{r migrationRateGermany, fig.height=2.5}
ggplot(posterior %>% filter(src == "Germany")) +
  geom_density(aes(value, color = dest, fill = dest), alpha = 0.3) +
  facet_grid(~epoch, scales = "free") +
  coord_cartesian(xlim = c(0,5), ylim = c(0,30)) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2) +
  theme(legend.position = "right")

stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), grepl("Germany_to", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## migrationRate Italy

**Prior:** `r t(prior %>% pull(analysis))`

\tiny
```{r migrationRateItaly, fig.height=2.5}
ggplot(posterior %>% filter(src == "Italy")) +
  geom_density(aes(value, color = dest, fill = dest), alpha = 0.3) +
  facet_grid(~epoch) +
  coord_cartesian(xlim = c(0,5), ylim = c(0,30)) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2) +
  theme(legend.position = "right")

stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), grepl("Italy_to", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## migrationRate Other European

**Prior:** `r t(prior %>% pull(analysis))`

\tiny
```{r migrationRateOtherEuropean, fig.height=2.5}
ggplot(posterior %>% filter(src == "OtherEuropean")) +
  geom_density(aes(value, color = dest, fill = dest), alpha = 0.3) +
  facet_grid(~epoch) +
  coord_cartesian(xlim = c(0,20), ylim = c(0,30)) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2) +
  theme(legend.position = "right")

stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), grepl("OtherEuropean_to", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## migrationRate Spain

**Prior:** `r t(prior %>% pull(analysis))`

\tiny
```{r migrationRateSpain, fig.height=2.5}
ggplot(posterior %>% filter(src == "Spain")) +
  geom_density(aes(value, color = dest, fill = dest), alpha = 0.3) +
  facet_grid(~epoch) +
  coord_cartesian(xlim = c(0,5), ylim = c(0,40)) +
  scale_color_manual(values = dcolors) +
  scale_fill_manual(values = dcolors) +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dlnorm(x, meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                         sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2) +
  theme(legend.position = "right")

stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), grepl("Spain_to", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)
```

## scalerParamGLM

```{r scalerParamPrior}
param = "scalerParamGLM"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", grepl(param, BEASTobject))
```

**Prior:** `r t(prior %>% pull(analysis))`

```{r scalerParamPosterior, fig.height=3}
posterior <- trace %>%
  select(Sample, contains(param) & !contains("global")) %>%
  pivot_longer(-Sample, names_to = "param", values_to = "value")

ggplot(posterior) +
  geom_density(aes(value, color = param, fill = param), alpha = 0.3) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  stat_function(aes(x = seq(0,0.8)), 
                fun = function(x) dnorm(x, mean = as.numeric(prior %>% filter(input == "Mean") %>% pull(analysis)),
                                         sd = as.numeric(prior %>% filter(input == "Sigma") %>% pull(analysis))), 
                linetype = 2)

stats <- trace_summary %>%
  filter(grepl(param, item),!grepl("endtime", item), !grepl("Among", item)) %>%
  select(item, mean, 5:7, ESS)
knitr::kable(stats)

```

## globalScalerParamGLM

```{r globalScalerParamPrior}
param = "globalScalerParamGLM"
prior <- analysis_specs %>%
  filter(BEASTtag == "Prior", BEASTobject == param)
```

**Prior:** `r t(prior %>% pull(analysis))`

```{r globalScalerParamPosterior, fig.height=3}
ggplot(trace) +
  geom_density(aes_string(param), color = colors[3], fill = colors[3], alpha = 0.3) +
  stat_function(aes(x = seq(0,0.8)), fun = function(x) dlnorm(x, 
                                                              meanlog = as.numeric(prior %>% filter(input == "M") %>% pull(analysis)),
                                                              sdlog = as.numeric(prior %>% filter(input == "S") %>% pull(analysis))), 
                linetype = 2)

stats <- trace_summary %>%
  filter(item == param) %>%
  select(5:7, ESS)
knitr::kable(stats)
```

## indicatorParamGLM

<!-- Fixed to included. -->

```{r indicatorParamGLM}
param = "indicatorParamGLM"
initial_value <- analysis_specs %>%
  filter(BEASTtag == "State", BEASTobject == param)

stats <- trace_summary %>%
  filter(grepl(param, item)) %>%
  select(item, mean, 5:7, ESS)
```

- **Initial value:** `r t(initial_value %>% pull(analysis))`
- **Inclusion probability:** `r stats %>% pull(mean)`

### 2021-03-22 scripts/analyze_trajectories.Rmd
analyze traj v2

---
title: "Trajectory Mapping Results"
subtitle: "Analysis Europe10 300 particles"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dove"
    fonttheme: "structurebold"
    highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
params:
  input: traj_file.traj
  burnin: 0
  metadata: metadata.csv
  demes: demes_yaml
  n: 1000
  output_figure: path
  output_table: path
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(knitr.kable.NA = '')
```

```{r libraries}
# Load libraries ---------------------------------------------------------------
library(argparse)
library(yaml)
library(tidyverse)
library(lubridate)
library(scales)
library(ggpubr)
library(ggsci)
library(ggridges)
library(patchwork)
library(hrbrthemes)
library(circlize)
library(chorddiag) 
library(ggforce)
library(ComplexHeatmap)
library(maps)
library(sf)
library(wpp2019)
library(gridExtra)
library(grid)
# Source files -----------------------------------------------------------------
source("./scripts/trajProcessing.R")
source("./scripts/utils.R")
source("./scripts/flight_data.R")
```

```{r parser}
# Parser -----------------------------------------------------------------------
# parser <- argparse::ArgumentParser()
# parser$add_argument("--input", type = "character", help="trajectory file")
# parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
# parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
# parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
# parser$add_argument("--n", type = "integer", help = "Number of trajectories to analyze")
# parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
# parser$add_argument("--output_table", type = "character", help = "Output path for the tables")
# args <- parser$parse_args()
```

```{r debugging}
# Debugging
# args$burnin = 0.1
# args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
# #params$input = "./results/europeC/201211_europe6p2.TL.traj"
# params$input = "/Users/maceci/210205_europe10/results/europeC/210205_europe10.10000.TL.traj"
# args$metadata =  "./results/europeC/final_metadata.tsv"
# args$n = 100
# demes = args$demes
# demes$exclude_country[5] = "France,Germany,Italy,Spain"
# demes <- demes %>% arrange(deme)
```

## Deme configuration
\tiny
```{r demeconfig}
# Load demes configuration and sequence metadata -------------------------------
demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
                    exclude_country = NA, exclude_division = NA,
                    min_date = NA, max_date = NA, seq_per_deme = NA) %>%
  bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(params$demes, collapse = " ")),
                          data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
  group_by(deme) %>%
  mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
  distinct() %>%
  filter_all(any_vars(!is.na(.))) %>%
  mutate(deme = str_to_title(deme)) %>%
  mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme))

print("DEME CONFIG")
knitr::kable(demes)
# 
# china_deme <- demes$deme[demes$region == "Asia"]
# if (china_deme == "Hubei-China") china_deme = "China"
```

```{r plotopts}
# Plot options -----------------------------------------------------------------------
set_plotopts()
dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
dcolors <- c("#E64B35FF", "#4DBBD5FF", "#00A087FF", "#3C5488FF", "#ffaa4fFF", "#7E6148FF")
#dcolors <- c("#E64B35FF", "#7E6148FF", "#00A087FF", "#3C5488FF", "#4DBBD5FF", "#ffc954")
#dcolors <- dcolors[1:nrow(demes)]
names(dcolors) <- sort(demes$deme)
dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
names(dark_dcolors) <- names(dcolors)
ecolors <- c(B = "#1e3d59", M = "#f5f0e1", IM = "#ff6e40", OM = "#ffc13b")
```

```{r map, fig.height=2, fig.align="center"}
plot_map(demes)
knitr::kable(demes %>% select(-c("demeID", "exclude_division", "prob")), booktabs = TRUE)
```

## ECDC Case count data

```{r ecdc-data, results='hide'}
# Case data information from ECDC ----------------------------------------------
# Most recent sample value
metadata <- read_tsv(params$metadata)
MRS <- max(metadata$date)
mrs <- ymd_hms(paste(MRS, '23:59:00 UTC'))

case_data <- get_cases(demes, from = ymd("2019-09-01"), to = MRS) 
```

::: {.columns}

:::: {.column width="40%"}
\small
```{r ecdc-table, echo=FALSE}
knitr::kable(case_data %>% filter(date == MRS) %>% select(deme, source, total_confirmed)%>% pivot_wider(names_from = "source", values_from = total_confirmed) %>% mutate_if(is.numeric, function (x) round(x,1)), booktabs = TRUE, caption = paste("Total number of cases reported to ECDC on March 8, 2020"))
```
::::

:::: {.column width="60%"}
```{r ecdc-plot, fig.height = 3, fig.width=4, fig.cap="ECDC case counts for each deme from the beginning of the pandemic to March 8, 2020"}

ggplot(case_data %>% filter(source == "ecdc")) +
  geom_line(aes(x = date, y = total_confirmed, color = deme)) +
  geom_point(aes(x = date, y = total_confirmed, color = deme)) +
  #facet_wrap(~source) +
  scale_color_manual(values = dcolors, name = "") +
  scale_x_date(limits = c(ymd("2019-12-30"), ymd("2020-03-18")), date_breaks = "1 month", date_labels = "%b %d") +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    ylab("Total confirmed cases") +
    theme(axis.title.x=element_blank(),
          legend.position = c(0.2, 0.7),
          legend.box="horizontal",
          panel.grid.major.y =  element_line(colour = "grey80", size = 0.3, linetype = 2),
          panel.grid.minor.y =  element_line(colour = "grey80", size = 0.2,  linetype = 2),
          panel.background =  element_blank())
```

::::

:::

## Epidemic trajectory data

From the Stochastic Trajectory Mapping analysis, we sample with importance sampling one epidemic trajectory per set of parameters + typed node tree from the set of simulated trajectories.

The processing of the trajectory data includes the generation of two different datasets:

- `states`: We have the total number of inferred cases by trajectory, deme and time.
- `events`: We have each event that happened in a epidemic trajectory, with its type (origin, birth, death or migration), the source/destination deme and time.

## Epidemic trajectory data

We use the events dataset to compute quantities of interest:

- `B`: transmissions (births) events 
- `D`: becoming unifectious (deaths) events
- `IM`: migrations into the deme 
- `OM`: migrations out of the deme
- `S`: sampling events
- `O`: origin
- `in_pop`: origin + transmissions + incoming migrations 
- `out_pop`: deaths + outgoing migration + sampling events
- `active_pop`: origin + transmissions + incoming migrations - (deaths + outgoing migration + sampling events)

```{r traj-data, warnings = FALSE, message = FALSE, cache = FALSE}
# Load trajectory data ---------------------------------------------------------
df <- loadTrajectories(params$input, burninFrac = params$burnin, subsample = params$n)

# Data wrangling ---------------------------------------------------------------
df_traj <- processEvents(df$events, demes, "China", mrs)
```

## Epidemic trajectory data

```{r table-events}
knitr::kable(head(df_traj %>% filter(value != 0), 5), booktabs = TRUE, caption = "Trajectories dataset")
```

\normalsize
To have a feasible time of analysis of the epidemic trajectories we take a random subsample of 500 trajectories.

To facilitate visualization and summarise the results, we take a grid time of 1 day and summarise the number of events that day as the sum of the events in the corresponding time interval; and the number of inferred cases as the maximum of the interval.

```{r traj-processing}
# Add reported case counts
df_traj_cases <- df_traj %>%
    mutate(traj = as.character(traj)) %>%
    bind_rows(case_data %>%
                mutate(var = source, value = new_confirmed, cumvalue = total_confirmed) %>%
                select(deme, var, date, value, cumvalue))

# Summarise trajectories, compute mean, median and 95% HPD interval for each variable
traj_summary <- df_traj %>%
  group_by(var, deme, partner, date) %>%
  summarise(l95_value = HDInterval::hdi(value)[[1]],
            h95_value = HDInterval::hdi(value)[[2]],
            median_value = median(value),
            mean_value = mean(value),
            l95_cumvalue = HDInterval::hdi(cumvalue)[[1]],
            h95_cumvalue = HDInterval::hdi(cumvalue)[[2]],
            #l95_cumvalue = quantile(cumvalue, 0.025),
            #h95_cumvalue = quantile(cumvalue, 0.975),
            median_cumvalue = median(cumvalue),
            mean_cumvalue = mean(cumvalue)) %>% ungroup

```

```{r}
# 1. Total 
# 1.1 Median and 95% CCI of Total cases on the full time period. Fraction median
#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases

# Total values March 8, 2020 join with reported cases
traj_total <- traj_summary %>%
  filter(date == MRS) %>%
  select(var, date, deme, partner, matches("cumvalue"))  %>%
  left_join(case_data %>% filter(source == "ecdc")) %>%
  select(var, deme, matches("cumvalue"), total_confirmed) %>%
  mutate(ltimes = l95_cumvalue/total_confirmed,
         htimes = h95_cumvalue/total_confirmed,
         mtimes = median_cumvalue/total_confirmed) %>%
  mutate_if(is.numeric, function(x) round(x,2))

#knitr::kable(t(traj_total %>% filter(var == "D"))[-1,])
```

## Summarised epidemic trajectories
```{r trajs ribbon, fig.height = 4}
# 1.1 Median and 95% CCI by deme in log scale
gribbon <- ggplot(traj_summary %>% filter(var == "active_pop")) +
  geom_ribbon(aes(date, ymin = l95_cumvalue, ymax = h95_cumvalue, fill = deme), alpha = 0.5) +
  geom_line(aes(date, median_cumvalue, colour = deme)) +
  ylab("Population size") +
  xlab("Date") +
  #labs(title = "SARS-CoV-2 early epidemics in Europe and Hubei",
       #subtitle = "Case trajectories Median and 95% CCI by deme") +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b %d") +
  scale_fill_manual(name = "", values = dcolors) +
  scale_color_manual(name = "", values = dcolors) +
  theme(legend.position = c(0.2, 0.7),
        legend.box="horizontal",
        axis.title.x=element_blank())

gribbon

ggexport(gribbon,
         filename = paste0(params$output_figure, "01.png"),
         width = 1500, height = 1000, res = 300)
```

```{r trajs}
# 1.2 All trajectories single deme in log scale with reported cases date
trajs <- lapply(demes$deme, function(d) {
df_traj_cases%>%
  # 1. Prepare data
  filter(deme == d, 
         var %in% c("D", "ecdc"), 
         is.na(partner)) %>%
  # 2. Plot
  ggplot() +
  geom_line(aes(date, cumvalue, group = traj,
                color = var == "ecdc",
                alpha = var == "ecdc",
                linetype = var == "ecdc",
                size = var == "ecdc")) +
  scale_color_manual(name = "",
                     labels = c(paste0(d, " inferred\n population trajectories"),
                                paste0(ifelse(d == china_deme, "China", as.character(d)),
                                       " total reported cases")),
                     values = c(dcolors[[d]], dark_dcolors[[d]])) +
  scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
  scale_linetype_manual(name = "", values = c(1, 5),
                        labels = c(paste0(d, " inferred\n population trajectories"),
                                   paste0(ifelse(d == china_deme, "China", as.character(d)),
                                          " total reported cases"))) +
  scale_size_manual(name = "", values = c(0.5, 1.5),
                    labels = c(paste0(d, " inferred\n population trajectories"),
                               paste0(ifelse(d == china_deme, "China", as.character(d)),
                                      " total reported cases"))) +
  scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month",
               date_labels = "%b %y") +
  scale_y_log10(limits= c(1,10^7.5), expand = c(0, 0),labels = trans_format("log10", math_format(10^.x))) +
  labs(title = d) +
  ylab("COVID-19 cases") +
  theme(axis.title.x=element_blank(),
        legend.position = "none",
        legend.box="horizontal",
        panel.grid.major.y =  element_line(colour = "grey80", size = 0.3, linetype = 2),
        panel.grid.minor.y =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
        panel.background =  element_blank(),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10)) 
  })

names(trajs) <- demes$deme
ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1:length(trajs)), nrow = 3, ncol = 2),
         filename = paste0(params$output_figure, "02.png"),
         width = 2300, height = 3000, res = 300)
```

\tiny
```{r trajs-plotv2, results='asis', fig.height = 4}
for (d in demes$deme) {
  cat("\n## Inferred case counts - ", d, "\n")
  grid.arrange(trajs[[d]], tableGrob(traj_total %>% filter(deme == d, var == "D") %>% t, theme = ttheme_minimal()), ncol = 2, widths = c(2,1))
  cat("\n")
}
```

## First introduction

```{r first, fig.height=4}
# 2. First introduction distribution density by deme and time of first official reported case to ECDC
first_ecdc <- case_data %>% 
  filter(source == "ecdc", total_confirmed != 0) %>% 
  group_by(deme) %>% 
  arrange(date) %>% 
  slice(1)
#first_reported$min_date[first_reported$dest == china_deme] <- ymd("2019-12-12")

first_times <- df_traj_cases %>%
  # 1. Prepare data
  filter(var == "IM" | var == "O", value != 0) %>%
  group_by(deme, traj) %>%
  arrange(date) %>%
  slice(1) %>%
  ungroup %>%
  bind_rows(df_traj_cases %>%
              filter(var == "IM", value != 0) %>%
              group_by(traj) %>%
              arrange(date) %>%
              slice(1)  %>%
              mutate(deme = "Europe")) %>%
  mutate(deme = fct_rev(deme)) %>%
  # 2. Plot
  ggplot() +
  geom_density_ridges(aes(x = date, y = deme, fill = deme), quantile_lines = TRUE, quantiles = c(0.025, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) +
  #stat_density_ridges(, alpha = 0.7)
  #geom_histogram(aes(x = date, fill = deme), bandwidth=5, alpha = 0.7, size = 0.2) +
  #geom_vline(aes(xintercept = min_date, color = dest), linetype = 2) +
  geom_point(data = first_ecdc, 
             aes(date, deme, fill = deme), shape = 23, size = 4) +
  scale_fill_manual(values = c(dcolors, "Europe" = "grey")) +
  scale_color_manual(values = c(dcolors, "Europe" = "grey")) +
  scale_x_date(limits = c(ymd("2019-09-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %Y", expand = c(0,0)) +
  scale_y_discrete(expand = c(0.05,0), "probability") +
  #theme_ridges() +
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        panel.grid.major.x =  element_line(colour = "grey80", size = 0.3, linetype = 2),
        #panel.grid.minor.x =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
        #axis.title.y=element_blank(),
        panel.background = element_rect(fill = "white", colour = "white"))

first_times

ggexport(first_times,
         filename = paste0(params$output_figure, "03.png"),
         width = 2000, height = 800, res = 300)
```

## First introduction into Europe, location
```{r first-eu, fig.height=4}
# 2.2b First introduction into Europe, distribution density destination
m = 1
first_df1 <- df_traj %>%
  # 1. Prepare data
  filter(var == "IM", !is.na(partner),  value != 0) %>%
  group_by(traj) %>%
  arrange(date, partner) %>%
  slice(1:m)

first_eu <- ggplot(first_df1, aes(x = date, y = ..count../(m*params$n))) +
  geom_density_line(
    data = select(first_df1, -deme), aes(fill = NA),
    color = "transparent", alpha = 0.8, bw = 2) + 
  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
  facet_wrap(~deme, nrow = 1) +
  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
  ylab("Destination country\nprobability") +
  labs(title = "First imported case into Europe") +
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.text.x = element_text(angle = 90))
first_eu
ggexport(first_eu,
         filename = paste0(params$output_figure, "04.png"),
         width = 2100, height = 600, res = 300)
```

## First 10 introductions into Europe, location
```{r first-eu10, fig.height=4}
# 2.2b First 10 introduction into Europe, distribution density destination
m = 10
first_df10 <- df_traj %>%
  # 1. Prepare data
  filter(var == "IM", !is.na(partner),  value != 0) %>%
  group_by(traj) %>%
  arrange(date, partner) %>%
  slice(1:m)

first_dest <- ggplot(first_df10, aes(x = date, y = ..count../(m*params$n))) +
  geom_density_line(
    data = select(first_df10, -deme), aes(fill = NA),
    color = "transparent", alpha = 0.8, bw = 2
  ) + 
  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
  facet_wrap(~deme, nrow = 1) +
  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
  ylab("Destination country\nprobability") +
  labs(title = "First 10 imported cases into Europe") +
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

first_source <- ggplot(first_df10, aes(x = date, y = ..count../(m*params$n))) +
  geom_density_line(
    data = select(first_df10, -deme), aes(fill = NA),
    color = "transparent", alpha = 0.8, bw = 2
  ) + 
  geom_density_line(aes(fill = partner), bw = 2, color = "transparent", alpha = 0.8) +
  facet_wrap(~deme, nrow = 1) +
  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
  ylab("Source country\nprobability") +
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.text.x = element_text(angle = 90))

first_10eu <- ggarrange(first_dest, first_source, labels = chartr("123456789", "ABCDEFGHI", 1:3), nrow = 2, ncol = 1)
first_10eu
ggexport(first_10eu,
         filename = paste0(params$output_figure, "05.png"),
         width = 2000, height = 1200, res = 300)
```

## First introductions into each country, source
```{r first-bycountry, fig.height=4}
# 2.2b First introduction into each country, distribution density destination
m = 1
first_country <- df_traj %>%
  # 1. Prepare data
  filter(var == "IM", !is.na(partner),  value != 0, deme != "China") %>%
  group_by(traj, deme) %>%
  arrange(date, partner) %>%
  slice(1:m)

first_countryplot <- ggplot(first_country, aes(x = date, y = ..count../(m*params$n))) +
  geom_density_line(
    data = select(first_country, -deme), aes(fill = NA),
    color = "transparent", alpha = 0.8, bw = 3
  ) + 
  geom_density_line(aes(fill = partner), bw = 3, color = "transparent", alpha = 0.6) +
  geom_density_line(aes(fill = partner, color = partner), bw = 3, alpha = 0.6) +
  labs(title = "First imported case into each country") +
  facet_wrap(~deme, nrow = 1) +
  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
  scale_color_manual(name = "", values = dcolors, na.value = "grey") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
  ylab("Source country\nprobability") +
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.text.x = element_text(angle = 90))
first_countryplot
ggexport(first_countryplot,
         filename = paste0(params$output_figure, "06.png"),
         width = 2000, height = 1200, res = 300)
```

## First 10 introductions into each country, source

```{r first-10bycountry, fig.height=4}
# 2.2b First 10 introductions into each country, distribution density destination
m = 10
first_10country <- df_traj %>%
  # 1. Prepare data
  filter(var == "IM", !is.na(partner),  value != 0, deme != "China") %>%
  group_by(traj, deme) %>%
  arrange(date, partner) %>%
  slice(1:m)

first_10countryplot <- ggplot(first_10country, aes(x = date, y = ..count../(m*params$n))) +
  geom_density_line(
    data = select(first_10country, -deme), aes(fill = NA),
    color = "transparent", alpha = 0.8, bw = 3
  ) + 
  geom_density_line(aes(fill = partner), bw = 3, color = "transparent", alpha = 0.6) +
  geom_density_line(aes(fill = partner, color = partner), bw = 3, alpha = 0.6) +
  labs(title = "First 10 imported cases into each country") +
  facet_wrap(~deme, nrow = 1) +
  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
  scale_color_manual(name = "", values = dcolors, na.value = "grey") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
  ylab("Source country\nprobability") +
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.text.x = element_text(angle = 90))
first_10countryplot
ggexport(first_10countryplot,
         filename = paste0(params$output_figure, "07.png"),
         width = 2000, height = 1200, res = 300)
```

## Local transmission vs imported cases

```{r bvsm, fig.height=4}
h = df_traj %>%
  filter(var %in% c("B", "IM"), is.na(partner)) %>%
  select(-cumvalue) %>%
  group_by(deme, traj) %>%
  pivot_wider(names_from = "var", values_from = "value") %>%
  filter(IM != 0) %>%
  #na_if(0) %>% replace_na(list(B = 1e-8)) %>%
  mutate(R = B/IM) %>%
  mutate(R = ifelse(R == Inf, NA, R)) %>%
  select(traj, deme, date, R) %>%
  arrange(date) #%>%
  #pivot_wider(names_from = date, values_from = R) %>%
  #arrange(traj)

inpop = df_traj %>% filter(var == "in_pop", date == "2020-03-07") %>% select(traj, cumvalue)

hweek = df_traj %>%
  filter(var %in% c("B", "IM"), is.na(partner)) %>%
  select(-cumvalue) %>%
  group_by(deme, traj, var, year_week = ceiling_date(date, "1 week")) %>%
  summarize(value = sum(value)) %>%
  pivot_wider(names_from = "var", values_from = "value") %>%
  filter(IM != 0) %>%
  #na_if(0) %>% replace_na(list(B = 1e-8)) %>%
  mutate(R = B/IM) %>%
  mutate(R = ifelse(R == Inf, NA, R)) %>%
  select(traj, deme, year_week, R) %>%
  arrange(year_week)

hsummary = hweek %>% group_by(deme, year_week) %>% 
  summarise(mR = median(R),
            lR  = HDInterval::hdi(R)[[1]],
            hR  = HDInterval::hdi(R)[[2]])

ratio_plots <- lapply(demes$deme, function(d){
  hweek %>% left_join(inpop) %>% filter(deme == d) %>%
    ggplot() +
    geom_line(aes(year_week, R, group=traj, color = log(cumvalue)), alpha = 0.5) +
    geom_line(data = hsummary %>% filter(deme == d),aes(year_week, mR), size = 1) +
    geom_vline(aes(xintercept=ymd("2020-01-23")), linetype = "dashed") +
    geom_vline(aes(xintercept=ymd("2020-03-01")), linetype = "dashed") +
    #facet_wrap(~deme, scales = "free") +
    coord_cartesian(ylim = c(0,200)) +
    xlab("") +
    ylab(ifelse(d %in% c("China", "Italy"), "local/imported cases\n by week", "")) +
    labs(title = d) +
    #scale_colour_gradient(low = "yellow", high = "red", na.value = NA)
    scale_colour_viridis_c(name = "log cases", direction = -1) +
    theme(legend.position = c(0.2, 0.8),
          legend.key.size = unit(0.2, "cm"),
          title = element_text(size = 10))
})

ratiobm = ggarrange(plotlist = ratio_plots, nrow = 2, ncol = 3)

ratiobm
ggexport(ratiobm,
           filename = paste0(params$output_figure, "08.png"),
           width = 2000, height = 1200, res = 300)
```

```{r circoplot}
plot_chord <- function(df, min_date, max_date) {
  # df <- df %>%
  #   filter(date > min_date, date <= max_date, var == "IM") %>%
  #   group_by(partner, deme, date) %>%
  #   # Take mean value over all trajectories
  #   summarise(Nmean = mean(value), .groups = "drop_last") %>%
  #   # Sum mean values of each day over all the time period
  #   summarise(N = sum(Nmean), .groups = "drop") %>%
  #   drop_na()

  df <- df %>%
    filter(date > min_date, date <= max_date, var == "IM") %>%
    group_by(partner, deme, traj) %>%
    # Take mean value over all trajectories
    summarise(N = sum(value), .groups = "drop_last") %>%
    #summarise(Nm = median(N)/as.numeric(max_date - min_date), .groups = "drop") %>%
    summarise(Nm = median(N), .groups = "drop") %>%
    drop_na()

  # parameters
  cplot <- circos.clear()
  cplot <- circos.par(start.degree = 90, gap.degree = 4, track.margin = c(-0.1, 0.1), 
                      points.overflow.warning = FALSE)
  par(mar = rep(0, 4))
  # Base plot
  chordDiagram(
    x = df, 
    grid.col = dcolors,
    transparency = 0.25,
    directional = 1,
    direction.type = c("arrows", "diffHeight"), 
    diffHeight  = -0.04,
    annotationTrack = "grid", 
    #annotationTrackHeight = c(0.05, 0.1),
    link.arr.type = "big.arrow", 
    link.sort = TRUE, 
    link.largest.ontop = TRUE)
  # Add text and axis
  circos.trackPlotRegion(
    track.index = 1, 
    bg.border = NA, 
    panel.fun = function(x, y) {
      xlim = get.cell.meta.data("xlim")
      sector.index = get.cell.meta.data("sector.index")
      # Add names to the sector. 
      circos.text(
        x = mean(xlim), 
        y = 3.2, 
        labels = sector.index, 
        facing = "bending", 
        cex = 0.9
      )
      # Add graduation on axis
      circos.axis(
        h = "top", 
        major.at = seq(from = 0, to = xlim[2], by = case_when(
          xlim[2] > 50500 ~ 20000,
          xlim[2] > 20500 ~ 20000,
          xlim[2] > 10500 ~ 5000,
          xlim[2] > 5500 ~ 5000,
          xlim[2] > 2500 ~ 2000,
          xlim[2] > 2000 ~ 2000,
          xlim[2] > 1500 ~ 1000,
          xlim[2] > 800 ~ 500,
          xlim[2] > 500 ~ 500,
          xlim[2] > 200 ~ 100,
          xlim[2] > 100 ~ 50,
          xlim[2] > 50 ~ 25,
          xlim[2] > 20 ~ 15,
          xlim[2] > 10 ~ 10,
          xlim[2] > 1 ~ 1,
          TRUE ~ 0.01)), 
        minor.ticks = 1, 
        major.tick.length = 0.5,
        labels.niceFacing = FALSE,
        labels.cex = 0.6)
    }
  )
  return(cplot)
}

```

## Migrations - Period 1

```{r, fig.height=2.5, fig.align='center'}
# Period 1
plot_chord(df_traj, ymd("2019-09-01"), ymd("2020-01-23"))

png(paste0(params$output_figure, "09a.png"), width = 1000, height = 1000, res = 200)
plot_chord(df_traj, ymd("2019-09-01"), ymd("2020-01-23")) 
text(label = "A", x = -1, y = 1, cex = 1.5, font = 2)
dev.off()
```

## Migrations - Period 2

```{r, fig.height=2.5, fig.align='center'}
# Period 2
plot_chord(df_traj, ymd("2020-01-23"), ymd("2020-02-28"))

png(paste0(params$output_figure, "09b.png"), width = 1000, height = 1000, res = 200)
plot_chord(df_traj, ymd("2020-01-23"), ymd("2020-02-28"))
mtext(text = "B", at = -1, cex = 1.5, font = 2)
text(label = "B", x = -1, y = 1, cex = 1.5, font = 2)
#dev.off()
```

## Migrations - Period 3

```{r, fig.height=2.5, fig.align='center'}
# Period 3
plot_chord(df_traj, ymd("2020-02-28"), ymd("2020-03-08"))

png(paste0(params$output_figure, "09c.png"), width = 1000, height = 1000, res = 200)
plot_chord(df_traj, ymd("2020-02-28"), ymd("2020-03-08"))
text(label = "C", x = -1, y = 1, cex = 1.5, font = 2)
dev.off()
```

## Daily flight passengers vs daily imported cases

```{r flight-pass, fig.height=4}
demes1 <- c(China = "CN", France = "FR", Germany = "DE", Italy = "IT", OtherEuropean = "OE", Spain = "ES")
flight_df <- get_flightData(countries = demes1)

demes2 <- setNames(names(demes1), demes1)
flight_df_3e <- flight_df %>%
  filter((month(time) %in% 1:3 & year(time) == 2020) | (month(time) == 12 & year(time) == 2019)) %>%
  mutate(epoch = case_when(
    month(time) == 2 ~ 2,
    month(time) == 3 ~ 3,
    TRUE ~ 1),
    pred = 1) %>%
  rename(src = geo, dest = partner) %>%
  group_by(src, dest, epoch, pred) %>%
  summarise(N = n(),
         values = sum(dvalues/N)) %>%
  ungroup %>%
  select(epoch, src, dest, values) %>%
  mutate(src = demes2[src], dest = demes2[dest])

by_epoch = df_traj %>%
  filter(var == "IM", cumvalue != 0, !is.na(partner)) %>%
  mutate(epoch = case_when(
    month(date) == 2 ~ 2,
    month(date) == 3 ~ 3,
    TRUE ~ 1)) %>%
  rename(src = partner, dest = deme) %>%
  group_by(src, dest, epoch, traj) %>%
  summarise(N = n(),
         evalues = sum(value)/N) %>%
  select(traj, epoch, src, dest, evalues)

df_f = left_join(by_epoch, flight_df_3e) %>%
  group_by(src, dest, epoch) %>%
  summarise(v = median(values),
            ev = median(evalues),
            lev = HDInterval::hdi(evalues)[[1]],
            hev = HDInterval::hdi(evalues)[[2]])

imflights <- ggplot(df_f, aes(ev, v, group=interaction(src, dest))) +
  #geom_line(aes(color = src)) + 
  geom_line(color = "grey80", size = 1) +
  geom_pointrange(shape = 21, size = 0.4, linetype = 1, alpha = 0.7, aes(x = ev, y = v, xmin = lev, xmax = hev, fill=factor(epoch), color=factor(epoch))) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_x_log10(labels = trans_format("log10", math_format(10^.x))) +
  ylab("daily flight passengers") +
  xlab("daily imported cases") +
  facet_grid(~dest) +
  scale_color_viridis_d(labels=c("period 1", "period 2", "period 3")) +
  scale_fill_viridis_d(labels=c("period 1", "period 2", "period 3")) +
   ggrepel::geom_text_repel(
    data = df_f %>% filter(epoch == 3),
    aes(ev, v, label = demes1[src]),
    segment.alpha = 0, point.padding = 0.25,
    #segment.size = 5,
    #box.padding = .5,
    force = 1,
    size = 8/.pt,
    nudge_x = 1,
    #nudge_y = 0.5,
    color = "grey30"
  ) +
  theme(panel.grid.major.y =  element_line(colour = "grey80", size = 0.2, linetype = 3),
        panel.grid.minor.y = element_blank(),
        axis.text.y = element_text(size = 8),
        panel.grid.major.x =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
        #legend.position = c(0.8, 0.1),
        legend.position = "bottom",
        legend.background = element_rect(fill = "white", size=0.1, linetype="solid", 
                                  colour ="white"),
        legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing = unit(0.4, "mm"),
        legend.direction = "horizontal",
        legend.box.margin = margin(c(0,0,0,0)),
        legend.margin = margin(c(0,0,0,0)),
        panel.background = element_rect(fill = "white", colour = "grey80"))
imflights
ggexport(imflights,
         filename = paste0(params$output_figure, "10.png"),
         width = 2000, height = 1200, res = 300)
```

## Weekly confirmed cases proportion

```{r report-prop, fig.height=4}
df_report <- traj_summary %>%
  filter(var %in% c("D"), deme != "China") %>%
  left_join(case_data %>% filter(source == "ecdc")) %>%
  select(deme, date, l95_value, h95_value, median_value, new_confirmed) %>%
  group_by(deme, year_week = ceiling_date(date, "1 week")) %>%
  summarise(l95_value = sum(l95_value),
            h95_value = sum(h95_value),
            median_value = sum(median_value),
            new_confirmed = sum(new_confirmed)) %>%
  mutate(median_report = new_confirmed/median_value,
         h95_report = new_confirmed/l95_value,
         l95_report = new_confirmed/h95_value)

report_plot <- ggplot(df_report) +
  geom_line(aes(x = year_week, y = median_report, color = deme), stat="identity") +
  geom_point(aes(x = year_week, y = median_report, color = deme), stat="identity") +
  geom_ribbon(aes(year_week, ymin = l95_report, ymax = h95_report,fill = deme), alpha = 0.1) +
  ylab("Reporting rate") +
  scale_color_manual(values = dcolors, name = "") +
  scale_fill_manual(values = dcolors, name = "") +
  scale_x_date(date_breaks = "5 days", date_labels = "%b %d", limits=c(ymd("2020-01-30"), ymd("2020-03-08"))) +
  #scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b %y") +
  coord_cartesian(ylim = c(0,1)) +
  theme(#legend.position = "none", 
        axis.title.x=element_blank()) #+
  #facet_wrap(~deme, scales = "free")

report_plot
ggexport(report_plot,
         filename = paste0(params$output_figure, "11.png"),
         width = 2000, height = 1200, res = 300)
```

## Weekly incidence population

```{r incidence, fig.height=4}
library(wpp2019)
data(pop)
pop <- pop %>% filter(name %in% demes$deme) %>% rename(deme = name) %>% mutate(population = `2020` * 1000) %>% select(deme, population)

df_inc <- traj_summary %>%
  filter(var %in% c("D")) %>%
  left_join(pop) %>%
  select(deme, date, l95_cumvalue, h95_cumvalue, median_cumvalue, population) %>%
  group_by(deme, population, date) %>%
  summarise(l95_cumvalue = sum(l95_cumvalue),
            h95_cumvalue = sum(h95_cumvalue),
            median_cumvalue = sum(median_cumvalue)) %>%
  mutate(median_inc = median_cumvalue/population  * 100,
         h95_inc = l95_cumvalue/population  * 100,
         l95_inc = h95_cumvalue/population  * 100)

inc_plot <- ggplot(df_inc) +
  geom_line(aes(x = date, y = median_inc, color = deme), stat="identity") +
  geom_point(aes(x = date, y = median_inc, color = deme), stat="identity") +
  geom_ribbon(aes(date, ymin = l95_inc, ymax = h95_inc,fill = deme), alpha = 0.1) +
  ylab("Reporting rate") +
  scale_color_manual(values = dcolors, name = "") +
  scale_fill_manual(values = dcolors, name = "") +
  scale_x_date(date_breaks = "5 days", date_labels = "%b %d", limits=c(ymd("2020-01-30"), ymd("2020-03-08"))) +
  #scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b %y") +
  #coord_cartesian(ylim = c(0,1)) +
  theme(#legend.position = "none", 
        axis.title.x=element_blank()) #+
inc_plot

ggexport(inc_plot,
         filename = paste0(params$output_figure, "12.png"),
         width = 2000, height = 1200, res = 300)
```

## Events timing

```{r events-timing, fig.height=4}
df_eventsttiming <- df_traj_cases %>%
  filter(var %in% c("O", "B", "D", "IM", "OM"), value != 0, is.na(partner)) %>%
  group_by(traj, var, deme, partner) %>%
  mutate(var = ifelse(var == "O", "B", var)) %>%
  arrange(date) %>%
   slice(1) #%>%
  # ungroup(traj) %>%
  # summarise(m = median(date),
  #           l = quantile(date, 0.025, type = 1),
  #           h = quantile(date, 0.925, type = 1))

events_timingplot <- ggplot(df_eventsttiming) +
  geom_density_ridges(aes(x = date, y = factor(var, level =  c("OM", "IM", "D", "B")), fill = var), color = "transparent", quantile_lines = TRUE, quantiles = c(0.025, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) + 
  geom_vline(data = first_ecdc, aes(xintercept = date, linetype = "first confirmed case to ECDC")) +
  scale_linetype_manual(values = 3, name = "") +
  facet_wrap(~deme) +
  scale_fill_viridis_d(guide = FALSE) +
  scale_y_discrete(breaks = c("B", "D", "IM", "OM"), labels = c("first local case", "first uninfectious", "first imported case", "first exported case")) + 
  ylab("") +
  xlab("") +
  theme(legend.position = "top")

events_timingplot
ggexport(events_timingplot,
         filename = paste0(params$output_figure, "13.png"),
         width = 2000, height = 1200, res = 300)

```

### 2021-03-22 scripts/case_data.R

@@ -3,9 +3,20 @@ library(tidyverse)
 
 ecdc <- get_dataECDC()
 who <- get_dataWHO()
-jh <- get_dataJH()
+# jh <- get_dataJH() Same than owid
 owid <- get_dataOWID() #%>% select(-new_cases) %>% rename(new_cases = new_cases_smoothed)
+covid19od <- get_dataC19OD()
 
+case_data <- bind_rows(ecdc, owid, covid19od, who) %>%
+  filter(country %in% c("China", "France", "Germany", "Italy", "Spain"),
+         date <= "2020-03-07")
+
+ggplot(case_data) +
+  geom_line(aes(date, total_confirmed, color = source)) +
+  facet_wrap(~country, scales = "free")
+  
+  
+  
 ecdc2 <- ecdc %>%
   mutate(new_cases = zoo::rollmean(new_cases, k = 7, fill = NA, align = "center"),
          source = "ecdc2")
@@ -37,11 +48,11 @@ sp <- case_data %>%
   left_join(samples) %>%
   mutate(sp = s/cases)
 
-spain <- read.csv("https://cnecovid.isciii.es/covid19/resources/casos_tecnica_ccaa.csv") %>%
+spain <- read.csv("https://cnecovid.isciii.es/covid19/resources/casos_diag_ccaadecl.csv") %>%
   group_by(fecha) %>%
   summarise(new_cases = sum(num_casos)) %>%
   mutate(country = "Spain",
-         date = ymd(fecha) + 6,
+         date = ymd(fecha),
          source = "RENAVE")
 
 italy <- read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-andamento-nazionale/dpc-covid19-ita-andamento-nazionale.csv") %>%

### 2021-03-22 scripts/subsample.R

@@ -50,26 +50,38 @@ subsample <- function(alignment, metadata, include = NA, exclude = NA,
   if (prob == "deaths") to <- ymd(to) + 14
   
   # Get case counts for deme
-  case_data <- get_dataECDC()
+  case_data <- get_dataOWID()
+  death_data <- get_dataECDC()
   cases_deme <- get_cases_deme(case_data, "deme", region_name, country_name, division_name, 
                                exclude_country, exclude_division, 
-                               from, to)  
+                               from, to) 
+  deaths_deme <- get_cases_deme(death_data, "deme", region_name, country_name, division_name, 
+                               exclude_country, exclude_division, 
+                               from, to)
   
   # Account for lag cases - deaths
-  if (prob == "deaths") cases_deme <- cases_deme %>% mutate(date = date - 14)
+  if (prob == "deaths") {
+    deaths_deme <- deaths_deme %>% mutate(date = date - 14)
+    by_date <- metadata_deme %>%
+      count(date, country) %>%
+      left_join(deaths_deme, by = c("date", "country")) %>%
+      replace_na(list(new_deaths = 0)) %>%
+      mutate(p_death = (deaths + 1)/sum(deaths),
+             w_date = sum(n)/n,
+             p_date_deaths = (p_death * w_date)/sum(p_death * w_date)) %>%
+      select(date, country, p_death, w_date,  p_date_deaths)
+  }
   
-  if (prob %in% c("cases", "cases_and_location", "deaths")){
+  if (prob %in% c("cases", "cases_and_location")){
     # Compute weights and probabilities for each date
     by_date <- metadata_deme %>%
       count(date, country) %>%
       left_join(cases_deme, by = c("date", "country")) %>%
-      replace_na(list(cases = 0, cumcases = 0, deaths = 0, cumdeaths = 0)) %>%
+      replace_na(list(new_confirmed = 0)) %>%
       mutate(p_case = (cases + 1)/sum(cases),
-             p_death = (deaths + 1)/sum(deaths),
              w_date = sum(n)/n,
-             p_date_cases = (p_case * w_date)/sum(p_case * w_date),
-             p_date_deaths = (p_death * w_date)/sum(p_death * w_date)) %>%
-      select(date, country, p_case, p_death, w_date, p_date_cases, p_date_deaths) 
+             p_date_cases = (p_case * w_date)/sum(p_case * w_date)) %>%
+      select(date, country, p_case, w_date, p_date_cases) 
   }
   
   if (prob %in% c("location", "cases_and_location")) {

### 2021-03-22 scripts/trajProcessing.R

@@ -134,3 +134,75 @@ gridEventsByAge <- function(trajEvents, ages) {
   
   return(df_grid)
 }
+
+processEvents <- function(events, demes, source = NA, mrs) {
+  events <- events %>%
+    mutate(src = factor(src,
+                        levels = 0:(nrow(demes) - 1),
+                        labels = sort(demes$deme)),
+           dest = (factor(dest,
+                          levels = 0:(nrow(demes) - 1),
+                          labels = sort(demes$deme))))
+  events[events$event == "O", "src"] = source
+  events[events$event == "O", "mult"] =  1
+  
+  df_traj0 <- events %>%
+    mutate(date = date(date_decimal(decimal_date(mrs) - age))) %>%
+    select(-time, -age) %>%
+    group_by_at(vars(-mult)) %>%
+    summarise(N = sum(mult), .groups = "drop") %>%
+    mutate(srctmp = src, destmp = dest) %>%
+    pivot_longer(src:dest, names_to = "role", values_to = "deme") %>%
+    mutate(event = ifelse(event == "M" & role == "src", "OM",
+                          ifelse(event == "M" & role == "dest", "IM", event)),
+           partner = ifelse(event == "OM", as.character(destmp),
+                            ifelse(event == "IM", as.character(srctmp), NA))) %>% 
+    select(-c("role", "srctmp", "destmp")) %>%
+    filter(!is.na(deme)) %>%
+    group_by_at(vars(-N)) %>%
+    summarise(value = sum(N), .groups = "drop") %>%
+    group_by(traj, deme, partner, event) %>%
+    arrange(traj, date, deme) %>%
+    mutate(cumvalue = cumsum(value)) %>%
+    rename(var = event) 
+  
+  mig_cum = df_traj0 %>%
+    filter(var %in% c("IM", "OM")) %>%
+    group_by(traj, var, date, deme) %>%
+    summarise(value = sum(value), .groups = "drop") %>%
+    group_by(traj, var, deme) %>%
+    arrange(traj, date, deme) %>%
+    mutate(cumvalue = cumsum(value),
+           partner = NA) 
+  
+  inout_pop = df_traj0 %>% 
+    mutate(var = ifelse(var %in% c("O", "B", "IM"), "in_pop", "out_pop")) %>%
+    group_by(traj, deme, var, date) %>%
[…]

### 2021-03-22 scripts/utils.R

@@ -7,22 +7,42 @@
 library(countrycode)
 library(ggpubr)
 library(lubridate)
+library(maps)
+library(sf)
+library(wpp2019)
 
 set_plotopts <- function() {
   theme_set(theme_minimal())
   theme_update(#strip.background = element_rect(fill = "grey95", color = "grey95", size = 1),
                panel.grid = element_blank(),
-               panel.background = element_rect(fill = "white", colour = "grey80"),
+               #panel.background = element_rect(fill = "white", colour = "grey80"),
                axis.ticks = element_line(),
                legend.position = "bottom",
                legend.box="vertical",
-               strip.background = element_rect(fill = "grey80", colour = "grey80"),
-               strip.text = element_text(color = "grey30"),
+               legend.text=element_text(size = 8),
+               legend.title=element_text(size = 8),
+               strip.background = element_rect(fill = "white", colour = "grey80"),
+               strip.text = element_text(color = "grey30", face = "bold"),
                #legend.title = element_text(size = 9, face = "bold"),
-               axis.text.x = element_text(size = 8),
-               axis.text.y = element_text(size = 8),
                axis.title.y = element_text(size = 9, face = "bold", margin = margin(r = 10)),
-               axis.title.x = element_text(size = 9, face = "bold", margin = margin(t = 10)))
+               axis.title.x = element_text(size = 9, face = "bold", margin = margin(t = 10)),
+               #axis.line.x = element_line(color = "grey30"),
+               # axis.text.x.bottom = element_text(hjust = 0, vjust = .5,
+               #                            margin = margin(0, 0, 0, 0),
+               #                            color = "grey30",
+               #                            lineheight = 0.8),
+               #axis.ticks.x = element_blank(),
+               #axis.line.y.left = element_line(color = "grey30"),
+               # axis.text.y.right = element_text(hjust = 0, vjust = .5,
+               #                           margin = margin(0, 0, 0, 0),
+               #                           color = "grey30",
+               #                           lineheight = 0.8),
[…]

### 2021-03-22 scripts/analyze_trajectories.Rmd
traj report working

@@
 tables: yes
-params:
-  input: traj_file.traj
-  burnin: 0
-  metadata: metadata.csv
-  demes: demes_yaml
-  n: 1000
-  output_figure: path
-  output_table: path
 ---
+
+```{r parser, echo = FALSE}
+library(tidyverse)
+# Parser -----------------------------------------------------------------------
+argsR <- commandArgs(TRUE)
+args_list = str_split(paste(argsR, collapse=""), pattern="--")
+args_llist = lapply(args_list, function(x) str_split(x, pattern="="))
+argsDF = as.data.frame(do.call("rbind", args_llist[[1]]))
+args <- as.list(as.character(argsDF$V2))
+names(args) <- argsDF$V1
+args$burnin = as.integer(args$burnin)
+args$n = as.integer(args$n)
+```
 
@@
 
-```{r libraries}
+```{r libraries, results='hide', message=FALSE, warning=FALSE}
 # Load libraries ---------------------------------------------------------------
-library(argparse)
 library(yaml)
-library(tidyverse)
 library(lubridate)
@@
 library(ggforce)
-library(ComplexHeatmap)
+#library(ComplexHeatmap)
 library(maps)
@@
 
-```{r parser}
-# Parser -----------------------------------------------------------------------
-# parser <- argparse::ArgumentParser()
-# parser$add_argument("--input", type = "character", help="trajectory file")
-# parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
-# parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
-# parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
-# parser$add_argument("--n", type = "integer", help = "Number of trajectories to analyze")
-# parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
-# parser$add_argument("--output_table", type = "character", help = "Output path for the tables")
-# args <- parser$parse_args()
-```
-
 ```{r debugging}
@@
 # args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
-# #params$input = "./results/europeC/201211_europe6p2.TL.traj"
-# params$input = "/Users/maceci/210205_europe10/results/europeC/210205_europe10.10000.TL.traj"
+# #args$input = "./results/europeC/201211_europe6p2.TL.traj"
+# args$input = "/Users/maceci/210205_europe10/results/europeC/210205_europe10.10000.TL.traj"
 # args$metadata =  "./results/europeC/final_metadata.tsv"
@@
                     min_date = NA, max_date = NA, seq_per_deme = NA) %>%
-  bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(params$demes, collapse = " ")),
+  bind_rows(bind_rows(lapply(yaml::yaml.load(string = gsub(",", ", ", gsub(":", ": ", args$demes))),
                           data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
@@
 
-print("DEME CONFIG")
-knitr::kable(demes)
-# 
-# china_deme <- demes$deme[demes$region == "Asia"]
-# if (china_deme == "Hubei-China") china_deme = "China"
+china_deme <- demes$deme[demes$region == "Asia"]
+if (china_deme == "Hubei-China") china_deme = "China"
 ```
@@
 plot_map(demes)
-knitr::kable(demes %>% select(-c("demeID", "exclude_division", "prob")), booktabs = TRUE)
+knitr::kable(demes %>% select(-c("exclude_division", "prob")), booktabs = TRUE)
 ```
@@
 # Most recent sample value
-metadata <- read_tsv(params$metadata)
+metadata <- read_tsv(args$metadata)
 MRS <- max(metadata$date)
@@
 # Load trajectory data ---------------------------------------------------------
-df <- loadTrajectories(params$input, burninFrac = params$burnin, subsample = params$n)
+df <- loadTrajectories(args$args, burninFrac = args$burnin, subsample = args$n)
 
@@
 ggexport(gribbon,
-         filename = paste0(params$output_figure, "01.png"),
+         filename = paste0(args$output_figure, "01.png"),
          width = 1500, height = 1000, res = 300)
@@
 ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1:length(trajs)), nrow = 3, ncol = 2),
-         filename = paste0(params$output_figure, "02.png"),
+         filename = paste0(args$output_figure, "02.png"),
          width = 2300, height = 3000, res = 300)
@@
 ggexport(first_times,
-         filename = paste0(params$output_figure, "03.png"),
+         filename = paste0(args$output_figure, "03.png"),
          width = 2000, height = 800, res = 300)
@@
 
-first_eu <- ggplot(first_df1, aes(x = date, y = ..count../(m*params$n))) +
+first_eu <- ggplot(first_df1, aes(x = date, y = ..count../(m*args$n))) +
   geom_density_line(
@@
 ggexport(first_eu,
-         filename = paste0(params$output_figure, "04.png"),
+         filename = paste0(args$output_figure, "04.png"),
          width = 2100, height = 600, res = 300)
@@
 
-first_dest <- ggplot(first_df10, aes(x = date, y = ..count../(m*params$n))) +
+first_dest <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
   geom_density_line(
@@
 
-first_source <- ggplot(first_df10, aes(x = date, y = ..count../(m*params$n))) +
+first_source <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
   geom_density_line(
@@
 ggexport(first_10eu,
-         filename = paste0(params$output_figure, "05.png"),
+         filename = paste0(args$output_figure, "05.png"),
          width = 2000, height = 1200, res = 300)
@@
 
-first_countryplot <- ggplot(first_country, aes(x = date, y = ..count../(m*params$n))) +
+first_countryplot <- ggplot(first_country, aes(x = date, y = ..count../(m*args$n))) +
   geom_density_line(
@@
 ggexport(first_countryplot,
-         filename = paste0(params$output_figure, "06.png"),
+         filename = paste0(args$output_figure, "06.png"),
          width = 2000, height = 1200, res = 300)
@@
 
-first_10countryplot <- ggplot(first_10country, aes(x = date, y = ..count../(m*params$n))) +
+first_10countryplot <- ggplot(first_10country, aes(x = date, y = ..count../(m*args$n))) +
   geom_density_line(
@@
 ggexport(first_10countryplot,
-         filename = paste0(params$output_figure, "07.png"),
+         filename = paste0(args$output_figure, "07.png"),
          width = 2000, height = 1200, res = 300)
@@
 ggexport(ratiobm,
-           filename = paste0(params$output_figure, "08.png"),
+         filename = paste0(args$output_figure, "08.png"),
          width = 2000, height = 1200, res = 300)
@@
 
-png(paste0(params$output_figure, "09a.png"), width = 1000, height = 1000, res = 200)
+png(paste0(args$output_figure, "09a.png"), width = 1000, height = 1000, res = 200)
 plot_chord(df_traj, ymd("2019-09-01"), ymd("2020-01-23")) 
@@
 
-png(paste0(params$output_figure, "09b.png"), width = 1000, height = 1000, res = 200)
+png(paste0(args$output_figure, "09b.png"), width = 1000, height = 1000, res = 200)
 plot_chord(df_traj, ymd("2020-01-23"), ymd("2020-02-28"))
@@
 
-png(paste0(params$output_figure, "09c.png"), width = 1000, height = 1000, res = 200)
+png(paste0(args$output_figure, "09c.png"), width = 1000, height = 1000, res = 200)
 plot_chord(df_traj, ymd("2020-02-28"), ymd("2020-03-08"))
@@
 ggexport(imflights,
-         filename = paste0(params$output_figure, "10.png"),
+         filename = paste0(args$output_figure, "10.png"),
          width = 2000, height = 1200, res = 300)
@@
 ggexport(report_plot,
-         filename = paste0(params$output_figure, "11.png"),
+         filename = paste0(args$output_figure, "11.png"),
          width = 2000, height = 1200, res = 300)
@@
 ggexport(inc_plot,
-         filename = paste0(params$output_figure, "12.png"),
+         filename = paste0(args$output_figure, "12.png"),
          width = 2000, height = 1200, res = 300)
@@
 ggexport(events_timingplot,
-         filename = paste0(params$output_figure, "13.png"),
+         filename = paste0(args$output_figure, "13.png"),
          width = 2000, height = 1200, res = 300)

### 2021-03-22 scripts/analyze_trajectories.R

@@ -45,7 +45,7 @@ print(args)
 df <- loadTrajectories(args$input, burninFrac = args$burnin, subsample = args$n)
 cat("\nNumber of analyzed trajectories (subsampled):", length(unique(df$states$traj)))
 
-
+print(args$demes)
 # Load demes configuration and sequence metadata -------------------------------
 demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
                     exclude_country = NA, exclude_division = NA,

### 2021-03-22 scripts/utils.R

@@ -301,7 +301,7 @@ plot_map <- function(demes) {
     #filter(country %in% demes$country | continent == "Europe") %>%
     mutate(deme = case_when(
       country %in% demes$country ~ country,
-      continent == "Europe" ~ "OtherEuropean"))
+      continent == "Europe" ~ "Other European"))
   
   zoom_to <- c(20, 50) 
   zoom_level <- 2.5

### 2021-03-22 scripts/analyze_trajectories.Rmd
dynamic title traj report

@@
 title: "Trajectory Mapping Results"
-subtitle: "Analysis Europe10 300 particles"
 author: "Cecilia Valenzuela"
@@
 argsR <- commandArgs(TRUE)
-args_list = str_split(paste(argsR, collapse=""), pattern="--")
-args_llist = lapply(args_list, function(x) str_split(x, pattern="="))
-argsDF = as.data.frame(do.call("rbind", args_llist[[1]]))
+args_list <- str_split(paste(argsR, collapse=""), pattern="--")
+args_llist <- lapply(args_list, function(x) str_split(x, pattern="="))
+argsDF <- as.data.frame(do.call("rbind", args_llist[[1]]))
 args <- as.list(as.character(argsDF$V2))
 names(args) <- argsDF$V1
-args$burnin = as.integer(args$burnin)
-args$n = as.integer(args$n)
+args$burnin <- as.integer(args$burnin)
+args$n <- as.integer(args$n)
+title <- paste("Analysis", gsub("\\.", " ", gsub(".TL.traj", " particles", str_split(args$args, "/")[[1]][3])))
 ```
+
+---
+subtitle: "`r title`"
+---
 

### 2021-03-23 reports/mt-thesis/discussion.tex
wip mt report

@@
 
+Once the first european imported cases, everything happen quickly.
+If we look at the evolution of thte epidedmic in China, longer ttime between first case andd exportedd cases.
+Greater connectivity among Europe spread faster the epidemic.
+
+Origin of Spain very close to official reporting date, while now the officiall case counts in Spain have been updadted to included cases detected early on by the surveillance system.
+This coulld be highly dependant on our samples and the dates of the sequences.
+
 \item Migration patterns other studies.

### 2021-03-23 reports/mt-thesis/results.tex
wip mt report

@@
-\graphicspath{{../../results/}}
-
 \chapter{Results}
 
-% \todo{Modify tables in R to have one value per cell, nr rate and times}
-
-\todo[inline]{The figures are from analysis Europe3: subsampled datasets according to cases/day and constant migration rate (no GLM), same xml specifications than Sarah's analysis.}
-
-\subsubsection*{Burden of SARS-CoV-2 infections in Europe}
-
-The inferred epidemic trajectories contain the information about the total number of cases until 8th of March.
-For the European countries, we obtain an inferred number of total cases above the number of confirmed cases to ECDC, consistent with known limited test availability of the first wave and  previous studies results \cite{Li2020} \cite{Wu2020}.
-These cases counts correspond to \todo{Include values} x-y times higher than the number of reported cases.
-Italy is the country with the highest infered number of cases x, followed by Spain, France and Germany.
-The infered number of cases for China is below the reported number of cases.
-\todo{limitations of the model, sequence information, partial outbreak dynamics}\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.8\textwidth]{201030_europe3_figtraj01.png}
-    \caption{Inferred population size summary statistics for each deme over time.
-The line represents the median population trajectory and the interval is the 95\%  credible interval in log scale from a random subsampled set of inferred epidemic trajectories.}
-    \label{fig:gribbon}
-    \todo[inline]{Specify that the random subset are 500 trajectories in all figure legends}
-\end{figure}
+\textbf{Origin of the first wave in Europe} 
+While the first cases in Europe were reported in mid-January, it is likely that the first European wave started earrlier and due to the lack of knowledge confusson  of the disease it was not detected by the surveillance systems.
+We assume the origin case to be an imported case from China into Europe.
+We analyze the time and destination of the first imported case into any of the European countries of our infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI).
+Location Italy or France.
+Figure \ref{fig:firstEUcase}.
+It is possible that several imported cases from China started the local transmission in the different Euorpean countries that a single imported case.
+So we cna look into the first imported cases in Europe and from its probability distribution the first scenario looks more likely, with most of these first introduction witht origin in China and like when looking at just one case they were more to Italy and France, but alos Germany.
+Earlly introductions in Spain have a very low probabilty, looking like Spain was not a key player in the orrigin of SARS-CoV-2 in Europe.
 
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj02.png}
-    \caption{Inferred epidemic trajectories over time.
-A random subsampled set of 500 trajectories is plotted.
-In each subplot, the trajectories (solid lines) are compared with the ECDC cumulative case count data (dashed line) in log scale.}
-    \label{fig:trajs}
-\end{figure}
-
-In Figure \ref{fig:gribbon} and/or \ref{fig:trajs} we compare the total number of inferred cases by day to the total cumulative number of cases that have been reported to ECDC that same day.
-Our inferred case counts follow a exponential growth earlier in time than the reported curve and with higher number of cases, being the difference bigger for later times in the epidemic.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.9\textwidth]{201030_europe3_figtraj03.png}
-    \caption{5-days reporting rate, calculated as the cumulative number of ECDC reported cases by the median number of cumulative inferred cases in intervals of 5 days until 8th of March. }
-    \label{fig:reported}
-\end{figure}
-
-We can think of a reporting rate as the number of reported cases relative to the total number of cases inferred by the model.
-This reporting rate decreases with time for all European countries, except for Italy that increases again around March, Figure \ref{fig:reported}.
-
-\subsubsection*{First introductions into European countries}
-
-\begin{figure}[p]
-    \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj04.png}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj0405.png}
     \caption{First introductions of SARS-CoV-2.
@@
 \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
-    \label{fig:first}
+    \label{fig:firstEUcase}
     \todo[inline]{ticks for every month}
 \end{figure}
+
+\textbf{First cases in European countries}
+
+Since in our model we consider the different European countries, we can also estimate from the infered epidemic trajectories the origin of the epidedmic in each particular country.
+In Figure we have the origin time for each country compared to the ddate of the first reporrted case to ECDC by that country.
+In average, we predict a delay in detection of x days for the Eurropean countries, out of the $95\% CCI$ for France, Germany and Italy.
+Origin in Spain close to the official detection day.
+We estimate that the epidemic in any of the European countries was seeded by an imported case from China as seen in Figurem with some small prrobability for Germany and Italy of having orriginatetdd from a case from each other, and mostly half of the rpobability in the case of Spain from the epidemdic started with a case imported from one of the others Euroopean countries.
+Same than before, we can look into the first 10 imported cases into each country and now we see that the transmission between european countries adquire much moer importance,  so once SARS-CoV-2 entered Europe, the transmission between european countrries were soon more important than the cases coming from China.
+This coulld relate also to the decrease in tthe trravel to and from China seen in figure.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj05.png}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj06.png}
     \caption{Countries ordered by the time of its first introduction, i.e.
@@
     \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj06.png}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj07.png}
     \caption{Order of countries by the time of its first migration out of the country, i.e.
@@
 
-\subsubsection*{Migration vs within-region transmission}
+\textbf{Detection timing and response}
+
+\begin{figure}[p]
+    \centering
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:first}
+    \todo[inline]{ticks for every month}
+\end{figure}
+
+The time between the first case in the country, i.e.
+first incoming migration event and the first case from within-region transmission is of x  (y-z) with similar values for all demes?.
+The time from the first incoming migration to the first outgoing migration is longer with a median of x (y-z).
+(This could be interesting to say if we should focus or not the screening and testing capacities to detect incoming migrations or if when we have evidence of cases in the population we shoud follow a more general strategy to find cases in the population according to the model.
+Is it different for each country?)\\
+
+(How well did the countries detecting the first cases, there were already within region transmision?) We compare the date of the first reported rate of each country with the date in which within-region transmission for that country started according to the model.
+We see some differences among countries, France and Germany had in all inferred epidemic trajectories ongoing within-region transmission when first cases werre reported, while x\% of epidemic trajectories did not have had a within region transmission case when the first case in Spain was reported.\\
+
+\begin{figure}[p]
+    \centering
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj13.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:first}
+    \todo[inline]{ticks for every month}
+\end{figure}
+
+We can also compare the date of the first reported case with the date of the first outgoing migration from the country.
+(This could be interesting to say if a extreme measure closing borders with the first case could be effective to impede transmission to other countries: percentage of trajectories whre transmission to Europe would have been avoided.
+For other countries we can look at how many migrations events could have been avoided (and how many not) if the country closed borders after first reported case according to the model.
+Not realistic measure, extreme case.)\\
+
+\textbf{Burden of SARS-CoV-2 infections in Europe}
+The inferred epidemic trajectories contain the information about the total number of cases until 8th of March.
+For the European countries, we obtain an inferred number of total cases above the number of confirmed cases to ECDC, consistent with known limited test availability of the first wave and  previous studies results \cite{Li2020} \cite{Wu2020}.
+These cases counts correspond to \todo{Include values} x-y times higher than the number of reported cases.
+Italy is the country with the highest infered number of cases x, followed by Spain, France and Germany.
+The infered number of cases for China is below the reported number of cases.
+\todo{limitations of the model, sequence information, partial outbreak dynamics}\\
 
@@
     \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj07.png}
+    \includegraphics[width=0.8\textwidth]{210205_europe10_figtraj01.png}
+    \caption{Inferred population size summary statistics for each deme over time.
+The line represents the median population trajectory and the interval is the 95\%  credible interval in log scale from a random subsampled set of inferred epidemic trajectories.}
+    \label{fig:gribbon}
+    \todo[inline]{Specify that the random subset are 500 trajectories in all figure legends}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj02.png}
+    \caption{Inferred epidemic trajectories over time.
+A random subsampled set of 500 trajectories is plotted.
+In each subplot, the trajectories (solid lines) are compared with the ECDC cumulative case count data (dashed line) in log scale.}
+    \label{fig:trajs}
+\end{figure}
+
+In Figure \ref{fig:gribbon} and/or \ref{fig:trajs} we compare the total number of inferred cases by day to the total cumulative number of cases that have been reported to ECDC that same day.
+Our inferred case counts follow a exponential growth earlier in time than the reported curve and with higher number of cases, being the difference bigger for later times in the epidemic.\\
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj11.png}
+    \caption{5-days reporting rate, calculated as the cumulative number of ECDC reported cases by the median number of cumulative inferred cases in intervals of 5 days until 8th of March. }
+    \label{fig:reported}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj12.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:first}
+    \todo[inline]{ticks for every month}
+\end{figure}
+
+We can think of a reporting rate as the number of reported cases relative to the total number of cases inferred by the model.
+This reporting rate decreases with time for all European countries, except for Italy that increases again around March, Figure \ref{fig:reported}.
+
+\textbf{Imported cases vs local transmission}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj08.png}
     \caption{Median and 95\% credible interval for the cumulative number of events (within-region transmissions and migrations to the country (incoming) and from the country (outgoing) over time.}
@@
 
-The time between the first case in the country, i.e.
-first incoming migration event and the first case from within-region transmission is of x  (y-z) with similar values for all demes?.
-The time from the first incoming migration to the first outgoing migration is longer with a median of x (y-z).
-(This could be interesting to say if we should focus or not the screening and testing capacities to detect incoming migrations or if when we have evidence of cases in the population we shoud follow a more general strategy to find cases in the population according to the model.
-Is it different for each country?)\\
-
-(How well did the countries detecting the first cases, there were already within region transmision?) We compare the date of the first reported rate of each country with the date in which within-region transmission for that country started according to the model.
-We see some differences among countries, France and Germany had in all inferred epidemic trajectories ongoing within-region transmission when first cases werre reported, while x\% of epidemic trajectories did not have had a within region transmission case when the first case in Spain was reported.\\
-
-We can also compare the date of the first reported case with the date of the first outgoing migration from the country.
-(This could be interesting to say if a extreme measure closing borders with the first case could be effective to impede transmission to other countries: percentage of trajectories whre transmission to Europe would have been avoided.
-For other countries we can look at how many migrations events could have been avoided (and how many not) if the country closed borders after first reported case according to the model.
-Not realistic measure, extreme case.)\\
-
-\subsubsection*{Migrations patterns}
+\textbf{Interational transmission patterns}
 Figures \ref{fig:migs_srcdest} and \ref{fig:migs}.\\
@@
   \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201030_europe3_figtraj09a.png}
+  \includegraphics[width=\textwidth]{210205_europe10_figtraj09a.png}
   \label{fig:migs1}
@@
   \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201030_europe3_figtraj09b.png}
+  \includegraphics[width=\textwidth]{210205_europe10_figtraj09b.png}
   \label{fig:migs2}
@@
   \begin{minipage}[t]{0.4\textwidth}
-  \includegraphics[width=\textwidth]{201030_europe3_figtraj09c.png}
+  \includegraphics[width=\textwidth]{210205_europe10_figtraj09c.png}
   \label{fig:migs3}
@@
 
-\begin{figure}[]
+\begin{figure}[p]
     \centering
-    \includegraphics[width=\textwidth]{201030_europe3_figtraj08.png}
-    \caption{}
-    \label{fig:migs_srcdest}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj10.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:first}
+    \todo[inline]{ticks for every month}
 \end{figure}

### 2021-03-23 scripts/analyze_trajectories.Rmd
wip mt report

@@
 # Debugging
+# args = list()
 # args$burnin = 0.1
 # args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
-# #args$input = "./results/europeC/201211_europe6p2.TL.traj"
-# args$input = "/Users/maceci/210205_europe10/results/europeC/210205_europe10.10000.TL.traj"
+# args$args = "/Users/maceci/210205_europe10/results/europeC/210205_europe10.10000.TL.traj"
 # args$metadata =  "./results/europeC/final_metadata.tsv"
@@
   mutate(deme = str_to_title(deme)) %>%
-  mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme))
+  mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme)) %>%
+  arrange(deme)
 
@@
   geom_line(aes(date, median_cumvalue, colour = deme)) +
-  ylab("Population size") +
-  xlab("Date") +
+  ylab("active SARS-CoV-2 infections") +
+  #xlab("Date") +
   #labs(title = "SARS-CoV-2 early epidemics in Europe and Hubei",
@@
                        values = c(dcolors[[d]], dark_dcolors[[d]])) +
-    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
+    scale_alpha_manual(name = "", values = c(0.1, 1), guide = FALSE) +
     scale_linetype_manual(name = "", values = c(1, 5),
@@
                                             " total reported cases"))) +
-    scale_size_manual(name = "", values = c(0.5, 1.5),
+    scale_size_manual(name = "", values = c(0.2, 1.5),
                       labels = c(paste0(d, " inferred\n population trajectories"),
@@
     labs(title = d) +
-    ylab("COVID-19 cases") +
+    ylab(ifelse(d %in% c("China", "Germany", "Other European"), "SARS-CoV-2 cases", "")) +
     theme(axis.title.x=element_blank(),
@@
   ggplot() +
-  geom_density_ridges(aes(x = date, y = deme, fill = deme), quantile_lines = TRUE, quantiles = c(0.025, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) +
+  geom_density_ridges(aes(x = date, y = deme, fill = deme), color = "grey50", quantile_lines = TRUE, quantiles = c(0.025, 0.5, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) +
   #stat_density_ridges(, alpha = 0.7)
@@
   geom_point(data = first_ecdc, 
-             aes(date, deme, fill = deme), shape = 23, size = 4) +
+             aes(date, deme, fill = deme), color = "grey50", shape = 23, size = 3) +
   scale_fill_manual(values = c(dcolors, "Europe" = "grey")) +
   scale_color_manual(values = c(dcolors, "Europe" = "grey")) +
-  scale_x_date(limits = c(ymd("2019-09-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %Y", expand = c(0,0)) +
+  scale_x_date(limits = c(ymd("2019-10-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %Y", expand = c(0,0)) +
   scale_y_discrete(expand = c(0.05,0), "probability") +
@@
         panel.grid.major.x =  element_line(colour = "grey80", size = 0.3, linetype = 2),
+        panel.grid.major.y =  element_line(colour = "grey80", size = 0.3, linetype = 1),
         #panel.grid.minor.x =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
@@
 
-## First introduction into Europe, location
+## First imported cases into Europe, time and location
 ```{r first-eu, fig.height=4}
@@
 m = 1
+facet_names <- demes$deme
+names(facet_names) <- demes$deme
+facet_names["OtherEuropean"] <- "Other EU"
 first_df1 <- df_traj %>%
@@
   geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_wrap(~deme, nrow = 1) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE,
+             labeller = labeller(deme = facet_names)) +
   scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
-  ylab("Destination country\nprobability") +
-  labs(title = "First imported case into Europe") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First imported case into Europe, \ndestination") +
   theme(legend.position = "none",
-        axis.title.x=element_blank(),
-        axis.text.x = element_text(angle = 90))
-first_eu
+        axis.title.x=element_blank())
 ggexport(first_eu,
@@
 
-## First 10 introductions into Europe, location
 ```{r first-eu10, fig.height=4}
@@
 
-first_dest <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
+first_eu10d <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
   geom_density_line(
     data = select(first_df10, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2
-  ) + 
+    color = "transparent", alpha = 0.8, bw = 2) + 
   geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_wrap(~deme, nrow = 1) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
   scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
-  ylab("Destination country\nprobability") +
-  labs(title = "First 10 imported cases into Europe") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into Europe, \ndestination") +
   theme(legend.position = "none",
-        axis.title.x=element_blank(),
-        axis.text.x = element_blank(),
-        axis.ticks.x = element_blank())
+        axis.title.x=element_blank())
 
-first_source <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
+first_eu10s <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
   geom_density_line(
-    data = select(first_df10, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2
-  ) + 
+    data = select(first_df10, -partner), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 2) + 
   geom_density_line(aes(fill = partner), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_wrap(~deme, nrow = 1) +
+  facet_grid(rows = vars(partner), switch = "y", drop = FALSE, labeller = labeller(partner = facet_names)) +
   scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
-  ylab("Source country\nprobability") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by source country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into Europe, \nsource") +
   theme(legend.position = "none",
-        axis.title.x=element_blank(),
-        strip.background = element_blank(),
-        strip.text = element_blank(),
-        axis.text.x = element_text(angle = 90))
+        axis.title.x=element_blank())
 
-first_10eu <- ggarrange(first_dest, first_source, labels = chartr("123456789", "ABCDEFGHI", 1:3), nrow = 2, ncol = 1)
-first_10eu
-ggexport(first_10eu,
+first_eu10 <- ggarrange(first_eu10d, first_eu10s, labels = chartr("123456789", "ABCDEFGHI", 1:3), font.label = list(size = 8), nrow = 1, ncol = 2, hjust = -7, vjust = 2)
+ggexport(first_eu10,
          filename = paste0(args$output_figure, "05.png"),
          width = 2000, height = 1200, res = 300)
+
+first_eu_all <- ggarrange(first_eu, first_eu10d, first_eu10s, labels = chartr("123456789", "ABCDEFGHI", 1:3), font.label = list(size = 9), nrow = 1, ncol = 3, hjust = -6, vjust = 2)
+first_eu_all
+ggexport(first_eu_all,
+         filename = paste0(args$output_figure, "0405.png"),
+         width = 2500, height = 1500, res = 300)
+
 ```
@@
   geom_density_line(aes(fill = partner, color = partner), bw = 3, alpha = 0.6) +
-  labs(title = "First imported case into each country") +
+  #labs(title = "First imported case into each country") +
   facet_wrap(~deme, nrow = 1) +
@@
   geom_density_line(aes(fill = partner, color = partner), bw = 3, alpha = 0.6) +
-  labs(title = "First 10 imported cases into each country") +
+  #labs(title = "First 10 imported cases into each country") +
   facet_wrap(~deme, nrow = 1) +
@@
         h = "top", 
-        major.at = seq(from = 0, to = xlim[2], by = case_when(
+        major.at = seq(from = 0, to = xlim[2], by =
+        case_when(
           xlim[2] > 50500 ~ 20000,
@@
           xlim[2] > 10500 ~ 5000,
-          xlim[2] > 5500 ~ 5000,
+          #xlim[2] > 5500 ~ 5000,
           xlim[2] > 2500 ~ 2000,
@@
           xlim[2] > 800 ~ 500,
-          xlim[2] > 500 ~ 500,
+          xlim[2] > 500 ~ 250,
           xlim[2] > 200 ~ 100,
@@
           xlim[2] > 1 ~ 1,
-          TRUE ~ 0.01)), 
+          round(xlim[2]) == 0 ~ 0.25,
+          TRUE ~ round(xlim[2])/4)),
         minor.ticks = 1, 
@@
             evalues = sum(value)/N) %>%
+  #mutate(evalues = value) %>%
   select(traj, epoch, src, dest, evalues)
@@
   geom_ribbon(aes(year_week, ymin = l95_report, ymax = h95_report,fill = deme), alpha = 0.1) +
-  ylab("Reporting rate") +
+  ylab("weekly reporting proportion") +
   scale_color_manual(values = dcolors, name = "") +
@@
   geom_ribbon(aes(date, ymin = l95_inc, ymax = h95_inc,fill = deme), alpha = 0.1) +
-  ylab("Reporting rate") +
+  ylab("weekly incidence %") +
   scale_color_manual(values = dcolors, name = "") +

### 2021-03-23 scripts/utils.R

@@ -13,35 +13,26 @@ library(wpp2019)
 
 set_plotopts <- function() {
   theme_set(theme_minimal())
-  theme_update(#strip.background = element_rect(fill = "grey95", color = "grey95", size = 1),
-               panel.grid = element_blank(),
-               #panel.background = element_rect(fill = "white", colour = "grey80"),
-               axis.ticks = element_line(),
+  theme_update(panel.grid = element_blank(),
                legend.position = "bottom",
                legend.box="vertical",
-               legend.text=element_text(size = 8),
-               legend.title=element_text(size = 8),
+               legend.text=element_text(size = 7),
+               legend.title=element_text(size = 7),
                strip.background = element_rect(fill = "white", colour = "grey80"),
-               strip.text = element_text(color = "grey30", face = "bold"),
-               #legend.title = element_text(size = 9, face = "bold"),
-               axis.title.y = element_text(size = 9, face = "bold", margin = margin(r = 10)),
-               axis.title.x = element_text(size = 9, face = "bold", margin = margin(t = 10)),
-               #axis.line.x = element_line(color = "grey30"),
-               # axis.text.x.bottom = element_text(hjust = 0, vjust = .5,
-               #                            margin = margin(0, 0, 0, 0),
-               #                            color = "grey30",
-               #                            lineheight = 0.8),
-               #axis.ticks.x = element_blank(),
-               #axis.line.y.left = element_line(color = "grey30"),
-               # axis.text.y.right = element_text(hjust = 0, vjust = .5,
-               #                           margin = margin(0, 0, 0, 0),
-               #                           color = "grey30",
-               #                           lineheight = 0.8),
-               panel.grid.major.y =  element_line(colour = "grey80", size = 0.3, linetype = 2),
-               panel.grid.minor.y =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
+               strip.text = element_text(size = 8, color = "grey30"),
+               axis.title.y = element_text(size = 8, #face = "bold", 
+                                           margin = margin(r = 10)),
+               axis.title.x = element_text(size = 8, #face = "bold", 
+                                           margin = margin(t = 10)),
+               panel.grid.major.y =  element_line(colour = "grey80", size = 0.2, linetype = 2),
+               #panel.grid.minor.y =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
                panel.background =  element_blank(),
[…]

### 2021-03-24 reports/mt-thesis/results.tex
results mt wip

@@
 
-\textbf{Origin of the first wave in Europe} 
-While the first cases in Europe were reported in mid-January, it is likely that the first European wave started earrlier and due to the lack of knowledge confusson  of the disease it was not detected by the surveillance systems.
-We assume the origin case to be an imported case from China into Europe.
-We analyze the time and destination of the first imported case into any of the European countries of our infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI).
-Location Italy or France.
-Figure \ref{fig:firstEUcase}.
-It is possible that several imported cases from China started the local transmission in the different Euorpean countries that a single imported case.
-So we cna look into the first imported cases in Europe and from its probability distribution the first scenario looks more likely, with most of these first introduction witht origin in China and like when looking at just one case they were more to Italy and France, but alos Germany.
+\textbf{The origin of SARS-CoV-2 in Europe} 
+While the first cases in Europe were reported end of January, the first European wave is suspected to have started earlier but been undetected, mistaken for other common repiratory diseases in winter.
+In our model, we assume the origin of SARS-CoV-2 in Europe to be an imported case from China into any of the European countries.
+We analyze the time and destination of the first imported case in the infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI).
+As shown in Figure \ref{fig:firstEUcase} A, we place the destination country of this first imported case in Italy or France, with x of probability each.
+
+Likewise, we analyze the first ten imported cases into Europe and obtain the probability distribution of the source and destination country, Figure \ref{fig:firstEUcase} B and C.
+The role of Italy as the receptor of these firts cases is highlighted even more looking at several cases.
 Earlly introductions in Spain have a very low probabilty, looking like Spain was not a key player in the orrigin of SARS-CoV-2 in Europe.
+China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China.
 
@@
 
-\textbf{First cases in European countries}
+\textbf{The early spread of SARS-CoV-2 among the European countries}
 
@@
 
-We would like to answer the question of when and where the first case in Europe occured.
-We can look to the model inferred epidemic trajectories and analyze the time distribution of the first case in Europe and in each deme.
-China is defined as the origin of the epidemic in the analysis and this is reflected in Figure \ref{fig:first} A.
-The occurrence of the origin of the epidemic, i.e.
-first case in China, is almost one month before the epidemic starts in any European country, ranging \todo{Include values}  x-y.
-The inferred date of the first case in Europe is inferred to be between a 95\% interval x-y.
 Among the European countries, France and Germany have the earlier time of introduction, followed by Italy, Other European deme and Spain.
@@
 In all cases, the reported day was later than the median of the inferred distribution, but it is inside the 95\% interval.\\
-
-The first case in Europe for each trajectoy corresponds to a migration from China to any European country.
-We can analyze the destination of this first migration into Europe as is shown in Figure \ref{fig:first} B.
-According to our model, the probability that this first European case was in Spain is very low.
-The destination of the first case amonth the rest of European demes is equally probable and there is no clear difference in the time distribution for each destination country, with a maximum probability around x.\\
-\todo{explain this better, compute the probability of the first case in Europe destination}
 
@@
 
-\textbf{Detection timing and response}
+\textbf{Detection timing and response to the European epidemics}
 
@@
 
-\textbf{Interational transmission patterns}
+\textbf{International transmission patterns}
 Figures \ref{fig:migs_srcdest} and \ref{fig:migs}.\\

### 2021-03-24 reports/mt-thesis/results.tex
wip results mt

@@
 In our model, we assume the origin of SARS-CoV-2 in Europe to be an imported case from China into any of the European countries.
-We analyze the time and destination of the first imported case in the infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI).
+We analyze the time and destination of the first imported case in the infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI)  \todo{with final results}.
 As shown in Figure \ref{fig:firstEUcase} A, we place the destination country of this first imported case in Italy or France, with x of probability each.
@@
 Likewise, we analyze the first ten imported cases into Europe and obtain the probability distribution of the source and destination country, Figure \ref{fig:firstEUcase} B and C.
-The role of Italy as the receptor of these firts cases is highlighted even more looking at several cases.
-Earlly introductions in Spain have a very low probabilty, looking like Spain was not a key player in the orrigin of SARS-CoV-2 in Europe.
+The role of Italy as the receptor of these first cases is highlighted even more looking at several cases.
+Early introductions in Spain have a very low probabilty, looking like Spain was not a key player in the origin of SARS-CoV-2 in Europe.
 China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China.
+\todo{with final results}
 

### 2021-03-24 scripts/analyze_trajectories.Rmd
wip results mt

@@
 # args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
-# args$args = "/Users/maceci/210205_europe10/results/europeC/210205_europe10.10000.TL.traj"
+# args$args = "/Users/maceci/europe10last/210205_europe10.10000.TL.traj"
 # args$metadata =  "./results/europeC/final_metadata.tsv"
@@
 
-## First introduction
+## Origin of SARS-CoV-2 in Europe
 
-```{r first, fig.height=4}
-# 2. First introduction distribution density by deme and time of first official reported case to ECDC
+```{r first-eu, fig.height=4}
+first_df1 <- df_traj %>%
+  # 1. Prepare data
+  filter(var == "IM", !is.na(partner),  value != 0) %>%
+  group_by(traj) %>%
+  arrange(date, partner) %>%
+  slice(1)
+
 first_ecdc <- case_data %>% 
@@
   slice(1)
-#first_reported$min_date[first_reported$dest == china_deme] <- ymd("2019-12-12")
 
+first_eui <- first_df1 %>% 
+  ggplot(aes(date)) +
+  geom_density(aes(y = ..count..), fill = "#21908CFF", alpha = .5, col = NA) + 
+  geom_histogram(binwidth = 1, alpha = .5, color = "grey60") +
+  geom_vline(aes(xintercept = median(first_df1$date)), linetype = 1) +
+  geom_vline(aes(xintercept = min((first_ecdc %>% filter(deme != "China"))$date)), linetype = 2) +
+  ylab("epidemic trajectory count") +
+  scale_x_date(limits = c(ymd("2019-12-15"), ymd("2020-02-01")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  theme(axis.title.x = element_blank())
+
+first_euii <- first_df1 %>% 
+  ggplot(aes(date)) +
+  geom_density(fill = "#21908CFF", alpha = .5, col = NA) + 
+  geom_histogram(binwidth = 1, alpha = .5, color = "grey60", aes( y = ..density..)) +
+  geom_vline(aes(xintercept = median(first_df1$date)), linetype = 1) +
+  geom_vline(aes(xintercept = min((first_ecdc %>% filter(deme != "China"))$date)), linetype = 2) +
+  ylab("probability density") +
+  scale_y_continuous(position = "right") +
+  scale_x_date(limits = c(ymd("2019-12-15"), ymd("2020-02-01")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  theme(axis.title.x = element_blank())
+
+ggexport(first_eui,
+         filename = paste0(args$output_figure, "03i.png"),
+         width = 2100, height = 600, res = 300)
+ggexport(first_euii,
+         filename = paste0(args$output_figure, "03ii.png"),
+         width = 2100, height = 600, res = 300)
+```
+
+## First imported cases into Europe, time and location
+```{r first-eu-destsrc, fig.height=4}
+# 2.2b First introduction into Europe, distribution density destination
+facet_names <- demes$deme
+names(facet_names) <- demes$deme
+facet_names["OtherEuropean"] <- "Other EU"
+
+first_eu1d <- ggplot(first_df1, aes(x = date, y = ..count../(args$n))) +
+  geom_density_line(
+    data = select(first_df1, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 2) + 
+  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE,
+             labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First imported case into Europe, \ndestination") +
+  theme(legend.position = "none",
+        axis.title.x=element_blank())
+
+# 2.2b First 10 introduction into Europe, distribution density destination
+m = 10
+first_df10 <- df_traj %>%
+  # 1. Prepare data
+  filter(var == "IM", !is.na(partner),  value != 0) %>%
+  group_by(traj) %>%
+  arrange(date, partner) %>%
+  slice(1:m)
+
+first_eu10d <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_df10, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 2) + 
+  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into Europe, \ndestination") +
+  theme(legend.position = "none",
+        axis.title.x=element_blank())
+
+first_eu10s <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_df10, -partner), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 2) + 
+  geom_density_line(aes(fill = partner), bw = 2, color = "transparent", alpha = 0.8) +
+  facet_grid(rows = vars(partner), switch = "y", drop = FALSE, labeller = labeller(partner = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by source country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into Europe, \nsource") +
+  theme(legend.position = "none",
+        axis.title.x=element_blank())
+
+first_eu_ds <- ggarrange(first_eu1d, first_eu10d, first_eu10s, labels = chartr("123456789", "ABCDEFGHI", 1:3), font.label = list(size = 9), nrow = 1, ncol = 3, hjust = -6, vjust = 2)
+first_eu_ds
+ggexport(first_eu_ds,
+         filename = paste0(args$output_figure, "04.png"),
+         width = 2500, height = 1500, res = 300)
+
+```
+
+## First introductions into each country, source
+```{r first-bycountry, fig.height=4}
+# 2.2b First introduction into each country, distribution density destination
+m = 1
+first_country <- df_traj %>%
+  # 1. Prepare data
+  filter(var == "IM", !is.na(partner),  value != 0) %>%
+  group_by(traj, deme) %>%
+  arrange(date, partner) %>%
+  slice(1:m)
+
+first_countryplot <- ggplot(first_country, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_country, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 3
+  ) + 
+  geom_density_line(aes(fill = partner), color = "transparent", bw = 3, alpha = 0.6) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(#limits = c(ymd("2020-01-01"), ymd("2020-02-15")), 
+    date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First imported case into the country, \nsource") +
+  theme(#legend.position = "none",
+        legend.key.size = unit(0.5, "cm"),
+        axis.title.x=element_blank()) +
+  guides(fill = guide_legend(nrow = 1))
+
+# 2.2b First 10 introductions into each country, distribution density destination
+m = 10
+first_10country <- df_traj %>%
+  # 1. Prepare data
+  filter(var == "IM", !is.na(partner),  value != 0) %>%
+  group_by(traj, deme) %>%
+  arrange(date, partner) %>%
+  slice(1:m)
+
+first_10countryplot <- ggplot(first_10country, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_10country, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 3
+  ) + 
+  geom_density_line(aes(fill = partner), color = "transparent", bw = 3, alpha = 0.6) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(#limits = c(ymd("2020-01-01"), ymd("2020-02-15")), 
+    date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into the country, \nsource") +
+  theme(#legend.position = "none",
+        legend.key.size = unit(0.5, "cm"),
+        axis.title.x=element_blank()) +
+  guides(fill = guide_legend(nrow = 1))
+
+first_countryds <- ggarrange(first_countryplot, first_10countryplot, labels = chartr("123456789", "ABCDEFGHI", 1:2), font.label = list(size = 9), nrow = 1, ncol = 2, hjust = -6, vjust = 2, common.legend = TRUE, legend = "bottom")
+first_countryds
+
+ggexport(first_countryds,
+         filename = paste0(args$output_figure, "05.png"),
+         width = 2000, height = 1200, res = 300)
+```
+
+```{r first-times, fig.height=4}
+# 2. First introduction distribution density by deme and time of first official reported case to ECDC
 first_times <- df_traj_cases %>%
@@
          width = 2000, height = 800, res = 300)
-```
-
-## First imported cases into Europe, time and location
-```{r first-eu, fig.height=4}
-# 2.2b First introduction into Europe, distribution density destination
-m = 1
-facet_names <- demes$deme
-names(facet_names) <- demes$deme
-facet_names["OtherEuropean"] <- "Other EU"
-first_df1 <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0) %>%
-  group_by(traj) %>%
-  arrange(date, partner) %>%
-  slice(1:m)
-
-first_eu <- ggplot(first_df1, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_df1, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2) + 
-  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_grid(rows = vars(deme), switch = "y", drop = FALSE,
-             labeller = labeller(deme = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by destination country") +
-  xlab("date") +
-  labs(title = "First imported case into Europe, \ndestination") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank())
-ggexport(first_eu,
-         filename = paste0(args$output_figure, "04.png"),
-         width = 2100, height = 600, res = 300)
-```
-
-```{r first-eu10, fig.height=4}
-# 2.2b First 10 introduction into Europe, distribution density destination
-m = 10
-first_df10 <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0) %>%
-  group_by(traj) %>%
-  arrange(date, partner) %>%
-  slice(1:m)
-
-first_eu10d <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_df10, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2) + 
-  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by destination country") +
-  xlab("date") +
-  labs(title = "First 10 imported cases into Europe, \ndestination") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank())
-
-first_eu10s <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_df10, -partner), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2) + 
-  geom_density_line(aes(fill = partner), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_grid(rows = vars(partner), switch = "y", drop = FALSE, labeller = labeller(partner = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by source country") +
-  xlab("date") +
-  labs(title = "First 10 imported cases into Europe, \nsource") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank())
-
-first_eu10 <- ggarrange(first_eu10d, first_eu10s, labels = chartr("123456789", "ABCDEFGHI", 1:3), font.label = list(size = 8), nrow = 1, ncol = 2, hjust = -7, vjust = 2)
-ggexport(first_eu10,
-         filename = paste0(args$output_figure, "05.png"),
-         width = 2000, height = 1200, res = 300)
-
-first_eu_all <- ggarrange(first_eu, first_eu10d, first_eu10s, labels = chartr("123456789", "ABCDEFGHI", 1:3), font.label = list(size = 9), nrow = 1, ncol = 3, hjust = -6, vjust = 2)
-first_eu_all
-ggexport(first_eu_all,
-         filename = paste0(args$output_figure, "0405.png"),
-         width = 2500, height = 1500, res = 300)
-
-```
-
-## First introductions into each country, source
-```{r first-bycountry, fig.height=4}
-# 2.2b First introduction into each country, distribution density destination
-m = 1
-first_country <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0, deme != "China") %>%
-  group_by(traj, deme) %>%
-  arrange(date, partner) %>%
-  slice(1:m)
-
-first_countryplot <- ggplot(first_country, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_country, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 3
-  ) + 
-  geom_density_line(aes(fill = partner), bw = 3, color = "transparent", alpha = 0.6) +
-  geom_density_line(aes(fill = partner, color = partner), bw = 3, alpha = 0.6) +
-  #labs(title = "First imported case into each country") +
-  facet_wrap(~deme, nrow = 1) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_color_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
-  ylab("Source country\nprobability") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank(),
-        axis.text.x = element_text(angle = 90))
-first_countryplot
-ggexport(first_countryplot,
-         filename = paste0(args$output_figure, "06.png"),
-         width = 2000, height = 1200, res = 300)
-```
-
-## First 10 introductions into each country, source
-
-```{r first-10bycountry, fig.height=4}
-# 2.2b First 10 introductions into each country, distribution density destination
-m = 10
-first_10country <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0, deme != "China") %>%
-  group_by(traj, deme) %>%
-  arrange(date, partner) %>%
-  slice(1:m)
-
-first_10countryplot <- ggplot(first_10country, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_10country, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 3
-  ) + 
-  geom_density_line(aes(fill = partner), bw = 3, color = "transparent", alpha = 0.6) +
-  geom_density_line(aes(fill = partner, color = partner), bw = 3, alpha = 0.6) +
-  #labs(title = "First 10 imported cases into each country") +
-  facet_wrap(~deme, nrow = 1) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_color_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(date_breaks = "1 week", date_labels = "%b %d", expand = c(0,0)) +
-  ylab("Source country\nprobability") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank(),
-        axis.text.x = element_text(angle = 90))
-first_10countryplot
-ggexport(first_10countryplot,
-         filename = paste0(args$output_figure, "07.png"),
-         width = 2000, height = 1200, res = 300)
 ```

### 2021-03-24 reports/mt-thesis/results.tex
results mt wip

@@
 \textbf{The origin of SARS-CoV-2 in Europe} 
-While the first cases in Europe were reported end of January, the first European wave is suspected to have started earlier but been undetected, mistaken for other common repiratory diseases in winter.
+While the first case in Europe was reported to ECDC by France authorities on January 25, 2020, the first European wave is suspected to have started earlier but been undetected, mistaken for other common repiratory diseases in winter.
 In our model, we assume the origin of SARS-CoV-2 in Europe to be an imported case from China into any of the European countries.
-We analyze the time and destination of the first imported case in the infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI)  \todo{with final results}.
-As shown in Figure \ref{fig:firstEUcase} A, we place the destination country of this first imported case in Italy or France, with x of probability each.
+We analyze the time and destination of the first imported case in the infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI), Figure \ref{fig:firstEU} \todo{with final results}.
+The actual reported date to ECDC is in the tail of the $95\%$ credible interval.
+As shown in Figure \ref{fig:firstEUcountry} A, we place the destination country of this first imported case in Italy or France, with x of probability each.
 
-Likewise, we analyze the first ten imported cases into Europe and obtain the probability distribution of the source and destination country, Figure \ref{fig:firstEUcase} B and C.
-The role of Italy as the receptor of these first cases is highlighted even more looking at several cases.
-Early introductions in Spain have a very low probabilty, looking like Spain was not a key player in the origin of SARS-CoV-2 in Europe.
-China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China.
-\todo{with final results}
-
-\begin{figure}[p]
+\begin{figure}[h]
     \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj0405.png}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
     \caption{First introductions of SARS-CoV-2.
@@
 \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
-    \label{fig:firstEUcase}
+    \label{fig:firstEU}
+    \todo[inline]{ticks for every month}
+\end{figure}
+
+Likewise, we analyze the first ten imported cases into Europe and obtain the probability distribution of the source and destination country, Figure \ref{fig:firstEUcountry} B and C.
+The role of Italy as the receptor of these first cases is highlighted even more looking at several cases.
+Early introductions in Spain have a very low probabilty, looking like Spain was not a key player in the origin of SARS-CoV-2 in Europe.
+China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China.
+\todo{with final results}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj04.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:firstEUcountry}
     \todo[inline]{ticks for every month}
@@
 Same than before, we can look into the first 10 imported cases into each country and now we see that the transmission between european countries adquire much moer importance,  so once SARS-CoV-2 entered Europe, the transmission between european countrries were soon more important than the cases coming from China.
-This coulld relate also to the decrease in tthe trravel to and from China seen in figure.
+This could relate also to the decrease in tthe trravel to and from China seen in figure.
 
@@
 However, in our analysis we obtain a much higher support for China being the most probable source of the first case for all European demes, Figure \ref{fig:first} C.\\
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj05.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:firstEUcountry}
+    \todo[inline]{ticks for every month}
+\end{figure}
 
@@
     \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj13.png}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj07.png}
     \caption{First introductions of SARS-CoV-2.

### 2021-03-24 scripts/analyze_trajectories.Rmd
results mt wip

@@
 # Debugging
-# args = list()
-# args$burnin = 0.1
-# args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
-# args$args = "/Users/maceci/europe10last/210205_europe10.10000.TL.traj"
-# args$metadata =  "./results/europeC/final_metadata.tsv"
-# args$n = 100
-# demes = args$demes
-# demes$exclude_country[5] = "France,Germany,Italy,Spain"
-# demes <- demes %>% arrange(deme)
+args = list()
+args$burnin = 0.1
+args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
+args$args = "/Users/maceci/europe10last/210205_europe10.10000.TL.traj"
+args$metadata =  "./results/europeC/final_metadata.tsv"
+args$n = 100
+demes = args$demes
+demes$exclude_country[5] = "France,Germany,Italy,Spain"
+demes <- demes %>% arrange(deme)
 ```
@@
 
+## Origin of SARS-CoV-2 in Europe
+
+```{r first-eu, fig.height=4}
+first_df1 <- df_traj %>%
+  # 1. Prepare data
+  filter(var == "IM", !is.na(partner),  value != 0) %>%
+  group_by(traj) %>%
+  arrange(date, partner) %>%
+  slice(1)
+
+first_ecdc <- case_data %>% 
+  filter(source == "ecdc", total_confirmed != 0) %>% 
+  group_by(deme) %>% 
+  arrange(date) %>% 
+  slice(1)
+
+first_eui <- first_df1 %>% 
+  ggplot(aes(date)) +
+  geom_density(aes(y = ..count..), fill = "grey", alpha = .5, col = NA) + 
+  geom_histogram(binwidth = 1, alpha = .5, fill = "#21908CFF", color = "#21908CFF") +
+  geom_vline(aes(xintercept = median(first_df1$date), linetype = "first case median infered date")) +
+  geom_vline(aes(xintercept = min((first_ecdc %>% filter(deme != "China"))$date), linetype = "first confirmed case to ECDC")) +
+  ylab("epidemic trajectory count") +
+  scale_x_date(limits = c(ymd("2019-12-15"), ymd("2020-02-01")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  scale_linetype_manual(name = "", values = c(1,2)) +
+  theme(axis.title.x = element_blank())
+
+first_euii <- first_df1 %>% 
+  ggplot(aes(date)) +
+  geom_density(fill = "grey", alpha = .5, col = NA) + 
+  geom_histogram(binwidth = 1, alpha = .5, fill = "#21908CFF", color = "#21908CFF", aes( y = ..density..)) +
+  geom_vline(aes(xintercept = median(first_df1$date), linetype = "first case median infered date")) +
+  geom_vline(aes(xintercept = min((first_ecdc %>% filter(deme != "China"))$date), linetype = "first confirmed case to ECDC")) +
+  ylab("probability density") +
+  scale_y_continuous(position = "right") +
+  scale_x_date(limits = c(ymd("2019-12-15"), ymd("2020-02-01")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  scale_linetype_manual(name = "", values = c(1,2)) +
+  theme(axis.title.x = element_blank())
+
+ggexport(first_eui,
+         filename = paste0(args$output_figure, "03i.png"),
+         width = 2100, height = 800, res = 300)
+ggexport(first_euii,
+         filename = paste0(args$output_figure, "03ii.png"),
+         width = 2100, height = 800, res = 300)
+```
+
+## First imported cases into Europe, time and location
+```{r first-eu-destsrc, fig.height=4}
+# 2.2b First introduction into Europe, distribution density destination
+facet_names <- demes$deme
+names(facet_names) <- demes$deme
+facet_names["OtherEuropean"] <- "Other EU"
+
+first_eu1d <- ggplot(first_df1, aes(x = date, y = ..count../(args$n))) +
+  geom_density_line(
+    data = select(first_df1, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 2) + 
+  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE,
+             labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First imported case into Europe, \ndestination") +
+  theme(legend.position = "none",
+        axis.title.x=element_blank())
+
+# 2.2b First 10 introduction into Europe, distribution density destination
+m = 10
+first_df10 <- df_traj %>%
+  # 1. Prepare data
+  filter(var == "IM", !is.na(partner),  value != 0) %>%
+  group_by(traj) %>%
+  arrange(date, partner) %>%
+  slice(1:m)
+
+first_eu10d <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_df10, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 2) + 
+  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into Europe, \ndestination") +
+  theme(legend.position = "none",
+        axis.title.x=element_blank())
+
+first_eu10s <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_df10, -partner), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 2) + 
+  geom_density_line(aes(fill = partner), bw = 2, color = "transparent", alpha = 0.8) +
+  facet_grid(rows = vars(partner), switch = "y", drop = FALSE, labeller = labeller(partner = facet_names)) +
+  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
+  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  ylab("probability density by source country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into Europe, \nsource") +
+  theme(legend.position = "none",
+        axis.title.x=element_blank())
+
+first_eu_ds <- ggarrange(first_eu1d, first_eu10d, first_eu10s, labels = chartr("123456789", "ABCDEFGHI", 1:3), font.label = list(size = 9), nrow = 1, ncol = 3, hjust = -6, vjust = 2)
+first_eu_ds
+ggexport(first_eu_ds,
+         filename = paste0(args$output_figure, "04.png"),
+         width = 2500, height = 1500, res = 300)
+
+```
+
+## First introductions into each country, source
+```{r first-bycountry, fig.height=4}
+# 2.2b First introduction into each country, distribution density destination
+m = 1
+first_country <- df_traj %>%
+  mutate(partner = ifelse(var == "O", "China", partner)) %>%
+  # 1. Prepare data
+  filter(var == "IM" & deme != "China"| var == "O" & deme == "China", !is.na(partner), value != 0) %>%
+  group_by(traj, deme) %>%
+  arrange(date, partner) %>%
+  slice(1:m)
+
+first_countryplot <- ggplot(first_country, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_country, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 3
+  ) + 
+  geom_density_line(aes(fill = partner), color = "transparent", bw = 3, alpha = 0.6) +
+  geom_vline(data = first_ecdc, aes(xintercept = date, linetype = "first confirmed case to ECDC")) +
+  geom_vline(data = first_country %>% group_by(deme) %>% summarise(median = median(date)), aes(xintercept = median, linetype = "first case median infered date")) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "source country", values = dcolors, na.value = "grey") +
+  scale_x_date(#limits = c(ymd("2020-01-01"), ymd("2020-02-15")), 
+    date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  scale_linetype_manual(name = "", values = c(1,2)) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First SARS-CoV-2 case in the country") +
+  theme(#legend.position = "none",
+        legend.key.size = unit(0.5, "cm"),
+        axis.title.x=element_blank()) +
+  guides(fill = guide_legend(nrow = 1))
+
+# 2.2b First 10 introductions into each country, distribution density destination
+m = 10
+first_10country <- df_traj %>%
+  # 1. Prepare data
+  filter(var == "IM", !is.na(partner),  value != 0) %>%
+  group_by(traj, deme) %>%
+  arrange(date, partner) %>%
+  slice(1:m)
+
+first_10countryplot <- ggplot(first_10country, aes(x = date, y = ..count../(m*args$n))) +
+  geom_density_line(
+    data = select(first_10country, -deme), aes(fill = NA),
+    color = "transparent", alpha = 0.8, bw = 3
+  ) + 
+  geom_density_line(aes(fill = partner), color = "transparent", bw = 3, alpha = 0.6) +
+  geom_vline(data = first_ecdc, aes(xintercept = date, linetype = "first confirmed case to ECDC")) +
+  #geom_vline(data = first_10country %>% group_by(deme) %>% summarise(median = median(date)), aes(xintercept = median, linetype = "first 10 imported cases median date")) +
+  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
+  scale_fill_manual(name = "source country", values = dcolors, na.value = "grey") +
+  scale_x_date(#limits = c(ymd("2020-01-01"), ymd("2020-02-15")), 
+    date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
+  scale_linetype_manual(name = "", values = 2) +
+  ylab("probability density by destination country") +
+  xlab("date") +
+  labs(title = "First 10 imported cases into the country") +
+  theme(#legend.position = "none",
+        legend.key.size = unit(0.5, "cm"),
+        axis.title.x=element_blank()) +
+  guides(fill = guide_legend(nrow = 1))
+
+first_countryds <- ggarrange(first_countryplot, first_10countryplot, labels = chartr("123456789", "ABCDEFGHI", 1:2), font.label = list(size = 9), nrow = 1, ncol = 2, hjust = -6, vjust = 2, common.legend = TRUE, legend = "bottom")
+first_countryds
+
+ggexport(first_countryds,
+         filename = paste0(args$output_figure, "05.png"),
+         width = 2500, height = 1700, res = 300)
+```
+
+```{r first-times, fig.height=4}
+# 2. First introduction distribution density by deme and time of first official reported case to ECDC
+first_times <- df_traj_cases %>%
+  # 1. Prepare data
+  filter(var == "IM" | var == "O", value != 0) %>%
+  group_by(deme, traj) %>%
+  arrange(date) %>%
+  slice(1) %>%
+  ungroup %>%
+  bind_rows(df_traj_cases %>%
+              filter(var == "IM", value != 0) %>%
+              group_by(traj) %>%
+              arrange(date) %>%
+              slice(1)  %>%
+              mutate(deme = "Europe")) %>%
+  mutate(deme = fct_rev(deme)) %>%
+  # 2. Plot
+  ggplot() +
+  geom_density_ridges(aes(x = date, y = deme, fill = deme), color = "grey50", quantile_lines = TRUE, quantiles = c(0.025, 0.5, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) +
+  #stat_density_ridges(, alpha = 0.7)
+  #geom_histogram(aes(x = date, fill = deme), bandwidth=5, alpha = 0.7, size = 0.2) +
+  #geom_vline(aes(xintercept = min_date, color = dest), linetype = 2) +
+  geom_point(data = first_ecdc, 
+             aes(date, deme, fill = deme), color = "grey50", shape = 23, size = 3) +
+  scale_fill_manual(values = c(dcolors, "Europe" = "grey")) +
+  scale_color_manual(values = c(dcolors, "Europe" = "grey")) +
+  scale_x_date(limits = c(ymd("2019-10-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %Y", expand = c(0,0)) +
+  scale_y_discrete(expand = c(0.05,0), "probability") +
+  #theme_ridges() +
+  theme(legend.position = "none",
+        axis.title.x=element_blank(),
+        panel.grid.major.x =  element_line(colour = "grey80", size = 0.3, linetype = 2),
+        panel.grid.major.y =  element_line(colour = "grey80", size = 0.3, linetype = 1),
+        #panel.grid.minor.x =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
+        #axis.title.y=element_blank(),
+        panel.background = element_rect(fill = "white", colour = "white"))
+
+first_times
+
+```
+
+## Events timing
+
+```{r events-timing, fig.height=4}
+df_eventsttiming <- df_traj_cases %>%
+  filter(var %in% c("O", "B", "D", "IM", "OM"), value != 0, is.na(partner)) %>%
+  group_by(traj, var, deme, partner) %>%
+  mutate(var = ifelse(var == "O", "B", var)) %>%
+  arrange(date) %>%
+  slice(1) #%>%
+# ungroup(traj) %>%
+# summarise(m = median(date),
+#           l = quantile(date, 0.025, type = 1),
+#           h = quantile(date, 0.925, type = 1))
+
+events_timingplot <- ggplot(df_eventsttiming) +
+  geom_density_ridges(aes(x = date, y = factor(var, level =  c("OM", "IM", "D", "B")), fill = var), color = "transparent", quantile_lines = TRUE, quantiles = c(0.025, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) +
+  geom_vline(data = first_ecdc, aes(xintercept = date, linetype = "first confirmed case to ECDC")) +
+  scale_linetype_manual(values = 2, name = "") +
+  facet_wrap(~deme) +
+  scale_fill_viridis_d(guide = FALSE) +
+  scale_y_discrete(breaks = c("B", "D", "IM", "OM"), labels = c("first local case", "first death/recovered/isolated", "first imported case", "first exported case")) + 
+  ylab("") +
+  xlab("")
+
+events_timingplot
+ggexport(events_timingplot,
+         filename = paste0(args$output_figure, "13.png"),
+         width = 2000, height = 1200, res = 300)
+
+```
+
 ```{r traj-processing}
@@
 }
-```
-
-## Origin of SARS-CoV-2 in Europe
-
-```{r first-eu, fig.height=4}
-first_df1 <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0) %>%
-  group_by(traj) %>%
-  arrange(date, partner) %>%
-  slice(1)
-
-first_ecdc <- case_data %>% 
-  filter(source == "ecdc", total_confirmed != 0) %>% 
-  group_by(deme) %>% 
-  arrange(date) %>% 
-  slice(1)
-
-first_eui <- first_df1 %>% 
-  ggplot(aes(date)) +
-  geom_density(aes(y = ..count..), fill = "#21908CFF", alpha = .5, col = NA) + 
-  geom_histogram(binwidth = 1, alpha = .5, color = "grey60") +
-  geom_vline(aes(xintercept = median(first_df1$date)), linetype = 1) +
-  geom_vline(aes(xintercept = min((first_ecdc %>% filter(deme != "China"))$date)), linetype = 2) +
-  ylab("epidemic trajectory count") +
-  scale_x_date(limits = c(ymd("2019-12-15"), ymd("2020-02-01")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  theme(axis.title.x = element_blank())
-
-first_euii <- first_df1 %>% 
-  ggplot(aes(date)) +
-  geom_density(fill = "#21908CFF", alpha = .5, col = NA) + 
-  geom_histogram(binwidth = 1, alpha = .5, color = "grey60", aes( y = ..density..)) +
-  geom_vline(aes(xintercept = median(first_df1$date)), linetype = 1) +
-  geom_vline(aes(xintercept = min((first_ecdc %>% filter(deme != "China"))$date)), linetype = 2) +
-  ylab("probability density") +
-  scale_y_continuous(position = "right") +
-  scale_x_date(limits = c(ymd("2019-12-15"), ymd("2020-02-01")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  theme(axis.title.x = element_blank())
-
-ggexport(first_eui,
-         filename = paste0(args$output_figure, "03i.png"),
-         width = 2100, height = 600, res = 300)
-ggexport(first_euii,
-         filename = paste0(args$output_figure, "03ii.png"),
-         width = 2100, height = 600, res = 300)
-```
-
-## First imported cases into Europe, time and location
-```{r first-eu-destsrc, fig.height=4}
-# 2.2b First introduction into Europe, distribution density destination
-facet_names <- demes$deme
-names(facet_names) <- demes$deme
-facet_names["OtherEuropean"] <- "Other EU"
-
-first_eu1d <- ggplot(first_df1, aes(x = date, y = ..count../(args$n))) +
-  geom_density_line(
-    data = select(first_df1, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2) + 
-  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_grid(rows = vars(deme), switch = "y", drop = FALSE,
-             labeller = labeller(deme = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by destination country") +
-  xlab("date") +
-  labs(title = "First imported case into Europe, \ndestination") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank())
-
-# 2.2b First 10 introduction into Europe, distribution density destination
-m = 10
-first_df10 <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0) %>%
-  group_by(traj) %>%
-  arrange(date, partner) %>%
-  slice(1:m)
-
-first_eu10d <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_df10, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2) + 
-  geom_density_line(aes(fill = deme), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by destination country") +
-  xlab("date") +
-  labs(title = "First 10 imported cases into Europe, \ndestination") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank())
-
-first_eu10s <- ggplot(first_df10, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_df10, -partner), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 2) + 
-  geom_density_line(aes(fill = partner), bw = 2, color = "transparent", alpha = 0.8) +
-  facet_grid(rows = vars(partner), switch = "y", drop = FALSE, labeller = labeller(partner = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(limits = c(ymd("2020-01-01"), ymd("2020-02-15")), date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by source country") +
-  xlab("date") +
-  labs(title = "First 10 imported cases into Europe, \nsource") +
-  theme(legend.position = "none",
-        axis.title.x=element_blank())
-
-first_eu_ds <- ggarrange(first_eu1d, first_eu10d, first_eu10s, labels = chartr("123456789", "ABCDEFGHI", 1:3), font.label = list(size = 9), nrow = 1, ncol = 3, hjust = -6, vjust = 2)
-first_eu_ds
-ggexport(first_eu_ds,
-         filename = paste0(args$output_figure, "04.png"),
-         width = 2500, height = 1500, res = 300)
-
-```
-
-## First introductions into each country, source
-```{r first-bycountry, fig.height=4}
-# 2.2b First introduction into each country, distribution density destination
-m = 1
-first_country <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0) %>%
-  group_by(traj, deme) %>%
-  arrange(date, partner) %>%
-  slice(1:m)
-
-first_countryplot <- ggplot(first_country, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_country, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 3
-  ) + 
-  geom_density_line(aes(fill = partner), color = "transparent", bw = 3, alpha = 0.6) +
-  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(#limits = c(ymd("2020-01-01"), ymd("2020-02-15")), 
-    date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by destination country") +
-  xlab("date") +
-  labs(title = "First imported case into the country, \nsource") +
-  theme(#legend.position = "none",
-        legend.key.size = unit(0.5, "cm"),
-        axis.title.x=element_blank()) +
-  guides(fill = guide_legend(nrow = 1))
-
-# 2.2b First 10 introductions into each country, distribution density destination
-m = 10
-first_10country <- df_traj %>%
-  # 1. Prepare data
-  filter(var == "IM", !is.na(partner),  value != 0) %>%
-  group_by(traj, deme) %>%
-  arrange(date, partner) %>%
-  slice(1:m)
-
-first_10countryplot <- ggplot(first_10country, aes(x = date, y = ..count../(m*args$n))) +
-  geom_density_line(
-    data = select(first_10country, -deme), aes(fill = NA),
-    color = "transparent", alpha = 0.8, bw = 3
-  ) + 
-  geom_density_line(aes(fill = partner), color = "transparent", bw = 3, alpha = 0.6) +
-  facet_grid(rows = vars(deme), switch = "y", drop = FALSE, labeller = labeller(deme = facet_names)) +
-  scale_fill_manual(name = "", values = dcolors, na.value = "grey") +
-  scale_x_date(#limits = c(ymd("2020-01-01"), ymd("2020-02-15")), 
-    date_breaks = "1 week", date_labels = "%b %d, %Y", expand = c(0,0)) +
-  ylab("probability density by destination country") +
-  xlab("date") +
-  labs(title = "First 10 imported cases into the country, \nsource") +
-  theme(#legend.position = "none",
-        legend.key.size = unit(0.5, "cm"),
-        axis.title.x=element_blank()) +
-  guides(fill = guide_legend(nrow = 1))
-
-first_countryds <- ggarrange(first_countryplot, first_10countryplot, labels = chartr("123456789", "ABCDEFGHI", 1:2), font.label = list(size = 9), nrow = 1, ncol = 2, hjust = -6, vjust = 2, common.legend = TRUE, legend = "bottom")
-first_countryds
-
-ggexport(first_countryds,
-         filename = paste0(args$output_figure, "05.png"),
-         width = 2000, height = 1200, res = 300)
-```
-
-```{r first-times, fig.height=4}
-# 2. First introduction distribution density by deme and time of first official reported case to ECDC
-first_times <- df_traj_cases %>%
-  # 1. Prepare data
-  filter(var == "IM" | var == "O", value != 0) %>%
-  group_by(deme, traj) %>%
-  arrange(date) %>%
-  slice(1) %>%
-  ungroup %>%
-  bind_rows(df_traj_cases %>%
-              filter(var == "IM", value != 0) %>%
-              group_by(traj) %>%
-              arrange(date) %>%
-              slice(1)  %>%
-              mutate(deme = "Europe")) %>%
-  mutate(deme = fct_rev(deme)) %>%
-  # 2. Plot
-  ggplot() +
-  geom_density_ridges(aes(x = date, y = deme, fill = deme), color = "grey50", quantile_lines = TRUE, quantiles = c(0.025, 0.5, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) +
-  #stat_density_ridges(, alpha = 0.7)
-  #geom_histogram(aes(x = date, fill = deme), bandwidth=5, alpha = 0.7, size = 0.2) +
-  #geom_vline(aes(xintercept = min_date, color = dest), linetype = 2) +
-  geom_point(data = first_ecdc, 
-             aes(date, deme, fill = deme), color = "grey50", shape = 23, size = 3) +
-  scale_fill_manual(values = c(dcolors, "Europe" = "grey")) +
-  scale_color_manual(values = c(dcolors, "Europe" = "grey")) +
-  scale_x_date(limits = c(ymd("2019-10-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %Y", expand = c(0,0)) +
-  scale_y_discrete(expand = c(0.05,0), "probability") +
-  #theme_ridges() +
-  theme(legend.position = "none",
-        axis.title.x=element_blank(),
-        panel.grid.major.x =  element_line(colour = "grey80", size = 0.3, linetype = 2),
-        panel.grid.major.y =  element_line(colour = "grey80", size = 0.3, linetype = 1),
-        #panel.grid.minor.x =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
-        #axis.title.y=element_blank(),
-        panel.background = element_rect(fill = "white", colour = "white"))
-
-first_times
-
-ggexport(first_times,
-         filename = paste0(args$output_figure, "03.png"),
-         width = 2000, height = 800, res = 300)
 ```
@@
 ```
-
-## Events timing
-
-```{r events-timing, fig.height=4}
-df_eventsttiming <- df_traj_cases %>%
-  filter(var %in% c("O", "B", "D", "IM", "OM"), value != 0, is.na(partner)) %>%
-  group_by(traj, var, deme, partner) %>%
-  mutate(var = ifelse(var == "O", "B", var)) %>%
-  arrange(date) %>%
-  slice(1) #%>%
-# ungroup(traj) %>%
-# summarise(m = median(date),
-#           l = quantile(date, 0.025, type = 1),
-#           h = quantile(date, 0.925, type = 1))
-
-events_timingplot <- ggplot(df_eventsttiming) +
-  geom_density_ridges(aes(x = date, y = factor(var, level =  c("OM", "IM", "D", "B")), fill = var), color = "transparent", quantile_lines = TRUE, quantiles = c(0.025, 0.975), bandwidth=2, alpha = 0.7, size = 0.2) +
-  geom_vline(data = first_ecdc, aes(xintercept = date, linetype = "first confirmed case to ECDC")) +
-  scale_linetype_manual(values = 3, name = "") +
-  facet_wrap(~deme) +
-  scale_fill_viridis_d(guide = FALSE) +
-  scale_y_discrete(breaks = c("B", "D", "IM", "OM"), labels = c("first local case", "first uninfectious", "first imported case", "first exported case")) + 
-  ylab("") +
-  xlab("") +
-  theme(legend.position = "top")
-
-events_timingplot
-ggexport(events_timingplot,
-         filename = paste0(args$output_figure, "13.png"),
-         width = 2000, height = 1200, res = 300)
-
-```

### 2021-03-24 scripts/comparison_nParticles.R

@@ -1,0 +1,49 @@
+# Comparison Number of particles
+
+library(tidyverse)
+source("./scripts/trajProcessing.R")
+source("./scripts/utils.R")
+
+particles <-  c(300, 1000, 3000, 10000)
+file_path <- "/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/210205_europe10."
+file_list <- paste0(file_path, particles, ".TL.traj")
+n = 1000
+
+df_trajs <- lapply(file_list, function(file) {
+  df <- loadTrajectories(file, burninFrac = 0, subsample = n)
+  df_traj <- processEvents(df$events, demes, "China", mrs)
+  return(df_traj)
+})
+names(df_trajs) <- particles
+
+df <- bind_rows(df_trajs, .id = 'particles') %>%
+  filter(var %in% c("active_pop", "B", "D", "IM"),
+    date == max(date), is.na(partner)) %>%
+  group_by(particles, var, traj) %>%
+  summarise(cumvalue = sum(cumvalue), .groups = "drop_last") %>%
+  summarise(m = median(cumvalue),
+            l = quantile(cumvalue, 0.25),
+            h = quantile(cumvalue, 0.75),
+            l95 = HDInterval::hdi(cumvalue)[[1]],
+            h95 = HDInterval::hdi(cumvalue)[[2]],)
+
+facet_names <- c("Active infections", "No longer infectious cases", 
+                 "Imported cases", "Local transmissions")
+names(facet_names) <- unique(df$var)
+
+plot_particles <- ggplot(df) +
+  geom_point(aes(x = as.numeric(particles), y = m)) +
+  geom_line(aes(x = as.numeric(particles), y = m)) +
+  geom_ribbon(aes(x = as.numeric(particles), ymin = l95, ymax = h95), alpha = 0.5) +
+  facet_wrap(~var, scales = "free", nrow = 1, labeller = labeller(var = facet_names))+
+  ylab("number of events") +
+  xlab("number of particles") +
+  scale_x_continuous(breaks = c(300, 1000, 3000, 10000)) +
+  theme(axis.text.x = element_text(angle = 45))
+  #scale_y_log10(labels = trans_format("log10", math_format(10^.x))) 
+
+ggexport(plot_particles,
+         filename = "reports/mt-thesis/figures/particles_comparison.png",
+         width = 2500, height = 1000, res = 300)
+
+

### 2021-03-24 reports/mt-thesis/results.tex
wip results mt

@@
 \textbf{The origin of SARS-CoV-2 in Europe} 
-While the first case in Europe was reported to ECDC by France authorities on January 25, 2020, the first European wave is suspected to have started earlier but been undetected, mistaken for other common repiratory diseases in winter.
-In our model, we assume the origin of SARS-CoV-2 in Europe to be an imported case from China into any of the European countries.
-We analyze the time and destination of the first imported case in the infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI), Figure \ref{fig:firstEU} \todo{with final results}.
+While the first case in Europe was reported to ECDC by France authorities on January 25, the first European wave is suspected to have started earlier but been undetected, mistaken for other common repiratory diseases in winter.
+In our model, we assume the origin of SARS-CoV-2 in Europe to be an imported case from China to any of the European countries.
+We can select a single event of interest from the inferred epidemic trajectories, and summarise its time and location.
+To study the origin of SARS-CoV-2 in Europe, we analyze the time and destination of the first imported case and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI), Figure \ref{fig:firstEU} \todo{with final results}.
 The actual reported date to ECDC is in the tail of the $95\%$ credible interval.
@@
     \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj01.png}
+    \caption{\textbf{First case of SARS-CoV-2 in Europe.} In the histogram, we count how many of the inferred epidemic trajectories started the European epidemic each day.
+From these counts, we estimate the probability density of the origin in Europe (in grey).
+The median date is drawn with the solid vertical line, and the date of the first reported European case to ECDC with a dashed vertical line.}
     \label{fig:firstEU}
-    \todo[inline]{ticks for every month}
 \end{figure}
 
-Likewise, we analyze the first ten imported cases into Europe and obtain the probability distribution of the source and destination country, Figure \ref{fig:firstEUcountry} B and C.
+Likewise, we analyze the first ten imported cases to Europe to estimate the main source of exported SARS-CoV-2 cases early on and their destination.
+China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China, Figure \ref{fig:firstEUcountry} B.
 The role of Italy as the receptor of these first cases is highlighted even more looking at several cases.
 Early introductions in Spain have a very low probabilty, looking like Spain was not a key player in the origin of SARS-CoV-2 in Europe.
-China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China.
-\todo{with final results}
+\todo{with final results}  Figure \ref{fig:firstEUcountry} C.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj04.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj02.png}
+    \caption{\textbf{First imported cases of SARS-CoV-2 to Europe.} \textbf{A} We estimate the probability distribution of the time for the first imported case to Europe (in grey).
+Each row highlights the probability of this first case going to the indicated country (in color).
+\textbf{B} In a similar way, we considered the time of the ten first imported cases into any of the European country, and plot the time probability distribution (in grey).
+In the rows, we plot the probability of each country being the destination country of any of the first 10 imported cases to Europe (in color).
+\textbf{C} Same probability distribution of the time for the 10 first imported cases to any European country (in grey) than in B, but in each row we highlight in color the
+    probability of these cases coming from a specific country.}
     \label{fig:firstEUcountry}
@@
 We estimate that the epidemic in any of the European countries was seeded by an imported case from China as seen in Figurem with some small prrobability for Germany and Italy of having orriginatetdd from a case from each other, and mostly half of the rpobability in the case of Spain from the epidemdic started with a case imported from one of the others Euroopean countries.
-Same than before, we can look into the first 10 imported cases into each country and now we see that the transmission between european countries adquire much moer importance,  so once SARS-CoV-2 entered Europe, the transmission between european countrries were soon more important than the cases coming from China.
+Same than before, we can look to the first 10 imported cases to each country and now we see that the transmission between european countries adquire much moer importance,  so once SARS-CoV-2 entered Europe, the transmission between european countrries were soon more important than the cases coming from China.
 This could relate also to the decrease in tthe trravel to and from China seen in figure.
@@
     \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj05.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
-    \label{fig:firstEUcountry}
-    \todo[inline]{ticks for every month}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
+    \caption{\textbf{Country-level origin of SARS-CoV-2 epidemic.} For China, we consider the origin of the epidemic as the first local case while for the European countries we select the first imported case into the country.
+\textbf{A} We estimate the probability distribution of the date for the first case in each of the locations included in the analysis (in grey).
+We highlight the conditional probability for the source of the first case in each country (in color).
+\textbf{B} The ten first imported cases into the country are selected, and we estimate the probability distribution of their occurrence.
+As in A, we show the conditional probability of the source country.}
+    \label{fig:firstcountry}
 \end{figure}
@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj06.png}
-    \caption{Countries ordered by the time of its first introduction, i.e.
-first case in the country.
-Each row is the order for one of the subsampled epidemic trajectories and each column represents the position relative to the other countries first introduction, e.g.
-in first position for all epidemic trajectories is China since it is the origin of the epidemic.}
-    \label{fig:first_imig}
-\end{figure}
-
 Along the same lines, we could ask if this order of countries is mantained when instead of looking at the first case in the country we look into at first case exported (migration) from that country to other European country.
 In Figure \ref{fig:first_omig} we observe a similar pattern to the first cases order, with Germany, France and Italy being the countries in the first positions in more than half of the inferred trajectories and never in the last position, and Spain as the country in last position in almost every epidemic trajectory.\\
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj07.png}
-    \caption{Order of countries by the time of its first migration out of the country, i.e.
-first exported case to other country.
-Each row is the order for one of the subsampled epidemic trajectories and each column represents the position relative to the other countries first migration, e.g.
-in first position for all epidemic trajectories is China since it was the first country with exported cases of SARS-CoV-2 to other regions.}
-    \label{fig:first_omig}
-    \todo[inline]{I don't like much these "order" plots but maybe they could be useful to detect interesting patterns?}
-\end{figure}
 
@@
     \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
-    \label{fig:first}
-    \todo[inline]{ticks for every month}
+    \includegraphics[width=\textwidth]{210205_europe10_figtraj04.png}
+    \caption{\textbf{Timing of key first epidemic events.} For each country, we select the key epidemic events: first local transmission case, first death, recovery or isolated case (same event category for the birth-death model), first imported and first exported case.
+We summarise the timing of these events from the inferred trajectories and obtain a probability distribution for each of them.
+The first detection of SARS-CoV-2 in each country reported to ECDC is shown with a dashed line.}
+    \label{fig:first_events}
 \end{figure}
@@
 We see some differences among countries, France and Germany had in all inferred epidemic trajectories ongoing within-region transmission when first cases werre reported, while x\% of epidemic trajectories did not have had a within region transmission case when the first case in Spain was reported.\\
-
-\begin{figure}[p]
-    \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj07.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
-    \label{fig:first}
-    \todo[inline]{ticks for every month}
-\end{figure}
 
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{210205_europe10_figtraj01.png}
-    \caption{Inferred population size summary statistics for each deme over time.
-The line represents the median population trajectory and the interval is the 95\%  credible interval in log scale from a random subsampled set of inferred epidemic trajectories.}
+    \includegraphics[width=0.8\textwidth]{210205_europe10_figtraj05.png}
+    \caption{\textbf{Active SARS-CoV-2 infections by country from the origin of the pandemic to March 8, 2020}.
+We compute the median number and 95\%  credible interval of active SARS-CoV-2 infections for each analysis day by country.
+We plot the median value for the country with a coloured line and the 95\% credible interval of active infections with a shaded area.}
     \label{fig:gribbon}
-    \todo[inline]{Specify that the random subset are 500 trajectories in all figure legends}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj02.png}
-    \caption{Inferred epidemic trajectories over time.
-A random subsampled set of 500 trajectories is plotted.
-In each subplot, the trajectories (solid lines) are compared with the ECDC cumulative case count data (dashed line) in log scale.}
+    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj06.png}
+    \caption{\textbf{SARS-CoV-2 inferred cases and ECDC confirmed cases by country over time.} To compare with the official ECDC confirmed cases in a analogous time frame, we consider the no longer infectious cases inferred by the model that appear after a mean period of ten days post-infection.
+Each line represent the cumulative number of inferred cases in log scale for each of the 500 analysed epidemic trajectories with time in the x-axis.
+The ECDC confirmed case counts are shown with dashed lines.}
+    \todo[figure]{Why is till March 20?}
     \label{fig:trajs}
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj11.png}
-    \caption{5-days reporting rate, calculated as the cumulative number of ECDC reported cases by the median number of cumulative inferred cases in intervals of 5 days until 8th of March. }
+    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj07.png}
+    \caption{\textbf{Weekly reporting proportion of new SARS-CoV-2 cases.} We compute the number of new confirmed cases by week reported to ECDC and divide it by the median number of inferred cases (no longer infectious cases) in the same week.
+For each week, the point represents the median reporting proportion, and the shaded area the 95\% credible interval.}
     \label{fig:reported}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \includegraphics[width=\textwidth]{210205_europe10_figtraj12.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
-    \label{fig:first}
-    \todo[inline]{ticks for every month}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{210205_europe10_figtraj08.png}
-    \caption{Median and 95\% credible interval for the cumulative number of events (within-region transmissions and migrations to the country (incoming) and from the country (outgoing) over time.}
+    \caption{\textbf{Evolution of the ratio of local transmissions by imported cases for the SARS-CoV-2 early epidemic in China, France, Germany, Italy, Spain and other European countries}.
+We compute for each day the ratio of new local infections by new imported cases for each of the analysed epidemic trajectories.
+Each line depicts the evolution of this ratio in one epidemic trajectory, coloured by the total number of estimated infections by the model.
+The black line is the median value for the ratio local/imported cases.}
     \label{fig:events}
@@
   \end{minipage}
-  \caption{Migration flux among demes over the three periods defined in the analysis.}
+  \caption{\textbf{International imported case dynamics by analysis constat migration rate periods.} One chord diagram is drawn for each of the main periods of the analysis defined by the change times of the migration rates in the model.
+Each flux represents the infections exported from one country to another during the full period.
+The median value of infections for each flux is indicated in the plot.}
   \label{fig:migs}
@@
 
-\subsubsection*{GLM predictors}
-\todo[inline]{TODO}
-
-Which channels were the main sources of transmission across national borders.
-
 \subsubsection*{Epidemiological parameters}
@@
 Not sure if it is necessary. But maybe include, briefly, the values of estimate R0, migration and sampling rates?
+Maybe a table in the appendix? Or the plots with the priors.
 
@@
 \todo[inline]{TODO, or into the discussion?? }
+
+\begin{figure}[p]
+    \centering
+    \includegraphics[width=\textwidth]{particles_comparison.png}
+    \caption{First introductions of SARS-CoV-2.
+From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
+\textbf{A} Probability density of the time of first introduction for each deme.
+Each dotted line represents the first date when cases where reported to ECDC by deme color.
+In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
+For the other five demes, the distribution of the time of first migration into the deme is shown.
+\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
+This first case corresponds to the first migration event from China to any of the European countries.
+\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \label{fig:particles}
+    \todo[inline]{ticks for every month}
+\end{figure}

### 2021-03-24 scripts/analyze_trajectories.Rmd
analize traj rmd with tables

@@
 # Parser -----------------------------------------------------------------------
-argsR <- commandArgs(TRUE)
-args_list <- str_split(paste(argsR, collapse=""), pattern="--")
-args_llist <- lapply(args_list, function(x) str_split(x, pattern="="))
-argsDF <- as.data.frame(do.call("rbind", args_llist[[1]]))
-args <- as.list(as.character(argsDF$V2))
-names(args) <- argsDF$V1
-args$burnin <- as.integer(args$burnin)
-args$n <- as.integer(args$n)
-title <- paste("Analysis", gsub("\\.", " ", gsub(".TL.traj", " particles", str_split(args$args, "/")[[1]][3])))
-```
-
----
-subtitle: "`r title`"
----
+# argsR <- commandArgs(TRUE)
+# args_list <- str_split(paste(argsR, collapse=""), pattern="--")
+# args_llist <- lapply(args_list, function(x) str_split(x, pattern="="))
+# argsDF <- as.data.frame(do.call("rbind", args_llist[[1]]))
+# args <- as.list(as.character(argsDF$V2))
+# names(args) <- argsDF$V1
+# args$burnin <- as.integer(args$burnin)
+# args$n <- as.integer(args$n)
+# title <- paste("Analysis", gsub("\\.", " ", gsub(".TL.traj", " particles", str_split(args$args, "/")[[1]][3])))
+# ---
+# subtitle: "`r title`"
+# ---
+```
 
@@
 args$metadata =  "./results/europeC/final_metadata.tsv"
-args$n = 100
+args$n = 10
 demes = args$demes
@@
 # Load demes configuration and sequence metadata -------------------------------
-demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
-                    exclude_country = NA, exclude_division = NA,
-                    min_date = NA, max_date = NA, seq_per_deme = NA) %>%
-  bind_rows(bind_rows(lapply(yaml::yaml.load(string = gsub(",", ", ", gsub(":", ": ", args$demes))),
-                          data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
-  group_by(deme) %>%
-  mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
-  distinct() %>%
-  filter_all(any_vars(!is.na(.))) %>%
-  mutate(deme = str_to_title(deme)) %>%
-  mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme)) %>%
-  arrange(deme)
+# demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
+#                     exclude_country = NA, exclude_division = NA,
+#                     min_date = NA, max_date = NA, seq_per_deme = NA) %>%
+#   bind_rows(bind_rows(lapply(yaml::yaml.load(string = gsub(",", ", ", gsub(":", ": ", args$demes))),
+#                           data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
+#   group_by(deme) %>%
+#   mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
+#   distinct() %>%
+#   filter_all(any_vars(!is.na(.))) %>%
+#   mutate(deme = str_to_title(deme)) %>%
+#   mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme)) %>%
+#   arrange(deme)
 
@@
 # Data wrangling ---------------------------------------------------------------
-df_traj <- processEvents(df$events, demes, "China", mrs)
+df_traj <- processEvents(df$events, demes, "China", MRS)
 ```
@@
 To facilitate visualization and summarise the results, we take a grid time of 1 day and summarise the number of events that day as the sum of the events in the corresponding time interval; and the number of inferred cases as the maximum of the interval.
+
+```{r traj-processing}
+# Add reported case counts
+df_traj_cases <- df_traj %>%
+  mutate(traj = as.character(traj)) %>%
+  bind_rows(case_data %>%
+              mutate(var = source, value = new_confirmed, cumvalue = total_confirmed) %>%
+              select(deme, var, date, value, cumvalue))
+
+# Summarise trajectories, compute mean, median and 95% HPD interval for each variable
+traj_summary <- df_traj %>%
+  group_by(var, deme, partner, date) %>%
+  summarise(l95_value = HDInterval::hdi(value)[[1]],
+            h95_value = HDInterval::hdi(value)[[2]],
+            median_value = median(value),
+            mean_value = mean(value),
+            l95_cumvalue = HDInterval::hdi(cumvalue)[[1]],
+            h95_cumvalue = HDInterval::hdi(cumvalue)[[2]],
+            #l95_cumvalue = quantile(cumvalue, 0.025),
+            #h95_cumvalue = quantile(cumvalue, 0.975),
+            median_cumvalue = median(cumvalue),
+            mean_cumvalue = mean(cumvalue)) %>% ungroup
+
+```
+
+```{r}
+# 1. Total 
+# 1.1 Median and 95% CCI of Total cases on the full time period. Fraction median
+#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases
+
+# Total values March 8, 2020 join with reported cases
+traj_total <- traj_summary %>%
+  filter(date == MRS) %>%
+  select(var, date, deme, partner, matches("cumvalue"))  %>%
+  left_join(case_data %>% filter(source == "ecdc")) %>%
+  select(var, deme, matches("cumvalue"), total_confirmed) %>%
+  mutate(ltimes = l95_cumvalue/total_confirmed,
+         htimes = h95_cumvalue/total_confirmed,
+         mtimes = median_cumvalue/total_confirmed) %>%
+  mutate_if(is.numeric, function(x) round(x,2))
+
+#knitr::kable(t(traj_total %>% filter(var == "D"))[-1,])
+```
 
@@
   arrange(date) %>% 
-  slice(1)
+  slice(1) %>%
+  ungroup()
 
@@
   ggplot(aes(date)) +
-  geom_density(aes(y = ..count..), fill = "grey", alpha = .5, col = NA) + 
+  geom_density(aes(y = ..count..), fill = "grey", alpha = .5, col = NA, bw = 2) + 
   geom_histogram(binwidth = 1, alpha = .5, fill = "#21908CFF", color = "#21908CFF") +
@@
   ggplot(aes(date)) +
-  geom_density(fill = "grey", alpha = .5, col = NA) + 
+  geom_density(fill = "grey", alpha = .5, col = NA, bw = 2) + 
   geom_histogram(binwidth = 1, alpha = .5, fill = "#21908CFF", color = "#21908CFF", aes( y = ..density..)) +
@@
 
+first_eui
+
 ggexport(first_eui,
-         filename = paste0(args$output_figure, "03i.png"),
+         filename = paste0(args$output_figure, "01i.png"),
          width = 2100, height = 800, res = 300)
 ggexport(first_euii,
-         filename = paste0(args$output_figure, "03ii.png"),
+         filename = paste0(args$output_figure, "01ii.png"),
          width = 2100, height = 800, res = 300)
@@
 
-## First imported cases into Europe, time and location
+*Median date:* `r median(first_df1$date)`\
+*95% CCI:* (`r quantile(first_df1$date, 0.025, type = 1)`, `r quantile(first_df1$date, 0.975, type = 1)`)\
+*ECDC first confirmed case:* `r min((first_ecdc %>% filter(deme != "China"))$date)`\
+
+## First imported cases to Europe, time and location
 ```{r first-eu-destsrc, fig.height=4}
-# 2.2b First introduction into Europe, distribution density destination
+# 2.2b First introduction to Europe, distribution density destination
 facet_names <- demes$deme
@@
   xlab("date") +
-  labs(title = "First imported case into Europe, \ndestination") +
+  labs(title = "First imported case to Europe, \ndestination") +
   theme(legend.position = "none",
@@
 
-# 2.2b First 10 introduction into Europe, distribution density destination
+# 2.2b First 10 introduction to Europe, distribution density destination
 m = 10
@@
   xlab("date") +
-  labs(title = "First 10 imported cases into Europe, \ndestination") +
+  labs(title = "First 10 imported cases to Europe, \ndestination") +
   theme(legend.position = "none",
@@
   xlab("date") +
-  labs(title = "First 10 imported cases into Europe, \nsource") +
+  labs(title = "First 10 imported cases to Europe, \nsource") +
   theme(legend.position = "none",
@@
 ggexport(first_eu_ds,
-         filename = paste0(args$output_figure, "04.png"),
+         filename = paste0(args$output_figure, "02.png"),
          width = 2500, height = 1500, res = 300)
@@
 
-## First introductions into each country, source
+## First imported cases to Europe, time and location
+```{r first-eu-table}
+knitr::kable(first_df1 %>%
+               ungroup %>%
+               count(deme, name = "n_1d") %>%
+               mutate(p_1d = n_1d/args$n) %>%
+               left_join(first_df10 %>%
+                           ungroup %>%
+                           count(deme,  name = "n_10d") %>%
+                           mutate(p_10d = n_10d/(10*args$n)) %>%
+                           left_join(first_df10 %>%
+                                       ungroup %>%
+                                       select(-deme) %>% rename(deme = partner) %>%
+                                       count(deme,  name = "n_10s") %>%
+                                       mutate(p_10s = n_10s/(10*args$n)))))
+```
+
+## First introductions to each country, source
 ```{r first-bycountry, fig.height=4}
@@
   xlab("date") +
-  labs(title = "First 10 imported cases into the country") +
+  labs(title = "First 10 imported cases to the country") +
   theme(#legend.position = "none",
@@
 ggexport(first_countryds,
-         filename = paste0(args$output_figure, "05.png"),
+         filename = paste0(args$output_figure, "03.png"),
          width = 2500, height = 1700, res = 300)
 ```
+
+## First introductions to each country, source
+
+::: {.columns}
+
+:::: {.column width="50%"}
+\small
+```{r first-country-table1}
+knitr::kable(first_country %>%
+               group_by(deme) %>%
+               summarise(m = median(date),
+                         l = quantile(date, 0.025, type = 1),
+                         h = quantile(date, 0.925, type = 1)))
+```
+::::
+
+:::: {.column width="50%"}
+\small
+```{r first-country-table2}
+
+knitr::kable(first_country %>%
+               ungroup %>%
+               group_by(deme) %>%
+               count(partner, name = "n_1s") %>%
+               mutate(p_1s = n_1s/args$n))
+```
+
+::::
+
+:::
+
+## First introductions to each country, source
+
+::: {.columns}
+
+:::: {.column width="50%"}
+\small
+```{r first-country-table3}
+knitr::kable((first_10country %>%
+                ungroup %>%
+                group_by(deme) %>%
+                count(partner, name = "n_10s") %>%
+                mutate(p_10s = n_10s/(10 * args$n)))[1:15,])
+```
+::::
+
+:::: {.column width="50%"}
+\small
+```{r first-country-table4}
+knitr::kable((first_10country %>%
+                ungroup %>%
+                group_by(deme) %>%
+                count(partner, name = "n_10s") %>%
+                mutate(p_10s = n_10s/(10 * args$n)))[16:30,])
+```
+
+::::
+
+:::
 
@@
   scale_y_discrete(breaks = c("B", "D", "IM", "OM"), labels = c("first local case", "first death/recovered/isolated", "first imported case", "first exported case")) + 
+  scale_x_date(#limits = c(ymd("2020-01-01"), ymd("2020-02-15")), 
+    date_breaks = "1 month", date_labels = "%b, %Y", expand = c(0,0)) +
   ylab("") +
@@
 ggexport(events_timingplot,
-         filename = paste0(args$output_figure, "13.png"),
+         filename = paste0(args$output_figure, "04.png"),
          width = 2000, height = 1200, res = 300)
@@
 
-```{r traj-processing}
-# Add reported case counts
-df_traj_cases <- df_traj %>%
-  mutate(traj = as.character(traj)) %>%
-  bind_rows(case_data %>%
-              mutate(var = source, value = new_confirmed, cumvalue = total_confirmed) %>%
-              select(deme, var, date, value, cumvalue))
-
-# Summarise trajectories, compute mean, median and 95% HPD interval for each variable
-traj_summary <- df_traj %>%
-  group_by(var, deme, partner, date) %>%
-  summarise(l95_value = HDInterval::hdi(value)[[1]],
-            h95_value = HDInterval::hdi(value)[[2]],
-            median_value = median(value),
-            mean_value = mean(value),
-            l95_cumvalue = HDInterval::hdi(cumvalue)[[1]],
-            h95_cumvalue = HDInterval::hdi(cumvalue)[[2]],
-            #l95_cumvalue = quantile(cumvalue, 0.025),
-            #h95_cumvalue = quantile(cumvalue, 0.975),
-            median_cumvalue = median(cumvalue),
-            mean_cumvalue = mean(cumvalue)) %>% ungroup
-
-```
-
-```{r}
-# 1. Total 
-# 1.1 Median and 95% CCI of Total cases on the full time period. Fraction median
-#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases
-
-# Total values March 8, 2020 join with reported cases
-traj_total <- traj_summary %>%
-  filter(date == MRS) %>%
-  select(var, date, deme, partner, matches("cumvalue"))  %>%
-  left_join(case_data %>% filter(source == "ecdc")) %>%
-  select(var, deme, matches("cumvalue"), total_confirmed) %>%
-  mutate(ltimes = l95_cumvalue/total_confirmed,
-         htimes = h95_cumvalue/total_confirmed,
-         mtimes = median_cumvalue/total_confirmed) %>%
-  mutate_if(is.numeric, function(x) round(x,2))
-
-#knitr::kable(t(traj_total %>% filter(var == "D"))[-1,])
-```
+## Events timing
+
+::: {.columns}
+
+:::: {.column width="50%"}
+\small
+```{r events-timing-table1}
+knitr::kable((df_eventsttiming %>%
+                group_by(deme, var) %>%
+                summarise(m = median(date),
+                          l = quantile(date, 0.025, type = 1),
+                          h = quantile(date, 0.925, type = 1)))[1:12,])
+```
+::::
+
+:::: {.column width="50%"}
+\small
+```{r events-timing-table2}
+knitr::kable((df_eventsttiming %>%
+                group_by(deme, var) %>%
+                summarise(m = median(date),
+                          l = quantile(date, 0.025, type = 1),
+                          h = quantile(date, 0.925, type = 1)))[12:24,])
+```
+
+::::
+
+:::
 
 ## Summarised epidemic trajectories
+
 ```{r trajs ribbon, fig.height = 4}
@@
   scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
-  scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b %d") +
+  scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b, %Y", expand = c(0,0)) +
   scale_fill_manual(name = "", values = dcolors) +
@@
 ggexport(gribbon,
-         filename = paste0(args$output_figure, "01.png"),
+         filename = paste0(args$output_figure, "05.png"),
          width = 1500, height = 1000, res = 300)
@@
                                         " total reported cases"))) +
-    scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month",
-                 date_labels = "%b %y") +
+    scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b, %Y", expand = c(0,0)) +
     scale_y_log10(limits= c(1,10^7.5), expand = c(0, 0),labels = trans_format("log10", math_format(10^.x))) +
@@
 ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1:length(trajs)), nrow = 3, ncol = 2),
-         filename = paste0(args$output_figure, "02.png"),
+         filename = paste0(args$output_figure, "06.png"),
          width = 2300, height = 3000, res = 300)
@@
 }
+```
+
+## Weekly confirmed cases proportion
+
+```{r report-prop, fig.height=4}
+df_report <- traj_summary %>%
+  filter(var %in% c("D"), deme != "China") %>%
+  left_join(case_data %>% filter(source == "ecdc")) %>%
+  select(deme, date, l95_value, h95_value, median_value, new_confirmed) %>%
+  group_by(deme, year_week = ceiling_date(date, "1 week")) %>%
+  summarise(l95_value = sum(l95_value),
+            h95_value = sum(h95_value),
+            median_value = sum(median_value),
+            new_confirmed = sum(new_confirmed)) %>%
+  mutate(median_report = new_confirmed/median_value,
+         h95_report = new_confirmed/l95_value,
+         l95_report = new_confirmed/h95_value)
+
+report_plot <- ggplot(df_report) +
+  geom_line(aes(x = year_week, y = median_report, color = deme), stat="identity") +
+  geom_point(aes(x = year_week, y = median_report, color = deme), stat="identity") +
+  geom_ribbon(aes(year_week, ymin = l95_report, ymax = h95_report,fill = deme), alpha = 0.1) +
+  ylab("weekly reporting proportion") +
+  scale_color_manual(values = dcolors, name = "") +
+  scale_fill_manual(values = dcolors, name = "") +
+  scale_x_date(date_breaks = "5 days", date_labels = "%b %d", limits=c(ymd("2020-01-30"), ymd("2020-03-08"))) +
+  #scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b %y") +
+  coord_cartesian(ylim = c(0,1)) +
+  theme(#legend.position = "none", 
+    axis.title.x=element_blank()) #+
+#facet_wrap(~deme, scales = "free")
+
+report_plot
+ggexport(report_plot,
+         filename = paste0(args$output_figure, "07.png"),
+         width = 2000, height = 1200, res = 300)
+```
+
+\small
+```{r report-prop-table}
+knitr::kable(traj_summary %>%
+               filter(var %in% c("D")) %>%
+               left_join(case_data %>% filter(source == "ecdc")) %>%
+               select(deme, date, l95_cumvalue, h95_cumvalue, median_cumvalue, total_confirmed) %>%
+               group_by(deme) %>%
+               filter(date == max(date)) %>%
+               mutate(median_report = total_confirmed/median_cumvalue,
+                      h95_report = total_confirmed/l95_cumvalue,
+                      l95_report = total_confirmed/h95_cumvalue))
 ```
@@
     xlab("") +
-    ylab(ifelse(d %in% c("China", "Italy"), "local/imported cases\n by week", "")) +
+    ylab(ifelse(d %in% c("China", "Italy"), "ratio local/imported\nnew cases by week", "")) +
     labs(title = d) +
     #scale_colour_gradient(low = "yellow", high = "red", na.value = NA)
+    scale_x_date(date_labels = "%b %d, %Y") +
     scale_colour_viridis_c(name = "log cases", direction = -1) +
@@
 
+```{r bvsm-table, results = 'asis'}
+for (d in demes$deme) {
+  cat("\n## Local transmission vs imported cases - ", d, "\n")
+  knitr::kable(hsummary %>% filter(deme == d))
+  cat("\n")
+}
+```
+
 ```{r circoplot}
 plot_chord <- function(df, min_date, max_date) {
-  # df <- df %>%
-  #   filter(date > min_date, date <= max_date, var == "IM") %>%
-  #   group_by(partner, deme, date) %>%
-  #   # Take mean value over all trajectories
-  #   summarise(Nmean = mean(value), .groups = "drop_last") %>%
-  #   # Sum mean values of each day over all the time period
-  #   summarise(N = sum(Nmean), .groups = "drop") %>%
-  #   drop_na()
-
   df <- df %>%
-    filter(date > min_date, date <= max_date, var == "IM") %>%
+    filter(date > min_date, date <= max_date, var == "IM", !is.na(partner)) %>%
     group_by(partner, deme, traj) %>%
@@
   cplot <- circos.clear()
-  cplot <- circos.par(start.degree = 90, gap.degree = 4, track.margin = c(-0.1, 0.1), 
+  cplot <- circos.par(start.degree = 90, gap.degree = 4, track.margin = c(-0.3, 0.3), 
                       points.overflow.warning = FALSE)
-  par(mar = rep(0, 4))
+  par(mar = rep(0.5, 4))
   # Base plot
@@
         facing = "bending", 
-        cex = 0.9
+        cex = 0.8
       )
@@
         labels.niceFacing = FALSE,
-        labels.cex = 0.6)
+        labels.cex = 0.7)
     }
   )
-  return(cplot)
+  return(list(df = df, cplot = cplot))
 }
@@
 
-```{r, fig.height=2.5, fig.align='center'}
+::: {.columns}
+
+:::: {.column width="50%"}
+```{r mig1, fig.height=2.5, fig.align='center'}
 # Period 1
-plot_chord(df_traj, ymd("2019-09-01"), ymd("2020-01-23"))
+mig1 <- plot_chord(df_traj, ymd("2019-09-01"), ymd("2020-01-23"))
 
@@
 text(label = "A", x = -1, y = 1, cex = 1.5, font = 2)
+text("Period 1: origin - January 22, 2020", x = 0, y = 1, cex = 0.8, font = 2)
 dev.off()
 ```
+::::
+
+:::: {.column width="50%"}
+\small
+```{r mig1-table1}
+knitr::kable(mig1$df %>% filter(Nm != 0))
+```
+
+::::
+
+:::
 
@@
 
-```{r, fig.height=2.5, fig.align='center'}
+::: {.columns}
+
+:::: {.column width="50%"}
+```{r mig2, fig.height=2.5, fig.align='center'}
 # Period 2
-plot_chord(df_traj, ymd("2020-01-23"), ymd("2020-02-28"))
+mig2 <- plot_chord(df_traj, ymd("2020-01-23"), ymd("2020-02-28"))
 
@@
 plot_chord(df_traj, ymd("2020-01-23"), ymd("2020-02-28"))
-mtext(text = "B", at = -1, cex = 1.5, font = 2)
 text(label = "B", x = -1, y = 1, cex = 1.5, font = 2)
+text("Period 2: January 23, 2020 - February 29, 2020", x = 0, y = 1, cex = 0.8, font = 2)
 #dev.off()
 ```
+::::
+
+:::: {.column width="50%"}
+\tiny
+```{r mig2-table1}
+knitr::kable(mig2$df %>% filter(Nm != 0))
+```
+
+::::
+
+:::
 
@@
 
-```{r, fig.height=2.5, fig.align='center'}
+```{r mig3, fig.height=2.5, fig.align='center'}
 # Period 3
-plot_chord(df_traj, ymd("2020-02-28"), ymd("2020-03-08"))
+mig3 <- plot_chord(df_traj, ymd("2020-02-28"), ymd("2020-03-08"))
 
@@
 text(label = "C", x = -1, y = 1, cex = 1.5, font = 2)
+text("Period 3: March 1, 2020 - March 8, 2020", x = 0, y = 1, cex = 0.8, font = 2)
+text("total number of exported-imported cases", x = 0.3, y = -1, cex = 0.8, font = 1)
 dev.off()
 ```
+::::
+
+:::: {.column width="50%"}
+\tiny
+```{r mig3-table1}
+knitr::kable(mig3$df %>% filter(Nm != 0))
+```
+
+::::
+
+:::
 
@@
 
-```{r flight-pass, fig.height=4}
+```{r flight-cases, fig.height=4}
 demes1 <- c(China = "CN", France = "FR", Germany = "DE", Italy = "IT", OtherEuropean = "OE", Spain = "ES")
@@
 
-## Weekly confirmed cases proportion
-
-```{r report-prop, fig.height=4}
-df_report <- traj_summary %>%
-  filter(var %in% c("D"), deme != "China") %>%
-  left_join(case_data %>% filter(source == "ecdc")) %>%
-  select(deme, date, l95_value, h95_value, median_value, new_confirmed) %>%
-  group_by(deme, year_week = ceiling_date(date, "1 week")) %>%
-  summarise(l95_value = sum(l95_value),
-            h95_value = sum(h95_value),
-            median_value = sum(median_value),
-            new_confirmed = sum(new_confirmed)) %>%
-  mutate(median_report = new_confirmed/median_value,
-         h95_report = new_confirmed/l95_value,
-         l95_report = new_confirmed/h95_value)
-
-report_plot <- ggplot(df_report) +
-  geom_line(aes(x = year_week, y = median_report, color = deme), stat="identity") +
-  geom_point(aes(x = year_week, y = median_report, color = deme), stat="identity") +
-  geom_ribbon(aes(year_week, ymin = l95_report, ymax = h95_report,fill = deme), alpha = 0.1) +
-  ylab("weekly reporting proportion") +
-  scale_color_manual(values = dcolors, name = "") +
-  scale_fill_manual(values = dcolors, name = "") +
-  scale_x_date(date_breaks = "5 days", date_labels = "%b %d", limits=c(ymd("2020-01-30"), ymd("2020-03-08"))) +
-  #scale_x_date(limits = c(ymd("2019-10-01"), ymd(MRS)), date_breaks = "1 month", date_labels = "%b %y") +
-  coord_cartesian(ylim = c(0,1)) +
-  theme(#legend.position = "none", 
-    axis.title.x=element_blank()) #+
-#facet_wrap(~deme, scales = "free")
-
-report_plot
-ggexport(report_plot,
-         filename = paste0(args$output_figure, "11.png"),
-         width = 2000, height = 1200, res = 300)
+## Daily flight passengers vs daily imported cases
+\tiny
+```{r flight-cases-table}
+knitr::kable(left_join(by_epoch, flight_df_3e) %>%
+               group_by(dest, epoch, src) %>%
+               summarise(v = median(values),
+                         ev = median(evalues),
+                         lev = HDInterval::hdi(evalues)[[1]],
+                         hev = HDInterval::hdi(evalues)[[2]], .groups = "drop_last") %>%
+               summarise(v = sum(v),
+                         ev = sum(ev),
+                         lev = sum(lev),
+                         hev = sum(hev)))
 ```
@@
 ggexport(inc_plot,
-         filename = paste0(args$output_figure, "12.png"),
+         filename = paste0(args$output_figure, "11.png"),
          width = 2000, height = 1200, res = 300)
 ```
+
+## Weekly incidence population %
+
+\small
+```{r incidence-table}
+knitr::kable(df_inc %>% filter(date == max(date)))
+```

### 2021-03-24 scripts/comparison_nParticles.R

@@ -4,10 +4,14 @@ library(tidyverse)
 source("./scripts/trajProcessing.R")
 source("./scripts/utils.R")
 
+set_plotopts()
+
 particles <-  c(300, 1000, 3000, 10000)
 file_path <- "/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/210205_europe10."
 file_list <- paste0(file_path, particles, ".TL.traj")
-n = 1000
+n <- 500
+demes <- read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
+mrs <- ymd("2020-03-07")
 
 df_trajs <- lapply(file_list, function(file) {
   df <- loadTrajectories(file, burninFrac = 0, subsample = n)
@@ -44,6 +48,6 @@ plot_particles <- ggplot(df) +
 
 ggexport(plot_particles,
          filename = "reports/mt-thesis/figures/particles_comparison.png",
-         width = 2500, height = 1000, res = 300)
+         width = 3000, height = 1000, res = 300)
 
 

### 2021-03-24 scripts/trajProcessing.R

@@ -147,7 +147,7 @@ processEvents <- function(events, demes, source = NA, mrs) {
   events[events$event == "O", "mult"] =  1
   
   df_traj0 <- events %>%
-    mutate(date = date(date_decimal(decimal_date(mrs) - age))) %>%
+    mutate(date = date(date_decimal(decimal_date(ymd(mrs)) - age))) %>%
     select(-time, -age) %>%
     group_by_at(vars(-mult)) %>%
     summarise(N = sum(mult), .groups = "drop") %>%
@@ -199,7 +199,7 @@ processEvents <- function(events, demes, source = NA, mrs) {
   min_date <- min(df_traj$date)
   df_traj_complete <- df_traj%>%
     group_by(traj, var, deme, partner) %>%
-    complete(date = seq.Date(min_date, MRS, by = "day")) %>%
+    complete(date = seq.Date(min_date, ymd(mrs), by = "day")) %>%
     replace_na(list(value = 0)) %>%
     arrange(traj, date, deme) %>%
     mutate(cumvalue = cumsum(value))

### 2021-03-24 reports/mt-thesis/results.tex
wip mt

@@
     \label{fig:firstEUcountry}
-    \todo[inline]{ticks for every month}
 \end{figure}
@@
 The ECDC confirmed case counts are shown with dashed lines.}
-    \todo[figure]{Why is till March 20?}
+    \todo[inline]{Why is till March 20?}
     \label{fig:trajs}
@@
 The black line is the median value for the ratio local/imported cases.}
-    \label{fig:events}
-    \todo[inline]{change legend outgoing migration}
+    \label{fig:localvsimported}
 \end{figure}
@@
   \end{minipage}
-  \caption{\textbf{International imported case dynamics by analysis constat migration rate periods.} One chord diagram is drawn for each of the main periods of the analysis defined by the change times of the migration rates in the model.
+  \caption{\textbf{International imported case dynamics by analysis constant-migration rate period.} One chord diagram is drawn for each of the main periods of the analysis defined by the change times of the migration rates in the model.
 Each flux represents the infections exported from one country to another during the full period.
@@
     \includegraphics[width=\textwidth]{210205_europe10_figtraj10.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
-    \label{fig:first}
-    \todo[inline]{ticks for every month}
+    \caption{\textbf{Evolution of daily imported cases and flight passengers by constant migration rate period of the analysis.} In each subfigure, we take the median number of daily imported cases by period in the country (x-axis) and the average of daily flight passengers from Eurostat for each source country.}
+    \label{fig:migsflights}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{particles_comparison.png}
-    \caption{First introductions of SARS-CoV-2.
-From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted.
-\textbf{A} Probability density of the time of first introduction for each deme.
-Each dotted line represents the first date when cases where reported to ECDC by deme color.
-In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1.
-For the other five demes, the distribution of the time of first migration into the deme is shown.
-\textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme.
-This first case corresponds to the first migration event from China to any of the European countries.
-\textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
+    \caption{\textbf{Impact of the number of particles on the stochastic simulation of epidemic trajectories.} We simulate epidemic trajectories with a tau-leaping algorithm from our infered parameters of the birth-death model and the phylogenetic trees.
+We use 300, 1000, 3000 and 10000 particles, i.e.
+number of trajectories simulated for each set of estimated parameters.
+In the y-axis, We plot the median number of total active infections, no longer infectious cases, imported cases and local transmissions, and the 95\% credible interval (shaded area) computed from the epidemic trajectories events and the number of particles in the x-axis.
+}
     \label{fig:particles}
-    \todo[inline]{ticks for every month}
 \end{figure}

### 2021-03-24 scripts/analyze_trajectories.Rmd
wip mt

@@
 # Parser -----------------------------------------------------------------------
-# argsR <- commandArgs(TRUE)
-# args_list <- str_split(paste(argsR, collapse=""), pattern="--")
-# args_llist <- lapply(args_list, function(x) str_split(x, pattern="="))
-# argsDF <- as.data.frame(do.call("rbind", args_llist[[1]]))
-# args <- as.list(as.character(argsDF$V2))
-# names(args) <- argsDF$V1
-# args$burnin <- as.integer(args$burnin)
-# args$n <- as.integer(args$n)
-# title <- paste("Analysis", gsub("\\.", " ", gsub(".TL.traj", " particles", str_split(args$args, "/")[[1]][3])))
-# ---
-# subtitle: "`r title`"
-# ---
-```
+argsR <- commandArgs(TRUE)
+args_list <- str_split(paste(argsR, collapse=""), pattern="--")
+args_llist <- lapply(args_list, function(x) str_split(x, pattern="="))
+argsDF <- as.data.frame(do.call("rbind", args_llist[[1]]))
+args <- as.list(as.character(argsDF$V2))
+names(args) <- argsDF$V1
+args$burnin <- as.integer(args$burnin)
+args$n <- as.integer(args$n)
+title <- paste("Analysis", gsub("\\.", " ", gsub(".TL.traj", " particles", str_split(args$args, "/")[[1]][3])))
+```
+
+---
+subtitle: "`r title`"
+---
 
@@
 # Debugging
-args = list()
-args$burnin = 0.1
-args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
-args$args = "/Users/maceci/europe10last/210205_europe10.10000.TL.traj"
-args$metadata =  "./results/europeC/final_metadata.tsv"
-args$n = 10
-demes = args$demes
-demes$exclude_country[5] = "France,Germany,Italy,Spain"
-demes <- demes %>% arrange(deme)
+# args = list()
+# args$burnin = 0.1
+# args$demes = read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
+# args$args = "/Users/maceci/europe10last/210205_europe10.10000.TL.traj"
+# args$metadata =  "./results/europeC/final_metadata.tsv"
+# args$n = 10
+# demes = args$demes
+# demes$exclude_country[5] = "France,Germany,Italy,Spain"
+# demes <- demes %>% arrange(deme)
 ```
@@
 # Load demes configuration and sequence metadata -------------------------------
-# demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
-#                     exclude_country = NA, exclude_division = NA,
-#                     min_date = NA, max_date = NA, seq_per_deme = NA) %>%
-#   bind_rows(bind_rows(lapply(yaml::yaml.load(string = gsub(",", ", ", gsub(":", ": ", args$demes))),
-#                           data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
-#   group_by(deme) %>%
-#   mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
-#   distinct() %>%
-#   filter_all(any_vars(!is.na(.))) %>%
-#   mutate(deme = str_to_title(deme)) %>%
-#   mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme)) %>%
-#   arrange(deme)
+demes <- data.frame(deme = NA, region = NA, country = NA, division = NA,
+                    exclude_country = NA, exclude_division = NA,
+                    min_date = NA, max_date = NA, seq_per_deme = NA) %>%
+  bind_rows(bind_rows(lapply(yaml::yaml.load(string = gsub(",", ", ", gsub(":", ": ", args$demes))),
+                          data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
+  group_by(deme) %>%
+  mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
+  distinct() %>%
+  filter_all(any_vars(!is.na(.))) %>%
+  mutate(deme = str_to_title(deme)) %>%
+  mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme)) %>%
+  arrange(deme)
 
@@
 ## First introductions to each country, source
+
 ```{r first-bycountry, fig.height=4}
@@
 
-```{r first-times, fig.height=4}
+```{r first-times, fig.height=4, include=FALSE}
 # 2. First introduction distribution density by deme and time of first official reported case to ECDC
@@
   cat("\n## Local transmission vs imported cases - ", d, "\n")
-  knitr::kable(hsummary %>% filter(deme == d))
+  print(hsummary %>% filter(deme == d))
   cat("\n")
@@
 
+::: {.columns}
+
+:::: {.column width="50%"}
+
 ```{r mig3, fig.height=2.5, fig.align='center'}
@@
 
-## Weekly incidence population %
+## Weekly incidence population percentage
 

### 2021-03-25 scripts/adjust_names.R
initializing repo

@@ -1,73 +1,0 @@
-##------------------------------------------------------------------------------
-## Adjust metadata and final alignment sequence names.
-## Replace names of sequences with complete names for 
-## analysis (GISAID ref|deme|date) based on script by 
-## Sarah Nadeau (2019-nCov Data > data > sequences > 2020-04-01)
-## Script for snakemake workflow.
-## 
-## 2020-10-05 Cecilia Valenzuela
-##------------------------------------------------------------------------------
-
-# Load libraries ---------------------------------------------------------------
-library(argparse)
-library(yaml)
-library(ape)
-library(dplyr)
-
-# Parser -----------------------------------------------------------------------
-parser <- argparse::ArgumentParser()
-parser$add_argument("--metadata", type="character", help="Metadata  file")
-parser$add_argument("--alignment", type="character", help="Alignment file")
-parser$add_argument("--demes", type = "character", nargs="+", help = "Demes configuration")
-parser$add_argument("--output_alignment", type = "character",
-                    help = "Output file for updated alignment")
-parser$add_argument("--output_metadata", type = "character", 
-                    help = "Output file for updated metadata")
-
-args <- parser$parse_args()
-
-# Read files -------------------------------------------------------------------
-ALIGNMENT <- args$alignment
-METADATA <- args$metadata
-DEMES <- args$demes
-OUTPUT_ALIGNMENT <- args$output_alignment
-OUTPUT_METADATA <- args$output_metadata
-
-print(paste("metadata: ", METADATA))
-print(paste("alignment: ", ALIGNMENT))
-print(paste("demes: ", "Demes specification in config file"))
-print(paste("output alignment: ", OUTPUT_ALIGNMENT))
-print(paste("output metadata: ", OUTPUT_METADATA))
-
-alignment <- ape::read.FASTA(file = ALIGNMENT)
-full_metadata <- read.delim(file = METADATA)
-demes <- bind_rows(lapply(yaml::yaml.load(string = paste(DEMES, collapse = " ")),
-                          data.frame, stringsAsFactors = FALSE), .id = "deme")
-
-# Adjust metadata --------------------------------------------------------------
-metadata <- full_metadata %>%
-# 1. Filter metadata for only sequences in the alignment
-  filter(strain %in% names(alignment)) %>% 
-# 2. Add deme and focal column
-  left_join(demes, by = c("region")) %>%
[…]

### 2021-03-25 scripts/adjust_regional_meta.py

@@ -1,52 +1,0 @@
-"""
-Add column to metadata to denote 'focal' samples based on supplied region
-Rewrite location, division and country for non-focal samples to be region
-Rewrite division_exposure and country_exposure for non-focal samples to be region_exposure
-from 2020-09-11
-"""
-
-import argparse
-import pandas as pd
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser(
-        description="Add column to metadata to denote 'focal' samples based on supplied region",
-        formatter_class=argparse.ArgumentDefaultsHelpFormatter
-    )
-    parser.add_argument("--metadata", type = str, required=True, help="metadata")
-    parser.add_argument("--region", type=str, required=False, help="focal region")
-    parser.add_argument("--country", type=str, required=False, help="focal country")
-    parser.add_argument("--division", type=str, required=False, help="focal division")
-    parser.add_argument("--location", type=str, required=False, help="focal location")
-    parser.add_argument("--composite", type=str, required=False, help="composite sampling")
-    parser.add_argument("--output", type=str, required=True, help="adjusted metadata")
-    args = parser.parse_args()
-
-    region_list = ["Asia", "Africa", "Europe", "North America", "Oceania", "South America"]
-
-    metadata = pd.read_csv(args.metadata, delimiter='\t')
-
-    # if in region list, then do the fixing
-    if args.region in region_list:
-        focal_region = args.region
-    else: # otherwise just write out metadata as is, and proceed
-        metadata.to_csv(args.output, index=False, sep="\t")
-        exit()
-
-    print("Adjusting metadata for focal region", args.region)
-
-
-    metadata.insert(12, 'focal', True)
-
-    metadata.loc[metadata.region != focal_region, 'focal'] = False
-    metadata.loc[metadata.region != focal_region, 'location'] = ""
-    metadata.loc[metadata.region != focal_region, 'division'] = metadata.region
-    metadata.loc[metadata.region != focal_region, 'country'] = metadata.region
-    metadata.loc[metadata.region != focal_region, 'division_exposure'] = metadata.region_exposure
-    metadata.loc[metadata.region != focal_region, 'country_exposure'] = metadata.region_exposure
-    metadata.loc[(metadata.region == focal_region) & (metadata.region_exposure != focal_region), 'division_exposure'] = metadata.region_exposure
[…]

### 2021-03-25 scripts/analyze_subsample.R

@@ -1,231 +1,0 @@
-##-----------------------------------------------------
-## Analyze subsample 
-## TODO
-## - Sequences vs cases
-## - Maps and histograms of available sequences and included sequences by deme
-## - Temporal distribution of sequences and MRS 
-##
-## 2020-10-21 Cecilia Valenzuela
-##------------------------------------------------------
-
-library(argparse)
-library(ape)
-library(tidyverse)
-
-parser <- argparse::ArgumentParser()
-parser$add_argument("--metadata", type="character", help="Metadata  file")
-parser$add_argument("--alignment", type="character", help="Alignment file")
-parser$add_argument("--subsample", type = "character", help = "Metadata sumsample file")
-parser$add_argument("--output_figure", type = "character", help = "Output file for histogram of sequences")
-parser$add_argument("--mrs", type = "character",help = "Output file for most recent sample date")
-
-args <- parser$parse_args()
-
-# Read files
-METADATA <- args$metadata
-ALIGNMENT <- args$alignment
-SUBSAMPLE <- args$subsample
-OUTPUT_FIGURE <- args$output_figure
-MRS <- args$mrs
-
-print(paste("metadata:", METADATA))
-print(paste("alignment:", ALIGNMENT))
-print(paste("metadata subsample: ", SUBSAMPLE))
-print(paste("output figure:", OUTPUT_FIGURE))
-print(paste("output most recent sample date:", MRS))
-
-cat("Number of sequences: ", length(alignment), "\n")
-cat("By deme:\n")
-knitr::kable(metadata%>%dplyr::filter(new_seq_name%in%names(alignment))%>%
-               dplyr::group_by(deme)%>%
-               dplyr::mutate(first_sample=min(as.Date(date), na.rm=TRUE),
-                             last_sample=max(as.Date(date), na.rm=TRUE))%>%
-               dplyr::group_by(deme,first_sample,last_sample)%>%
-               dplyr::summarise(n = n())%>%
-               dplyr::arrange(desc(n)))
-
-cat("\nDetail OtherEuropean:")
-kable(metadata%>%dplyr::filter(new_seq_name%in%names(alignment) & deme == 'OtherEuropean')%>%
-        dplyr::count(country, sort=TRUE))
-
-png(filename = OUTPUT_FIGURE, width = 600)
-ggplot(data = metadata%>%dplyr::filter(new_seq_name%in%names(alignment)), 
[…]

### 2021-03-25 scripts/analyze_trajectories.R

@@ -1,798 +1,0 @@
-##------------------------------------------------------------------------------
-## Script for plotting and manipulating trajectory data using functions from 
-## BDMM-Prime project from T.Vaughan to parse trajectories.
-## Script for snakemake workflow.
-## 
-## 2020-10-12 Cecilia Valenzuela
-##------------------------------------------------------------------------------
-
-# Load libraries ---------------------------------------------------------------
-library(argparse)
-library(yaml)
-library(tidyverse)
-library(lubridate)
-library(scales)
-library(ggpubr)
-library(ggsci)
-library(ggridges)
-library(patchwork)
-library(hrbrthemes)
-library(circlize)
-library(chorddiag) 
-library(ggforce)
-library(ComplexHeatmap)
-
-# Source files -----------------------------------------------------------------
-source("./scripts/trajProcessing.R")
-source("./scripts/utils.R")
-
-
-# Parser -----------------------------------------------------------------------
-parser <- argparse::ArgumentParser()
-parser$add_argument("--input", type = "character", help="trajectory file")
-parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
-parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
-parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
-parser$add_argument("--n", type = "integer", help = "Number of trajectories to analyze")
-parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
-parser$add_argument("--output_table", type = "character", help = "Output path for the tables")
-args <- parser$parse_args()
-
-print(args)
-
-
-# Load trajectory data ---------------------------------------------------------
-df <- loadTrajectories(args$input, burninFrac = args$burnin, subsample = args$n)
-cat("\nNumber of analyzed trajectories (subsampled):", length(unique(df$states$traj)))
-
-print(args$demes)
-# Load demes configuration and sequence metadata -------------------------------
-demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
[…]

### 2021-03-25 scripts/case_data.R

@@ -1,96 +1,0 @@
-source("./scripts/utils.R")
-library(tidyverse)
-
-ecdc <- get_dataECDC()
-who <- get_dataWHO()
-# jh <- get_dataJH() Same than owid
-owid <- get_dataOWID() #%>% select(-new_cases) %>% rename(new_cases = new_cases_smoothed)
-covid19od <- get_dataC19OD()
-
-case_data <- bind_rows(ecdc, owid, covid19od, who) %>%
-  filter(country %in% c("China", "France", "Germany", "Italy", "Spain"),
-         date <= "2020-03-07")
-
-ggplot(case_data) +
-  geom_line(aes(date, total_confirmed, color = source)) +
-  facet_wrap(~country, scales = "free")
-  
-  
-  
-ecdc2 <- ecdc %>%
-  mutate(new_cases = zoo::rollmean(new_cases, k = 7, fill = NA, align = "center"),
-         source = "ecdc2")
-case_data <- bind_rows(bind_rows(bind_rows(ecdc, jh), owid), ecdc2)
-
-
-case_data2 <- case_data %>%
-  group_by(country, source) %>%
-  mutate(new_cases2 = zoo::rollmean(new_cases, k = 7, fill = NA)) %>%
-  select(country, date, source, new_cases, new_cases2)
-
-#bind_rows(case_data, china) %>%
-case_data  %>%
-  filter(date > ymd("2019-12-01"),
-         date <= ymd("2020-03-18"),
-         country %in% c("China", "France", "Germany", "Italy", "Spain")) %>%
-  ggplot() +
-  geom_line(aes(date, new_cases, color = source, linetype = source)) +
-  facet_wrap(~country, scales = "free")
-
-samples <- tibble(country = c("China", "France", "Germany", "Italy", "OtherEuropean", "Spain"), s = c(60, 60, 50, 60, 50, 60))
-sp <- case_data %>%
-  filter(date > ymd("2019-12-01"),
-         (date <= ymd("2020-03-08") & country != "China")
-         | (date <= ymd("2020-01-23") & country == "China"),
-         country %in% c("China", "France", "Germany", "Italy", "Spain")) %>%
-  group_by(country, source) %>%
-  summarise(cases = sum(new_cases, na.rm = TRUE)) %>%
-  left_join(samples) %>%
-  mutate(sp = s/cases)
-
[…]

### 2021-03-25 scripts/chains_diagnostic.R

@@ -1,58 +1,0 @@
-##------------------------------------------------------------------------------
-## Check ESS Values > 200 in trace files from BEAST analysis
-## Create file with input file names that pass the check 
-## to combine them with LogCombiner v1.8.2, 2002-2015, 
-## ETH euler version. Script for snakemake workflow.
-## 
-## 2020-10-27 Cecilia Valenzuela
-##------------------------------------------------------------------------------
-
-# Load libraries ---------------------------------------------------------------
-library(argparse)
-library(dplyr)
-library(stringr)
-library(readr)
-
-# Parser -----------------------------------------------------------------------
-parser <- argparse::ArgumentParser()
-parser$add_argument("--input", nargs="+", help="Input log summary file")
-parser$add_argument("--ess", type="integer", help="ESS value cutoff")
-parser$add_argument("--burnin", type="integer", help="Burning percentage")
-parser$add_argument("--length", type="integer", help="Chain length")
-parser$add_argument("--output", type="character", help="Output diagnostic file")
-args <- parser$parse_args()
-
-ESS_cutoff <- args$ess
-
-
-
-# Diagnostics ------------------------------------------------------------------
-diagnostic <- data.frame()
-for (fname in args$input){
-  if (file.size(fname) != 0) {
-    tb <- read_table(fname) %>%
-    filter(`95%HPDlo` != `95%HPDup`) # Filter fixed values
-    if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
-      cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
-      diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
-                                                 seed = str_split(fname, pattern ="\\.")[[1]][2],
-                                                 burnin = args$burnin,
-                                                 length = args$length,
-                                                 min_ESS = ESS_cutoff, 
-                                                 included = 1))
-    } else {
-      cat("\n\n\nItems:\n\n", tb %>% filter(ESS < ESS_cutoff) %>% pull(item),
-          "\n\nhave ESS value < ", ESS_cutoff, " in log file ", fname, "\n\nLog file not included in analysis.\n")
[…]

### 2021-03-25 scripts/check_log.R

@@ -1,59 +1,0 @@
-##-----------------------------------------------------
-## Check ESS Values > 200 in trace files from BEAST analysis
-## Create file with input file names that pass the check 
-## to combine them with LogCombiner v1.8.2, 2002-2015, 
-## ETH euler version. Script for snakemake workflow.
-## 
-## 2020-10-06 Cecilia Valenzuela
-##------------------------------------------------------
-
-library(argparse)
-library(tidyverse)
-
-parser <- argparse::ArgumentParser()
-parser$add_argument("--input", nargs="+", help="")
-parser$add_argument("--include_all", type="logical", help="")
-parser$add_argument("--output_log", type="character", help="")
-parser$add_argument("--output_tree", type="character", help="")
-parser$add_argument("--output_traj", type="character", help="")
-args <- parser$parse_args()
-
-INPUT <- args$input
-ALL <- args$include_all
-OUTPUT_LOG <- args$output_log
-OUTPUT_TREE <- args$output_tree
-OUTPUT_TRAJ <- args$output_traj
-
-print(paste("log summary:", INPUT))
-print(paste("Include all logs:", ALL))
-print(paste("output log file names:", OUTPUT_LOG))
-print(paste("output tree file names:", OUTPUT_TREE))
-print(paste("output traj file names:", OUTPUT_TREE))
-
-for (fname in INPUT){
-  tb <- read_table2(fname, skip=1)
-  
-  if (all(tb$ESS >= 200, na.rm = TRUE)) {
-    cat("All ESS values > 200 in log file ", fname, "\nLog file included in analysis.")
-    log_file = gsub(".logsummary.txt", ".log", fname)
-    tree_file = gsub(".logsummary.txt", ".typed.node.trees", fname)
-    traj_file = gsub(".logsummary.txt", ".TL.traj", fname)
-    write(log_file, file=OUTPUT_LOG, append=TRUE)
-    write(tree_file, file=OUTPUT_TREE, append=TRUE)
-    write(traj_file, file=OUTPUT_TRAJ, append=TRUE)
-  } else {
-    if (!ALL){
-      cat("Items:\n", tb%>%dplyr::filter(ESS < 200)%>%dplyr::pull(statistic),
-        "\nhave ESS value < 200 in log file ", fname, "\nLog file not included in analysis.")
-    }else{
[…]

### 2021-03-25 scripts/combine-and-dedup-fastas.py

@@ -1,25 +1,0 @@
-# from 2020-09-11
-
-from Bio import SeqIO
-import argparse
-from augur.align import read_sequences
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser(
-        description="Combine and dedup FASTAs",
-        formatter_class=argparse.ArgumentDefaultsHelpFormatter
-    )
-
-    parser.add_argument('--input', type=str,  nargs="+", metavar="FASTA", required=True, help="input FASTAs")
-    parser.add_argument('--output', type=str, metavar="FASTA", required=True, help="output FASTA")
-    args = parser.parse_args()
-
-    # Read sequences with augur to benefit from additional checks for duplicates.
-    sequences = read_sequences(*args.input)
-
-    # Convert dictionary of sequences by id to a list, for compatibility with
-    # augur versions <9.0.0.
-    if isinstance(sequences, dict):
-        sequences = list(sequences.values())
-
-    SeqIO.write(sequences, args.output, 'fasta')

### 2021-03-25 scripts/comparison_nParticles.R

@@ -1,53 +1,0 @@
-# Comparison Number of particles
-
-library(tidyverse)
-source("./scripts/trajProcessing.R")
-source("./scripts/utils.R")
-
-set_plotopts()
-
-particles <-  c(300, 1000, 3000, 10000)
-file_path <- "/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/210205_europe10."
-file_list <- paste0(file_path, particles, ".TL.traj")
-n <- 500
-demes <- read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
-mrs <- ymd("2020-03-07")
-
-df_trajs <- lapply(file_list, function(file) {
-  df <- loadTrajectories(file, burninFrac = 0, subsample = n)
-  df_traj <- processEvents(df$events, demes, "China", mrs)
-  return(df_traj)
-})
-names(df_trajs) <- particles
-
-df <- bind_rows(df_trajs, .id = 'particles') %>%
-  filter(var %in% c("active_pop", "B", "D", "IM"),
-    date == max(date), is.na(partner)) %>%
-  group_by(particles, var, traj) %>%
-  summarise(cumvalue = sum(cumvalue), .groups = "drop_last") %>%
-  summarise(m = median(cumvalue),
-            l = quantile(cumvalue, 0.25),
-            h = quantile(cumvalue, 0.75),
-            l95 = HDInterval::hdi(cumvalue)[[1]],
-            h95 = HDInterval::hdi(cumvalue)[[2]],)
-
-facet_names <- c("Active infections", "No longer infectious cases", 
-                 "Imported cases", "Local transmissions")
-names(facet_names) <- unique(df$var)
-
-plot_particles <- ggplot(df) +
-  geom_point(aes(x = as.numeric(particles), y = m)) +
-  geom_line(aes(x = as.numeric(particles), y = m)) +
-  geom_ribbon(aes(x = as.numeric(particles), ymin = l95, ymax = h95), alpha = 0.5) +
-  facet_wrap(~var, scales = "free", nrow = 1, labeller = labeller(var = facet_names))+
-  ylab("number of events") +
-  xlab("number of particles") +
-  scale_x_continuous(breaks = c(300, 1000, 3000, 10000)) +
-  theme(axis.text.x = element_text(angle = 45))
-  #scale_y_log10(labels = trans_format("log10", math_format(10^.x))) 
-
-ggexport(plot_particles,
[…]

### 2021-03-25 scripts/diagnostic.py

@@ -1,146 +1,0 @@
-# MIT License
-# Copyright (c) 2020 Nextstrain, full text https://github.com/nextstrain/ncov/blob/master/LICENSE
-# from 2020-09-11
-
-"""
-Mask initial bases from alignment FASTA
-"""
-import argparse, gzip, sys
-sys.path.insert(0,'.')
-from collections import defaultdict
-import numpy as np
-from Bio.SeqIO.FastaIO import SimpleFastaParser
-from Bio.Seq import Seq
-from Bio import AlignIO, SeqIO
-from priorities import sequence_to_int_array
-from augur.utils import read_metadata
-from datetime import datetime
-
-tmrca = datetime(2019, 12, 1).toordinal()
-
-def expected_divergence(date, rate_per_day = 25/365):
-    try:
-        return (datetime.strptime(date, '%Y-%m-%d').toordinal() - tmrca)*rate_per_day
-    except:
-        return np.nan
-
-
-def analyze_divergence(sequences, metadata, reference, mask_5p=0, mask_3p=0):
-    int_ref = sequence_to_int_array(reference, fill_gaps=False)
-    diagnostics = defaultdict(dict)
-    fill_value = 110
-    gap_value = 45
-    ws = 50
-    known_true_clusters = [(28880,28883)]
-    known_true_cluster_array = np.ones_like(int_ref, dtype=int)
-    for b,e in known_true_clusters:
-        known_true_cluster_array[b:e]=0
-
-    cluster_cut_off = 5
-    with open(sequences) as fasta:
-        for h,s in SimpleFastaParser(fasta):
-            left_gaps = len(s) - len(s.lstrip('-'))
-            right_gaps = len(s) - len(s.rstrip('-'))
-            s = sequence_to_int_array(s, fill_value=fill_value, fill_gaps=False)
-            # mask from both ends to avoid exclusion for problems at sites that will be masked anyway
-            if mask_5p:
-                s[:mask_5p] = fill_value
-            if mask_3p:
-                s[-mask_3p:] = fill_value
-
-            # fill terminal gaps -- those will be filled anyway
-            if left_gaps:
-                s[:left_gaps] = fill_value
-            if right_gaps:
-                s[-right_gaps:] = fill_value
-
-            # determine non-gap non-N mismatches
-            snps = (int_ref!=s) & (s!=fill_value) & (s!=gap_value)
-            # determine N positions
-            filled = s==fill_value
[…]

### 2021-03-25 scripts/flight_data.R

@@ -1,72 +1,0 @@
- ## Flight data
-
-library(tidyverse)
-library(lubridate)
-library(eurostat)
-source("./scripts/utils.R")
-
-# Returns dayly passengers in a given month 
-get_flightData <- function(intraUE_data, extraUE_data, countries) {
-  
-  # Get datasets from eurostats, montthly information
-  intraUE_data <- eurostat::get_eurostat("avia_paincc", select_time = "M")
-  extraUE_data <- eurostat::get_eurostat("avia_paexcc", select_time = "M")
-
-  if ("OE" %in% countries) countries <- c(countries, "EU27_2020", "EU28", "UK")
-  
-  df <- bind_rows(intraUE_data, extraUE_data) %>%
-    # Data from passengers
-    filter(unit == "PAS", geo %in% countries, 
-           partner %in% countries) %>%
-    # Select passengers carried, departures for all countries (if we consider
-    # departures and arrivals for each country we have duplicated values, one
-    # for each reporting country. Recommendation from eurostats is to take only
-    # the departures) and departures and arrivals for China
-    filter(tra_meas == "PAS_CRD_DEP" | 
-             (tra_meas == "PAS_CRD_ARR" & partner == "CN")) %>%
-    mutate(partner = ifelse(tra_meas == "PAS_CRD_ARR" & partner == "CN", geo, partner),
-           geo = ifelse(tra_meas == "PAS_CRD_ARR", "CN", geo),
-           tra_meas = ifelse(tra_meas == "PAS_CRD_ARR", "PAS_CRD_DEP", tra_meas)) %>%
-    filter(!is.na(values)) 
-  
-  # In 2020 UK is out of the european aggregated value
-  # Correct for UK Brexit, EU28 -> EU27 in 2020
-  # Only for months 2020
-  
-  df1 <- df %>%
-    mutate(geo = ifelse(geo %in% c("EU27_2020", "UK"), "EU28", geo),
-           partner = ifelse(partner %in% c("EU27_2020", "UK"), "EU28", partner)) %>%
-    group_by(geo, partner, time) %>%
-    mutate(values = sum(values)) %>%
-    unique()
-  
-  # Create Other European = EU28 - (France, Germany, Italy, Spain)
-  oe_geo <- df1 %>%
-    filter(!geo %in% c("EU28", "CN"), partner != "EU28") %>%
-    group_by(partner, time) %>%
[…]

### 2021-03-25 scripts/mask-alignment.py

@@ -1,56 +1,0 @@
-# MIT License
-# Copyright (c) 2020 Nextstrain, full text https://github.com/nextstrain/ncov/blob/master/LICENSE
-# from 2020-09-11
-
-"""
-Mask initial bases from alignment FASTA
-"""
-import argparse
-import Bio
-import Bio.SeqIO
-from Bio.Seq import Seq
-
-def mask_terminal_gaps(seq):
-    L = len(seq)
-    seq_trimmed = seq.lstrip('-')
-    left_gaps = L - len(seq_trimmed)
-    seq_trimmed = seq_trimmed.rstrip('-')
-    right_gaps = L - len(seq_trimmed) - left_gaps
-    return "N"*left_gaps + seq_trimmed + "N"*right_gaps 
-
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser(
-        description="Mask initial bases from alignment FASTA",
-        formatter_class=argparse.ArgumentDefaultsHelpFormatter
-    )
-    parser.add_argument("--alignment", required=True, help="FASTA file of alignment")
-    parser.add_argument("--mask-terminal-gaps", action='store_true', help="fill all terminal gaps with N as they likely represent missing data")
-    parser.add_argument("--mask-from-beginning", type = int, required=True, help="number of bases to mask from start")
-    parser.add_argument("--mask-from-end", type = int, help="number of bases to mask from end")
-    parser.add_argument("--mask-sites", nargs='+', type = int,  help="list of sites to mask")
-    parser.add_argument("--output", required=True, help="FASTA file of output alignment")
-    args = parser.parse_args()
-
-    begin_length = 0
-    if args.mask_from_beginning:
-        begin_length = args.mask_from_beginning
-    end_length = 0
-    if args.mask_from_end:
-        end_length = args.mask_from_end
-
-    with open(args.output, 'w') as outfile:
-        for record in Bio.SeqIO.parse(args.alignment, 'fasta'):
-            seq = str(record.seq)
-            if args.mask_terminal_gaps:
-                seq = mask_terminal_gaps(seq)
-
-            start = "N" * begin_length
-            middle = seq[begin_length:-end_length]
-            end = "N" * end_length
-            seq_list = list(start + middle + end)
-            if args.mask_sites:
-                for site in args.mask_sites:
-                    seq_list[site-1] = "N"
-            record.seq = Seq("".join(seq_list))
-            Bio.SeqIO.write(record, outfile, 'fasta')

### 2021-03-25 scripts/partition-sequences.py

@@ -1,46 +1,0 @@
-# MIT License
-# Copyright (c) 2020 Nextstrain, full text https://github.com/nextstrain/ncov/blob/master/LICENSE
-# from 2020-09-11
-
-
-"""Split sequences for multiple sequence alignment into smaller chunks to speed up alignment.
-"""
-import argparse
-from augur.align import read_sequences
-from Bio import SeqIO
-from pathlib import Path
-
-
-if __name__ == "__main__":
-    parser = argparse.ArgumentParser()
-    parser.add_argument("--sequences", required=True, nargs="+", help="FASTA file of sequences to partition into smaller chunks")
-    parser.add_argument("--sequences-per-group", required=True, type=int, help="number of sequences to include in each group")
-    parser.add_argument("--output-dir", required=True, help="directory to write out partitioned sequences")
-
-    args = parser.parse_args()
-
-    # Read sequences with augur to benefit from additional checks for duplicates.
-    sequences = read_sequences(*args.sequences)
-
-    # Convert dictionary of sequences by id to a list, for compatibility with
-    # augur versions <9.0.0.
-    if isinstance(sequences, dict):
-        sequences = list(sequences.values())
-
-    # Create the requested output directory.
-    output_dir = Path(args.output_dir)
-    output_dir.mkdir(exist_ok=True)
-
-    # Determine partition indices.
-    indices = list(range(0, len(sequences), args.sequences_per_group))
-
-    # Add a final index to represent the end of the last partition.
-    if indices[-1] != len(sequences):
-        indices.append(len(sequences))
-
-    # Partition sequences into groups of no more than the requested number.
-    for i in range(len(indices) - 1):
-        # Save partitioned sequences to a new FASTA file named after the partition number.
-        print("Write out %i sequences for partition %i" % (indices[i + 1] - indices[i], i))
-        output_path = output_dir / Path("%i.fasta" % i)
-        SeqIO.write(sequences[indices[i]:indices[i + 1]], str(output_path), 'fasta')

### 2021-03-25 scripts/priorities.py

@@ -1,182 +1,0 @@
-"""
-Mask initial bases from alignment FASTA
-from 2020-09-11
-"""
-import argparse
-from random import shuffle
-from collections import defaultdict
-import Bio
-import numpy as np
-from Bio.SeqIO.FastaIO import SimpleFastaParser
-from Bio.Seq import Seq
-from Bio import AlignIO, SeqIO
-from scipy import sparse
-
-
-def compactify_sequences(sparse_matrix, sequence_names):
-    sequence_groups = defaultdict(list)
-    for s, snps in zip(sequence_names, sparse_matrix):
-        ind = snps.nonzero()
-        vals = np.array(snps[ind])
-        if len(ind[1]):
-            sequence_groups[tuple(zip(ind[1], vals[0]))].append(s)
-        else:
-            sequence_groups[tuple()].append(s)
-
-    return sequence_groups
-
-INITIALISATION_LENGTH = 1000000
-
-def sequence_to_int_array(s, fill_value=110, fill_gaps=True):
-    seq = np.frombuffer(str(s).lower().encode('utf-8'), dtype=np.int8).copy()
-    if fill_gaps:
-        seq[(seq!=97) & (seq!=99) & (seq!=103) & (seq!=116)] = fill_value
-    else:
-        seq[(seq!=97) & (seq!=99) & (seq!=103) & (seq!=116) & (seq!=45)] = fill_value
-    return seq
-
-# Function adapted from https://github.com/gtonkinhill/pairsnp-python
-def calculate_snp_matrix(fastafile, consensus=None, zipped=False, fill_value=110):
-    # This function generate a sparse matrix where differences to the consensus are coded as integers.
-
-    row = np.empty(INITIALISATION_LENGTH)
-    col = np.empty(INITIALISATION_LENGTH, dtype=np.int64)
-    val = np.empty(INITIALISATION_LENGTH, dtype=np.int8)
-
-    r = 0
-    n_snps = 0
-    nseqs = 0
-    seq_names = []
-    filled_positions = []
-    current_length = INITIALISATION_LENGTH
-    if zipped:
-        fh = gzip.open(fastafile, 'rt')
-    else:
-        fh = open(fastafile, 'rt')
-    with fh as fasta:
-        for h,s in SimpleFastaParser(fasta):
-            if consensus is None:
-                align_length = len(s)
-                # Take consensus as first sequence
-                consensus = sequence_to_int_array(s, fill_value=fill_value)
-            else:
[…]

### 2021-03-25 scripts/sampling_proportion.R

@@ -1,52 +1,0 @@
-##------------------------------------------------------------------------------
-## Script for plotting and manipulating trajectory data using functions from 
-## BDMM-Prime project from T.Vaughan to parse trajectories.
-## Script for snakemake workflow.
-## 
-## 2020-10-12 Cecilia Valenzuela
-##------------------------------------------------------------------------------
-
-# Load libraries ---------------------------------------------------------------
-library(argparse)
-library(yaml)
-library(tidyverse)
-
-# Source files -----------------------------------------------------------------
-source("./scripts/utils.R")
-
-# Parser -----------------------------------------------------------------------
-parser <- argparse::ArgumentParser()
-parser$add_argument("--input", type = "character", nargs = "+", help="Subsampling demes configuration")
-parser$add_argument("--output", type = "character", help = "Output path for the sampling proportions table")
-args <- parser$parse_args()
-
-print(args)
-
-# Load demes configuration -----------------------------------------------------
-demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
-                    exclude_country = NA, exclude_division = NA,
-                    min_date = NA, max_date = NA, seq_per_deme = NA) %>%
-  bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(args$input, collapse = " ")),
-                             data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
-  group_by(deme) %>% 
-  mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
-  distinct() %>%
-  filter_all(any_vars(!is.na(.)))
-
-cat("\nDeme configuration:\n")
-knitr::kable(demes)
-
-# Case data information from ECDC ----------------------------------------------
-case_data <- get_cases(demes) %>%
-  arrange(date) %>%
-  group_by(deme) %>%
-  summarise(totalcases = sum(cases))
-
-# Compute sampling proportions -------------------------------------------------
-sampling_proportions <- case_data%>%
-  left_join(demes) %>%
-  mutate(sampling = seq_per_deme/totalcases)
-
-# Save table -------------------------------------------------------------------
-write.table(sampling_proportions, args$output, sep = "\t")
-

### 2021-03-25 scripts/subsample.R

@@ -1,234 +1,0 @@
-##------------------------------------------------------------------------------
-## Subsample sequences for a deme.
-## Filter by region, country and division, and date interval. 
-## The probability of sampling a specific sequence is proportional to the 
-## probability of having a case that date, weighted by the inverse of the 
-## probability of having a sequence that day and also weighted by the inverse 
-## of the probability of having a sequence from that division.
-## 
-## In this way, we expect to account for the differences in sampling efforts 
-## across days and divisions; and to obtain a constant sampling rate 
-## (assumption of the model) by taking into account the total number of cases each day.
-##
-## Script for snakemake workflow.
-## 
-## 2020-10-10 Cecilia Valenzuela
-##------------------------------------------------------------------------------
-
-subsample <- function(alignment, metadata, include = NA, exclude = NA,
-                      region_name = NA, country_name = NA, division_name = NA,
-                      exclude_country = NA, exclude_division = NA,
-                      from, to, seq_per_deme, seed, prob = "time") {
-  set.seed(seed)
-  
-  # Read files
-  alignment <- read.FASTA(file = alignment)
-  metadata <- read.delim(file = metadata, stringsAsFactors = FALSE) %>%
-    filter(strain %in% names(alignment)) %>% # Keep only metadata about seqs in the alignment
-    mutate(date = ymd(date)) 
-  
-  if (!is.null(exclude)) {
-    exclude <- readLines(include)
-    metadata <- filter(metadata, !strain %in% exclude)
-  }
-  
-  # Filter metadata
-  metadata_deme <- metadata %>%
-    filter(region == region_name,
-           if (is.na(country_name)) TRUE else country %in% country_name,
-           if (is.na(division_name)) TRUE else division %in% division_name,
-           if (all(is.na(exclude_country))) TRUE else !country %in% exclude_country,
-           if (all(is.na(exclude_division))) TRUE else !division %in% exclude_division,
-           date >= ymd(from),
-           date <= ymd(to))
-  
-  if (nrow(metadata_deme) == 0) {
-    warning("No sequences for these specifications")
-    return()
-  }
-  
-  if (prob == "deaths") to <- ymd(to) + 14
-  
-  # Get case counts for deme
-  case_data <- get_dataOWID()
-  death_data <- get_dataECDC()
[…]

### 2021-03-25 scripts/trajProcessing.R

@@ -1,208 +1,0 @@
-## Scripts useful for manipulating trajectory data
-
-require(tidyverse)
-
-parseTrajectory <- function(trajStr) {
-  strValues <- str_split(str_split(trajStr, ",")[[1]], ":", simplify = TRUE)
-  values <- apply(strValues[,-2], 2, as.numeric)
-  time <- values[,1]
-  src <- values[,2]
-  dest <- values[,3]
-  mult <- values[,4]
-  N <- values[,-(1:4)]
-  event<- strValues[,2]
-
-  res <- list(time = time,
-              N = N,
-              event = event,
-              src = src,
-              dest = dest,
-              mult = mult)
-
-  return(res)
-}
-
-loadTrajectories <- function(filename, burninFrac=0.1, subsample=NA) {
-    states <- NULL
-    events <- NULL
-
-    message("Loading ", filename,"...", appendLF = FALSE)
-    df_in <- read_tsv(filename, col_types="ic")
-
-    if (burninFrac>0) {
-        n <- dim(df_in)[1]
-        df_in <- df_in[-(1:ceiling(burninFrac*n)),]
-    }
-
-    if (!is.na(subsample)) {
-        indices <- unique(round(seq(1, dim(df_in)[1], length.out=subsample)))
-        df_in <- df_in[indices,]
-    }
-    
-    for (row in 1:(dim(df_in)[1])) {
-        trajStr <- df_in[row,2]
-        trajStates <- parseTrajectory(trajStr)
-        Ndim <- dim(trajStates$N)
-
-        if (length(Ndim)==0) {
-            ntypes <- 1
-            states <- bind_rows(states,
-                                tibble(traj=row,
-                                       type=0,
-                                       time=trajStates$time,
-                                       N=trajStates$N))
-        } else {
-            ntypes <- dim(trajStates$N)[2]
-            for (s in 1:ntypes) {
-                
-                states <- bind_rows(states,
-                                    tibble(traj=row,
-                                           type=s-1,
-                                           time=trajStates$time,
-                                           N=trajStates$N[,s]))
-            }
-        }
-
-        events <- bind_rows(events,
-                            tibble(traj=row,
-                                   time=trajStates$time,
[…]

### 2021-03-25 scripts/utils.R

@@ -1,331 +1,0 @@
-##-----------------------------------------------------
-## Util functions for analysis workflow
-## 
-## 2020-10-14 Cecilia Valenzuela
-##------------------------------------------------------
-
-library(countrycode)
-library(ggpubr)
-library(lubridate)
-library(maps)
-library(sf)
-library(wpp2019)
-
-set_plotopts <- function() {
-  theme_set(theme_minimal())
-  theme_update(panel.grid = element_blank(),
-               legend.position = "bottom",
-               legend.box="vertical",
-               legend.text=element_text(size = 7),
-               legend.title=element_text(size = 7),
-               strip.background = element_rect(fill = "white", colour = "grey80"),
-               strip.text = element_text(size = 8, color = "grey30"),
-               axis.title.y = element_text(size = 8, #face = "bold", 
-                                           margin = margin(r = 10)),
-               axis.title.x = element_text(size = 8, #face = "bold", 
-                                           margin = margin(t = 10)),
-               panel.grid.major.y =  element_line(colour = "grey80", size = 0.2, linetype = 2),
-               #panel.grid.minor.y =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
-               panel.background =  element_blank(),
-               axis.text.x = element_text(size = 7, angle = 30, hjust = 1, vjust = 1),
-               axis.text.y = element_text(size = 7),
-               plot.title = element_text(size=8, face = "bold"),
-               axis.ticks.length.x=unit(0.1, "cm"),
-               axis.ticks.length.y=unit(0.1, "cm"),
-               axis.ticks = element_line(colour = "grey50", size = 0.2))
-               #plot.margin = margin(3, 7, 3, 1.5))
-}
-
-
-
-get_cases <- function(demes, from = NA, to = NA) {
-  # Get case data information from ECDC for specific demes (countries and regions)
-  # Set from and to dates with values from the function call instead of the ones from the demes info
-  #cat("\nLoading confirmed case counts from ECDC...\n")
-  #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
-  #else case_data <- get_dataECDC()
[…]

### 2021-03-25 scripts/analyze_trajectories.Rmd
initial commit

@@
 
+## Trajectory parameters
+
+\tiny
+```{r args-table}
+knitr::kable(argsDF)
+```
+
 ## Deme configuration
@@
 # Load trajectory data ---------------------------------------------------------
-df <- loadTrajectories(args$args, burninFrac = args$burnin, subsample = args$n)
+df <- loadTrajectories2(args$args, burninFrac = args$burnin, subsample = args$n)
 
 # Data wrangling ---------------------------------------------------------------
-df_traj <- processEvents(df$events, demes, "China", MRS)
+df_traj <- processEvents(df, demes, "China", mrs)
 ```
@@
                mutate(p_1d = n_1d/args$n) %>%
-               left_join(first_df10 %>%
+               full_join(first_df10 %>%
                            ungroup %>%
@@
                            mutate(p_10d = n_10d/(10*args$n)) %>%
-                           left_join(first_df10 %>%
+                           full_join(first_df10 %>%
                                        ungroup %>%
@@
 knitr::kable((df_eventsttiming %>% 
-                group_by(deme, var) %>%
-                summarise(m = median(date),
-                          l = quantile(date, 0.025, type = 1),
-                          h = quantile(date, 0.925, type = 1)))[1:12,])
+  left_join(first_ecdc %>% rename(ecdc_date = date)) %>%
+  group_by(var, deme) %>%
+  filter(date < ecdc_date) %>%
+  count() %>%
+  mutate(p = n/args$n))[1:12,])
 ```
@@
 knitr::kable((df_eventsttiming %>% 
-                group_by(deme, var) %>%
-                summarise(m = median(date),
-                          l = quantile(date, 0.025, type = 1),
-                          h = quantile(date, 0.925, type = 1)))[12:24,])
+  left_join(first_ecdc %>% rename(ecdc_date = date)) %>%
+  group_by(var, deme) %>%
+  filter(date < ecdc_date) %>%
+  count() %>%
+  mutate(p = n/args$n))[12:24,])
 ```
@@
 ```{r bvsm, fig.height=4}
-h = df_traj %>%
+h <- df_traj %>%
   filter(var %in% c("B", "IM"), is.na(partner)) %>%
@@
   filter(IM != 0) %>%
-  #na_if(0) %>% replace_na(list(B = 1e-8)) %>%
   mutate(R = B/IM) %>%
@@
   select(traj, deme, date, R) %>%
-  arrange(date) #%>%
-#pivot_wider(names_from = date, values_from = R) %>%
-#arrange(traj)
-
-inpop = df_traj %>% filter(var == "in_pop", date == "2020-03-07") %>% select(traj, cumvalue)
-
-hweek = df_traj %>%
+  arrange(date)
+
+inpop <- df_traj %>% filter(var == "in_pop", date == "2020-03-07") %>% select(traj, cumvalue)
+
+hweek <- df_traj %>%
   filter(var %in% c("B", "IM"), is.na(partner)) %>%
@@
   filter(IM != 0) %>%
-  #na_if(0) %>% replace_na(list(B = 1e-8)) %>%
   mutate(R = B/IM) %>%
   mutate(R = ifelse(R == Inf, NA, R)) %>%
-  select(traj, deme, year_week, R) %>%
+  select(traj, deme, year_week, R, B, IM) %>%
   arrange(year_week)
@@
             lR  = HDInterval::hdi(R)[[1]],
-            hR  = HDInterval::hdi(R)[[2]])
+            hR  = HDInterval::hdi(R)[[2]],
+            mB = median(B),
+            mIM = median(IM))
 
@@
   cat("\n## Local transmission vs imported cases - ", d, "\n")
-  print(hsummary %>% filter(deme == d))
+  print(knitr::kable(hsummary %>% filter(deme == d)))
   cat("\n")
-}
+  cat("\n")}
 ```
@@
 ```{r incidence-table}
-knitr::kable(df_inc %>% filter(date == max(date)))
-```
+knitr::kable(df_inc %>% filter(date == max(date)) %>% select(-date))
+```

### 2021-03-25 scripts/MigRateGLM.java

@@ -1,0 +1,78 @@
+package bdmmprime.util.priors;
+
+import beast.core.CalculationNode;
+import beast.core.Function;
+import beast.core.Input;
+import beast.core.parameter.RealParameter;
+
+public class MigRateGLM extends CalculationNode implements Function {
+
+    public Input<Function> covariateListInput = new Input<>("covariateList",
+            "List of matrix of covariates, e.g. numbers of flights between different locations",
+            Input.Validate.REQUIRED);
+
+    public Input<Function> scalerParamInput = new Input<>("scalerParam",
+            "Scaler parameter.", Input.Validate.REQUIRED);
+
+    public Input<Function> indicatorParamInput = new Input<>("indicatorParam",
+            "Indicator parameter.", Input.Validate.REQUIRED);
+
+    public Input<RealParameter> globalScalerParamInput = new Input<>("globalScalerParam",
+            "Global scaler parameter.", Input.Validate.OPTIONAL);
+
+//    final public Input<Double> scaleUpperLimit = new Input<>("upper", "Upper Limit of scale factor", 100.0);
+//    final public Input<Double> scaleLowerLimit = new Input<>("lower", "Lower limit of scale factor", 0.0);
+
+
+//    public Input<Function> errorInput = new Input<>("errorTerm",
+//            "Error term.", Input.Validate.OPTIONAL);
+
+    Function covariateList, scalerParam, indicatorParam, globalScalerParam;
+    int covariateSize;
+//    private double upper, lower;
+
+    @Override
+    public void initAndValidate() {
+        covariateList = covariateListInput.get();
+        scalerParam = scalerParamInput.get();
+        indicatorParam = indicatorParamInput.get();
+
+        globalScalerParam = globalScalerParamInput.get();
+//        errorTerm = errorInput.get();
+
+        covariateSize = covariateList.getDimension()/(scalerParam.getDimension());
+
+//        upper = scaleUpperLimit.get();
+//        lower = scaleLowerLimit.get();
+    }
+
+//    protected boolean outsideBounds(final double value) {
+//
+//        return (value < lower || value > upper);
+//        //return (l != null && value < l || h != null && value > h);
+//    }
+
+    @Override
+    public int getDimension() {
+        return covariateSize;
+    }
+
+    @Override
+    public double getArrayValue(int i) {
+        double lograte = 0;
+        for (int j = 0; j < scalerParam.getDimension(); j++) {
[…]

### 2021-03-25 scripts/adjust_names.R

@@ -1,0 +1,73 @@
+##------------------------------------------------------------------------------
+## Adjust metadata and final alignment sequence names.
+## Replace names of sequences with complete names for 
+## analysis (GISAID ref|deme|date) based on script by 
+## Sarah Nadeau (2019-nCov Data > data > sequences > 2020-04-01)
+## Script for snakemake workflow.
+## 
+## 2020-10-05 Cecilia Valenzuela
+##------------------------------------------------------------------------------
+
+# Load libraries ---------------------------------------------------------------
+library(argparse)
+library(yaml)
+library(ape)
+library(dplyr)
+
+# Parser -----------------------------------------------------------------------
+parser <- argparse::ArgumentParser()
+parser$add_argument("--metadata", type="character", help="Metadata  file")
+parser$add_argument("--alignment", type="character", help="Alignment file")
+parser$add_argument("--demes", type = "character", nargs="+", help = "Demes configuration")
+parser$add_argument("--output_alignment", type = "character",
+                    help = "Output file for updated alignment")
+parser$add_argument("--output_metadata", type = "character", 
+                    help = "Output file for updated metadata")
+
+args <- parser$parse_args()
+
+# Read files -------------------------------------------------------------------
+ALIGNMENT <- args$alignment
+METADATA <- args$metadata
+DEMES <- args$demes
+OUTPUT_ALIGNMENT <- args$output_alignment
+OUTPUT_METADATA <- args$output_metadata
+
+print(paste("metadata: ", METADATA))
+print(paste("alignment: ", ALIGNMENT))
+print(paste("demes: ", "Demes specification in config file"))
+print(paste("output alignment: ", OUTPUT_ALIGNMENT))
+print(paste("output metadata: ", OUTPUT_METADATA))
+
+alignment <- ape::read.FASTA(file = ALIGNMENT)
+full_metadata <- read.delim(file = METADATA)
+demes <- bind_rows(lapply(yaml::yaml.load(string = paste(DEMES, collapse = " ")),
+                          data.frame, stringsAsFactors = FALSE), .id = "deme")
+
+# Adjust metadata --------------------------------------------------------------
+metadata <- full_metadata %>%
+# 1. Filter metadata for only sequences in the alignment
+  filter(strain %in% names(alignment)) %>% 
+# 2. Add deme and focal column
+  left_join(demes, by = c("region")) %>%
[…]

### 2021-03-25 scripts/analyze_subsample.R

@@ -1,0 +1,231 @@
+##-----------------------------------------------------
+## Analyze subsample 
+## TODO
+## - Sequences vs cases
+## - Maps and histograms of available sequences and included sequences by deme
+## - Temporal distribution of sequences and MRS 
+##
+## 2020-10-21 Cecilia Valenzuela
+##------------------------------------------------------
+
+library(argparse)
+library(ape)
+library(tidyverse)
+
+parser <- argparse::ArgumentParser()
+parser$add_argument("--metadata", type="character", help="Metadata  file")
+parser$add_argument("--alignment", type="character", help="Alignment file")
+parser$add_argument("--subsample", type = "character", help = "Metadata sumsample file")
+parser$add_argument("--output_figure", type = "character", help = "Output file for histogram of sequences")
+parser$add_argument("--mrs", type = "character",help = "Output file for most recent sample date")
+
+args <- parser$parse_args()
+
+# Read files
+METADATA <- args$metadata
+ALIGNMENT <- args$alignment
+SUBSAMPLE <- args$subsample
+OUTPUT_FIGURE <- args$output_figure
+MRS <- args$mrs
+
+print(paste("metadata:", METADATA))
+print(paste("alignment:", ALIGNMENT))
+print(paste("metadata subsample: ", SUBSAMPLE))
+print(paste("output figure:", OUTPUT_FIGURE))
+print(paste("output most recent sample date:", MRS))
+
+cat("Number of sequences: ", length(alignment), "\n")
+cat("By deme:\n")
+knitr::kable(metadata%>%dplyr::filter(new_seq_name%in%names(alignment))%>%
+               dplyr::group_by(deme)%>%
+               dplyr::mutate(first_sample=min(as.Date(date), na.rm=TRUE),
+                             last_sample=max(as.Date(date), na.rm=TRUE))%>%
+               dplyr::group_by(deme,first_sample,last_sample)%>%
+               dplyr::summarise(n = n())%>%
+               dplyr::arrange(desc(n)))
+
+cat("\nDetail OtherEuropean:")
+kable(metadata%>%dplyr::filter(new_seq_name%in%names(alignment) & deme == 'OtherEuropean')%>%
+        dplyr::count(country, sort=TRUE))
+
+png(filename = OUTPUT_FIGURE, width = 600)
+ggplot(data = metadata%>%dplyr::filter(new_seq_name%in%names(alignment)), 
[…]

### 2021-03-25 scripts/case_data.R

@@ -1,0 +1,96 @@
+source("./scripts/utils.R")
+library(tidyverse)
+
+ecdc <- get_dataECDC()
+who <- get_dataWHO()
+# jh <- get_dataJH() Same than owid
+owid <- get_dataOWID() #%>% select(-new_cases) %>% rename(new_cases = new_cases_smoothed)
+covid19od <- get_dataC19OD()
+
+case_data <- bind_rows(ecdc, owid, covid19od, who) %>%
+  filter(country %in% c("China", "France", "Germany", "Italy", "Spain"),
+         date <= "2020-03-07")
+
+ggplot(case_data) +
+  geom_line(aes(date, total_confirmed, color = source)) +
+  facet_wrap(~country, scales = "free")
+  
+  
+  
+ecdc2 <- ecdc %>%
+  mutate(new_cases = zoo::rollmean(new_cases, k = 7, fill = NA, align = "center"),
+         source = "ecdc2")
+case_data <- bind_rows(bind_rows(bind_rows(ecdc, jh), owid), ecdc2)
+
+
+case_data2 <- case_data %>%
+  group_by(country, source) %>%
+  mutate(new_cases2 = zoo::rollmean(new_cases, k = 7, fill = NA)) %>%
+  select(country, date, source, new_cases, new_cases2)
+
+#bind_rows(case_data, china) %>%
+case_data  %>%
+  filter(date > ymd("2019-12-01"),
+         date <= ymd("2020-03-18"),
+         country %in% c("China", "France", "Germany", "Italy", "Spain")) %>%
+  ggplot() +
+  geom_line(aes(date, new_cases, color = source, linetype = source)) +
+  facet_wrap(~country, scales = "free")
+
+samples <- tibble(country = c("China", "France", "Germany", "Italy", "OtherEuropean", "Spain"), s = c(60, 60, 50, 60, 50, 60))
+sp <- case_data %>%
+  filter(date > ymd("2019-12-01"),
+         (date <= ymd("2020-03-08") & country != "China")
+         | (date <= ymd("2020-01-23") & country == "China"),
+         country %in% c("China", "France", "Germany", "Italy", "Spain")) %>%
+  group_by(country, source) %>%
+  summarise(cases = sum(new_cases, na.rm = TRUE)) %>%
+  left_join(samples) %>%
+  mutate(sp = s/cases)
+
[…]

### 2021-03-25 scripts/chains_diagnostic.R

@@ -1,0 +1,58 @@
+##------------------------------------------------------------------------------
+## Check ESS Values > 200 in trace files from BEAST analysis
+## Create file with input file names that pass the check 
+## to combine them with LogCombiner v1.8.2, 2002-2015, 
+## ETH euler version. Script for snakemake workflow.
+## 
+## 2020-10-27 Cecilia Valenzuela
+##------------------------------------------------------------------------------
+
+# Load libraries ---------------------------------------------------------------
+library(argparse)
+library(dplyr)
+library(stringr)
+library(readr)
+
+# Parser -----------------------------------------------------------------------
+parser <- argparse::ArgumentParser()
+parser$add_argument("--input", nargs="+", help="Input log summary file")
+parser$add_argument("--ess", type="integer", help="ESS value cutoff")
+parser$add_argument("--burnin", type="integer", help="Burning percentage")
+parser$add_argument("--length", type="integer", help="Chain length")
+parser$add_argument("--output", type="character", help="Output diagnostic file")
+args <- parser$parse_args()
+
+ESS_cutoff <- args$ess
+
+
+
+# Diagnostics ------------------------------------------------------------------
+diagnostic <- data.frame()
+for (fname in args$input){
+  if (file.size(fname) != 0) {
+    tb <- read_table(fname) %>%
+    filter(`95%HPDlo` != `95%HPDup`) # Filter fixed values
+    if (all(tb$ESS >= ESS_cutoff, na.rm = TRUE)) {
+      cat("\n\n\nAll ESS values > ", ESS_cutoff, " in log file ", fname, "\n\nLog file included in analysis.\n")
+      diagnostic <- rbind(diagnostic, data.frame(chain = gsub(".summary.txt", "", str_split(fname, pattern = "/", n = 2)[[1]][[2]]),
+                                                 seed = str_split(fname, pattern ="\\.")[[1]][2],
+                                                 burnin = args$burnin,
+                                                 length = args$length,
+                                                 min_ESS = ESS_cutoff, 
+                                                 included = 1))
+    } else {
+      cat("\n\n\nItems:\n\n", tb %>% filter(ESS < ESS_cutoff) %>% pull(item),
+          "\n\nhave ESS value < ", ESS_cutoff, " in log file ", fname, "\n\nLog file not included in analysis.\n")
[…]

### 2021-03-25 scripts/combine-and-dedup-fastas.py

@@ -1,0 +1,25 @@
+# from 2020-09-11
+
+from Bio import SeqIO
+import argparse
+from augur.align import read_sequences
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(
+        description="Combine and dedup FASTAs",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+
+    parser.add_argument('--input', type=str,  nargs="+", metavar="FASTA", required=True, help="input FASTAs")
+    parser.add_argument('--output', type=str, metavar="FASTA", required=True, help="output FASTA")
+    args = parser.parse_args()
+
+    # Read sequences with augur to benefit from additional checks for duplicates.
+    sequences = read_sequences(*args.input)
+
+    # Convert dictionary of sequences by id to a list, for compatibility with
+    # augur versions <9.0.0.
+    if isinstance(sequences, dict):
+        sequences = list(sequences.values())
+
+    SeqIO.write(sequences, args.output, 'fasta')

### 2021-03-25 scripts/comparison_nParticles.R

@@ -1,0 +1,53 @@
+# Comparison Number of particles
+
+library(tidyverse)
+source("./scripts/trajProcessing.R")
+source("./scripts/utils.R")
+
+set_plotopts()
+
+particles <-  c(300, 1000, 3000, 10000)
+file_path <- "/Users/maceci/Documents/CBB Master/HS20/Master Thesis/sars-cov-2-eu-phylodynamics/results/europeC/210205_europe10."
+file_list <- paste0(file_path, particles, ".TL.traj")
+n <- 500
+demes <- read.csv("./files/demes_China.csv", stringsAsFactors = FALSE)
+mrs <- ymd("2020-03-07")
+
+df_trajs <- lapply(file_list, function(file) {
+  df <- loadTrajectories(file, burninFrac = 0, subsample = n)
+  df_traj <- processEvents(df$events, demes, "China", mrs)
+  return(df_traj)
+})
+names(df_trajs) <- particles
+
+df <- bind_rows(df_trajs, .id = 'particles') %>%
+  filter(var %in% c("active_pop", "B", "D", "IM"),
+    date == max(date), is.na(partner)) %>%
+  group_by(particles, var, traj) %>%
+  summarise(cumvalue = sum(cumvalue), .groups = "drop_last") %>%
+  summarise(m = median(cumvalue),
+            l = quantile(cumvalue, 0.25),
+            h = quantile(cumvalue, 0.75),
+            l95 = HDInterval::hdi(cumvalue)[[1]],
+            h95 = HDInterval::hdi(cumvalue)[[2]],)
+
+facet_names <- c("Active infections", "No longer infectious cases", 
+                 "Imported cases", "Local transmissions")
+names(facet_names) <- unique(df$var)
+
+plot_particles <- ggplot(df) +
+  geom_point(aes(x = as.numeric(particles), y = m)) +
+  geom_line(aes(x = as.numeric(particles), y = m)) +
+  geom_ribbon(aes(x = as.numeric(particles), ymin = l95, ymax = h95), alpha = 0.5) +
+  facet_wrap(~var, scales = "free", nrow = 1, labeller = labeller(var = facet_names))+
+  ylab("number of events") +
+  xlab("number of particles") +
+  scale_x_continuous(breaks = c(300, 1000, 3000, 10000)) +
+  theme(axis.text.x = element_text(angle = 45))
+  #scale_y_log10(labels = trans_format("log10", math_format(10^.x))) 
+
+ggexport(plot_particles,
[…]

### 2021-03-25 scripts/diagnostic.py

@@ -1,0 +1,146 @@
+# MIT License
+# Copyright (c) 2020 Nextstrain, full text https://github.com/nextstrain/ncov/blob/master/LICENSE
+# from 2020-09-11
+
+"""
+Mask initial bases from alignment FASTA
+"""
+import argparse, gzip, sys
+sys.path.insert(0,'.')
+from collections import defaultdict
+import numpy as np
+from Bio.SeqIO.FastaIO import SimpleFastaParser
+from Bio.Seq import Seq
+from Bio import AlignIO, SeqIO
+from priorities import sequence_to_int_array
+from augur.utils import read_metadata
+from datetime import datetime
+
+tmrca = datetime(2019, 12, 1).toordinal()
+
+def expected_divergence(date, rate_per_day = 25/365):
+    try:
+        return (datetime.strptime(date, '%Y-%m-%d').toordinal() - tmrca)*rate_per_day
+    except:
+        return np.nan
+
+
+def analyze_divergence(sequences, metadata, reference, mask_5p=0, mask_3p=0):
+    int_ref = sequence_to_int_array(reference, fill_gaps=False)
+    diagnostics = defaultdict(dict)
+    fill_value = 110
+    gap_value = 45
+    ws = 50
+    known_true_clusters = [(28880,28883)]
+    known_true_cluster_array = np.ones_like(int_ref, dtype=int)
+    for b,e in known_true_clusters:
+        known_true_cluster_array[b:e]=0
+
+    cluster_cut_off = 5
+    with open(sequences) as fasta:
+        for h,s in SimpleFastaParser(fasta):
+            left_gaps = len(s) - len(s.lstrip('-'))
+            right_gaps = len(s) - len(s.rstrip('-'))
+            s = sequence_to_int_array(s, fill_value=fill_value, fill_gaps=False)
+            # mask from both ends to avoid exclusion for problems at sites that will be masked anyway
+            if mask_5p:
+                s[:mask_5p] = fill_value
+            if mask_3p:
+                s[-mask_3p:] = fill_value
+
+            # fill terminal gaps -- those will be filled anyway
+            if left_gaps:
+                s[:left_gaps] = fill_value
+            if right_gaps:
+                s[-right_gaps:] = fill_value
+
+            # determine non-gap non-N mismatches
+            snps = (int_ref!=s) & (s!=fill_value) & (s!=gap_value)
+            # determine N positions
+            filled = s==fill_value
[…]

### 2021-03-25 scripts/flight_data.R

@@ -1,0 +1,72 @@
+ ## Flight data
+
+library(tidyverse)
+library(lubridate)
+library(eurostat)
+source("./scripts/utils.R")
+
+# Returns dayly passengers in a given month 
+get_flightData <- function(intraUE_data, extraUE_data, countries) {
+  
+  # Get datasets from eurostats, montthly information
+  intraUE_data <- eurostat::get_eurostat("avia_paincc", select_time = "M")
+  extraUE_data <- eurostat::get_eurostat("avia_paexcc", select_time = "M")
+
+  if ("OE" %in% countries) countries <- c(countries, "EU27_2020", "EU28", "UK")
+  
+  df <- bind_rows(intraUE_data, extraUE_data) %>%
+    # Data from passengers
+    filter(unit == "PAS", geo %in% countries, 
+           partner %in% countries) %>%
+    # Select passengers carried, departures for all countries (if we consider
+    # departures and arrivals for each country we have duplicated values, one
+    # for each reporting country. Recommendation from eurostats is to take only
+    # the departures) and departures and arrivals for China
+    filter(tra_meas == "PAS_CRD_DEP" | 
+             (tra_meas == "PAS_CRD_ARR" & partner == "CN")) %>%
+    mutate(partner = ifelse(tra_meas == "PAS_CRD_ARR" & partner == "CN", geo, partner),
+           geo = ifelse(tra_meas == "PAS_CRD_ARR", "CN", geo),
+           tra_meas = ifelse(tra_meas == "PAS_CRD_ARR", "PAS_CRD_DEP", tra_meas)) %>%
+    filter(!is.na(values)) 
+  
+  # In 2020 UK is out of the european aggregated value
+  # Correct for UK Brexit, EU28 -> EU27 in 2020
+  # Only for months 2020
+  
+  df1 <- df %>%
+    mutate(geo = ifelse(geo %in% c("EU27_2020", "UK"), "EU28", geo),
+           partner = ifelse(partner %in% c("EU27_2020", "UK"), "EU28", partner)) %>%
+    group_by(geo, partner, time) %>%
+    mutate(values = sum(values)) %>%
+    unique()
+  
+  # Create Other European = EU28 - (France, Germany, Italy, Spain)
+  oe_geo <- df1 %>%
+    filter(!geo %in% c("EU28", "CN"), partner != "EU28") %>%
+    group_by(partner, time) %>%
[…]

### 2021-03-25 scripts/mask-alignment.py

@@ -1,0 +1,56 @@
+# MIT License
+# Copyright (c) 2020 Nextstrain, full text https://github.com/nextstrain/ncov/blob/master/LICENSE
+# from 2020-09-11
+
+"""
+Mask initial bases from alignment FASTA
+"""
+import argparse
+import Bio
+import Bio.SeqIO
+from Bio.Seq import Seq
+
+def mask_terminal_gaps(seq):
+    L = len(seq)
+    seq_trimmed = seq.lstrip('-')
+    left_gaps = L - len(seq_trimmed)
+    seq_trimmed = seq_trimmed.rstrip('-')
+    right_gaps = L - len(seq_trimmed) - left_gaps
+    return "N"*left_gaps + seq_trimmed + "N"*right_gaps 
+
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(
+        description="Mask initial bases from alignment FASTA",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+    parser.add_argument("--alignment", required=True, help="FASTA file of alignment")
+    parser.add_argument("--mask-terminal-gaps", action='store_true', help="fill all terminal gaps with N as they likely represent missing data")
+    parser.add_argument("--mask-from-beginning", type = int, required=True, help="number of bases to mask from start")
+    parser.add_argument("--mask-from-end", type = int, help="number of bases to mask from end")
+    parser.add_argument("--mask-sites", nargs='+', type = int,  help="list of sites to mask")
+    parser.add_argument("--output", required=True, help="FASTA file of output alignment")
+    args = parser.parse_args()
+
+    begin_length = 0
+    if args.mask_from_beginning:
+        begin_length = args.mask_from_beginning
+    end_length = 0
+    if args.mask_from_end:
+        end_length = args.mask_from_end
+
+    with open(args.output, 'w') as outfile:
+        for record in Bio.SeqIO.parse(args.alignment, 'fasta'):
+            seq = str(record.seq)
+            if args.mask_terminal_gaps:
+                seq = mask_terminal_gaps(seq)
+
+            start = "N" * begin_length
+            middle = seq[begin_length:-end_length]
+            end = "N" * end_length
+            seq_list = list(start + middle + end)
+            if args.mask_sites:
+                for site in args.mask_sites:
+                    seq_list[site-1] = "N"
+            record.seq = Seq("".join(seq_list))
+            Bio.SeqIO.write(record, outfile, 'fasta')

### 2021-03-25 scripts/partition-sequences.py

@@ -1,0 +1,46 @@
+# MIT License
+# Copyright (c) 2020 Nextstrain, full text https://github.com/nextstrain/ncov/blob/master/LICENSE
+# from 2020-09-11
+
+
+"""Split sequences for multiple sequence alignment into smaller chunks to speed up alignment.
+"""
+import argparse
+from augur.align import read_sequences
+from Bio import SeqIO
+from pathlib import Path
+
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--sequences", required=True, nargs="+", help="FASTA file of sequences to partition into smaller chunks")
+    parser.add_argument("--sequences-per-group", required=True, type=int, help="number of sequences to include in each group")
+    parser.add_argument("--output-dir", required=True, help="directory to write out partitioned sequences")
+
+    args = parser.parse_args()
+
+    # Read sequences with augur to benefit from additional checks for duplicates.
+    sequences = read_sequences(*args.sequences)
+
+    # Convert dictionary of sequences by id to a list, for compatibility with
+    # augur versions <9.0.0.
+    if isinstance(sequences, dict):
+        sequences = list(sequences.values())
+
+    # Create the requested output directory.
+    output_dir = Path(args.output_dir)
+    output_dir.mkdir(exist_ok=True)
+
+    # Determine partition indices.
+    indices = list(range(0, len(sequences), args.sequences_per_group))
+
+    # Add a final index to represent the end of the last partition.
+    if indices[-1] != len(sequences):
+        indices.append(len(sequences))
+
+    # Partition sequences into groups of no more than the requested number.
+    for i in range(len(indices) - 1):
+        # Save partitioned sequences to a new FASTA file named after the partition number.
+        print("Write out %i sequences for partition %i" % (indices[i + 1] - indices[i], i))
+        output_path = output_dir / Path("%i.fasta" % i)
+        SeqIO.write(sequences[indices[i]:indices[i + 1]], str(output_path), 'fasta')

### 2021-03-25 scripts/priorities.py

@@ -1,0 +1,182 @@
+"""
+Mask initial bases from alignment FASTA
+from 2020-09-11
+"""
+import argparse
+from random import shuffle
+from collections import defaultdict
+import Bio
+import numpy as np
+from Bio.SeqIO.FastaIO import SimpleFastaParser
+from Bio.Seq import Seq
+from Bio import AlignIO, SeqIO
+from scipy import sparse
+
+
+def compactify_sequences(sparse_matrix, sequence_names):
+    sequence_groups = defaultdict(list)
+    for s, snps in zip(sequence_names, sparse_matrix):
+        ind = snps.nonzero()
+        vals = np.array(snps[ind])
+        if len(ind[1]):
+            sequence_groups[tuple(zip(ind[1], vals[0]))].append(s)
+        else:
+            sequence_groups[tuple()].append(s)
+
+    return sequence_groups
+
+INITIALISATION_LENGTH = 1000000
+
+def sequence_to_int_array(s, fill_value=110, fill_gaps=True):
+    seq = np.frombuffer(str(s).lower().encode('utf-8'), dtype=np.int8).copy()
+    if fill_gaps:
+        seq[(seq!=97) & (seq!=99) & (seq!=103) & (seq!=116)] = fill_value
+    else:
+        seq[(seq!=97) & (seq!=99) & (seq!=103) & (seq!=116) & (seq!=45)] = fill_value
+    return seq
+
+# Function adapted from https://github.com/gtonkinhill/pairsnp-python
+def calculate_snp_matrix(fastafile, consensus=None, zipped=False, fill_value=110):
+    # This function generate a sparse matrix where differences to the consensus are coded as integers.
+
+    row = np.empty(INITIALISATION_LENGTH)
+    col = np.empty(INITIALISATION_LENGTH, dtype=np.int64)
+    val = np.empty(INITIALISATION_LENGTH, dtype=np.int8)
+
+    r = 0
+    n_snps = 0
+    nseqs = 0
+    seq_names = []
+    filled_positions = []
+    current_length = INITIALISATION_LENGTH
+    if zipped:
+        fh = gzip.open(fastafile, 'rt')
+    else:
+        fh = open(fastafile, 'rt')
+    with fh as fasta:
+        for h,s in SimpleFastaParser(fasta):
+            if consensus is None:
+                align_length = len(s)
+                # Take consensus as first sequence
+                consensus = sequence_to_int_array(s, fill_value=fill_value)
+            else:
[…]

### 2021-03-25 scripts/subsample.R

@@ -1,0 +1,234 @@
+##------------------------------------------------------------------------------
+## Subsample sequences for a deme.
+## Filter by region, country and division, and date interval. 
+## The probability of sampling a specific sequence is proportional to the 
+## probability of having a case that date, weighted by the inverse of the 
+## probability of having a sequence that day and also weighted by the inverse 
+## of the probability of having a sequence from that division.
+## 
+## In this way, we expect to account for the differences in sampling efforts 
+## across days and divisions; and to obtain a constant sampling rate 
+## (assumption of the model) by taking into account the total number of cases each day.
+##
+## Script for snakemake workflow.
+## 
+## 2020-10-10 Cecilia Valenzuela
+##------------------------------------------------------------------------------
+
+subsample <- function(alignment, metadata, include = NA, exclude = NA,
+                      region_name = NA, country_name = NA, division_name = NA,
+                      exclude_country = NA, exclude_division = NA,
+                      from, to, seq_per_deme, seed, prob = "time") {
+  set.seed(seed)
+  
+  # Read files
+  alignment <- read.FASTA(file = alignment)
+  metadata <- read.delim(file = metadata, stringsAsFactors = FALSE) %>%
+    filter(strain %in% names(alignment)) %>% # Keep only metadata about seqs in the alignment
+    mutate(date = ymd(date)) 
+  
+  if (!is.null(exclude)) {
+    exclude <- readLines(include)
+    metadata <- filter(metadata, !strain %in% exclude)
+  }
+  
+  # Filter metadata
+  metadata_deme <- metadata %>%
+    filter(region == region_name,
+           if (is.na(country_name)) TRUE else country %in% country_name,
+           if (is.na(division_name)) TRUE else division %in% division_name,
+           if (all(is.na(exclude_country))) TRUE else !country %in% exclude_country,
+           if (all(is.na(exclude_division))) TRUE else !division %in% exclude_division,
+           date >= ymd(from),
+           date <= ymd(to))
+  
+  if (nrow(metadata_deme) == 0) {
+    warning("No sequences for these specifications")
+    return()
+  }
+  
+  if (prob == "deaths") to <- ymd(to) + 14
+  
+  # Get case counts for deme
+  case_data <- get_dataOWID()
+  death_data <- get_dataECDC()
[…]

### 2021-03-25 scripts/trajProcessing.R

@@ -1,0 +1,242 @@
+## Scripts useful for manipulating trajectory data
+
+require(tidyverse)
+
+parseTrajectory <- function(trajStr) {
+  strValues <- str_split(str_split(trajStr, ",")[[1]], ":", simplify = TRUE)
+  values <- apply(strValues[,-2], 2, as.numeric)
+  time <- values[,1]
+  src <- values[,2]
+  dest <- values[,3]
+  mult <- values[,4]
+  N <- values[,-(1:4)]
+  event<- strValues[,2]
+
+  res <- list(time = time,
+              N = N,
+              event = event,
+              src = src,
+              dest = dest,
+              mult = mult)
+
+  return(res)
+}
+
+loadTrajectories <- function(filename, burninFrac=0.1, subsample=NA) {
+  states <- NULL
+  events <- NULL
+  
+  message("Loading ", filename,"...", appendLF = FALSE)
+  df_in <- read_tsv(filename, col_types="ic")
+  
+  if (burninFrac>0) {
+    n <- dim(df_in)[1]
+    df_in <- df_in[-(1:ceiling(burninFrac*n)),]
+  }
+  
+  if (!is.na(subsample)) {
+    indices <- unique(round(seq(1, dim(df_in)[1], length.out=subsample)))
+    df_in <- df_in[indices,]
+  }
+  
+  for (row in 1:(dim(df_in)[1])) {
+    trajStr <- df_in[row,2]
+    trajStates <- parseTrajectory(trajStr)
+    Ndim <- dim(trajStates$N)
+    
+    if (length(Ndim)==0) {
+      ntypes <- 1
+      states <- bind_rows(states,
+                          tibble(traj=row,
+                                 type=0,
+                                 time=trajStates$time,
+                                 N=trajStates$N))
+    } else {
+      ntypes <- dim(trajStates$N)[2]
+      for (s in 1:ntypes) {
+        
+        states <- bind_rows(states,
+                            tibble(traj=row,
+                                   type=s-1,
+                                   time=trajStates$time,
+                                   N=trajStates$N[,s]))
+      }
+    }
+    
+    events <- bind_rows(events,
+                        tibble(traj=row,
+                               time=trajStates$time,
+                               event=trajStates$event,
+                               src=trajStates$src,
[…]

### 2021-03-25 scripts/utils.R

@@ -1,0 +1,331 @@
+##-----------------------------------------------------
+## Util functions for analysis workflow
+## 
+## 2020-10-14 Cecilia Valenzuela
+##------------------------------------------------------
+
+library(countrycode)
+library(ggpubr)
+library(lubridate)
+library(maps)
+library(sf)
+library(wpp2019)
+
+set_plotopts <- function() {
+  theme_set(theme_minimal())
+  theme_update(panel.grid = element_blank(),
+               legend.position = "bottom",
+               legend.box="vertical",
+               legend.text=element_text(size = 7),
+               legend.title=element_text(size = 7),
+               strip.background = element_rect(fill = "white", colour = "grey80"),
+               strip.text = element_text(size = 8, color = "grey30"),
+               axis.title.y = element_text(size = 8, #face = "bold", 
+                                           margin = margin(r = 10)),
+               axis.title.x = element_text(size = 8, #face = "bold", 
+                                           margin = margin(t = 10)),
+               panel.grid.major.y =  element_line(colour = "grey80", size = 0.2, linetype = 2),
+               #panel.grid.minor.y =  element_line(colour = "grey80", size = 0.2,  linetype = 3),
+               panel.background =  element_blank(),
+               axis.text.x = element_text(size = 7, angle = 30, hjust = 1, vjust = 1),
+               axis.text.y = element_text(size = 7),
+               plot.title = element_text(size=8, face = "bold"),
+               axis.ticks.length.x=unit(0.1, "cm"),
+               axis.ticks.length.y=unit(0.1, "cm"),
+               axis.ticks = element_line(colour = "grey50", size = 0.2))
+               #plot.margin = margin(3, 7, 3, 1.5))
+}
+
+
+
+get_cases <- function(demes, from = NA, to = NA) {
+  # Get case data information from ECDC for specific demes (countries and regions)
+  # Set from and to dates with values from the function call instead of the ones from the demes info
+  #cat("\nLoading confirmed case counts from ECDC...\n")
+  #if (!is.na(division_name)) case_data <- get_dataJH() Hubei specific data, but only from 22 Jan
+  #else case_data <- get_dataECDC()
[…]
